### 2024-05-08 manuscript.Rmd
start writing

---
title             : "FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
shorttitle        : "FAIR THEORY"

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Andreas Brandmaier"
    affiliation: "2"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Felix Schönbrodt"
    affiliation: "2"
    role:
      - Conceptualization
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Other affiliations"

authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.

abstract: |
  Test test.
keywords          : "keywords"
wordcount         : "X"
bibliography      : "theory-specification.bib"
floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

<!-- QUOTES -->
Kurt Lewin: Nothing is as practical as a good theory
Paul Meehl: [Psychological theories are neither] refuted nor corroborated, but instead merely fade away as people lose interest [@meehlTheoreticalRisksTabular1978].
[Theories are] like toothbrushes — no self-respecting person wants to use anyone else’s [@mischelToothbrushProblem2008]
"Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger. He could explain why the study did not work, e.g. because you used the wrong font for the questionnaire.

<!-- END QUOTES -->

The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of *FAIR Theory*.

FAIR Theory incorporates theory into open science workflows,
facilitates scholarly communication about theories,
making it easier to share theories with less opportunity for ambiguity and misunderstanding.
FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
removes barriers for knowledge exchange with the global scholarly community,
opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.

## Theory and Scientific Progress

Theories reflect scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
This process is described in the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \@ref(fig:figec)).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.

```{r}
library(tidySEM)
library(ggplot2)
y_spacing <- .5
lo <- get_layout("", "A", "",
                 "D", "", "B",
                 "", "C", "", rows = 3)

edg <- data.frame(from = c("A", "B", "C", "D"),
                  to = c("B", "C", "D", "A")
)

p <- prepare_graph(layout = lo, edges = edg)
p$edges[] <- lapply(p$edges, unlist)
p$edges$connect_from <- list("right", "bottom", "left", "top")
p$edges$connect_to <- list("top", "right", "bottom", "left")
p$edges$curvature <- c(rep(60, 4))
p$edges$label <- c("deduction", "testing", "induction", "generalization")
p$edges$label_colour <- "gray50"
p$edges$colour <- "gray50"

p$nodes$label <- c("Theory", "Hypothesis", "Data", "Observed\npatterns")

p_old <- p
g <- plot(p)

topofplot <- max(p$nodes$node_ymax)

headers <- data.frame(
  x = c(3, 5),
  y = topofplot + 2*y_spacing,
  lab = c("Inductive phase", "Deductive phase")
)

letters <- p$nodes[, c("name", "x", "y")]
letters$y <- letters$y+.26
letters$x <- letters$x+.5
g <- g + geom_hline(yintercept = topofplot+y_spacing) +
  geom_vline(xintercept = median(p$nodes$x), linetype = 5) +
  geom_text(data = headers, aes(x = x, y = y, label = lab), vjust = 1) +
  #geom_label(data = letters, aes(x = x, y = y, label = name), fill = "lightblue") +
  scale_x_continuous(expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))
g$layers <- g$layers[c(9, 1:8, 10:length(g$layers))]
ggsave("empirical_cycle.pdf", g, device = "pdf", width = 200, height = 200, units = "mm")
```

```{r figec, fig.cap="A take on the empirical cycle by De Groot"}
knitr::include_graphics("empirical_cycle.pdf")
```

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.

When comparing common practice in contemporary psychological research to this idealized model, at first glance,
it appears that deductive research is overrepresented in the literature. 
An overwhelming 89.6% of published studies ostensibly tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged, or are forgotten [REF Meehl].

Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
The present paper applies, for the first time, open science principles to theory.

## Publication is not Enough

Merely publishing a theory does not make it open;
open theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made *Findable* via a DOI,
are Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.

## Adapting the FAIR Principles

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable. From their inception, these principles were developed with "other research resources" in mind. Since then, scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht]. The present paper further extends the FAIR principles' definition to theory. Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.

## Is Current Psychological Theory FAIR?

Is it findable? There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory). Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).

Is it accessible? Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.

Is it interoperable? Theories rarely change,

Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).

In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties. Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.

FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.

There are key distinctions between theory and other FAIR digital research artefacts. With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.

## The Role of Theory Formalization

Concerns about the state of theory are a recurring theme in the psychological literature,
but previous writing has focused on theory formalization as a solution for ambiguity in psychological theory.
Greater formality increases theories' *empirical content*,
making them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
Conceptually, theory formalization is orthogonal to FAIR theory.
FAIR Theory does not require theories to be formal, and formal theory can be represented in a way that is not FAIR.
It is - in principle - possible to represent a collection of verbal statements as a FAIR Theory.
While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.

# Examples

## Using FAIR Theory to Perform Causal Inference

Based on this example: https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/

<!-- NOTES -->

* Theory is the vehicle of cumulative knowledge acquisition
* According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights. When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena.
* At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory.
* One source of potential improvements of theory methodology that has not been previously considered is computer science.
* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.
* FAIR theory allows for algorithmic derivation of hypotheses implied by the theory.
* FAIR theory enables integration testing: researchers can build a "test suite" of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
* To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2024-05-08 manuscript.tex
start writing

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2, Andreas Brandmaier2, \& Felix Schönbrodt2},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2}, Andreas Brandmaier\textsuperscript{2}, \& Felix Schönbrodt\textsuperscript{2}}
\date{}

\shorttitle{FAIR THEORY}

\authornote{

This is a preprint paper, generated from Git Commit \# fab0d274.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.

Correspondence concerning this article should be addressed to Caspar J. Van Lissa, Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands. E-mail: \href{mailto:c.j.vanlissa@tilburguniversity.edu}{\nolinkurl{c.j.vanlissa@tilburguniversity.edu}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Other affiliations}

\abstract{%
Test test.
}

\begin{document}
\maketitle

Kurt Lewin: Nothing is as practical as a good theory
Paul Meehl: {[}Psychological theories are neither{]} refuted nor corroborated, but instead merely fade away as people lose interest (Meehl, 1978).
{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's (Mischel, 2008)
``Call Leon'': When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger. He could explain why the study did not work, e.g.~because you used the wrong font for the questionnaire.

The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021),
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, ``theory crisis''.
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of \emph{FAIR Theory}.

FAIR Theory incorporates theory into open science workflows,
facilitates scholarly communication about theories,
making it easier to share theories with less opportunity for ambiguity and misunderstanding.
FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
removes barriers for knowledge exchange with the global scholarly community,
opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.

\subsection{Theory and Scientific Progress}\label{theory-and-scientific-progress}

Theories reflect scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
This process is described in the \emph{empirical cycle} (de Groot, 1961),
a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.

\begin{figure}
\centering
\includegraphics{empirical_cycle.pdf}
\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
\end{figure}

In a progressive research program (Lakatos, 1971),
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.

When comparing common practice in contemporary psychological research to this idealized model, at first glance,
it appears that deductive research is overrepresented in the literature.
An overwhelming 89.6\% of published studies ostensibly tests hypotheses (Kühberger, Fritz, \& Scherndl, 2014).
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies even reference specific theories (McPhetres et al., 2021).
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged, or are forgotten {[}REF Meehl{]}.

Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
The present paper applies, for the first time, open science principles to theory.

\subsection{Publication is not Enough}\label{publication-is-not-enough}

Merely publishing a theory does not make it open;
open theory should adhere to established open science standards.
The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output (e.g., software Lamprecht et al., 2019).
We propose to apply the FAIR principles to digital representations of theory as well,
introducing a FAIR metadata format to represent (formal) theories.
The resulting theories are made \emph{Findable} via a DOI,
are Accessible in a machine- and human-readable filetype,
Interoperable within the data analysis environment,
and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.

\subsection{Adapting the FAIR Principles}\label{adapting-the-fair-principles}

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable. From their inception, these principles were developed with ``other research resources'' in mind. Since then, scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}. The present paper further extends the FAIR principles' definition to theory. Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.

\subsection{Is Current Psychological Theory FAIR?}\label{is-current-psychological-theory-fair}

Is it findable? There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory). Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).

Is it accessible? Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being ``top of mind'', as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.

Is it interoperable? Theories rarely change,

Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).

In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties. Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.

FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.

There are key distinctions between theory and other FAIR digital research artefacts. With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.

\subsection{The Role of Theory Formalization}\label{the-role-of-theory-formalization}

Concerns about the state of theory are a recurring theme in the psychological literature,
but previous writing has focused on theory formalization as a solution for ambiguity in psychological theory.
Greater formality increases theories' \emph{empirical content},
making them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
Conceptually, theory formalization is orthogonal to FAIR theory.
FAIR Theory does not require theories to be formal, and formal theory can be represented in a way that is not FAIR.
It is - in principle - possible to represent a collection of verbal statements as a FAIR Theory.
While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.

\section{Examples}\label{examples}

\subsection{Using FAIR Theory to Perform Causal Inference}\label{using-fair-theory-to-perform-causal-inference}

Based on this example: \url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}

\begin{itemize}
\tightlist
\item
  Theory is the vehicle of cumulative knowledge acquisition
\item
  According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights. When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena.
\item
  At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory.
\item
  One source of potential improvements of theory methodology that has not been previously considered is computer science.
\item
  The process of ``iteratively improving'' digital objects - in this case, computer code - is well understood.
\item
  Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
\item
  This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
\item
  We introduce ``FAIR theory'', a digital research artifact to represent formal social scientific theories
\item
  FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit ``pull requests'' to suggest modifications of a theory, or can ``fork'' existing theories to create a spin-off from an existing theory.
\item
  FAIR theory allows for algorithmic derivation of hypotheses implied by the theory.
\item
  FAIR theory enables integration testing: researchers can build a ``test suite'' of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
\item
  To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.
\end{itemize}

\newpage

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-degrootMethodologieGrondslagenVan1961}
de Groot, A. D. (1961). \emph{Methodologie: Grondslagen van onderzoek en denken in de gedragswetenschappen}. 's Gravenhage: Uitgeverij Mouton. Retrieved from \url{https://books.google.com?id=6hiBDwAAQBAJ}

\bibitem[\citeproctext]{ref-kuhbergerPublicationBiasPsychology2014}
Kühberger, A., Fritz, A., \& Scherndl, T. (2014). Publication {Bias} in {Psychology}: {A Diagnosis Based} on the {Correlation} between {Effect Size} and {Sample Size}. \emph{PLoS ONE}, \emph{9}(9), e105825. \url{https://doi.org/10.1371/journal.pone.0105825}

\bibitem[\citeproctext]{ref-lakatosHistoryScienceIts1971}
Lakatos, I. (1971). History of {Science} and its {Rational Reconstructions}. In R. C. Buck \& R. S. Cohen (Eds.), \emph{{PSA} 1970: {In Memory} of {Rudolf Carnap Proceedings} of the 1970 {Biennial Meeting Philosophy} of {Science Association}} (pp. 91--136). Dordrecht: Springer Netherlands. \url{https://doi.org/10.1007/978-94-010-3142-4_7}

\bibitem[\citeproctext]{ref-lamprechtFAIRPrinciplesResearch2019}
Lamprecht, A.-L., Garcia, L., Kuzak, M., Martinez, C., Arcila, R., Martin Del Pico, E., \ldots{} Capella-Gutierrez, S. (2019). Towards {FAIR} principles for research software. \emph{Data Science}, 1--23. \url{https://doi.org/10.3233/DS-190026}

\bibitem[\citeproctext]{ref-lavelleWhenCrisisBecomes2021}
Lavelle, J. S. (2021). When a {Crisis Becomes} an {Opportunity}: {The Role} of {Replications} in {Making Better Theories}. \emph{The British Journal for the Philosophy of Science}, 714812. \url{https://doi.org/10.1086/714812}

\bibitem[\citeproctext]{ref-mcphetresDecadeTheoryReflected2021}
McPhetres, J., Albayrak-Aydemir, N., Mendes, A. B., Chow, E. C., Gonzalez-Marquez, P., Loukras, E., \ldots{} Volodko, K. (2021). A decade of theory as reflected in {Psychological Science} (2009--2019). \emph{PLOS ONE}, \emph{16}(3), e0247986. \url{https://doi.org/10.1371/journal.pone.0247986}

\bibitem[\citeproctext]{ref-meehlTheoreticalRisksTabular1978}
Meehl, P. E. (1978). Theoretical {Risks} and {Tabular Asterisks}: {Sir Karl}, {Sir Ronald}, and the {Slow Progress} of {Soft Psychology}. \emph{Journal of Consulting \& Clinical Psychology}, \emph{46}(4), 806--834.

\bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}

\bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015a}
Nosek, B. A., Alter, G., Banks, G. C., Borsboom, D., Bowman, S. D., Breckler, S. J., \ldots{} Yarkoni, T. (2015). Promoting an open research culture. \emph{Science}, \emph{348}(6242), 1422--1425. \url{https://doi.org/10.1126/science.aab2374}

\bibitem[\citeproctext]{ref-oberauerAddressingTheoryCrisis2019}
Oberauer, K., \& Lewandowsky, S. (2019). Addressing the theory crisis in psychology. \emph{Psychonomic Bulletin \& Review}, \emph{26}(5), 1596--1618. \url{https://doi.org/10.3758/s13423-019-01645-2}

\bibitem[\citeproctext]{ref-robinaughInvisibleHandsFine2021}
Robinaugh, D. J., Haslbeck, J. M. B., Ryan, O., Fried, E. I., \& Waldorp, L. J. (2021). Invisible {Hands} and {Fine Calipers}: {A Call} to {Use Formal Theory} as a {Toolkit} for {Theory Construction}. \emph{Perspectives on Psychological Science}, \emph{16}(4), 725--743. \url{https://doi.org/10.1177/1745691620974697}

\bibitem[\citeproctext]{ref-scheelWhyMostPsychological2022}
Scheel, A. M. (2022). Why most psychological research findings are not even wrong. \emph{Infant and Child Development}, \emph{31}(1), e2295. \url{https://doi.org/10.1002/icd.2295}

\bibitem[\citeproctext]{ref-scheelExcessPositiveResults2021}
Scheel, A. M., Schijen, M. R. M. J., \& Lakens, D. (2021). An {Excess} of {Positive Results}: {Comparing} the {Standard Psychology Literature With Registered Reports}. \emph{Advances in Methods and Practices in Psychological Science}, \emph{4}(2), 25152459211007467. \url{https://doi.org/10.1177/25152459211007467}

\bibitem[\citeproctext]{ref-scheelWhyHypothesisTesters2021}
Scheel, A. M., Tiokhin, L., Isager, P. M., \& Lakens, D. (2021). Why {Hypothesis Testers Should Spend Less Time Testing Hypotheses}. \emph{Perspectives on Psychological Science}, \emph{16}(4), 744--755. \url{https://doi.org/10.1177/1745691620966795}

\bibitem[\citeproctext]{ref-szollosiArrestedTheoryDevelopment2021}
Szollosi, A., \& Donkin, C. (2021). Arrested theory development: {The} misguided distinction between exploratory and confirmatory research. \emph{Perspectives on Psychological Science}, \emph{16}(4), 717--724. \url{https://doi.org/10.1177/1745691620966796}

\end{CSLReferences}

\end{document}

### 2024-05-14 manuscript.Rmd
proposed keywords and automated word count

@@
   - name: "Aaron Peikert"
-    affiliation: "2"
+    affiliation: "2,3"
     role:
@@
       - Writing – review & editing
-  - name: "Andreas Brandmaier"
-    affiliation: "2"
+  - name: "Andreas M. Brandmaier"
+    affiliation: "2,3,4"
     role:
@@
   - name: "Felix Schönbrodt"
-    affiliation: "2"
+    affiliation: "5"
     role:
@@
   - id            : "2"
+    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
+  - id            : "3"
+    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
+  - id            : "4"
+    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
+  - id            : "5"
     institution   : "Other affiliations"
-
 authornote: |
@@
   Test test.
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "meta theory, theory formation, cumulative science, formal models"
+wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
 bibliography      : "theory-specification.bib"

### 2024-05-15 manuscript.Rmd
update 2024-05-15 16:18:17.547173

--- 2024-05-08 manuscript.Rmd
@@
 
-Based on this example: https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/
+This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
+We could use that to illustrate updating FAIR theory:
+
+https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45
+
+We can find more examples of causal inference with DAGs in these tutorials:
+
+https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/
+
+https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/
 

### 2024-05-16 manuscript.Rmd
theories, frameworks, paradigms?!
Do we want to discuss this?

--- 2024-05-14 manuscript.Rmd
@@
 
+In our experience, there is no clear delineation between paradigms, frameworks and theories. 
+We think it is helpful to view these terms as discrete labels on a continuum from broad to narrow, from general to specific, when organizing knowledge.
+Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
+Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
+
 ## Publication is not Enough

### 2024-09-08 manuscript.Rmd
update 2024-09-08 08:51:02.338884

--- 2024-05-16 manuscript.Rmd
@@
 
+The FAIR Guiding Principles (hereafter: FAIR principles) were initially established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
+Since they were first introduced, scholars have demonstrated the relevance of these criteria for making other digital research artefacts more open.
+This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
+We introduce "FAIR theory", a digital representation of theory that can complies with the FAIR principles.
+
 <!-- QUOTES -->
@@
 The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
+
 Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
@@
 
-When comparing common practice in contemporary psychological research to this idealized model, at first glance,
-it appears that deductive research is overrepresented in the literature. 
-An overwhelming 89.6% of published studies ostensibly tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
+Contemporary psychological research falls short of this idealized model in several ways.
+Firstly, because deductive research is overrepresented in the literature. 
+According to one estimate, 89.6% of published studies tests hypotheses, which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
 Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
@@
 Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
-Existing theories either persist unchanged, or are forgotten [REF Meehl].
+Existing theories either persist unchanged for decades, or are forgotten [REF Meehl].
 

### 2024-09-13 manuscript.Rmd
update 2024-09-13 10:38:59.00593

@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were initially established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
-Since they were first introduced, scholars have demonstrated the relevance of these criteria for making other digital research artefacts more open.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
+Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
-We introduce "FAIR theory", a digital representation of theory that can complies with the FAIR principles.
+To this end, we introduce the concept of "FAIR theory",
+a digital representation of theory compliant with the FAIR principles.
 
-<!-- QUOTES -->
+<!-- QUOTES
+
 Kurt Lewin: Nothing is as practical as a good theory
@@
 
-<!-- END QUOTES -->
+END QUOTES -->
 
 The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-
 Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 But have these reforms met their goal?
-Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
-which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
-If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.
+Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021].
+One plausible explanation is that the replication crisis is symptomatic of an underlying,
+and more fundamental, "theory crisis".
+If psychological theories are insufficiently precise to derive testable hypotheses,
+or sufficiently vague to explain contradictory findings,
+then preregistration and replication will only serve to highlight those shortcomings.
 
 Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
-A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
-The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of *FAIR Theory*.
+One contributing factor is a lack of theory formalization:
+social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
+A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
+The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
+and introducing the concept of *FAIR Theory*.
 
@@
 
-Theories reflect scientists' understanding of phenomena.
+According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
+a philosophical model of cumulative knowledge acquisition,
+research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
+In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
+In this model, theories are the vehicle of scientists' understanding of phenomena.
 Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
-This process is described in the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
-a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \@ref(fig:figec)).
-In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
 
@@
 the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-
-Contemporary psychological research falls short of this idealized model in several ways.
-Firstly, because deductive research is overrepresented in the literature. 
-According to one estimate, 89.6% of published studies tests hypotheses, which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
+There are clear indications that contemporary psychology falls short of this idealized model, however.
+Firstly, because deductive research is over-represented in the literature. 
+According to one estimate, 89.6% of published studies tests hypotheses,
+which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
 Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
-Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
-Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
-Existing theories either persist unchanged for decades, or are forgotten [REF Meehl].
+Only 15% of deductive studies reference theory at all [@mcphetresDecadeTheoryReflected2021].
+This raises the question where the other hypotheses come from,
+and what consequence their refutation would have for theory.
+These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
+The majority of papers ostensibly test hypotheses,
+but these are rarely derived from theory,
+and test results rarely contribute to the improvement of existing theories.
+Consequently, theories either persist unchanged for decades, or are forgotten [REF Meehl].
 
@@
 Merely publishing a theory does not make it open;
-open theory should adhere to established open science standards.
+to be open, theory should adhere to established open science standards.
 The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
@@
 The resulting theories are made *Findable* via a DOI,
-are Accessible in a machine- and human-readable filetype,
+Accessible in a machine- and human-readable filetype,
 Interoperable within the data analysis environment,
-and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.
+and Reusable in the practical and legal sense, so that they may be improved over time.
 
@@
 From their inception, these principles were developed with "other research resources" in mind.
-Since then, scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
-The present paper further extends the FAIR principles' definition to theory.
-Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.
+Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
+The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
+
+```{r tabfair}
+tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
+names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
+papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
+```
 
@@
 
-Is it findable?
-There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory).
-Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).
+While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
+we provide a narrative review of examples and counterexamples for each criterion.
 
-Is it accessible?
-Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.
+One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
+A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
+or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
+Modular publishing could ameliorate the Findability of psychological theory.
+In this approach, distinct resources are individually published as citable academic output.
+Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
+Another effort to improve theories' findability is post-hoc curation.
+For example, Gray and colleagues introduced a format for representing theories,
+and post many examples on their website.
+Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).
 
-Is it interoperable? Theories rarely change,
+With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
+Open Access publishing increases the accessibility of all academic output, including theory.
+Another factor curtailing the accessibility of theory is ambiguity;
+a lack of theory formalization introduces a dependency on the original author for clarification.
+The discourse on "Great Man Theorizing" touches upon the problems this introduces [REF Guest et al moral theory].
+For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
+Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
+To some extent, this relates to the problem of translation [REF Duhem]:
+it is not possible to entirely formalize an idea to enable unambiguous interpretation.
+Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
+
+ <!-- (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.
+-->
+
+The primary indication that psychological theory has limited interoperability ? Theories rarely change,
 

### 2024-09-13 manuscript.tex
update 2024-09-13 10:38:59.00593

@@
 {}{\PatchFailed}
-\keywords{keywords\newline\indent Word count: X}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 1808}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
   pdftitle={FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2, Andreas Brandmaier2, \& Felix Schönbrodt2},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix Schönbrodt5},
   pdflang={en-EN},
-  pdfkeywords={keywords},
+  pdfkeywords={meta theory, theory formation, cumulative science, formal models},
   hidelinks,
@@
 \title{FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
-\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2}, Andreas Brandmaier\textsuperscript{2}, \& Felix Schönbrodt\textsuperscript{2}}
+\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix Schönbrodt\textsuperscript{5}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# fab0d274.
+This is a preprint paper, generated from Git Commit \# 50b18562.
 
@@
 Caspar J.
-Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.
+Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M.
+Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Other affiliations}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Other affiliations}
 
@@
 
-Kurt Lewin: Nothing is as practical as a good theory
-Paul Meehl: {[}Psychological theories are neither{]} refuted nor corroborated, but instead merely fade away as people lose interest (Meehl, 1978).
-{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's (Mischel, 2008)
-``Call Leon'': When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger.
-He could explain why the study did not work, e.g.~because you used the wrong font for the questionnaire.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
+Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
+This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
+To this end, we introduce the concept of ``FAIR theory'',
+a digital representation of theory compliant with the FAIR principles.
 
@@
 But have these reforms met their goal?
-Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021),
-which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, ``theory crisis''.
-If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.
+Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
+One plausible explanation is that the replication crisis is symptomatic of an underlying,
+and more fundamental, ``theory crisis''.
+If psychological theories are insufficiently precise to derive testable hypotheses,
+or sufficiently vague to explain contradictory findings,
+then preregistration and replication will only serve to highlight those shortcomings.
 
 Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
-A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
-The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of \emph{FAIR Theory}.
+One contributing factor is a lack of theory formalization:
+social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
+A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
+The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
+and introducing the concept of \emph{FAIR Theory}.
 
@@
 
-Theories reflect scientists' understanding of phenomena.
+According to the \emph{empirical cycle} (de Groot, 1961),
+a philosophical model of cumulative knowledge acquisition,
+research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
+In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
+In this model, theories are the vehicle of scientists' understanding of phenomena.
 Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
-This process is described in the \emph{empirical cycle} (de Groot, 1961),
-a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \ref{fig:figec}).
-In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
 
@@
 the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-
-When comparing common practice in contemporary psychological research to this idealized model, at first glance,
-it appears that deductive research is overrepresented in the literature.
-An overwhelming 89.6\% of published studies ostensibly tests hypotheses (Kühberger, Fritz, \& Scherndl, 2014).
+There are clear indications that contemporary psychology falls short of this idealized model, however.
+Firstly, because deductive research is over-represented in the literature.
+According to one estimate, 89.6\% of published studies tests hypotheses,
+which suggests that the literature is predominantly comprised of deductive research (Kühberger, Fritz, \& Scherndl, 2014).
 Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
-Only 15\% of deductive studies even reference specific theories (McPhetres et al., 2021).
-Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
-Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
-Existing theories either persist unchanged, or are forgotten {[}REF Meehl{]}.
+Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
+This raises the question where the other hypotheses come from,
+and what consequence their refutation would have for theory.
+These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
+The majority of papers ostensibly test hypotheses,
+but these are rarely derived from theory,
+and test results rarely contribute to the improvement of existing theories.
+Consequently, theories either persist unchanged for decades, or are forgotten {[}REF Meehl{]}.
 
@@
 Merely publishing a theory does not make it open;
-open theory should adhere to established open science standards.
+to be open, theory should adhere to established open science standards.
 The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output (e.g., software Lamprecht et al., 2019).
@@
 The resulting theories are made \emph{Findable} via a DOI,
-are Accessible in a machine- and human-readable filetype,
+Accessible in a machine- and human-readable filetype,
 Interoperable within the data analysis environment,
-and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.
+and Reusable in the practical and legal sense, so that they may be improved over time.
 
@@
 From their inception, these principles were developed with ``other research resources'' in mind.
-Since then, scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
-The present paper further extends the FAIR principles' definition to theory.
-Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.
+Scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
+The present paper further extends the FAIR principles' definition to theory, see Table \ref{tab:tabfair}.
+
+\begin{lltable}
+
+\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:tabfair}}\\
+\toprule
+Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
+\toprule
+Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
+\midrule
+\endhead
+F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory and its associated metadata has a global, unique and persistent identifier for each version (using semantic versioning) & Rephrased\\
+F2 & Data are described with rich metadata & Theory is described with rich metadata & \textasciitilde{}same\\
+F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & \textasciitilde{}same\\
+F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & NEEDS WORK!
+GitHub is indexed by Google I believe, but ideally, we'd like our theories to show up in Google Scholar, or even dedicated academic search enginges.
+Where could we put them to realize this?\\
+A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & \textasciitilde{}same\\
+A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Same\\
+A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Do\ \ we need this?\\
+A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Do\ \ we need this?\\
+I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased\\
+I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & NEEDS WORK!
+I think this is where we explain the value of e.g.
+universal graph languages like Aaron and Max'\\
+I2S.1 & - &  & \\
+I2S.2 & - &  & \\
+I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Rephrased.
+I envision a LinkList-like structure where each theory version references its ancestor\\
+I4S & - &  & Discard\\
+R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Needs work.
+How do we envision this?
+Keywords?
+ISBN-like codes for variable type?\\
+R1.1 & (Meta)data are released with a clear and accessible data usage license & (Meta)data are released with a clear and accessible license & Needs work: which license is good for theory?\\
+R1.2 & (Meta)data are associated with detailed provenance & (Meta)data are associated with detailed provenance & Relates to I3; I think this point relates more to the theory's ancestry, and I3 relates to e.g.
+incorporating other theories within a theory (e.g., theory of measurement inside of structural theory)\\
+R1.3 & (Meta)data meet domain-relevant community standards & Theory metadata and documentation meet domain-relevant community standards & These standards do not yet exist; we can take a first step towards developing them and recommend that this be an active area of development\\
+\bottomrule
+\end{longtable}
+
+\end{lltable}
 
@@
 
-Is it findable?
-There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory).
-Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).
+While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
+we provide a narrative review of examples and counterexamples for each criterion.
+One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
+A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
+or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
+The modular publishing movement may ameliorate the Findability of psychological theory.
+In this approach, distinct resources are individually published as citable academic output.
+Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
+Some have sought to address limitations of theories'findability through post-hoc curation.
+For example, Gray and colleagues introduced a format for representing theories,
+and post many examples on their website.
+Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).
 
-Is it accessible?
-Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being ``top of mind'', as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.
+With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
+Open Access publishing is a practice that increases the accessibility of all academic output, including theory.
+Another factor curtailing the accessibility of theory is ambiguity;
+a lack of formalization introduces a dependency on the original author for clarification.
+The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
+Dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
+Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
+This also relates to the problem of translation {[}REF Duhem{]}:
+it is not possible to entirely formalize an idea to enable unambiguous interpretation.
+Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
 
@@
 
-Based on this example: \url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}
+This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
+We could use that to illustrate updating FAIR theory:
+
+\url{https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45}
+
+We can find more examples of causal inference with DAGs in these tutorials:
+
+\url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}
+
+\url{https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/}
 
@@
 
-\bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
-Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}
-
 \bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015a}

### 2024-09-13 manuscript.Rmd
update 2024-09-13 15:29:39.292327

@@
 Kurt Lewin: Nothing is as practical as a good theory
-Paul Meehl: [Psychological theories are neither] refuted nor corroborated, but instead merely fade away as people lose interest [@meehlTheoreticalRisksTabular1978].
-[Theories are] like toothbrushes — no self-respecting person wants to use anyone else’s [@mischelToothbrushProblem2008]
+
 "Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger.
@@
 
-The primary indication that psychological theory has limited interoperability ? Theories rarely change,
+<!-- The data usually need to be integrated with other data. In addition, the data need to interoperate with applications or workflows for analysis, storage, and processing. -->
 
+There are clear indications that psychological theory has limited interoperability. 
+For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
+There even appears to be a norm *against* theory interoperability, as evident from the quip that 
+*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008]
+To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
+For example, it should be possible to integrate theory directly into analysis workflows:
+to derive hypotheses, select control variables, and guide model specification.
+A lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense. 
+Additionally, theories should be interoperable with each other:
+for instance, it should be possible to embed a specific theory about the process of emotion regulation [REF Gross] within a theory of emotion regulation development [REF Morris].
+
+<!-- The ultimate goal of FAIR is to optimise the reuse of data. To achieve this, metadata and data should be well-described so that they can be replicated and/or combined in different settings. -->
 Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).

### 2024-09-16 manuscript.Rmd
update 2024-09-16 09:57:53.8483

@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
-Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
+Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
-To this end, we introduce the concept of "FAIR theory",
+To this end, we introduce "FAIR theory":
 a digital representation of theory compliant with the FAIR principles.
+FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
-There even appears to be a norm *against* theory interoperability, as evident from the quip that 
-*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008]
 To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
@@
 to derive hypotheses, select control variables, and guide model specification.
-A lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense. 
+The aforementioned lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense.
 Additionally, theories should be interoperable with each other:
@@
 
-<!-- The ultimate goal of FAIR is to optimise the reuse of data. To achieve this, metadata and data should be well-described so that they can be replicated and/or combined in different settings. -->
-Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).
+The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
+While theory is ostensibly held in high regard in the social sciences,
+there also appears to be a norm *against* theory reuse.
+This is evident from the quip that 
+*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
+In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
+and "who determines how a theory may be reused or changed"?
+Licensing theories for reuse provides an unambiguous answer to such questions.
+
+One important consideration is that there may be legal limits to the rights authors can claim over theory.
+An important concept in copyright law is the *idea-expression dichotomy* [REF].
+
+o.
+Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts.
+Separating an idea from the expression or manifestation of that idea is known in copyright law as the
+Copyright explicitly does not
+*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*,
+but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
+Copyright pertaining to recipes offer an interesting perspective:
+Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright.
+Extrapolating to FAIR Theory, it would seem that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
+If copyright does not cover (formal) theory,
+then it would be pointless to adopt a license that assumes copyright protection.
+Furtermore, even if copyright would apply, research is covered under "fair use" excemptions to copyright law, so copyright would impose few restrictions to theory reuse in scholarly communication.
+In sum, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
+it explicitly waives all rights and encourages reuse.
+Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [REF TOP guidelines],
+and the spirit of good scholarship,
+it is essential that scholars do comprehensively cite theory,
+including prior work that new theories are based on.
+
+## FAIR Theory and Recognition & Rewards
 
@@
 
+## Version Control
+
+* One source of potential improvements of theory methodology that has not been previously considered is computer science.
+* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
+* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
+* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
+* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
+* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
+Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.
+
 # Examples
+
+## Formalizing the Empirical Cycle
+
+In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
+As several authors have taken inspiration from the work by De Groot,
+we compare our interpretation of the original theory to the interpretation of others.
+Originally, the theory has the following structure:
+
+```
+
+digraph {
+
+  observation;
+  induction;
+  deduction;
+  test;
+  evaluation;
+
+  observation -> induction;
+  induction -> deduction;
+  deduction -> test;
+  test -> evaluation;
+  evaluation -> observation;
+
+}
+```
+
+Subsequently, Wagenmakers and colleagues modified the theory by *"[adding the] Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*:
+
+```
+digraph {
+
+  subgraph cluster_discovery {
+    label="Discovery";
+    hypothesis [label="New hypothesis"];
+    prediction [label="New prediction"];
+  }
+  data  [label="Old knowledge and old data"];      
+  subgraph cluster_justification {
+    label="Justification";
+    test [label="Test on new data"];
+    evaluation;
+  }
+
+  data -> hypothesis [label="Speculate & explore"];
+  hypothesis -> prediction  [label="Deduce"];
+  prediction -> test  [label="Design new experiment"];
+  test -> evaluation  [label="Statistical analysis"];
+  evaluation -> data  [label="Knowledge accumulation"];
+
+}
+```
+
+Note, however, that there appear to be further changes:
+the phases of the cycle have been renamed,
+and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
+Moreover, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
+As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
+
+Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:
+
+```
+digraph {
+
+  theory;
+  prediction;
+  test [label="inferential procedure"];
+  observation;
+
+  theory -> prediction [label="deduction"];
+  prediction -> test;
+  test -> observation;
+  observation -> theory [label="generalization"];
+
+}
+```
+
+In our representation,
+induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
+For example, the refutation of a hypothesized effect,
+or the serendipitous observation of some pattern in data,
+might be a reason to revise or construct theory.
+Induction, incidentally, also occurs within the link from prediction to testing:
+in the form of the inductive bias of methods used to perform the test,
+and auxiliary assumptions that must be made to address remaining theoretical ambiguities.
+
+<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced.
+-->
+<!-- ``` -->
 
@@
 
-This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
+Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
+According to Pearl and colleagues,
+explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
+Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
+and could even be integrated into the analysis environment.
+
+In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
 We could use that to illustrate updating FAIR theory:

### 2024-09-16 manuscript.tex
update 2024-09-16 09:57:53.8483

@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 1808}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 2657}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-This is a preprint paper, generated from Git Commit \# 50b18562.
+This is a preprint paper, generated from Git Commit \# d1a82fc7.
 
@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
-Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
+Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
-To this end, we introduce the concept of ``FAIR theory'',
+To this end, we introduce ``FAIR theory'':
 a digital representation of theory compliant with the FAIR principles.
+FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 we provide a narrative review of examples and counterexamples for each criterion.
+
 One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
@@
 or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
-The modular publishing movement may ameliorate the Findability of psychological theory.
+Modular publishing could ameliorate the Findability of psychological theory.
 In this approach, distinct resources are individually published as citable academic output.
 Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
-Some have sought to address limitations of theories'findability through post-hoc curation.
+Another effort to improve theories' findability is post-hoc curation.
 For example, Gray and colleagues introduced a format for representing theories,
@@
 With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
-Open Access publishing is a practice that increases the accessibility of all academic output, including theory.
+Open Access publishing increases the accessibility of all academic output, including theory.
 Another factor curtailing the accessibility of theory is ambiguity;
-a lack of formalization introduces a dependency on the original author for clarification.
+a lack of theory formalization introduces a dependency on the original author for clarification.
 The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
-Dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
+For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
 Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
-This also relates to the problem of translation {[}REF Duhem{]}:
+To some extent, this relates to the problem of translation {[}REF Duhem{]}:
 it is not possible to entirely formalize an idea to enable unambiguous interpretation.
@@
 
-Is it interoperable? Theories rarely change,
+There are clear indications that psychological theory has limited interoperability.
+For one, theories are rarely \emph{``refuted nor corroborated, but instead merely fade away as people lose interest''} (Meehl, 1978).
+To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
+For example, it should be possible to integrate theory directly into analysis workflows:
+to derive hypotheses, select control variables, and guide model specification.
+The aforementioned lack of formalization {[}REF Robinaugh{]} and ambiguity {[}REF Frankenhuis{]} prevent interoperability in such a practical sense.
+Additionally, theories should be interoperable with each other:
+for instance, it should be possible to embed a specific theory about the process of emotion regulation {[}REF Gross{]} within a theory of emotion regulation development {[}REF Morris{]}.
 
-Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).
+There appears to be a norm \emph{against} the reuse of theory,
+as evident from the quip that
+\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
+Moreover, a legal basis for theory reuse is often absent.
+Questions that often come up in our workshops on FAIR Theory are ``who owns theory'',
+and ``who determines how a theory may be changed''?
+Such questions require a clear answer, even if that answer may vary across theories.
+Licensing theories for reuse provides clarity about how a theory can be reused.
+A crucial legal consideration is that,
+while specific (FAIR) representations of theory might be protected by copyright,
+the underlying ideas are not.
+This is another argument in favor of publishing specific instantiations of theory along with licenses detailing how they may be reused.
+
+\subsection{FAIR Theory and Recognition \& Rewards}\label{fair-theory-and-recognition-rewards}
 
@@
 
+\subsection{Version Control}\label{version-control}
+
+\begin{itemize}
+\tightlist
+\item
+  One source of potential improvements of theory methodology that has not been previously considered is computer science.
+\item
+  The process of ``iteratively improving'' digital objects - in this case, computer code - is well understood.
+\item
+  Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
+\item
+  This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
+\item
+  We introduce ``FAIR theory'', a digital research artifact to represent formal social scientific theories
+\item
+  FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
+Version control also enables diffuse collaboration in theory development, as other researchers can submit ``pull requests'' to suggest modifications of a theory, or can ``fork'' existing theories to create a spin-off from an existing theory.
+\end{itemize}
+
 \section{Examples}\label{examples}
+
+\subsection{Formalizing the Empirical Cycle}\label{formalizing-the-empirical-cycle}
+
+In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
+As several authors have taken inspiration from the work by De Groot,
+we compare our interpretation of the original theory to the interpretation of others.
+Originally, the theory has the following structure:
+
+\begin{verbatim}
+
+digraph {
+
+  observation;
+  induction;
+  deduction;
+  test;
+  evaluation;
+
+  observation -> induction;
+  induction -> deduction;
+  deduction -> test;
+  test -> evaluation;
+  evaluation -> observation;
+
+}
+\end{verbatim}
+
+Subsequently, Wagenmakers and colleagues modified the theory by \emph{``{[}adding the{]} Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}:
+
+\begin{verbatim}
+digraph {
+
+  subgraph cluster_discovery {
+    label="Discovery";
+    hypothesis [label="New hypothesis"];
+    prediction [label="New prediction"];
+  }
+  data  [label="Old knowledge and old data"];      
+  subgraph cluster_justification {
+    label="Justification";
+    test [label="Test on new data"];
+    evaluation;
+  }
+
+  data -> hypothesis [label="Speculate & explore"];
+  hypothesis -> prediction  [label="Deduce"];
+  prediction -> test  [label="Design new experiment"];
+  test -> evaluation  [label="Statistical analysis"];
+  evaluation -> data  [label="Knowledge accumulation"];
+
+}
+\end{verbatim}
+
+Note, however, that there appear to be further changes:
+the phases of the cycle have been renamed,
+and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
+Moreover, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
+As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
+
+Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:
+
+\begin{verbatim}
+digraph {
+
+  theory;
+  prediction;
+  test [label="inferential procedure"];
+  observation;
+
+  theory -> prediction [label="deduction"];
+  prediction -> test;
+  test -> observation;
+  observation -> theory [label="generalization"];
+
+}
+\end{verbatim}
+
+In our representation,
+induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
+For example, the refutation of a hypothesized effect,
+or the serendipitous observation of some pattern in data,
+might be a reason to revise or construct theory.
+Induction, incidentally, also occurs within the link from prediction to testing:
+in the form of the inductive bias of methods used to perform the test,
+and auxiliary assumptions that must be made to address remaining theoretical ambiguities.
 
@@
 
-This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
+Some have argued that \emph{causal explanations} are a property of good theory {[}REF Meehl, etc?{]}.
+According to Pearl and colleagues,
+explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
+Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
+and could even be integrated into the analysis environment.
+
+In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
 We could use that to illustrate updating FAIR theory:
@@
 
+\bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
+Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}
+
 \bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015a}

### 2024-09-30 manuscript.Rmd
update 2024-09-30 10:40:00.745403

@@
 
-The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
-But have these reforms met their goal?
-Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021].
-One plausible explanation is that the replication crisis is symptomatic of an underlying,
-and more fundamental, "theory crisis".
-If psychological theories are insufficiently precise to derive testable hypotheses,
-or sufficiently vague to explain contradictory findings,
-then preregistration and replication will only serve to highlight those shortcomings.
+## The Need for FAIR Theory
+
+The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
+Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+These various practices ensure transparent and repeated testing of hypotheses.
+However, recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021].
+Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+psychological theories rarely produce hypotheses that are found to be true,
+and are often sufficiently vague to explain contradictory findings.
 
 Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One contributing factor is a lack of theory formalization:
-social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
-A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
-The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
+One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as *formalization* [@szollosiArrestedTheoryDevelopment2021].
+A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
+The present paper focuses on the second concern,
+advancing transparent communication about theory by applying open science principles to psychological theory for the first time
 and introducing the concept of *FAIR Theory*.
-
-FAIR Theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers for knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
 
@@
 the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-There are clear indications that contemporary psychology falls short of this idealized model, however.
+There are indications that contemporary psychology falls short of this idealized model, however.
 Firstly, because deductive research is over-represented in the literature. 
@@
 Only 15% of deductive studies reference theory at all [@mcphetresDecadeTheoryReflected2021].
-This raises the question where the other hypotheses come from,
-and what consequence their refutation would have for theory.
+This invites the question where the other hypotheses come from,
+and to what extent testing an ad-hoc hypothesis not grounded in theory can advance our principled understanding of the studied phenomena.
 These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
@@
 and test results rarely contribute to the improvement of existing theories.
-Consequently, theories either persist unchanged for decades, or are forgotten [REF Meehl].
+Consequently, theories either persist unchanged, immune to refutation,
+or are forgotten [REF Meehl].
 
-Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
-The present paper applies, for the first time, open science principles to theory.
+<!-- Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory. -->
+<!-- The present paper applies, for the first time, open science principles to theory. -->
 
-## Publication is not Enough
+## The Solution
 
-Merely publishing a theory does not make it open;
-to be open, theory should adhere to established open science standards.
-The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
-We propose to apply the FAIR principles to digital representations of theory as well,
-introducing a FAIR metadata format to represent (formal) theories.
-The resulting theories are made *Findable* via a DOI,
-Accessible in a machine- and human-readable filetype,
-Interoperable within the data analysis environment,
-and Reusable in the practical and legal sense, so that they may be improved over time.
+Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition for several reasons.
 
-## Adapting the FAIR Principles
+### Findability
 
-The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
-From their inception, these principles were developed with "other research resources" in mind.
-Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
-The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
+One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
+The primary unit of dissemination and search in psychology is often the academic paper.
+A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
+or standardized metadata to signal the presence of theory within a work,
+nor even a standardized keyword (model, framework, etc are often used interchangeably with theory).
+Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
+This would allow researchers to easily identify relevant theories to inform their hypotheses,
+grounding their work in established theoretical foundations.
+Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
+For example, it becomes possible to study theories (the same way libraries and search engines have enabled scholars to study the literature via systematic reviews);
+theories might have transdisciplinary impact;
+and the structure of theories from one discipline might inspire analogical modeling in other fields, the way - for example, predator-prey models have inspired theories of XXX and the Eysenck model of atomic magnetism has inspired a network theory of depression.
 
-```{r tabfair}
-tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
-names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
-papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
-```
-
-## Is Current Psychological Theory FAIR?
-
-While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
-we provide a narrative review of examples and counterexamples for each criterion.
-
-One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
-A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
-or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
-Modular publishing could ameliorate the Findability of psychological theory.
-In this approach, distinct resources are individually published as citable academic output.
-Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
-Another effort to improve theories' findability is post-hoc curation.
+One effort that improves theories' findability is post-hoc curation.
 For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website.
-Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).
+and post many examples on their website [REF GRAY].
+Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain) [REF BORSBOOM].
+Another method that could ameliorate theory Findability is modular publishing.
+In this approach, distinct assets are individually published as citable academic output.
+Such output can be linked; for example, a FAIR theory could have an associated paper.
 
@@
 including prior work that new theories are based on.
+
+2. Accessibility: Promoting Open and Transparent Theoretical Development
+Accessibility involves ensuring that theories are openly available to all researchers, not locked behind paywalls or isolated in specific journals.
+By making theories FAIR, they would be accessible under clear, standardized, and open conditions, using platforms that allow both humans and machines to retrieve them effortlessly.
+
+How it helps: Openly accessible theories would allow for greater scrutiny and replication by the research community, fostering transparent scholarly communication.
+This ensures that researchers worldwide can contribute to, refine, and build upon existing theories, thus accelerating the pace of scientific discovery and enhancing the collective understanding of a given phenomenon.
+3. Interoperability: Facilitating Integration Across Fields
+In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks.
+The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.
+
+How it helps: If theories are made interoperable, they can be used in multi-disciplinary research and can better interface with diverse fields.
+This reduces duplication of effort and enables researchers from different areas to combine and compare theories, bridging gaps between disciplines and fostering innovative, cross-disciplinary approaches.
+Researchers can easily synthesize theoretical perspectives, generating new hypotheses that are more comprehensive and grounded in a wider range of evidence.
+4. Reusability: Enabling Continuous Improvement of Theories
+For knowledge to accumulate meaningfully, theories need to be adaptable and subject to continuous revision as new data emerge.
+The reusability principle calls for thorough documentation and community standards to ensure that data (or theories) can be applied in new contexts.
+
+How it helps: Reusable theories would provide clear metadata, version histories, and supporting evidence, allowing other researchers to test, refine, or extend them as new empirical findings emerge.
+This improves the feedback loop between theory and experimentation, ensuring that test results inform future theoretical development.
+It encourages the use of theories not just for their original purpose but in new and evolving contexts, thus ensuring that the theory remains dynamic and continuously refined over time.
+Conclusion: Making Theories FAIR Accelerates Scientific Progress
+Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
+By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
+This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
+
+## Publication is not Enough
+
+Merely publishing a theory does not make it open;
+to be open, theory should adhere to established open science standards.
+The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
+We propose to apply the FAIR principles to digital representations of theory as well,
+introducing a FAIR metadata format to represent (formal) theories.
+The resulting theories are made *Findable* via a DOI,
+Accessible in a machine- and human-readable filetype,
+Interoperable within the data analysis environment,
+and Reusable in the practical and legal sense, so that they may be improved over time.
+
+## Adapting the FAIR Principles
+
+The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
+From their inception, these principles were developed with "other research resources" in mind.
+Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
+The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
+
+```{r tabfair}
+tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
+names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
+papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
+```
 
@@
 
+# Discussion
+
+## Future Directions
+
+One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
+Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
+it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
+Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
+the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
+FAIR Theory can help address the measurement crisis:
+since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
+
+FAIR Theory incorporates theory into open science workflows,
+facilitates scholarly communication about theories,
+making it easier to share theories with less opportunity for ambiguity and misunderstanding.
+FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
+More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
+removes barriers for knowledge exchange with the global scholarly community,
+opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
+
 \newpage

### 2024-11-19 manuscript.Rmd
update 2024-11-19 16:28:15.604972

@@
 There are indications that contemporary psychology falls short of this idealized model, however.
-Firstly, because deductive research is over-represented in the literature. 
-According to one estimate, 89.6% of published studies tests hypotheses,
-which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
-Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies reference theory at all [@mcphetresDecadeTheoryReflected2021].
-This invites the question where the other hypotheses come from,
-and to what extent testing an ad-hoc hypothesis not grounded in theory can advance our principled understanding of the studied phenomena.
+Firstly, because deductive research is over-represented in the literature: 
+According to one estimate, 89.6% of paper published in psychology tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
+Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+Only 15% of deductive studies referenced any theory, and theory was often not references in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
+This invites the question where hypotheses come from in the remaining 85% of deductive studies,
+and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
 These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
@@
 and test results rarely contribute to the improvement of existing theories.
-Consequently, theories either persist unchanged, immune to refutation,
-or are forgotten [REF Meehl].
+The paradoxical role of theory in psychology may help explain the observation that "theories either persist unchanged, immune to refutation,
+or are forgotten" [check quote REF Meehl].
 
@@
 A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
-or standardized metadata to signal the presence of theory within a work,
-nor even a standardized keyword (model, framework, etc are often used interchangeably with theory).
+nor standardized metadata or even a keyword to signal the presence of theory within a paper.
 Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
@@
 Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
-For example, it becomes possible to study theories (the same way libraries and search engines have enabled scholars to study the literature via systematic reviews);
-theories might have transdisciplinary impact;
-and the structure of theories from one discipline might inspire analogical modeling in other fields, the way - for example, predator-prey models have inspired theories of XXX and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+For example, it becomes possible to conduct meta-research on theories, 
+in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
+Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
+In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
 
@@
 and post many examples on their website [REF GRAY].
-Similarly, Borsboom and colleagues seek to establish a dictionary of psychological "phenomena" (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain) [REF BORSBOOM].
-Another method that could ameliorate theory Findability is modular publishing.
-In this approach, distinct assets are individually published as citable academic output.
-Such output can be linked; for example, a FAIR theory could have an associated paper.
+Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
+Post-hoc curation does not address the root cause of the lack of Findability, however.
+This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
+similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
+As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
 
-With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
+### Accessibility
+
+Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
+Publishing theories behind paywalls impedes this process.
 Open Access publishing increases the accessibility of all academic output, including theory.
-Another factor curtailing the accessibility of theory is ambiguity;
-a lack of theory formalization introduces a dependency on the original author for clarification.
-The discourse on "Great Man Theorizing" touches upon the problems this introduces [REF Guest et al moral theory].
-For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
+If theories were more accessible, researchers could reuse and refine them, thus accelerating cumulative knowledge acquisition.
+Making theories accessible also allows other stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena:
+More so than single studies, theories offer a top-down, big picture representation of the phenomena studied in a field.
+
+Beyond open access publishing, making theories accessible also requires communicating them clearly and explicitly.
+This relates to a property of good theories that has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
+But contemporary psychological theory is often difficult to understand due to its ambiguity [REF Frankenhuis].
+Ambiguity reduces the portability of theory by introducing a dependency on the original author for interpretation and clarification.
+This dependency can be an obstacle, for example, when a theory is insufficiently clear to derive hypotheses from it without first asking the original author what they meant - but it can also render theories immune to refutation.
+When confronted with falsifying evidence, the original author can always claim that the theory was misconstrued or misrepresented, thus making it a moving target.
+Dependency on the author relates to the discourse on "Great Man Theorizing" [REF Guest et al moral theory].
+For example, because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
 Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
-To some extent, this relates to the problem of translation [REF Duhem]:
-it is not possible to entirely formalize an idea to enable unambiguous interpretation.
-Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
+Dependency on the author also relates to the problem of translation [REF Duhem]:
+it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
+Despite this theoretical upper bound on formalization,
+efforts to formalize a theory will advance its Accessibility.
 
@@
 <!-- The data usually need to be integrated with other data. In addition, the data need to interoperate with applications or workflows for analysis, storage, and processing. -->
+
+### Interoperability
 
@@
 
-2. Accessibility: Promoting Open and Transparent Theoretical Development
-Accessibility involves ensuring that theories are openly available to all researchers, not locked behind paywalls or isolated in specific journals.
-By making theories FAIR, they would be accessible under clear, standardized, and open conditions, using platforms that allow both humans and machines to retrieve them effortlessly.
-
-How it helps: Openly accessible theories would allow for greater scrutiny and replication by the research community, fostering transparent scholarly communication.
-This ensures that researchers worldwide can contribute to, refine, and build upon existing theories, thus accelerating the pace of scientific discovery and enhancing the collective understanding of a given phenomenon.
 3. Interoperability: Facilitating Integration Across Fields

### 2024-11-19 manuscript.tex
update 2024-11-19 16:28:15.604972

@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 2657}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 3547}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-This is a preprint paper, generated from Git Commit \# d1a82fc7.
+This is a preprint paper, generated from Git Commit \# 4d935da1.
 
@@
 
-The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
-Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
-But have these reforms met their goal?
-Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
-One plausible explanation is that the replication crisis is symptomatic of an underlying,
-and more fundamental, ``theory crisis''.
-If psychological theories are insufficiently precise to derive testable hypotheses,
-or sufficiently vague to explain contradictory findings,
-then preregistration and replication will only serve to highlight those shortcomings.
+\subsection{The Need for FAIR Theory}\label{the-need-for-fair-theory}
+
+The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
+Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
+These various practices ensure transparent and repeated testing of hypotheses.
+However, recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
+Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+psychological theories rarely produce hypotheses that are found to be true,
+and are often sufficiently vague to explain contradictory findings.
 
 Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-One contributing factor is a lack of theory formalization:
-social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
-A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
-The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
+One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as \emph{formalization} (Szollosi \& Donkin, 2021).
+A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
+The present paper focuses on the second concern,
+advancing transparent communication about theory by applying open science principles to psychological theory for the first time
 and introducing the concept of \emph{FAIR Theory}.
-
-FAIR Theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers for knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
 
@@
 the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-There are clear indications that contemporary psychology falls short of this idealized model, however.
+There are indications that contemporary psychology falls short of this idealized model, however.
 Firstly, because deductive research is over-represented in the literature.
@@
 Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
-This raises the question where the other hypotheses come from,
-and what consequence their refutation would have for theory.
+This invites the question where the other hypotheses come from,
+and to what extent testing an ad-hoc hypothesis not grounded in theory can advance our principled understanding of the studied phenomena.
 These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
@@
 and test results rarely contribute to the improvement of existing theories.
-Consequently, theories either persist unchanged for decades, or are forgotten {[}REF Meehl{]}.
+Consequently, theories either persist unchanged, immune to refutation,
+or are forgotten {[}REF Meehl{]}.
 
-Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
-The present paper applies, for the first time, open science principles to theory.
+\subsection{The Solution}\label{the-solution}
+
+Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition for several reasons.
+
+\subsubsection{Findability}\label{findability}
+
+One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
+The primary unit of dissemination and search in psychology is often the academic paper.
+A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
+or standardized metadata to signal the presence of theory within a work,
+nor even a standardized keyword (model, framework, etc are often used interchangeably with theory).
+Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
+This would allow researchers to easily identify relevant theories to inform their hypotheses,
+grounding their work in established theoretical foundations.
+Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
+For example, it becomes possible to study theories (the same way libraries and search engines have enabled scholars to study the literature via systematic reviews);
+theories might have transdisciplinary impact;
+and the structure of theories from one discipline might inspire analogical modeling in other fields, the way - for example, predator-prey models have inspired theories of XXX and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+
+One effort that improves theories' findability is post-hoc curation.
+For example, Gray and colleagues introduced a format for representing theories,
+and post many examples on their website {[}REF GRAY{]}.
+Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain) {[}REF BORSBOOM{]}.
+Another method that could ameliorate theory Findability is modular publishing.
+In this approach, distinct resources (data, code, instruments, theory) are individually published as citable academic assets.
+Such output can be linked to a specific paper.
+
+\subsubsection{Accessibility}\label{accessibility}
+
+Advancing transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
+Publishing behind paywalls clearly creates obstacles.
+Open Access publishing increases the accessibility of all academic output, including theory.
+If theories were more accessible, researchers could reuse, refine, and relate their work to those theories, thus accelerating cumulative knowledge acquisition.
+Making theories more accessible also allows other stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena:
+More so than single studies, theories offer a top-down, big picture representation of the phenomena studied in a field.
+
+Beyond open access publishing, making theories accessible also requires communicating them clearly and explicitly.
+This property has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (\textbf{guestWhatMakesGood2024?}).
+Yet contemporary psychological theory is often ambiguous {[}REF Frankenhuis{]}.
+Ambiguity introduces a dependency on the original author for clarification,
+because the author can always claim that his ideas were misconstrued or misrepresented.
+This discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
+For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
+Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
+To some extent, this relates to the problem of translation {[}REF Duhem{]}:
+it is not possible to entirely formalize an idea to enable unambiguous interpretation.
+Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
+
+There are clear indications that psychological theory has limited interoperability.
+For one, theories are rarely \emph{``refuted nor corroborated, but instead merely fade away as people lose interest''} (Meehl, 1978).
+To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
+For example, it should be possible to integrate theory directly into analysis workflows:
+to derive hypotheses, select control variables, and guide model specification.
+The aforementioned lack of formalization {[}REF Robinaugh{]} and ambiguity {[}REF Frankenhuis{]} prevent interoperability in such a practical sense.
+Additionally, theories should be interoperable with each other:
+for instance, it should be possible to embed a specific theory about the process of emotion regulation {[}REF Gross{]} within a theory of emotion regulation development {[}REF Morris{]}.
+
+The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
+While theory is ostensibly held in high regard in the social sciences,
+there also appears to be a norm \emph{against} theory reuse.
+This is evident from the quip that
+\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
+In our workshops on FAIR Theory, we often hear questions such as ``who owns theory'',
+and ``who determines how a theory may be reused or changed''?
+Licensing theories for reuse provides an unambiguous answer to such questions.
+
+One important consideration is that there may be legal limits to the rights authors can claim over theory.
+An important concept in copyright law is the \emph{idea-expression dichotomy} {[}REF{]}.
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi}.}
+\setcounter{enumi}{14}
+\tightlist
+\item
+  Ideas, facts, and concepts are not protected by Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts.
+Separating an idea from the expression or manifestation of that idea is known in copyright law as the
+  Copyright explicitly does not
+  \emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''},
+  but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
+  Copyright pertaining to recipes offer an interesting perspective:
+  Whereas the recipes themselves are not protected, \emph{``substantial literary expression in the form of an explanation or directions''} can be protected by copyright.
+  Extrapolating to FAIR Theory, it would seem that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
+  If copyright does not cover (formal) theory,
+  then it would be pointless to adopt a license that assumes copyright protection.
+  Furtermore, even if copyright would apply, research is covered under ``fair use'' excemptions to copyright law, so copyright would impose few restrictions to theory reuse in scholarly communication.
+  In sum, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
+  it explicitly waives all rights and encourages reuse.
+  Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation {[}REF TOP guidelines{]},
+  and the spirit of good scholarship,
+  it is essential that scholars do comprehensively cite theory,
+  including prior work that new theories are based on.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Interoperability: Facilitating Integration Across Fields
+  In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks.
+The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.
+\end{enumerate}
+
+How it helps: If theories are made interoperable, they can be used in multi-disciplinary research and can better interface with diverse fields.
+This reduces duplication of effort and enables researchers from different areas to combine and compare theories, bridging gaps between disciplines and fostering innovative, cross-disciplinary approaches.
+Researchers can easily synthesize theoretical perspectives, generating new hypotheses that are more comprehensive and grounded in a wider range of evidence.
+4. Reusability: Enabling Continuous Improvement of Theories
+For knowledge to accumulate meaningfully, theories need to be adaptable and subject to continuous revision as new data emerge.
+The reusability principle calls for thorough documentation and community standards to ensure that data (or theories) can be applied in new contexts.
+
+How it helps: Reusable theories would provide clear metadata, version histories, and supporting evidence, allowing other researchers to test, refine, or extend them as new empirical findings emerge.
+This improves the feedback loop between theory and experimentation, ensuring that test results inform future theoretical development.
+It encourages the use of theories not just for their original purpose but in new and evolving contexts, thus ensuring that the theory remains dynamic and continuously refined over time.
+Conclusion: Making Theories FAIR Accelerates Scientific Progress
+Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
+By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
+This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
 
@@
 \end{lltable}
-
-\subsection{Is Current Psychological Theory FAIR?}\label{is-current-psychological-theory-fair}
-
-While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
-we provide a narrative review of examples and counterexamples for each criterion.
-
-One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
-A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
-or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
-Modular publishing could ameliorate the Findability of psychological theory.
-In this approach, distinct resources are individually published as citable academic output.
-Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
-Another effort to improve theories' findability is post-hoc curation.
-For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website.
-Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).
-
-With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
-Open Access publishing increases the accessibility of all academic output, including theory.
-Another factor curtailing the accessibility of theory is ambiguity;
-a lack of theory formalization introduces a dependency on the original author for clarification.
-The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
-For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
-Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
-To some extent, this relates to the problem of translation {[}REF Duhem{]}:
-it is not possible to entirely formalize an idea to enable unambiguous interpretation.
-Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
-
-There are clear indications that psychological theory has limited interoperability.
-For one, theories are rarely \emph{``refuted nor corroborated, but instead merely fade away as people lose interest''} (Meehl, 1978).
-To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
-For example, it should be possible to integrate theory directly into analysis workflows:
-to derive hypotheses, select control variables, and guide model specification.
-The aforementioned lack of formalization {[}REF Robinaugh{]} and ambiguity {[}REF Frankenhuis{]} prevent interoperability in such a practical sense.
-Additionally, theories should be interoperable with each other:
-for instance, it should be possible to embed a specific theory about the process of emotion regulation {[}REF Gross{]} within a theory of emotion regulation development {[}REF Morris{]}.
-
-There appears to be a norm \emph{against} the reuse of theory,
-as evident from the quip that
-\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
-Moreover, a legal basis for theory reuse is often absent.
-Questions that often come up in our workshops on FAIR Theory are ``who owns theory'',
-and ``who determines how a theory may be changed''?
-Such questions require a clear answer, even if that answer may vary across theories.
-Licensing theories for reuse provides clarity about how a theory can be reused.
-A crucial legal consideration is that,
-while specific (FAIR) representations of theory might be protected by copyright,
-the underlying ideas are not.
-This is another argument in favor of publishing specific instantiations of theory along with licenses detailing how they may be reused.
 
@@
 
+\section{Discussion}\label{discussion}
+
+\subsection{Future Directions}\label{future-directions}
+
+One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
+Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
+it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties {[}REF{]}.
+Additionally, the ``jingle-jangle'' fallacy is prevalent in the social sciences:
+the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
+FAIR Theory can help address the measurement crisis:
+since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
+
+FAIR Theory incorporates theory into open science workflows,
+facilitates scholarly communication about theories,
+making it easier to share theories with less opportunity for ambiguity and misunderstanding.
+FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
+More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
+removes barriers for knowledge exchange with the global scholarly community,
+opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
+
 \newpage

### 2024-11-20 manuscript.Rmd
update 2024-11-20 11:57:25.770558

@@
 
+Interoperability pertains to the property of digital research artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
+As the original definition of interoperability was narrow and somewhat abstract,
+the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
+
 There are clear indications that psychological theory has limited interoperability. 
 For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
+
 To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
+
 For example, it should be possible to integrate theory directly into analysis workflows:
@@
 for instance, it should be possible to embed a specific theory about the process of emotion regulation [REF Gross] within a theory of emotion regulation development [REF Morris].
+
+### Reusability
 

### 2024-11-20 manuscript.Rmd
update 2024-11-20 17:52:03.486257

--- 2024-11-19 manuscript.Rmd
@@
 
-There are clear indications that psychological theory has limited interoperability. 
-For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
-To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
-For example, it should be possible to integrate theory directly into analysis workflows:
-to derive hypotheses, select control variables, and guide model specification.
-The aforementioned lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense.
+Interoperability pertains to the property of digital research artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
+As the original definition of interoperability was somewhat narrow and abstract,
+the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
+This definition raises an interesting question, namely *interoperable for what?*
+Suitable answers for FAIR theory may be: this theory is X-interoperable for deriving testable null-hypotheses, or for the purpose of selecting relevant control variables,
+or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
+
+These definitions suggest that FAIR theory should, ad minimum,
+be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].
+Furthermore, it is important to acknowledge that interoperability is subject to the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970],
+which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
+
+Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
+The Kuhnian notion of "normal science" conducted within the context of a shared paradigm may provide sufficient shared background knowledge for mutual understanding [@kuhnStructureScientificRevolutions2009].
+From a pragmatic perspective, these considerations indicate that,
+when striving to make theory interoperable,
+it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions),
+while acknowledging that interoperability exists on a spectrum,
+and that it is impossible to eliminate all ambiguity.
+Rather, it may benefit scientific discourse to anticipate misunderstanding,
+and use it to drive further explication of theory.
+
+<!-- Do we want to say something about this?
+Interoperability of psychological theory may be limited when theories are instantiated in natural language without consideration of ontological complexities. -->
+<!-- or as visualizations without a straightforward interpretation -->
+<!-- What can go wrong?
+A well-formed sentence may be logically inconsistent (e.g., the theory that conceptualized father involvement simultaneously as mediator and moderator); a picture may seem sensible but there's no straightforward way to translate it to hypotheses, etc.
+-->
+<!-- which are not machine-readable and l -->
+
+According to Kurt Lewin, "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] - but as we argued bfore,
+there are clear signs that theory rarely plays a practical role in the work of contemporary psychologists.
+Interoperability makes it possible to integrate theory into scientists' day-to-day work in a practical way.
+If a theory is X-interoperable with software used for data analysis,
+it becomes possible to integrate theory directly into analysis workflows.
+For example, interoperable theory could be used to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021],
+to select control variables for causal inference [@cinelliCrashCourseGood2022],
+and to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021]
 Additionally, theories should be interoperable with each other:
-for instance, it should be possible to embed a specific theory about the process of emotion regulation [REF Gross] within a theory of emotion regulation development [REF Morris].
+for instance, it should be possible to embed a specific theory about the process of emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
+
+### Reusability
 
@@
 While theory is ostensibly held in high regard in the social sciences,
-there also appears to be a norm *against* theory reuse.
-This is evident from the quip that 
+some have described a norm *against* theory reuse:
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
+Given that theory is the vehicle of cumulative knowledge acquisition [@degrootMethodologieGrondslagenVan1961],
+such a norm impedes scientific progress.
+Even if scholars wish to diverge substantially from prior theory,
+explicitly refering back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
+Reusing and adapting existing theories makes it possible to study the provenance of a theory and investigate how well different iterations of the theory explain emprical evidence [@vanlissaUsingEndpointsCheck2023]
+
 In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
@@
 Licensing theories for reuse provides an unambiguous answer to such questions.
+These questions suggest that there is a norm against modifying theory without consent from the author.
+However, scholars might not be aware of the legal limits to the rights authors can claim over theory.
+An important concept in copyright law is the *idea-expression dichotomy* [@bently2010copyright].
 
-One important consideration is that there may be legal limits to the rights authors can claim over theory.
-An important concept in copyright law is the *idea-expression dichotomy* [REF].
-
-o.
+<!-- o.
 Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts.
-Separating an idea from the expression or manifestation of that idea is known in copyright law as the
+Separating an idea from the expression or manifestation of that idea is known in copyright law as the  -->
 Copyright explicitly does not
@@
 Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright.
-Extrapolating to FAIR Theory, it would seem that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
+Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
 If copyright does not cover (formal) theory,
 then it would be pointless to adopt a license that assumes copyright protection.
-Furtermore, even if copyright would apply, research is covered under "fair use" excemptions to copyright law, so copyright would impose few restrictions to theory reuse in scholarly communication.
-In sum, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
+Furtermore, even if copyright would apply, academic research is covered under "fair use" exemptions to copyright law,
+so copyright would impose few restrictions to theory reuse in scholarly communication.
+Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
 it explicitly waives all rights and encourages reuse.
-Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [REF TOP guidelines],
+Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [@aalbersbergMakingScienceTransparent2018],
 and the spirit of good scholarship,
 it is essential that scholars do comprehensively cite theory,
-including prior work that new theories are based on.
+including prior work that new theories are based on,
+even in absence of legal obligations to do so.
 

### 2024-11-21 manuscript.Rmd
update 2024-11-21 19:03:07.450896

@@
 Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-Publishing theories behind paywalls impedes this process.
+If theories are accessible, researchers can reuse and refine them,
+thus accelerating cumulative knowledge acquisition.
+Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+While isolated empirical findings can appear fragmented and contradictory [@dumas-malletPoorReplicationValidity2017],
+theories offer a top-down, big picture representation of the phenomena studied in a field.
+
+At present, there are several impediments to theories' accessibility.
+The first are paywalls erected by commercial publishers.
 Open Access publishing increases the accessibility of all academic output, including theory.
-If theories were more accessible, researchers could reuse and refine them, thus accelerating cumulative knowledge acquisition.
-Making theories accessible also allows other stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena:
-More so than single studies, theories offer a top-down, big picture representation of the phenomena studied in a field.
 
-Beyond open access publishing, making theories accessible also requires communicating them clearly and explicitly.
-This relates to a property of good theories that has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
-But contemporary psychological theory is often difficult to understand due to its ambiguity [REF Frankenhuis].
-Ambiguity reduces the portability of theory by introducing a dependency on the original author for interpretation and clarification.
-This dependency can be an obstacle, for example, when a theory is insufficiently clear to derive hypotheses from it without first asking the original author what they meant - but it can also render theories immune to refutation.
-When confronted with falsifying evidence, the original author can always claim that the theory was misconstrued or misrepresented, thus making it a moving target.
-Dependency on the author relates to the discourse on "Great Man Theorizing" [REF Guest et al moral theory].
-For example, because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
-Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
-Dependency on the author also relates to the problem of translation [REF Duhem]:
-it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
-Despite this theoretical upper bound on formalization,
-efforts to formalize a theory will advance its Accessibility.
+A second impediment is therefore the prevalence of (strategic) ambiguity,
+which renders theory difficult to understand [REF Frankenhuis].
+While open access publishing increases the availability of theories,
+accessibility further requires clear and explicit communication.
+This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
+It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
+which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
+Thus, it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
+This places a theoretical upper bound on theories' ability to be understood.
 
- <!-- (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.
--->
+Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
+The Kuhnian notion of "normal science", conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding [@kuhnStructureScientificRevolutions2009].
+From a pragmatic perspective, these considerations indicate that,
+when striving to make theory accessible,
+it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions),
+while acknowledging that accessibility exists on a spectrum,
+and that it is impossible to eliminate all ambiguity.
+Rather, it may benefit scientific discourse to anticipate misunderstanding,
+and use it to drive further explication of theory.
+In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.
+
+A third, and related, impediment is a phenomenon we will call "dependency on the author" (DOA).
+DOA occurs whenever a theory cannot be understood by independent scholars,
+thus requiring the original author for interpretation and clarification.
+We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
+as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
+DOA is an obstacle when theory is insufficiently clear to derive hypotheses from it without first asking the author what they meant.
+It also renders theories immune to refutation,
+because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target.
+DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate, which violates checks and balances of scientific research.
+
+DOA is also a problem because authors may not understand their own theories well.
+illustrated by instances where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
 
@@
 be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].
-Furthermore, it is important to acknowledge that interoperability is subject to the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970],
-which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
-
-Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
-The Kuhnian notion of "normal science" conducted within the context of a shared paradigm may provide sufficient shared background knowledge for mutual understanding [@kuhnStructureScientificRevolutions2009].
-From a pragmatic perspective, these considerations indicate that,
-when striving to make theory interoperable,
-it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions),
-while acknowledging that interoperability exists on a spectrum,
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
 
@@
 The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
+We establish guiding principles for instantiating theory as a FAIR digital research artifact,
+and provide applied examples to encourage their adoption.
 

### 2024-12-04 manuscript.Rmd
update 2024-12-04 17:35:40.415285

@@
 bibliography      : "theory-specification.bib"
-floatsintext      : no
+floatsintext      : yes
 linenumbers       : yes
@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
-Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
+Since their inception, scholars have demonstrated their relevance for making other information artefacts more open.
+This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
-a digital representation of theory compliant with the FAIR principles.
-FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.
+a digital representation of theory, compliant with the FAIR principles.
+By improving the efficiency of scholarly communication, FAIR Theory has the potential to accelerate cumulative knowledge acquisition and advance social scientific research.
 
@@
 Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as *formalization* [@szollosiArrestedTheoryDevelopment2021].
-A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
-The present paper focuses on the second concern,
-advancing transparent communication about theory by applying open science principles to psychological theory for the first time
-and introducing the concept of *FAIR Theory*.
+Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
+In other words, social scientific theories lack *formalization* .
+A second concern is the lack of transparent and democratic scholarly communication about psychological theory.
+
+Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+The equally critical inductive processes of theory construction and -improvement have been largely overlooked.
+The present paper restores balance by applying, for the first time,
+open science principles to psychological theory.
+We apply the FAIR principles to scientific theories,
+introducing the concept of *FAIR Theory* to 
+facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
-the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-There are indications that contemporary psychology falls short of this idealized model, however.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+There are, however, indications that contemporary psychology falls short of this ideal.
 Firstly, because deductive research is over-represented in the literature: 
@@
 Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies referenced any theory, and theory was often not references in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
-This invites the question where hypotheses come from in the remaining 85% of deductive studies,
-and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
-These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
+Only 15% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
+The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
+Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit [REF Rasoul].
+Or, perhaps the hypotheses are of no substantive interest [REF Hoijtink], and researchers are simply testing them as part of a cultural ritual [REF Gigerenzer].
+No matter what the explanation - testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+
+According to the aforementioned statistics, theory has an uncomfortable and paradoxical role in contemporary psychology:
 The majority of papers ostensibly test hypotheses,
 but these are rarely derived from theory,
-and test results rarely contribute to the improvement of existing theories.
-The paradoxical role of theory in psychology may help explain the observation that "theories either persist unchanged, immune to refutation,
-or are forgotten" [check quote REF Meehl].
+and test results do not routinely contribute to the improvement of existing theories.
+The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
 
-<!-- Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory. -->
-<!-- The present paper applies, for the first time, open science principles to theory. -->
-
-## The Solution
-
-Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition for several reasons.
+To date, scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
+The present paper applies, for the first time, open science principles to theory.
+Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition.
 
@@
 One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
-The primary unit of dissemination and search in psychology is often the academic paper.
-A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
-nor standardized metadata or even a keyword to signal the presence of theory within a paper.
-Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
+One contributing cause of theories' lack of Findability is the lack of standardized metadata or even a keyword to signal the presence of theory within a paper, and - consequently - the absence of a unified search engine for theory.
+Making theories more findable through rich metadata would allow them to be systematically indexed, tagged, and made searchable.
 This would allow researchers to easily identify relevant theories to inform their hypotheses,
@@
 
-One effort that improves theories' findability is post-hoc curation.
+Another contributing cause to the lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
+A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
+This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
+similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
+As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
+
+There have been notable efforts to improve theories' findability through post-hoc curation.
 For example, Gray and colleagues introduced a format for representing theories,
@@
 Post-hoc curation does not address the root cause of the lack of Findability, however.
-This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
-similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
-As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
+Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
 
@@
 
-A third, and related, impediment is a phenomenon we will call "dependency on the author" (DOA).
-DOA occurs whenever a theory cannot be understood by independent scholars,
+A third impediment arises when theories have a "dependency on the author" (DOA).
+DOA occurs when a theory cannot be understood by independent scholars,
 thus requiring the original author for interpretation and clarification.
@@
 as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
-DOA is an obstacle when theory is insufficiently clear to derive hypotheses from it without first asking the author what they meant.
-It also renders theories immune to refutation,
-because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target.
-DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate, which violates checks and balances of scientific research.
-
-DOA is also a problem because authors may not understand their own theories well.
-illustrated by instances where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
+DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
+which violates checks and balances of scientific research.
+DOA also renders theories immune to refutation,
+because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
+DOA is inherently problematic because authors may not understand their own theories well.
+This is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
+In short, DOA unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
 
@@
 
-Interoperability pertains to the property of digital research artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
+Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
 As the original definition of interoperability was somewhat narrow and abstract,
 the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
-This definition raises an interesting question, namely *interoperable for what?*
-Suitable answers for FAIR theory may be: this theory is X-interoperable for deriving testable null-hypotheses, or for the purpose of selecting relevant control variables,
+This definition invites the question: *interoperable for what purpose?*
+Suitable answers for FAIR theory may be:
+this theory is X-interoperable for deriving testable null-hypotheses,
+or for the purpose of selecting relevant control variables,
 or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-
 These definitions suggest that FAIR theory should, ad minimum,
 be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].
+Furthermore, theories may have specific properties that incur affordances in terms of X-interoperability;
+for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories.
+
+```{r}
+data.frame(
+  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
+  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
+) |>
+papaja::apa_table()
+```
 
@@
 
-According to Kurt Lewin, "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] - but as we argued bfore,
-there are clear signs that theory rarely plays a practical role in the work of contemporary psychologists.
-Interoperability makes it possible to integrate theory into scientists' day-to-day work in a practical way.
-If a theory is X-interoperable with software used for data analysis,
-it becomes possible to integrate theory directly into analysis workflows.
-For example, interoperable theory could be used to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021],
+Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] seems to imply that theory plays a practical role in the day-to-day work of psychological researchers.
+But, as we argued before, there are clear signs that this is not the case.
+The examples of X-interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
+For example, interoperable theory could be used
 to select control variables for causal inference [@cinelliCrashCourseGood2022],
-and to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021]
-Additionally, theories should be interoperable with each other:
-for instance, it should be possible to embed a specific theory about the process of emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
+or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
+or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
+Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
+For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
 
@@
 
-The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
-While theory is ostensibly held in high regard in the social sciences,
-some have described a norm *against* theory reuse:
+Some have argued that there is a norm *against* theory reuse:
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
-Given that theory is the vehicle of cumulative knowledge acquisition [@degrootMethodologieGrondslagenVan1961],
-such a norm impedes scientific progress.
-Even if scholars wish to diverge substantially from prior theory,
-explicitly refering back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
-Reusing and adapting existing theories makes it possible to study the provenance of a theory and investigate how well different iterations of the theory explain emprical evidence [@vanlissaUsingEndpointsCheck2023]
+Such a norm impedes scientific progress.
+Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
+Theories are made reusable by thorough documentation,
+appropriate licensing, 
+and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning).
 
@@
 Licensing theories for reuse provides an unambiguous answer to such questions.
-These questions suggest that there is a norm against modifying theory without consent from the author.
-However, scholars might not be aware of the legal limits to the rights authors can claim over theory.
-An important concept in copyright law is the *idea-expression dichotomy* [@bently2010copyright].
-
+These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
+However, copyright law limits the rights authors can claim over theory based on the *idea-expression dichotomy* [@bently2010copyright],
 <!-- o.
@@
 Separating an idea from the expression or manifestation of that idea is known in copyright law as the  -->
-Copyright explicitly does not
-*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*,
-but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
-Copyright pertaining to recipes offer an interesting perspective:
-Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright.
-Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
-If copyright does not cover (formal) theory,
-then it would be pointless to adopt a license that assumes copyright protection.
-Furtermore, even if copyright would apply, academic research is covered under "fair use" exemptions to copyright law,
-so copyright would impose few restrictions to theory reuse in scholarly communication.
+which holds that copyright explicitly does not
+*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
+It may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
+<!-- Copyright pertaining to recipes offer an interesting perspective: -->
+<!-- Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright. -->
+Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short in terms of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+If copyright limits Reusability and does not cover ideas in their purest form (like formal theory),
+then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
+Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
+so copyright would pose few restrictions to Reusability in scholarly communication.
 Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
 it explicitly waives all rights and encourages reuse.
-Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [@aalbersbergMakingScienceTransparent2018],
-and the spirit of good scholarship,
-it is essential that scholars do comprehensively cite theory,
+In principle, CC0 does not require attribution.
+Nevertheless, is essential that scholars do comprehensively cite theory,
 including prior work that new theories are based on,
-even in absence of legal obligations to do so.
+even in absence of legal obligations to do so,
+to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].
 
-3. Interoperability: Facilitating Integration Across Fields
-In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks.
-The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code.
+Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
+Git also facilitates diffuse and adversarial collaboration,
+as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
+In sum, version control using Git enables systematic, collaborative, and transparent theory development,
+enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
 
-How it helps: If theories are made interoperable, they can be used in multi-disciplinary research and can better interface with diverse fields.
-This reduces duplication of effort and enables researchers from different areas to combine and compare theories, bridging gaps between disciplines and fostering innovative, cross-disciplinary approaches.
-Researchers can easily synthesize theoretical perspectives, generating new hypotheses that are more comprehensive and grounded in a wider range of evidence.
-4. Reusability: Enabling Continuous Improvement of Theories
-For knowledge to accumulate meaningfully, theories need to be adaptable and subject to continuous revision as new data emerge.
-The reusability principle calls for thorough documentation and community standards to ensure that data (or theories) can be applied in new contexts.
+Even if scholars wish to diverge substantially from prior theory,
+explicitly referring back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
 
-How it helps: Reusable theories would provide clear metadata, version histories, and supporting evidence, allowing other researchers to test, refine, or extend them as new empirical findings emerge.
-This improves the feedback loop between theory and experimentation, ensuring that test results inform future theoretical development.
-It encourages the use of theories not just for their original purpose but in new and evolving contexts, thus ensuring that the theory remains dynamic and continuously refined over time.
-Conclusion: Making Theories FAIR Accelerates Scientific Progress
+From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
+
+Making Theories FAIR Accelerates Scientific Progress
 Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
@@
 From their inception, these principles were developed with "other research resources" in mind.
-Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
-The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
+Scholars have adapted the FAIR principles to, e.g., research software [REF Lamprecht].
+Yet there are key distinctions between theory and other FAIR information artefacts.
+With this in mind, the present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
 We establish guiding principles for instantiating theory as a FAIR digital research artifact,
@@
 
-In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
-From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
-Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.
-
 FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
-
-There are key distinctions between theory and other FAIR digital research artefacts.
-With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.
+In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging.
 
@@
 While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.
-
-## Version Control
-
-* One source of potential improvements of theory methodology that has not been previously considered is computer science.
-* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
-* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
-* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
-* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
-* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
-Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.
 

### 2024-12-04 manuscript.tex
update 2024-12-04 17:35:40.415285

--- 2024-09-16 manuscript.tex
@@
 \documentclass[
-  man]{apa6}
+  man,floatsintext]{apa6}
 \usepackage{amsmath,amssymb}
@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 2657}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 3877}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# d1a82fc7.
+This is a preprint paper, generated from Git Commit \# 4a12ff5b.
 
@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
-Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
+Since their inception, scholars have demonstrated their relevance for making other information artefacts more open.
+This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
 To this end, we introduce ``FAIR theory'':
-a digital representation of theory compliant with the FAIR principles.
-FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.
+a digital representation of theory, compliant with the FAIR principles.
+By improving the efficiency of scholarly communication, FAIR Theory has the potential to accelerate cumulative knowledge acquisition and advance social scientific research.
 
-The ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
-Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
-But have these reforms met their goal?
-Recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
-One plausible explanation is that the replication crisis is symptomatic of an underlying,
-and more fundamental, ``theory crisis''.
-If psychological theories are insufficiently precise to derive testable hypotheses,
-or sufficiently vague to explain contradictory findings,
-then preregistration and replication will only serve to highlight those shortcomings.
+\subsection{The Need for FAIR Theory}\label{the-need-for-fair-theory}
+
+The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
+Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
+These various practices ensure transparent and repeated testing of hypotheses.
+However, recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
+Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+psychological theories rarely produce hypotheses that are found to be true,
+and are often sufficiently vague to explain contradictory findings.
 
 Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-One contributing factor is a lack of theory formalization:
-social scientific theories often lack the precision and clarity of theories in the physical sciences (Szollosi \& Donkin, 2021).
-A second factor, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
-The present paper seeks to advance transparent communication about theory by applying open science principles to psychological theory for the first time,
+One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as \emph{formalization} (Szollosi \& Donkin, 2021).
+A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
+The present paper focuses on the second concern,
+advancing transparent communication about theory by applying open science principles to psychological theory for the first time
 and introducing the concept of \emph{FAIR Theory}.
-
-FAIR Theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers for knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
 
@@
 In a progressive research program (Lakatos, 1971),
-the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-There are clear indications that contemporary psychology falls short of this idealized model, however.
-Firstly, because deductive research is over-represented in the literature.
-According to one estimate, 89.6\% of published studies tests hypotheses,
-which suggests that the literature is predominantly comprised of deductive research (Kühberger, Fritz, \& Scherndl, 2014).
-Closer examination reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
-Only 15\% of deductive studies reference theory at all (McPhetres et al., 2021).
-This raises the question where the other hypotheses come from,
-and what consequence their refutation would have for theory.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+There are, however, indications that contemporary psychology falls short of this ideal.
+Firstly, because deductive research is over-represented in the literature:
+According to one estimate, 89.6\% of paper published in psychology tests hypotheses (Kühberger, Fritz, \& Scherndl, 2014).
+Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
+Only 15\% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis (McPhetres et al., 2021).
+This invites the question where hypotheses come from in the remaining 85\% of deductive studies,
+and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
 These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
@@
 and test results rarely contribute to the improvement of existing theories.
-Consequently, theories either persist unchanged for decades, or are forgotten {[}REF Meehl{]}.
+The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
 
-Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
+To date, scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
 The present paper applies, for the first time, open science principles to theory.
+Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition.
+
+\subsubsection{Findability}\label{findability}
+
+One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
+One contributing cause of theories' lack of Findability is the lack of standardized metadata or even a keyword to signal the presence of theory within a paper, and - consequently - the absence of a unified search engine for theory.
+Making theories more findable through rich metadata would allow them to be systematically indexed, tagged, and made searchable.
+This would allow researchers to easily identify relevant theories to inform their hypotheses,
+grounding their work in established theoretical foundations.
+Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
+For example, it becomes possible to conduct meta-research on theories,
+in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
+Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
+In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+
+Another contributing cause to the lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
+A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
+This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
+similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
+As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
+
+There have been notable efforts to improve theories' findability through post-hoc curation.
+For example, Gray and colleagues introduced a format for representing theories,
+and post many examples on their website {[}REF GRAY{]}.
+Similarly, Borsboom and colleagues seek to establish a database of psychological theories {[}REF BORSBOOM{]}.
+Post-hoc curation does not address the root cause of the lack of Findability, however.
+Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
+
+\subsubsection{Accessibility}\label{accessibility}
+
+Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
+If theories are accessible, researchers can reuse and refine them,
+thus accelerating cumulative knowledge acquisition.
+Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
+theories offer a top-down, big picture representation of the phenomena studied in a field.
+
+At present, there are several impediments to theories' accessibility.
+The first are paywalls erected by commercial publishers.
+Open Access publishing increases the accessibility of all academic output, including theory.
+
+A second impediment is therefore the prevalence of (strategic) ambiguity,
+which renders theory difficult to understand {[}REF Frankenhuis{]}.
+While open access publishing increases the availability of theories,
+accessibility further requires clear and explicit communication.
+This property of good theories has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
+It is important to acknowledge the \emph{indeterminacy of translation} (Quine, 1970):
+which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
+Thus, it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
+This places a theoretical upper bound on theories' ability to be understood.
+
+Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
+The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
+From a pragmatic perspective, these considerations indicate that,
+when striving to make theory accessible,
+it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
+while acknowledging that accessibility exists on a spectrum,
+and that it is impossible to eliminate all ambiguity.
+Rather, it may benefit scientific discourse to anticipate misunderstanding,
+and use it to drive further explication of theory.
+In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.
+
+A third impediment arises when theories have a ``dependency on the author'' (DOA).
+DOA occurs when a theory cannot be understood by independent scholars,
+thus requiring the original author for interpretation and clarification.
+We have heard DOA referred to apocryphally as the ``ask Leon'' phenomenon,
+as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
+DOA relates to the discourse on ``Great Man Theorizing'' (Guest, 2024) because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
+which violates checks and balances of scientific research.
+DOA also renders theories immune to refutation,
+because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target (Szollosi \& Donkin, 2021).
+DOA is inherently problematic because authors may not understand their own theories well.
+This is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
+In short, DOA unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
+
+\subsubsection{Interoperability}\label{interoperability}
+
+Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (Wilkinson et al., 2016).
+As the original definition of interoperability was somewhat narrow and abstract,
+the concept has recently been further refined in terms of facilitating ``successful communication between machines and between humans and machines'', where ``A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
+This definition invites the question: \emph{interoperable for what purpose?}
+Suitable answers for FAIR theory may be:
+this theory is X-interoperable for deriving testable null-hypotheses,
+or for the purpose of selecting relevant control variables,
+or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
+These definitions suggest that FAIR theory should, ad minimum,
+be instantiated in a human- and machine-readable file format, as previously recommended (Van Lissa et al., 2021).
+Furthermore, theories may have specific properties that incur affordances in terms of X-interoperability;
+for example, Table \ref{tab:tabmeehl} illustrates the affordances of Meehl's nine properties of strong theories.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:unnamed-chunk-2}}
+
+\begin{tabular}{ll}
+\toprule
+Property & \multicolumn{1}{c}{X-interoperability}\\
+\midrule
+1) Ontology & Variable selection\\
+2) Causal connections & Model specification, covariate selection, causal inference\\
+3-8) Functional Form & Deriving specific hypotheses\\
+9) Numerical Value & Simulating data\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) seems to imply that theory plays a practical role in the day-to-day work of psychological researchers.
+But, as we argued before, there are clear signs that this is not the case.
+The examples of X-interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
+For example, interoperable theory could be used
+to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
+or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert, Van Lissa, \& Brandmaier, 2021),
+or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
+Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
+For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris, Silk, Steinberg, Myers, \& Robinson, 2007).
+
+\subsubsection{Reusability}\label{reusability}
+
+Some have argued that there is a norm \emph{against} theory reuse:
+\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
+Such a norm impedes scientific progress.
+Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data (de Groot, 1961).
+Theories are made reusable by thorough documentation,
+appropriate licensing,
+and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning).
+
+In our workshops on FAIR Theory, we often hear questions such as ``who owns theory'',
+and ``who determines how a theory may be reused or changed''?
+Licensing theories for reuse provides an unambiguous answer to such questions.
+These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
+However, copyright law limits the rights authors can claim over theory based on the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010),
+which holds that copyright explicitly does not
+\emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''}.
+It may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
+Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short in terms of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+If copyright limits Reusability and does not cover ideas in their purest form (like formal theory),
+then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
+Furthermore, even if copyright would apply, academic research is covered under ``fair use'' exemptions,
+so copyright would pose few restrictions to Reusability in scholarly communication.
+Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
+it explicitly waives all rights and encourages reuse.
+In principle, CC0 does not require attribution.
+Nevertheless, is essential that scholars do comprehensively cite theory,
+including prior work that new theories are based on,
+even in absence of legal obligations to do so,
+to meet the definition of Reusability (R1.2, Table \ref{tab:tabfair} and to comply with other definitions of open scholarship (Aalbersberg et al., 2018).
+
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts, like computer code.
+Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
+Git also facilitates diffuse and adversarial collaboration,
+as independent researchers can create independent versions of existing theories through ``forking'', or suggest modifications to existing theories via ``pull requests''.
+In sum, version control using Git enables systematic, collaborative, and transparent theory development,
+enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence (Van Lissa, 2023).
+
+Even if scholars wish to diverge substantially from prior theory,
+explicitly referring back to it enables clear comparison of the differences (Ram, 2013).
+
+From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
+
+Making Theories FAIR Accelerates Scientific Progress
+Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
+By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
+This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
 
@@
 From their inception, these principles were developed with ``other research resources'' in mind.
-Scholars have translated the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
-The present paper further extends the FAIR principles' definition to theory, see Table \ref{tab:tabfair}.
+Scholars have adapted the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
+Yet there are key distinctions between theory and other FAIR information artefacts.
+With this in mind, the present paper further extends the FAIR principles' definition to theory, see Table \ref{tab:tabfair}.
+We establish guiding principles for instantiating theory as a FAIR digital research artifact,
+and provide applied examples to encourage their adoption.
 
@@
 
-\subsection{Is Current Psychological Theory FAIR?}\label{is-current-psychological-theory-fair}
-
-While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
-we provide a narrative review of examples and counterexamples for each criterion.
-
-One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
-A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
-or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
-Modular publishing could ameliorate the Findability of psychological theory.
-In this approach, distinct resources are individually published as citable academic output.
-Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
-Another effort to improve theories' findability is post-hoc curation.
-For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website.
-Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).
-
-With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
-Open Access publishing increases the accessibility of all academic output, including theory.
-Another factor curtailing the accessibility of theory is ambiguity;
-a lack of theory formalization introduces a dependency on the original author for clarification.
-The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
-For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
-Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
-To some extent, this relates to the problem of translation {[}REF Duhem{]}:
-it is not possible to entirely formalize an idea to enable unambiguous interpretation.
-Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.
-
-There are clear indications that psychological theory has limited interoperability.
-For one, theories are rarely \emph{``refuted nor corroborated, but instead merely fade away as people lose interest''} (Meehl, 1978).
-To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
-For example, it should be possible to integrate theory directly into analysis workflows:
-to derive hypotheses, select control variables, and guide model specification.
-The aforementioned lack of formalization {[}REF Robinaugh{]} and ambiguity {[}REF Frankenhuis{]} prevent interoperability in such a practical sense.
-Additionally, theories should be interoperable with each other:
-for instance, it should be possible to embed a specific theory about the process of emotion regulation {[}REF Gross{]} within a theory of emotion regulation development {[}REF Morris{]}.
-
-There appears to be a norm \emph{against} the reuse of theory,
-as evident from the quip that
-\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
-Moreover, a legal basis for theory reuse is often absent.
-Questions that often come up in our workshops on FAIR Theory are ``who owns theory'',
-and ``who determines how a theory may be changed''?
-Such questions require a clear answer, even if that answer may vary across theories.
-Licensing theories for reuse provides clarity about how a theory can be reused.
-A crucial legal consideration is that,
-while specific (FAIR) representations of theory might be protected by copyright,
-the underlying ideas are not.
-This is another argument in favor of publishing specific instantiations of theory along with licenses detailing how they may be reused.
-
 \subsection{FAIR Theory and Recognition \& Rewards}\label{fair-theory-and-recognition-rewards}
 
-In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
-From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
-Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.
-
 FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
-
-There are key distinctions between theory and other FAIR digital research artefacts.
-With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.
+In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging.
 
@@
 While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.
-
-\subsection{Version Control}\label{version-control}
-
-\begin{itemize}
-\tightlist
-\item
-  One source of potential improvements of theory methodology that has not been previously considered is computer science.
-\item
-  The process of ``iteratively improving'' digital objects - in this case, computer code - is well understood.
-\item
-  Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
-\item
-  This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
-\item
-  We introduce ``FAIR theory'', a digital research artifact to represent formal social scientific theories
-\item
-  FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
-Version control also enables diffuse collaboration in theory development, as other researchers can submit ``pull requests'' to suggest modifications of a theory, or can ``fork'' existing theories to create a spin-off from an existing theory.
-\end{itemize}
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-aalbersbergMakingScienceTransparent2018}
+Aalbersberg, Ij.
+J., Appleyard, T., Brookhart, S., Carpenter, T., Clarke, M., Curry, S., \ldots{} Vazire, S.
+(2018).
+\emph{Making {Science Transparent By Default}; {Introducing} the {TOP Statement}}.
+\url{https://doi.org/10.31219/osf.io/sm78t}
+
+\bibitem[\citeproctext]{ref-bently2010copyright}
+Bently, L., Davis, J., \& Ginsburg, J. C. (2010). \emph{Copyright and {Piracy}: {An} interdisciplinary critique} (Vol. 13). Cambridge University Press.
+
+\bibitem[\citeproctext]{ref-cinelliCrashCourseGood2022}
+Cinelli, C., Forney, A., \& Pearl, J. (2022). A {Crash Course} in {Good} and {Bad Controls}. \emph{Sociological Methods \& Research}, 00491241221099552. \url{https://doi.org/10.1177/00491241221099552}
+
 \bibitem[\citeproctext]{ref-degrootMethodologieGrondslagenVan1961}
@@
 
+\bibitem[\citeproctext]{ref-dumas-malletPoorReplicationValidity2017}
+Dumas-Mallet, E., Smith, A., Boraud, T., \& Gonon, F.
+(2017).
+Poor replication validity of biomedical association studies reported by newspapers.
+\emph{PLOS ONE}, \emph{12}(2), e0172650.
+\url{https://doi.org/10.1371/journal.pone.0172650}
+
+\bibitem[\citeproctext]{ref-grossEmotionRegulationCurrent2015}
+Gross, J. J. (2015). Emotion regulation: {Current} status and future prospects. \emph{Psychological Inquiry}, \emph{26}(1), 1--26. \url{https://doi.org/10.1080/1047840X.2014.940781}
+
+\bibitem[\citeproctext]{ref-guestWhatMakesGood2024}
+Guest, O. (2024). What {Makes} a {Good Theory}, and {How Do We Make} a {Theory Good}? \emph{Computational Brain \& Behavior}. \url{https://doi.org/10.1007/s42113-023-00193-2}
+
+\bibitem[\citeproctext]{ref-kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008}
+Kissner, J.
+(2008).
+{ON THE IDENTIFICATION OF A LOGICAL INCONSISTENCY IN THE GENERAL THEORY OF CRIME}.
+\emph{Journal of Crime and Justice}.
+Retrieved from \url{https://www.tandfonline.com/doi/abs/10.1080/0735648X.2008.9721251}
+
 \bibitem[\citeproctext]{ref-kuhbergerPublicationBiasPsychology2014}
@@
 \url{https://doi.org/10.1371/journal.pone.0105825}
+
+\bibitem[\citeproctext]{ref-kuhnStructureScientificRevolutions2009}
+Kuhn, T. S. (2009). \emph{The structure of scientific revolutions} (3. ed., {[}Nachdr.{]}). Chicago: Univ. of Chicago Press.
 
@@
 \url{https://doi.org/10.1007/978-94-010-3142-4_7}
+
+\bibitem[\citeproctext]{ref-lakensImprovingTransparencyFalsifiability2021}
+Lakens, D., \& DeBruine, L.
+M.
+(2021).
+Improving {Transparency}, {Falsifiability}, and {Rigor} by {Making Hypothesis Tests Machine-Readable}.
+\emph{Advances in Methods and Practices in Psychological Science}, \emph{4}(2), 2515245920970949.
+\url{https://doi.org/10.1177/2515245920970949}
 
@@
 \url{https://doi.org/10.1086/714812}
+
+\bibitem[\citeproctext]{ref-lewinPsychologyProcessGroup1943}
+Lewin, K. (1943). Psychology and the {Process} of {Group Living}. \emph{The Journal of Social Psychology}, \emph{17}(1), 113--131. \url{https://doi.org/10.1080/00224545.1943.9712269}
 
@@
 
+\bibitem[\citeproctext]{ref-morrisRoleFamilyContext2007}
+Morris, A.
+S., Silk, J.
+S., Steinberg, L., Myers, S.
+S., \& Robinson, L.
+R.
+(2007).
+The role of the family context in the development of emotion regulation.
+\emph{Social Development}, \emph{16}(2), 361--388.
+\url{https://doi.org/10.1111/j.1467-9507.2007.00389.x}
+
 \bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015a}
@@
 Oberauer, K., \& Lewandowsky, S. (2019). Addressing the theory crisis in psychology. \emph{Psychonomic Bulletin \& Review}, \emph{26}(5), 1596--1618. \url{https://doi.org/10.3758/s13423-019-01645-2}
+
+\bibitem[\citeproctext]{ref-peikertReproducibleResearchTutorial2021}
+Peikert, A., Van Lissa, C.
+J., \& Brandmaier, A.
+M.
+(2021).
+Reproducible {Research} in {R}: {A Tutorial} on {How} to {Do} the {Same Thing More Than Once}.
+\emph{Psych}, \emph{3}(4), 836--867.
+\url{https://doi.org/10.3390/psych3040053}
+
+\bibitem[\citeproctext]{ref-quineReasonsIndeterminacyTranslation1970}
+Quine, W. V. (1970). On the {Reasons} for {Indeterminacy} of {Translation}. \emph{The Journal of Philosophy}, \emph{67}(6), 178--183. \url{https://doi.org/10.2307/2023887}
+
+\bibitem[\citeproctext]{ref-ramGitCanFacilitate2013}
+Ram, K.
+(2013).
+Git can facilitate greater reproducibility and increased transparency in science.
+\emph{Source Code for Biology and Medicine}, \emph{8}(1), 7.
+\url{https://doi.org/10.1186/1751-0473-8-7}
 
@@
 
+\bibitem[\citeproctext]{ref-vanlissaUsingEndpointsCheck2023}
+Van Lissa, C. J. (2023). Using {Endpoints} to {Check Reproducibility} {[}Package Documentation{]}. Retrieved March 21, 2024, from \url{https://cjvanlissa.github.io/worcs/articles/endpoints.html}
+
+\bibitem[\citeproctext]{ref-vanlissaWORCSWorkflowOpen2021}
+Van Lissa, C.
+J., Brandmaier, A.
+M., Brinkman, L., Lamprecht, A.-L., Peikert, A., Struiksma, M.
+E., \& Vreede, B.
+M.
+I.
+(2021).
+{WORCS}: {A} workflow for open reproducible code in science.
+\emph{Data Science}, \emph{4}(1), 29--49.
+\url{https://doi.org/10.3233/DS-210031}
+
+\bibitem[\citeproctext]{ref-vogtFAIR20Extending2024}
+Vogt, L., Strömert, P., Matentzoglu, N., Karam, N., Konrad, M., Prinz, M., \& Baum, R.
+(2024, May 6).
+{FAIR} 2.0: {Extending} the {FAIR Guiding Principles} to {Address Semantic Interoperability}.
+Retrieved November 20, 2024, from \url{http://arxiv.org/abs/2405.03345}
+
+\bibitem[\citeproctext]{ref-wilkinsonFAIRGuidingPrinciples2016a}
+Wilkinson, M.
+D., Dumontier, M., Aalbersberg, Ij.
+J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
+(2016).
+The {FAIR Guiding Principles} for scientific data management and stewardship.
+\emph{Scientific Data}, \emph{3}(1), 160018.
+\url{https://doi.org/10.1038/sdata.2016.18}
+
 \end{CSLReferences}

### 2024-12-06 manuscript.Rmd
update 2024-12-06 10:08:25.896332

@@
 The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit [REF Rasoul].
-Or, perhaps the hypotheses are of no substantive interest [REF Hoijtink], and researchers are simply testing them as part of a cultural ritual [REF Gigerenzer].
-No matter what the explanation - testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
+Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
+Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
 
-According to the aforementioned statistics, theory has an uncomfortable and paradoxical role in contemporary psychology:
+Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
 The majority of papers ostensibly test hypotheses,
@@
 
-To date, scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
-The present paper applies, for the first time, open science principles to theory.
-Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition.
+The present paper advances transparent scholarly communication by applying the FAIR principles to scientific theories.
+<!-- First, we reflect on the differences between theory and data. <!-- CJ: Should we? --> -->
+<!-- Second, we adapt the FAIR principles for theory, respecting these differences. -->
+<!-- Third, we discuss the current state and future directions of psychological theory with respect to each criterion. -->
+<!-- Fourth, we present worked examples of theory made FAIR. -->
+<!-- We conclude with a summary and future directions for FAIR theory. -->
 
-### Findability
+## Making Theory FAIR
 
-One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
-One contributing cause of theories' lack of Findability is the lack of standardized metadata or even a keyword to signal the presence of theory within a paper, and - consequently - the absence of a unified search engine for theory.
-Making theories more findable through rich metadata would allow them to be systematically indexed, tagged, and made searchable.
-This would allow researchers to easily identify relevant theories to inform their hypotheses,
+There are different definitions of theory,
+and many of those definitions are consistent with the FAIR theory principles.
+This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward].
+<!-- Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried]. -->
+<!-- This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
+There is no principled difference, but a FAIR theory is the most general instance of a theory, whereas a model is highly specific - e.g., by assuming functional form and error families.
+We can make a distinction between statistical models - which make just enough assumptions to allow estimation of the remaining unknown parameters from data - and generative/computational models, which are completely parametrized to the point that a specific interpreter can use them to generate new data.
+To some extent, -->
+Meehl [-@meehlAppraisingAmendingTheories1990] provides guidance as to what kinds of "statements" such a theory might contain:
+statements about the types of entities postulated (i.e., ontology),
+statements about causal connections between those entities,
+statements about the functional form of those connections,
+and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
+In order to make theory FAIR, it must be instantiated as a type of data,
+in a format that is both human- and machine-readable with as few interpretative steps as possible.
+Verbal statements in plain text fit this description,
+as do mathematical formulae and statements expressed in some axiomatic system.
+Examples include pseudo-code,
+interpretable computer code,
+graphs described in the DOT language,
+and Gray's theory maps [@grayHowMapTheory2017].
+The common practice of instantiating theory as a bitmap image reproduced in a journal article falls short of this ideal;
+but the same image represented as a DOT graph (which can in turn be *reproduced* in the print version) does meet this ideal.
+Assuming that the theory is instantiated as human- and machine readable data, we can apply the principles of FAIR data to it.
+
+## Findability
+
+Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
 grounding their work in established theoretical foundations.
-Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
-For example, it becomes possible to conduct meta-research on theories, 
-in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
-Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
+Making theories Findable also increases the impact and reuse potential of theories across disciplines,
+either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
+or through analogical modeling.
 In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
 For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+Findability also enables meta-research on theories, 
+in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
+In a similar way, it would become possible to compare all theories of a specific phenomenon,
+or to study structural properties of theories.
 
-Another contributing cause to the lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
+The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
+First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+Of the many services that provide DOIs for scientific information artefacts,
+Zenodo and the Open Science Framework are commonly used in psychology.
+Second, Findable theory is described with rich metadata	(F2).
+This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
+It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
+ontology [@guyonMeasurementOntologyEpistemology2018],
+or catalog of psychological phenomena [REF Noah Denny].
+Metadata should also include identifiers for all the versions of the theory it describes	(F3);
+Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
+Finally, these metadata should be registered or indexed in a searchable registry (F4).
+This final criterion is less straightforward.
+Ideally, FAIR theories should be indexed in search engines used by academics, like Google Scholar.
+At present, however, these search engines are designed to index traditional print publications.
+The *data paper* solves this problem for research data;
+the idea is that scholars publish a paper (or even preprint) as documentation for the data resource [REF McGIllivray on data papers].
+The data paper is indexed by search engines, and in turn points to the relevant information artifact.
+The same solution could be applied to theories - but it seems superfluous to generate papers whose only purpose is to redirect to a specific resource.
+Another solution is to manually index FAIR theories,
+for example by adding them to one's Google Scholar profile,
+or entering them in PURE.
+<!-- CJ: Can we guarantee automatically indexing FAIR theory somehow? -->
+
+At present, theories have poor findability, which impedes cumulative knowledge acquisition.
+One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like "theory", "model", and "framework" are used interchangeably.
+To curb this trend, we suggest using the keyword `"FAIR theory"` for all resources that constitute or reference a FAIR theory.
+This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
+Another factor contributing to the present lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
 A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website [REF GRAY].
+and post many examples on their website [@grayHowMapTheory2017].
 Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
@@
 
-### Accessibility
+## Accessibility
 
@@
 theories offer a top-down, big picture representation of the phenomena studied in a field.
+In other words, theories are an important instrument in science communication.
+
+The Accessibility criteria serve to *regulate* access, not to maximize it.
+These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
+Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
+This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
+The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
+Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
+which is also achieved via long-term storage (e.g., on Zenodo).
+Git remote repositories allow for access control,
+and Zenodo allows for access control of individual files/resources.
 
 At present, there are several impediments to theories' accessibility.
-The first are paywalls erected by commercial publishers.
-Open Access publishing increases the accessibility of all academic output, including theory.
-
-A second impediment is therefore the prevalence of (strategic) ambiguity,
-which renders theory difficult to understand [REF Frankenhuis].
-While open access publishing increases the availability of theories,
-accessibility further requires clear and explicit communication.
+To the extent that theories are still contained within papers,
+paywalls erected by commercial publishers constitute a barrier.
+Open Access publishing thus increases the accessibility of all academic output, including theory.
+A second impediment is more indirect:
+While open access publishing increases practical access to theories,
+accessibility also requires clear and explicit communication.
 This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
+The current prevalence of strategic ambiguity renders psychological theory difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
 It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
-which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
-Thus, it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
+which holds that every communicative utterance has multiple alternative translations, with no *objective* means of choosing the correct one.
+It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
 This places a theoretical upper bound on theories' ability to be understood.
@@
 because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
-DOA is inherently problematic because authors may not understand their own theories well.
-This is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
-In short, DOA unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
+The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
+This demonstrates that original authors are not the ultimate authority on their theories.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
 

### 2024-12-06 manuscript.Rmd
update 2024-12-06 12:42:03.255242

@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
-In order to make theory FAIR, it must be instantiated as a type of data,
-in a format that is both human- and machine-readable with as few interpretative steps as possible.
-Verbal statements in plain text fit this description,
-as do mathematical formulae and statements expressed in some axiomatic system.
-Examples include pseudo-code,
-interpretable computer code,
-graphs described in the DOT language,
-and Gray's theory maps [@grayHowMapTheory2017].
-The common practice of instantiating theory as a bitmap image reproduced in a journal article falls short of this ideal;
-but the same image represented as a DOT graph (which can in turn be *reproduced* in the print version) does meet this ideal.
-Assuming that the theory is instantiated as human- and machine readable data, we can apply the principles of FAIR data to it.
 
@@
 
-### Interoperability
+## Interoperability
 
 Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
-As the original definition of interoperability was somewhat narrow and abstract,
+The original interoperability principles can be rephrased somewhat to apply to theory.
+Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
+The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
+Instead, FAIR theory should, ad minimum,
+be instantiated as 
+as a type of data that is human- and machine-readable with as few interpretative steps as possible,
+as previously recommended [@vanlissaWORCSWorkflowOpen2021].
+Depending on the level of formalization of the theory,
+different formats may be appropriate,
+such as verbal statements in plain text,
+mathematical formulae,
+and statements expressed in some axiomatic system.
+Examples of the latter include pseudo-code,
+interpretable computer code,
+and Gray's theory maps [@grayHowMapTheory2017].
+While a theory represented as a bitmap image is not very interoperable,
+the same image represented in the DOT language for representing graphs does meet this ideal.
+
+Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
+This is essentially a call to establish standardized ontologies,
+which are themselves a type of theory [@meehlAppraisingAmendingTheories1990]. 
+Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
+The first part of this principle allows for nested theories;
+for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
+This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
+The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
+This can be achieved by using Git for version control and Zenodo for archiving.
+
+As the original definition of interoperability was somewhat narrow [@wilkinsonFAIRGuidingPrinciples2016a],
 the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
-This definition invites the question: *interoperable for what purpose?*
+This definition invites the question: *interoperable for what?*
 Suitable answers for FAIR theory may be:
-this theory is X-interoperable for deriving testable null-hypotheses,
+this theory is X-interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
 or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-These definitions suggest that FAIR theory should, ad minimum,
-be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].
-Furthermore, theories may have specific properties that incur affordances in terms of X-interoperability;
-for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories.
+This revised definition implies that theories have specific properties that incur affordances in terms of X-interoperability;
+for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form).
 
@@
 
-Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] seems to imply that theory plays a practical role in the day-to-day work of psychological researchers.
-But, as we argued before, there are clear signs that this is not the case.
+With regard to the state of interoperability in contemporary psychology, 
+Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
+But, as we argued, this is not the case.
 The examples of X-interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
@@
 
-### Reusability
+## Reusability
 
-Some have argued that there is a norm *against* theory reuse:
+If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
+Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
+As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
+although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].
+Criterion R1.2 is met by version control with Git and archival on Zenodo.
+Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
+
+If we consider the current state of Reusability in psychological theory, there appears to be a norm *against* theory reuse:
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
-Such a norm impedes scientific progress.
+This norm impedes scientific progress.
 Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
-Theories are made reusable by thorough documentation,
-appropriate licensing, 
-and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning).
+<!-- Theories are made reusable by thorough documentation, -->
+<!-- appropriate licensing,  -->
+<!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
+In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
+reflected in questions such as "who owns theory",
+and "who determines how a theory may be reused or changed"?
+These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
 
-In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
-and "who determines how a theory may be reused or changed"?
 Licensing theories for reuse provides an unambiguous answer to such questions.
-These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
-However, copyright law limits the rights authors can claim over theory based on the *idea-expression dichotomy* [@bently2010copyright],
-<!-- o.
-Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts.
-Separating an idea from the expression or manifestation of that idea is known in copyright law as the  -->
+In determining what license is appropriate for theory,
+it is important to consider that copyright law limits authors' rights based on the *idea-expression dichotomy* [@bently2010copyright],
 which holds that copyright explicitly does not
 *"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
-It may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
-<!-- Copyright pertaining to recipes offer an interesting perspective: -->
-<!-- Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright. -->
-Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short in terms of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-If copyright limits Reusability and does not cover ideas in their purest form (like formal theory),
+Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
+It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
 then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
@@
 to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].
+
+## Some other stuff
 

### 2024-12-06 manuscript.Rmd
update 2024-12-06 13:12:45.606398

@@
 
-## FAIR Theory and Recognition & Rewards
+<!-- ## FAIR Theory and Recognition & Rewards -->
 
-FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
-In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging.
+<!-- FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory. -->
+<!-- In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging. -->
 

### 2024-12-07 manuscript.Rmd
update 2024-12-07 07:14:42.558463

@@
 
-The present paper advances transparent scholarly communication by applying the FAIR principles to scientific theories.
 <!-- First, we reflect on the differences between theory and data. <!-- CJ: Should we? --> -->
@@
 ## Making Theory FAIR
+
+The present paper addresses the lack of open science methods for theory development and improvement by applying the FAIR principles to scientific theories.
+Merely publishing theory does not make it open;
+to be open, theory should adhere to established open science standards.
+The FAIR principles, initially introduced for open research data,
+have since been applied to other information artifacts [@lamprechtFAIRPrinciplesResearch2019].
+We apply the FAIR principles to digital representations of theory,
+introducing a FAIR metadata format to make theories *Findable* via a DOI,
+*Accessible* in a machine- and human-readable filetype,
+*Interoperable* within the data analysis environment,
+and *Reusable* in the practical and legal sense, so that they may be improved over time.
+Following Lamprecht and colleagues,
+we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
+We reflect on the necessary changes (which are minor),
+as well as on the current state and future of FAIR theory in the social sciences.
+The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
+and we provide worked examples to encourage their adoption.
+
+```{r tabfair}
+tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
+names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
+papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
+```
 
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
+
+### The Role of Theory Formalization
+
+Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
+Greater formalization increases theories' *empirical content* [REF],
+making them easier to falsify,
+which necessitates revising them,
+thus advancing our principled understanding of the phenomena they describe.
+FAIR Theory does not require theories to be formal,
+and formal theory can be represented in a way that is not FAIR.
+For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
+Conversely, a directed acyclic graph (DAG) is a type of formal theory,
+but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
+then this formal theory is not FAIR.
+FAIR Theory is thus consistent with, but does not require, formal theory.
+There is more to say about this in light of the *Accessibility* principle,
+where we discuss theories' "ability to be understood" [@guestWhatMakesGood2024] and the *problem of translation* [@quineReasonsIndeterminacyTranslation1970].
 
@@
 
-## Publication is not Enough
-
-Merely publishing a theory does not make it open;
-to be open, theory should adhere to established open science standards.
-The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
-We propose to apply the FAIR principles to digital representations of theory as well,
-introducing a FAIR metadata format to represent (formal) theories.
-The resulting theories are made *Findable* via a DOI,
-Accessible in a machine- and human-readable filetype,
-Interoperable within the data analysis environment,
-and Reusable in the practical and legal sense, so that they may be improved over time.
-
-## Adapting the FAIR Principles
-
-The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
-From their inception, these principles were developed with "other research resources" in mind.
-Scholars have adapted the FAIR principles to, e.g., research software [REF Lamprecht].
-Yet there are key distinctions between theory and other FAIR information artefacts.
-With this in mind, the present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
-We establish guiding principles for instantiating theory as a FAIR digital research artifact,
-and provide applied examples to encourage their adoption.
-
-```{r tabfair}
-tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
-names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
-papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
-```
-
 <!-- ## FAIR Theory and Recognition & Rewards -->
@@
 <!-- In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging. -->
-
-## The Role of Theory Formalization
-
-Concerns about the state of theory are a recurring theme in the psychological literature,
-but previous writing has focused on theory formalization as a solution for ambiguity in psychological theory.
-Greater formality increases theories' *empirical content*,
-making them easier to falsify,
-which necessitates revising them,
-thus advancing our principled understanding of the phenomena they describe.
-Conceptually, theory formalization is orthogonal to FAIR theory.
-FAIR Theory does not require theories to be formal, and formal theory can be represented in a way that is not FAIR.
-It is - in principle - possible to represent a collection of verbal statements as a FAIR Theory.
-While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.
 

### 2024-12-07 manuscript.Rmd
update 2024-12-07 07:27:16.199069

@@
 then this formal theory is not FAIR.
-FAIR Theory is thus consistent with, but does not require, formal theory.
-There is more to say about this in light of the *Accessibility* principle,
-where we discuss theories' "ability to be understood" [@guestWhatMakesGood2024] and the *problem of translation* [@quineReasonsIndeterminacyTranslation1970].
+FAIR Theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
 
@@
 
-```{r}
+```{r tabmeehl}
 data.frame(

### 2024-12-07 manuscript.tex
update 2024-12-07 07:27:16.199069

@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 3877}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 4906}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 4a12ff5b.
+This is a preprint paper, generated from Git Commit \# aec0b761.
 
@@
 
-Scholars have raised concerns about the state of theory in social science for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as \emph{formalization} (Szollosi \& Donkin, 2021).
-A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
-The present paper focuses on the second concern,
-advancing transparent communication about theory by applying open science principles to psychological theory for the first time
-and introducing the concept of \emph{FAIR Theory}.
+Scholars have raised concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
+Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences (Szollosi \& Donkin, 2021).
+In other words, social scientific theories lack \emph{formalization} .
+A second concern is the lack of transparent and democratic scholarly communication about psychological theory.
+
+Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+The equally critical inductive processes of theory construction and -improvement have been largely overlooked.
+The present paper restores balance by applying, for the first time,
+open science principles to psychological theory.
+We apply the FAIR principles to scientific theories,
+introducing the concept of \emph{FAIR Theory} to
+facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 Only 15\% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis (McPhetres et al., 2021).
-This invites the question where hypotheses come from in the remaining 85\% of deductive studies,
-and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
-These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
+The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
+Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
+Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
+Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+
+Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
 The majority of papers ostensibly test hypotheses,
 but these are rarely derived from theory,
-and test results rarely contribute to the improvement of existing theories.
+and test results do not routinely contribute to the improvement of existing theories.
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Paul E.
+Meehl, 1978).
 
-To date, scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
-The present paper applies, for the first time, open science principles to theory.
-Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition.
+--\textgreater{}
 
-\subsubsection{Findability}\label{findability}
+\subsection{Making Theory FAIR}\label{making-theory-fair}
 
-One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
-One contributing cause of theories' lack of Findability is the lack of standardized metadata or even a keyword to signal the presence of theory within a paper, and - consequently - the absence of a unified search engine for theory.
-Making theories more findable through rich metadata would allow them to be systematically indexed, tagged, and made searchable.
-This would allow researchers to easily identify relevant theories to inform their hypotheses,
+The present paper addresses the lack of open science methods for theory development and improvement by applying the FAIR principles to scientific theories.
+Merely publishing theory does not make it open;
+to be open, theory should adhere to established open science standards.
+The FAIR principles, initially introduced for open research data,
+have since been applied to other information artifacts (Lamprecht et al., 2019).
+We apply the FAIR principles to digital representations of theory,
+introducing a FAIR metadata format to make theories \emph{Findable} via a DOI,
+\emph{Accessible} in a machine- and human-readable filetype,
+\emph{Interoperable} within the data analysis environment,
+and \emph{Reusable} in the practical and legal sense, so that they may be improved over time.
+Following Lamprecht and colleagues,
+we adapt the FAIR principles for theory, see Table \ref{tab:tabfair}.
+We reflect on the necessary changes (which are minor),
+as well as on the current state and future of FAIR theory in the social sciences.
+The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
+and we provide worked examples to encourage their adoption.
+
+\begin{lltable}
+
+\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
+\caption{\label{tab:tabfair}}\\
+\toprule
+Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
+\midrule
+\endfirsthead
+\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
+\toprule
+Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
+\midrule
+\endhead
+F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory (meta)data has a global, unique and persistent identifier & Rephrased\\
+F2 & Data are described with rich metadata & Theory is described with rich metadata & Rephrased\\
+F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & Rephrased and extended\\
+F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & Rephrased, needs work\\
+A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & Rephrased\\
+A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Remain the same\\
+A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Remain the same, but less relevant\\
+A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Rephrased, but less relevant\\
+I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
+I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
+I2S.1 & - &  & \\
+I2S.2 & - &  & \\
+I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
+I4S & - &  & \\
+R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
+R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
+R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
+R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
+\bottomrule
+\end{longtable}
+
+\end{lltable}
+
+There are different definitions of theory,
+and many of those definitions are consistent with the FAIR theory principles.
+This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data {[}bogen/woodward{]}.
+Meehl (1990) provides guidance as to what kinds of ``statements'' such a theory might contain:
+statements about the types of entities postulated (i.e., ontology),
+statements about causal connections between those entities,
+statements about the functional form of those connections,
+and statements about their specific numerical values Guest (2024).
+
+\subsubsection{The Role of Theory Formalization}\label{the-role-of-theory-formalization}
+
+Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
+Greater formalization increases theories' \emph{empirical content} {[}REF{]},
+making them easier to falsify,
+which necessitates revising them,
+thus advancing our principled understanding of the phenomena they describe.
+FAIR Theory does not require theories to be formal,
+and formal theory can be represented in a way that is not FAIR.
+For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
+Conversely, a directed acyclic graph (DAG) is a type of formal theory,
+but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
+then this formal theory is not FAIR.
+FAIR Theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).
+
+\subsection{Findability}\label{findability}
+
+Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
 grounding their work in established theoretical foundations.
-Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
-For example, it becomes possible to conduct meta-research on theories,
-in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
-Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
+Making theories Findable also increases the impact and reuse potential of theories across disciplines,
+either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
+or through analogical modeling.
 In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
 For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+Findability also enables meta-research on theories,
+in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
+In a similar way, it would become possible to compare all theories of a specific phenomenon,
+or to study structural properties of theories.
 
-Another contributing cause to the lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
+The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
+First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+Of the many services that provide DOIs for scientific information artefacts,
+Zenodo and the Open Science Framework are commonly used in psychology.
+Second, Findable theory is described with rich metadata (F2).
+This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
+It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
+ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
+or catalog of psychological phenomena {[}REF Noah Denny{]}.
+Metadata should also include identifiers for all the versions of the theory it describes (F3);
+Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
+Finally, these metadata should be registered or indexed in a searchable registry (F4).
+This final criterion is less straightforward.
+Ideally, FAIR theories should be indexed in search engines used by academics, like Google Scholar.
+At present, however, these search engines are designed to index traditional print publications.
+The \emph{data paper} solves this problem for research data;
+the idea is that scholars publish a paper (or even preprint) as documentation for the data resource {[}REF McGIllivray on data papers{]}.
+The data paper is indexed by search engines, and in turn points to the relevant information artifact.
+The same solution could be applied to theories - but it seems superfluous to generate papers whose only purpose is to redirect to a specific resource.
+Another solution is to manually index FAIR theories,
+for example by adding them to one's Google Scholar profile,
+or entering them in PURE.
+
+At present, theories have poor findability, which impedes cumulative knowledge acquisition.
+One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like ``theory'', ``model'', and ``framework'' are used interchangeably.
+To curb this trend, we suggest using the keyword \texttt{"FAIR\ theory"} for all resources that constitute or reference a FAIR theory.
+This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
+Another factor contributing to the present lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
 A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website {[}REF GRAY{]}.
+and post many examples on their website (Gray, 2017).
 Similarly, Borsboom and colleagues seek to establish a database of psychological theories {[}REF BORSBOOM{]}.
@@
 
-\subsubsection{Accessibility}\label{accessibility}
+\subsection{Accessibility}\label{accessibility}
 
@@
 theories offer a top-down, big picture representation of the phenomena studied in a field.
+In other words, theories are an important instrument in science communication.
+
+The Accessibility criteria serve to \emph{regulate} access, not to maximize it.
+These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
+Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
+This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
+The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via \texttt{https} or \texttt{git}).
+Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
+which is also achieved via long-term storage (e.g., on Zenodo).
+Git remote repositories allow for access control,
+and Zenodo allows for access control of individual files/resources.
 
 At present, there are several impediments to theories' accessibility.
-The first are paywalls erected by commercial publishers.
-Open Access publishing increases the accessibility of all academic output, including theory.
-
-A second impediment is therefore the prevalence of (strategic) ambiguity,
-which renders theory difficult to understand {[}REF Frankenhuis{]}.
-While open access publishing increases the availability of theories,
-accessibility further requires clear and explicit communication.
+To the extent that theories are still contained within papers,
+paywalls erected by commercial publishers constitute a barrier.
+Open Access publishing thus increases the accessibility of all academic output, including theory.
+A second impediment is more indirect:
+While open access publishing increases practical access to theories,
+accessibility also requires clear and explicit communication.
 This property of good theories has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
+The current prevalence of strategic ambiguity renders psychological theory difficult to understand (Frankenhuis et al., 2023).
 It is important to acknowledge the \emph{indeterminacy of translation} (Quine, 1970):
-which holds that every communicative utterance has multiple alternative translations, with no objective means of choosing the correct or best one.
-Thus, it is not possible to entirely formalize an idea to the point that it becomes unambiguously interpretable.
+which holds that every communicative utterance has multiple alternative translations, with no \emph{objective} means of choosing the correct one.
+It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
 This places a theoretical upper bound on theories' ability to be understood.
@@
 because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target (Szollosi \& Donkin, 2021).
-DOA is inherently problematic because authors may not understand their own theories well.
-This is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
-In short, DOA unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
+The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
+This demonstrates that original authors are not the ultimate authority on their theories.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
 
-\subsubsection{Interoperability}\label{interoperability}
+\subsection{Interoperability}\label{interoperability}
 
 Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (Wilkinson et al., 2016).
-As the original definition of interoperability was somewhat narrow and abstract,
+The original interoperability principles can be rephrased somewhat to apply to theory.
+Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
+The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
+Instead, FAIR theory should, ad minimum,
+be instantiated as
+as a type of data that is human- and machine-readable with as few interpretative steps as possible,
+as previously recommended (Van Lissa et al., 2021).
+Depending on the level of formalization of the theory,
+different formats may be appropriate,
+such as verbal statements in plain text,
+mathematical formulae,
+and statements expressed in some axiomatic system.
+Examples of the latter include pseudo-code,
+interpretable computer code,
+and Gray's theory maps (Gray, 2017).
+While a theory represented as a bitmap image is not very interoperable,
+the same image represented in the DOT language for representing graphs does meet this ideal.
+
+Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
+This is essentially a call to establish standardized ontologies,
+which are themselves a type of theory (Paul E. Meehl, 1990).
+Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
+The first part of this principle allows for nested theories;
+for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
+This can be achieved by linking the DOI of those nested theories ({``Contributing {Citations} and {References},''} n.d.).
+The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
+This can be achieved by using Git for version control and Zenodo for archiving.
+
+As the original definition of interoperability was somewhat narrow (Wilkinson et al., 2016),
 the concept has recently been further refined in terms of facilitating ``successful communication between machines and between humans and machines'', where ``A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
-This definition invites the question: \emph{interoperable for what purpose?}
+This definition invites the question: \emph{interoperable for what?}
 Suitable answers for FAIR theory may be:
-this theory is X-interoperable for deriving testable null-hypotheses,
+this theory is X-interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
 or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-These definitions suggest that FAIR theory should, ad minimum,
-be instantiated in a human- and machine-readable file format, as previously recommended (Van Lissa et al., 2021).
-Furthermore, theories may have specific properties that incur affordances in terms of X-interoperability;
-for example, Table \ref{tab:tabmeehl} illustrates the affordances of Meehl's nine properties of strong theories.
+This revised definition implies that theories have specific properties that incur affordances in terms of X-interoperability;
+for example, Table \ref{tab:tabmeehl} illustrates the affordances of Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form).
 
@@
 
-\caption{\label{tab:unnamed-chunk-2}}
+\caption{\label{tab:tabmeehl}}
 
@@
 
-Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) seems to imply that theory plays a practical role in the day-to-day work of psychological researchers.
-But, as we argued before, there are clear signs that this is not the case.
+With regard to the state of interoperability in contemporary psychology,
+Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
+But, as we argued, this is not the case.
 The examples of X-interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
@@
 
-\subsubsection{Reusability}\label{reusability}
+\subsection{Reusability}\label{reusability}
 
-Some have argued that there is a norm \emph{against} theory reuse:
+If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
+Applied to FAIR theory, reusability requires that theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
+As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
+although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements {[}REF TOP guidelines{]}.
+Criterion R1.2 is met by version control with Git and archival on Zenodo.
+Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
+
+If we consider the current state of Reusability in psychological theory, there appears to be a norm \emph{against} theory reuse:
 \emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
-Such a norm impedes scientific progress.
+This norm impedes scientific progress.
 Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data (de Groot, 1961).
-Theories are made reusable by thorough documentation,
-appropriate licensing,
-and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning).
+In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
+reflected in questions such as ``who owns theory'',
+and ``who determines how a theory may be reused or changed''?
+These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
 
-In our workshops on FAIR Theory, we often hear questions such as ``who owns theory'',
-and ``who determines how a theory may be reused or changed''?
 Licensing theories for reuse provides an unambiguous answer to such questions.
-These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
-However, copyright law limits the rights authors can claim over theory based on the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010),
+In determining what license is appropriate for theory,
+it is important to consider that copyright law limits authors' rights based on the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010),
 which holds that copyright explicitly does not
 \emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''}.
-It may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
-Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short in terms of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-If copyright limits Reusability and does not cover ideas in their purest form (like formal theory),
+Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
+It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
 then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
@@
 to meet the definition of Reusability (R1.2, Table \ref{tab:tabfair} and to comply with other definitions of open scholarship (Aalbersberg et al., 2018).
+
+\subsection{Some other stuff}\label{some-other-stuff}
 
@@
 This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
-
-\subsection{Publication is not Enough}\label{publication-is-not-enough}
-
-Merely publishing a theory does not make it open;
-to be open, theory should adhere to established open science standards.
-The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output (e.g., software Lamprecht et al., 2019).
-We propose to apply the FAIR principles to digital representations of theory as well,
-introducing a FAIR metadata format to represent (formal) theories.
-The resulting theories are made \emph{Findable} via a DOI,
-Accessible in a machine- and human-readable filetype,
-Interoperable within the data analysis environment,
-and Reusable in the practical and legal sense, so that they may be improved over time.
-
-\subsection{Adapting the FAIR Principles}\label{adapting-the-fair-principles}
-
-The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
-From their inception, these principles were developed with ``other research resources'' in mind.
-Scholars have adapted the FAIR principles to, e.g., research software {[}REF Lamprecht{]}.
-Yet there are key distinctions between theory and other FAIR information artefacts.
-With this in mind, the present paper further extends the FAIR principles' definition to theory, see Table \ref{tab:tabfair}.
-We establish guiding principles for instantiating theory as a FAIR digital research artifact,
-and provide applied examples to encourage their adoption.
-
-\begin{lltable}
-
-\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:tabfair}}\\
-\toprule
-Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
-\toprule
-Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
-\midrule
-\endhead
-F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory and its associated metadata has a global, unique and persistent identifier for each version (using semantic versioning) & Rephrased\\
-F2 & Data are described with rich metadata & Theory is described with rich metadata & \textasciitilde{}same\\
-F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & \textasciitilde{}same\\
-F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & NEEDS WORK!
-GitHub is indexed by Google I believe, but ideally, we'd like our theories to show up in Google Scholar, or even dedicated academic search enginges.
-Where could we put them to realize this?\\
-A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & \textasciitilde{}same\\
-A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Same\\
-A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Do\ \ we need this?\\
-A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Do\ \ we need this?\\
-I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased\\
-I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & NEEDS WORK!
-I think this is where we explain the value of e.g.
-universal graph languages like Aaron and Max'\\
-I2S.1 & - &  & \\
-I2S.2 & - &  & \\
-I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Rephrased.
-I envision a LinkList-like structure where each theory version references its ancestor\\
-I4S & - &  & Discard\\
-R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Needs work.
-How do we envision this?
-Keywords?
-ISBN-like codes for variable type?\\
-R1.1 & (Meta)data are released with a clear and accessible data usage license & (Meta)data are released with a clear and accessible license & Needs work: which license is good for theory?\\
-R1.2 & (Meta)data are associated with detailed provenance & (Meta)data are associated with detailed provenance & Relates to I3; I think this point relates more to the theory's ancestry, and I3 relates to e.g.
-incorporating other theories within a theory (e.g., theory of measurement inside of structural theory)\\
-R1.3 & (Meta)data meet domain-relevant community standards & Theory metadata and documentation meet domain-relevant community standards & These standards do not yet exist; we can take a first step towards developing them and recommend that this be an active area of development\\
-\bottomrule
-\end{longtable}
-
-\end{lltable}
-
-\subsection{FAIR Theory and Recognition \& Rewards}\label{fair-theory-and-recognition-rewards}
-
-FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
-In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging.
-
-\subsection{The Role of Theory Formalization}\label{the-role-of-theory-formalization}
-
-Concerns about the state of theory are a recurring theme in the psychological literature,
-but previous writing has focused on theory formalization as a solution for ambiguity in psychological theory.
-Greater formality increases theories' \emph{empirical content},
-making them easier to falsify,
-which necessitates revising them,
-thus advancing our principled understanding of the phenomena they describe.
-Conceptually, theory formalization is orthogonal to FAIR theory.
-FAIR Theory does not require theories to be formal, and formal theory can be represented in a way that is not FAIR.
-It is - in principle - possible to represent a collection of verbal statements as a FAIR Theory.
-While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.
 
@@
 
+\bibitem[\citeproctext]{ref-boscoMetaBUSVehicleFacilitating2017}
+Bosco, F.
+A., Uggerslev, K.
+L., \& Steel, P.
+(2017).
+{MetaBUS} as a vehicle for facilitating meta-analysis.
+\emph{Human Resource Management Review}, \emph{27}(1), 237--254.
+\url{https://doi.org/10.1016/j.hrmr.2016.09.013}
+
 \bibitem[\citeproctext]{ref-cinelliCrashCourseGood2022}
 Cinelli, C., Forney, A., \& Pearl, J. (2022). A {Crash Course} in {Good} and {Bad Controls}. \emph{Sociological Methods \& Research}, 00491241221099552. \url{https://doi.org/10.1177/00491241221099552}
+
+\bibitem[\citeproctext]{ref-ContributingCitationsReferences}
+Contributing {Citations} and {References}. (n.d.). Retrieved December 5, 2024, from \url{https://support.datacite.org/docs/data-citation}
 
@@
 
+\bibitem[\citeproctext]{ref-frankenhuisStrategicAmbiguitySocial2023}
+Frankenhuis, W.
+E., Panchanathan, K., \& Smaldino, P.
+E.
+(2023).
+Strategic {Ambiguity} in the {Social Sciences}.
+\emph{Social Psychological Bulletin}, \emph{18}, 1--25.
+\url{https://doi.org/10.32872/spb.9923}
+
+\bibitem[\citeproctext]{ref-friedTheoriesModelsWhat2020}
+Fried, E.
+I.
+(2020).
+Theories and {Models}: {What They Are}, {What They Are} for, and {What They Are About}.
+\emph{Psychological Inquiry}, \emph{31}(4), 336--344.
+\url{https://doi.org/10.1080/1047840X.2020.1854011}
+
+\bibitem[\citeproctext]{ref-gigerenzerNullRitualWhat2004}
+Gigerenzer, G., Krauss, S., \& Vitouch, O.
+(2004).
+The null ritual : {What} you always wanted to know about significance testing but were afraid to ask.
+In D.
+Kaplan (Ed.), \emph{The {Sage} handbook of quantitative methodology for the social sciences} (pp.
+391--408).
+Thousand Oaks: Sage.
+
+\bibitem[\citeproctext]{ref-grayHowMapTheory2017}
+Gray, K.
+(2017).
+How to {Map Theory}: {Reliable Methods Are Fruitless Without Rigorous Theory}.
+\emph{Perspectives on Psychological Science}, \emph{12}(5), 731--741.
+\url{https://doi.org/10.1177/1745691617691949}
+
 \bibitem[\citeproctext]{ref-grossEmotionRegulationCurrent2015}
@@
 Guest, O. (2024). What {Makes} a {Good Theory}, and {How Do We Make} a {Theory Good}? \emph{Computational Brain \& Behavior}. \url{https://doi.org/10.1007/s42113-023-00193-2}
+
+\bibitem[\citeproctext]{ref-guyonMeasurementOntologyEpistemology2018}
+Guyon, H., Kop, J.-L., Juhel, J., \& Falissard, B.
+(2018).
+Measurement, ontology, and epistemology: {Psychology} needs pragmatism-realism.
+\emph{Theory \& Psychology}, \emph{28}(2), 149--171.
+\url{https://doi.org/10.1177/0959354318761606}
 
@@
 \bibitem[\citeproctext]{ref-meehlTheoreticalRisksTabular1978}
-Meehl, P.
-E.
+Meehl, Paul E.
 (1978).
@@
 \emph{Journal of Consulting \& Clinical Psychology}, \emph{46}(4), 806--834.
+
+\bibitem[\citeproctext]{ref-meehlAppraisingAmendingTheories1990}
+Meehl, Paul E.
+(1990).
+Appraising and {Amending Theories}: {The Strategy} of {Lakatosian Defense} and {Two Principles} that {Warrant It}.
+\emph{Psychological Inquiry}, \emph{1}(2), 108--141.
+\url{https://doi.org/10.1207/s15327965pli0102_1}
 
@@
 \url{https://doi.org/10.1111/j.1467-9507.2007.00389.x}
+
+\bibitem[\citeproctext]{ref-norouziCapturingCausalClaims2024}
+Norouzi, R., Kleinberg, B., Vermunt, J., \& Van Lissa, C.
+J.
+(2024).
+\emph{Capturing {Causal Claims}: {A Fine Tuned Text Mining Model} for {Extracting Causal Sentences} from {Social Science Papers}}.
+Retrieved from \url{https://osf.io/kwtpm/download}
 
@@
 
+\bibitem[\citeproctext]{ref-vanlissaTeacherCornerEvaluating2020}
+Van Lissa, C.
+J., Gu, X., Mulder, J., Rosseel, Y., Zundert, C.
+V., \& Hoijtink, H.
+(2020).
+Teacher's {Corner}: {Evaluating Informative Hypotheses Using} the {Bayes Factor} in {Structural Equation Models}.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{0}(0), 1--10.
+\url{https://doi.org/10.1080/10705511.2020.1745644}
+
 \bibitem[\citeproctext]{ref-vogtFAIR20Extending2024}

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
-Since their inception, scholars have demonstrated their relevance for making other information artefacts more open.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
+Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprecht2020towards; @barker2022introducing], computational workflows [@wilkinson2024applying] or simulation studies [@amaro2024need].
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
-a digital representation of theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR Theory has the potential to accelerate cumulative knowledge acquisition and advance social scientific research.
+a digital representation of a scientific theory, compliant with the FAIR principles.
+By improving the efficiency of scholarly communication, FAIR Theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
 

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 These various practices ensure transparent and repeated testing of hypotheses.
-However, recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021].
+However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
 Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
 psychological theories rarely produce hypotheses that are found to be true,
-and are often sufficiently vague to explain contradictory findings.
+and are often sufficiently vague, such that contradictory findings do not falsify the central claims.
+<!--AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
 
-Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
+Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
 Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
+<!-- AB is it really a lack of clarity, ie.
+that everyone understands it?
+I think the core problem is a lack of precision, ie.
+accuracy in their predictions.
+Maybe a lot of theories are actually too clear but not precise; making them more precise also often makes them less clear, doesn't it?-->
 In other words, social scientific theories lack *formalization* .
-A second concern is the lack of transparent and democratic scholarly communication about psychological theory.
+A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
 
 Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and -improvement have been largely overlooked.
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
 The present paper restores balance by applying, for the first time,
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena by successively increasing the empirical content of a theory.
+In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
 There are, however, indications that contemporary psychology falls short of this ideal.

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 Firstly, because deductive research is over-represented in the literature: 
-According to one estimate, 89.6% of paper published in psychology tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
-Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
+According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
+Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
 The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
+In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+Or to adapt an analogy of Roji, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.
 

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-Or to adapt an analogy of Roji, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.
+Or to adapt an analogy of @van2021theory, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.
 

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 
-The present paper addresses the lack of open science methods for theory development and improvement by applying the FAIR principles to scientific theories.
-Merely publishing theory does not make it open;
+The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
+Merely publishing theory -- as in a classic research article -- does not make it open;
 to be open, theory should adhere to established open science standards.
-The FAIR principles, initially introduced for open research data,
-have since been applied to other information artifacts [@lamprechtFAIRPrinciplesResearch2019].
+<!--AB  I deleted this sentence because it was already in the introduction -->
 We apply the FAIR principles to digital representations of theory,
@@
 *Interoperable* within the data analysis environment,
-and *Reusable* in the practical and legal sense, so that they may be improved over time.
-Following Lamprecht and colleagues,
+and *Reusable* in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
+Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
+<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover-->
+Following the original proposal of Lamprecht and colleagues,
 we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 and many of those definitions are consistent with the FAIR theory principles.
-This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward].
+This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward]. <!-- what is this ref exactly?! -->
 <!-- Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried]. -->
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
+Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
+From this definition, it becomes clear that FAIR models are a special case of FAIR theory.
+Yet, theory intentionally has a broader scope and may contain one or more models as specific instances.
+For example, a comprehensive theory of disease spread and pandemics may cover various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety).
+This may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 Yet, theory intentionally has a broader scope and may contain one or more models as specific instances.
-For example, a comprehensive theory of disease spread and pandemics may cover various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety).
-This may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+For example, a comprehensive theory of disease spread and pandemics may cover various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
+The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' *empirical content* [REF],
-making them easier to falsify,
+Greater formalization increases theories' *empirical content* [REF] as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes  that would otherwise get away with their vague formulation in verbal theories.
+For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented [@lewandowsky2010computational].
+More precise theories makes them easier to falsify,
 which necessitates revising them,

### 2024-12-12 manuscript.Rmd
Update manuscript.Rmd

@@
 To curb this trend, we suggest using the keyword `"FAIR theory"` for all resources that constitute or reference a FAIR theory.
+<!-- this is problematic because many search engine will tear apart this term and interpret it as two keywords - I think we need a single word without space FAIRtheory maybe -->
 This would allow theoretical resources to be systematically indexed, tagged, and made searchable.

### 2024-12-13 manuscript.Rmd
Update manuscript.Rmd
add a sentence to tell people: don't panic

@@
 and formal theory can be represented in a way that is not FAIR.
+It is therefore important for us to emphasize that FAIR Theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
+The guidelines introduced by FAIR Theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
 For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.

### 2024-12-16 manuscript.Rmd
some more thoughts

@@
 Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
-Post-hoc curation does not address the root cause of the lack of Findability, however.
-Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
+Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
+Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
 
@@
 Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
+This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
@@
 and Zenodo allows for access control of individual files/resources.
+An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics).
+In general, it makes sense to keep outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
 
@@
 
-A third impediment arises when theories have a "dependency on the author" (DOA).
+A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
 DOA occurs when a theory cannot be understood by independent scholars,
@@
 such as verbal statements in plain text,
-mathematical formulae,
+mathematical formulae, <!-- AB: US readers may expect formulas here ? not sure-->
 and statements expressed in some axiomatic system.
@@
 While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language for representing graphs does meet this ideal.
+the same image represented in the DOT language <!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
 
@@
 Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
-As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
+As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved), <!-- AB : maybe add a caveat that there may be country-specific legislation, company property rights, etc.
+and that this is of course no legal advice-->
 although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].

### 2024-12-16 manuscript.Rmd
final thoughts

@@
 
-## Some other stuff
+## Additional considerations
 
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code.
-Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code (which we also have successfully applied in the domain of reproducible research findings, see XXX).
+<!--AB cite WORCS and Aarons tutorial on how to do the same thing again..-->
+Using version control systems, like git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
 Git also facilitates diffuse and adversarial collaboration,
@@
 
-Making Theories FAIR Accelerates Scientific Progress
+### Making Theories FAIR Accelerates Scientific Progress
 Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
@@
 In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
-As several authors have taken inspiration from the work by De Groot,
+As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
 we compare our interpretation of the original theory to the interpretation of others.
@@
 ```
+
+<!--AB TODO: Also add images of the graph syntax - not everyone will be able to read this in syntax and see the graph in their minds-->
 
@@
 One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
+Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].

### 2024-12-17 manuscript.Rmd
turn into list

--- 2024-12-13 manuscript.Rmd
@@
 bibliography      : "theory-specification.bib"
-floatsintext      : yes
+floatsintext      : no
 linenumbers       : yes
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
-Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprecht2020towards; @barker2022introducing], computational workflows [@wilkinson2024applying] or simulation studies [@amaro2024need].
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
-To this end, we introduce "FAIR theory":
-a digital representation of a scientific theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR Theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
-
-<!-- QUOTES
-
+<!-- QUOTES -->
 Kurt Lewin: Nothing is as practical as a good theory
-
+Paul Meehl: [Psychological theories are neither] refuted nor corroborated, but instead merely fade away as people lose interest [@meehlTheoreticalRisksTabular1978].
+[Theories are] like toothbrushes — no self-respecting person wants to use anyone else’s [@mischelToothbrushProblem2008]
 "Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger.
@@
 
-END QUOTES -->
+<!-- END QUOTES -->
 
-## The Need for FAIR Theory
+The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
+Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+But have these reforms met their goal?
+Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
+which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
+If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.
 
-The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
-These various practices ensure transparent and repeated testing of hypotheses.
-However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
-Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-psychological theories rarely produce hypotheses that are found to be true,
-and are often sufficiently vague, such that contradictory findings do not falsify the central claims.
-<!--AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
+Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
+There is increasing consensus that one contributing cause is a lack of theory formalization: social scientific theories often lack the precision and clarity of theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
+A second issue, which received less attention, is the a lack of transparent and democratic scholarly communication about psychological theory.
+The present paper seeks to address this issue by applying open science principles, for the first time, to psychological theory, and introducing the concept of *FAIR Theory*.
 
-Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
-<!-- AB is it really a lack of clarity, ie.
-that everyone understands it?
-I think the core problem is a lack of precision, ie.
-accuracy in their predictions.
-Maybe a lot of theories are actually too clear but not precise; making them more precise also often makes them less clear, doesn't it?-->
-In other words, social scientific theories lack *formalization* .
-A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
-
-Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
-The present paper restores balance by applying, for the first time,
-open science principles to psychological theory.
-We apply the FAIR principles to scientific theories,
-introducing the concept of *FAIR Theory* to 
-facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
+FAIR Theory incorporates theory into open science workflows,
+facilitates scholarly communication about theories,
+making it easier to share theories with less opportunity for ambiguity and misunderstanding.
+FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
+More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
+removes barriers for knowledge exchange with the global scholarly community,
+opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
 
@@
 
-According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
-a philosophical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
+Theories reflect scientists' understanding of phenomena.
+Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
+This process is described in the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
+a model that describes cumulative knowledge acquisition as a cyclical process with two phases (Figure \@ref(fig:figec)).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
-In this model, theories are the vehicle of scientists' understanding of phenomena.
-Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
 
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena by successively increasing the empirical content of a theory.
-In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
-There are, however, indications that contemporary psychology falls short of this ideal.
-Firstly, because deductive research is over-represented in the literature: 
-According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
-The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
-Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
-Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-Or to adapt an analogy of @van2021theory, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.
+the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
 
-Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
-The majority of papers ostensibly test hypotheses,
-but these are rarely derived from theory,
-and test results do not routinely contribute to the improvement of existing theories.
-The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
+When comparing common practice in contemporary psychological research to this idealized model, at first glance,
+it appears that deductive research is overrepresented in the literature. 
+An overwhelming 89.6% of published studies ostensibly tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
+Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
+Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
+Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
+Existing theories either persist unchanged, or are forgotten [REF Meehl].
 
-<!-- First, we reflect on the differences between theory and data. <!-- CJ: Should we? --> -->
-<!-- Second, we adapt the FAIR principles for theory, respecting these differences. -->
-<!-- Third, we discuss the current state and future directions of psychological theory with respect to each criterion. -->
-<!-- Fourth, we present worked examples of theory made FAIR. -->
-<!-- We conclude with a summary and future directions for FAIR theory. -->
+Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
+The present paper applies, for the first time, open science principles to theory.
 
-## Making Theory FAIR
+In our experience, there is no clear delineation between paradigms, frameworks and theories. 
+We think it is helpful to view these terms as discrete labels on a continuum from broad to narrow, from general to specific, when organizing knowledge.
+Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
+Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
 
-The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
-Merely publishing theory -- as in a classic research article -- does not make it open;
-to be open, theory should adhere to established open science standards.
-<!--AB  I deleted this sentence because it was already in the introduction -->
-We apply the FAIR principles to digital representations of theory,
-introducing a FAIR metadata format to make theories *Findable* via a DOI,
-*Accessible* in a machine- and human-readable filetype,
-*Interoperable* within the data analysis environment,
-and *Reusable* in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
-Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
-<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover-->
-Following the original proposal of Lamprecht and colleagues,
-we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
-We reflect on the necessary changes (which are minor),
-as well as on the current state and future of FAIR theory in the social sciences.
-The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
-and we provide worked examples to encourage their adoption.
+## Publication is not Enough
 
-```{r tabfair}
-tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
-names(tab_fair) <- c("Criterion", "Original", "Theory", "Action")
-papaja::apa_table(tab_fair, landscape = TRUE, align = "m{.1\\linewidth}m{.35\\linewidth}m{.35\\linewidth}m{.15\\linewidth}")
-```
+Merely publishing a theory does not make it open;
+open theory should adhere to established open science standards.
+The FAIR principles, initially introduced as a standard for open research data, have since been applied to other forms of digital scholarly output [e.g., software @lamprechtFAIRPrinciplesResearch2019].
+We propose to apply the FAIR principles to digital representations of theory as well,
+introducing a FAIR metadata format to represent (formal) theories.
+The resulting theories are made *Findable* via a DOI,
+are Accessible in a machine- and human-readable filetype,
+Interoperable within the data analysis environment,
+and Reusable in the practical and legal sense, using semantic versioning and permissive licensing.
 
-There are different definitions of theory,
-and many of those definitions are consistent with the FAIR theory principles.
-This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward]. <!-- what is this ref exactly?! -->
-<!-- Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried]. -->
-<!-- This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-There is no principled difference, but a FAIR theory is the most general instance of a theory, whereas a model is highly specific - e.g., by assuming functional form and error families.
-We can make a distinction between statistical models - which make just enough assumptions to allow estimation of the remaining unknown parameters from data - and generative/computational models, which are completely parametrized to the point that a specific interpreter can use them to generate new data.
-To some extent, -->
-Meehl [-@meehlAppraisingAmendingTheories1990] provides guidance as to what kinds of "statements" such a theory might contain:
-statements about the types of entities postulated (i.e., ontology),
-statements about causal connections between those entities,
-statements about the functional form of those connections,
-and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
-Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
-From this definition, it becomes clear that FAIR models are a special case of FAIR theory.
-Yet, theory intentionally has a broader scope and may contain one or more models as specific instances.
-For example, a comprehensive theory of disease spread and pandemics may cover various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
-The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+## Adapting the FAIR Principles
 
-### The Role of Theory Formalization
+The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
+From their inception, these principles were developed with "other research resources" in mind.
+Since then, scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
+The present paper further extends the FAIR principles' definition to theory.
+Doing so requires first representing theory as a digital research artefact, and then making it compliant with the FAIR principles.
 
-Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' *empirical content* [REF] as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes  that would otherwise get away with their vague formulation in verbal theories.
-For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented [@lewandowsky2010computational].
-More precise theories makes them easier to falsify,
+## Is Current Psychological Theory FAIR?
+
+* Is it findable?
+There is no unified search engine for theory, and not even an agreed-upon search term (model, framework, etc are used interchangeably with theory).
+Some have sought to address findability through post-hoc curation of theories (e.g., theory mapping website, Borsboom's dictionary of psychological phenomena).
+* Is it accessible?
+Paywalls, partly implicit (Great Man Theorizing / Call Leon), certainly not accessible in the sense of theory being "top of mind", as researchers appear to prefer ad-hoc hypothesizing rather than deriving hypotheses from specific theories.
+* Is it interoperable? Theories rarely change, 
+* Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).
+
+In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
+From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
+Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.
+
+FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
+
+There are key distinctions between theory and other FAIR digital research artefacts.
+With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.
+
+## The Role of Theory Formalization
+
+Concerns about the state of theory are a recurring theme in the psychological literature,
+but previous writing has focused on theory formalization as a solution for ambiguity in psychological theory.
+Greater formality increases theories' *empirical content*,
+making them easier to falsify,
 which necessitates revising them,
 thus advancing our principled understanding of the phenomena they describe.
-FAIR Theory does not require theories to be formal,
-and formal theory can be represented in a way that is not FAIR.
-It is therefore important for us to emphasize that FAIR Theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
-The guidelines introduced by FAIR Theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
-For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
-Conversely, a directed acyclic graph (DAG) is a type of formal theory,
-but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
-then this formal theory is not FAIR.
-FAIR Theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
-
-## Findability
-
-Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
-grounding their work in established theoretical foundations.
-Making theories Findable also increases the impact and reuse potential of theories across disciplines,
-either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
-or through analogical modeling.
-In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
-For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
-Findability also enables meta-research on theories, 
-in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
-In a similar way, it would become possible to compare all theories of a specific phenomenon,
-or to study structural properties of theories.
-
-The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
-First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
-Of the many services that provide DOIs for scientific information artefacts,
-Zenodo and the Open Science Framework are commonly used in psychology.
-Second, Findable theory is described with rich metadata	(F2).
-This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
-It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
-ontology [@guyonMeasurementOntologyEpistemology2018],
-or catalog of psychological phenomena [REF Noah Denny].
-Metadata should also include identifiers for all the versions of the theory it describes	(F3);
-Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
-Finally, these metadata should be registered or indexed in a searchable registry (F4).
-This final criterion is less straightforward.
-Ideally, FAIR theories should be indexed in search engines used by academics, like Google Scholar.
-At present, however, these search engines are designed to index traditional print publications.
-The *data paper* solves this problem for research data;
-the idea is that scholars publish a paper (or even preprint) as documentation for the data resource [REF McGIllivray on data papers].
-The data paper is indexed by search engines, and in turn points to the relevant information artifact.
-The same solution could be applied to theories - but it seems superfluous to generate papers whose only purpose is to redirect to a specific resource.
-Another solution is to manually index FAIR theories,
-for example by adding them to one's Google Scholar profile,
-or entering them in PURE.
-<!-- CJ: Can we guarantee automatically indexing FAIR theory somehow? -->
-
-At present, theories have poor findability, which impedes cumulative knowledge acquisition.
-One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like "theory", "model", and "framework" are used interchangeably.
-To curb this trend, we suggest using the keyword `"FAIR theory"` for all resources that constitute or reference a FAIR theory.
-<!-- this is problematic because many search engine will tear apart this term and interpret it as two keywords - I think we need a single word without space FAIRtheory maybe -->
-This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
-Another factor contributing to the present lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
-A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
-This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
-similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
-As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
-
-There have been notable efforts to improve theories' findability through post-hoc curation.
-For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website [@grayHowMapTheory2017].
-Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
-Post-hoc curation does not address the root cause of the lack of Findability, however.
-Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
-
-## Accessibility
-
-Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-If theories are accessible, researchers can reuse and refine them,
-thus accelerating cumulative knowledge acquisition.
-Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
-While isolated empirical findings can appear fragmented and contradictory [@dumas-malletPoorReplicationValidity2017],
-theories offer a top-down, big picture representation of the phenomena studied in a field.
-In other words, theories are an important instrument in science communication.
-
-The Accessibility criteria serve to *regulate* access, not to maximize it.
-These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
-Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
-The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
-Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
-which is also achieved via long-term storage (e.g., on Zenodo).
-Git remote repositories allow for access control,
-and Zenodo allows for access control of individual files/resources.
-
-At present, there are several impediments to theories' accessibility.
-To the extent that theories are still contained within papers,
-paywalls erected by commercial publishers constitute a barrier.
-Open Access publishing thus increases the accessibility of all academic output, including theory.
-A second impediment is more indirect:
-While open access publishing increases practical access to theories,
-accessibility also requires clear and explicit communication.
-This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
-The current prevalence of strategic ambiguity renders psychological theory difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
-It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
-which holds that every communicative utterance has multiple alternative translations, with no *objective* means of choosing the correct one.
-It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
-This places a theoretical upper bound on theories' ability to be understood.
-
-Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
-The Kuhnian notion of "normal science", conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding [@kuhnStructureScientificRevolutions2009].
-From a pragmatic perspective, these considerations indicate that,
-when striving to make theory accessible,
-it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions),
-while acknowledging that accessibility exists on a spectrum,
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.
-
-A third impediment arises when theories have a "dependency on the author" (DOA).
-DOA occurs when a theory cannot be understood by independent scholars,
-thus requiring the original author for interpretation and clarification.
-We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
-as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
-DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
-which violates checks and balances of scientific research.
-DOA also renders theories immune to refutation,
-because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
-The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
-This demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
-
-<!-- The data usually need to be integrated with other data. In addition, the data need to interoperate with applications or workflows for analysis, storage, and processing. -->
-
-## Interoperability
-
-Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
-The original interoperability principles can be rephrased somewhat to apply to theory.
-Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
-The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
-Instead, FAIR theory should, ad minimum,
-be instantiated as 
-as a type of data that is human- and machine-readable with as few interpretative steps as possible,
-as previously recommended [@vanlissaWORCSWorkflowOpen2021].
-Depending on the level of formalization of the theory,
-different formats may be appropriate,
-such as verbal statements in plain text,
-mathematical formulae,
-and statements expressed in some axiomatic system.
-Examples of the latter include pseudo-code,
-interpretable computer code,
-and Gray's theory maps [@grayHowMapTheory2017].
-While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language for representing graphs does meet this ideal.
-
-Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
-This is essentially a call to establish standardized ontologies,
-which are themselves a type of theory [@meehlAppraisingAmendingTheories1990]. 
-Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
-The first part of this principle allows for nested theories;
-for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
-The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
-This can be achieved by using Git for version control and Zenodo for archiving.
-
-As the original definition of interoperability was somewhat narrow [@wilkinsonFAIRGuidingPrinciples2016a],
-the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
-This definition invites the question: *interoperable for what?*
-Suitable answers for FAIR theory may be:
-this theory is X-interoperable for deriving testable hypotheses,
-or for the purpose of selecting relevant control variables,
-or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-This revised definition implies that theories have specific properties that incur affordances in terms of X-interoperability;
-for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form).
-
-```{r tabmeehl}
-data.frame(
-  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
-  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
-) |>
-papaja::apa_table()
-```
-
-<!-- Do we want to say something about this?
-Interoperability of psychological theory may be limited when theories are instantiated in natural language without consideration of ontological complexities. -->
-<!-- or as visualizations without a straightforward interpretation -->
-<!-- What can go wrong?
-A well-formed sentence may be logically inconsistent (e.g., the theory that conceptualized father involvement simultaneously as mediator and moderator); a picture may seem sensible but there's no straightforward way to translate it to hypotheses, etc.
--->
-<!-- which are not machine-readable and l -->
-
-With regard to the state of interoperability in contemporary psychology, 
-Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
-But, as we argued, this is not the case.
-The examples of X-interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
-For example, interoperable theory could be used
-to select control variables for causal inference [@cinelliCrashCourseGood2022],
-or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
-or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
-Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
-For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
-
-## Reusability
-
-If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
-As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
-although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].
-Criterion R1.2 is met by version control with Git and archival on Zenodo.
-Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
-
-If we consider the current state of Reusability in psychological theory, there appears to be a norm *against* theory reuse:
-*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
-This norm impedes scientific progress.
-Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
-<!-- Theories are made reusable by thorough documentation, -->
-<!-- appropriate licensing,  -->
-<!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
-In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
-reflected in questions such as "who owns theory",
-and "who determines how a theory may be reused or changed"?
-These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
-
-Licensing theories for reuse provides an unambiguous answer to such questions.
-In determining what license is appropriate for theory,
-it is important to consider that copyright law limits authors' rights based on the *idea-expression dichotomy* [@bently2010copyright],
-which holds that copyright explicitly does not
-*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
-Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
-It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
-then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
-Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
-so copyright would pose few restrictions to Reusability in scholarly communication.
-Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
-it explicitly waives all rights and encourages reuse.
-In principle, CC0 does not require attribution.
-Nevertheless, is essential that scholars do comprehensively cite theory,
-including prior work that new theories are based on,
-even in absence of legal obligations to do so,
-to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].
-
-## Some other stuff
-
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code.
-Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
-Git also facilitates diffuse and adversarial collaboration,
-as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
-In sum, version control using Git enables systematic, collaborative, and transparent theory development,
-enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
-
-Even if scholars wish to diverge substantially from prior theory,
-explicitly referring back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
-
-From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
-
-Making Theories FAIR Accelerates Scientific Progress
-Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
-By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
-This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
-
-<!-- ## FAIR Theory and Recognition & Rewards -->
-
-<!-- FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory. -->
-<!-- In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging. -->
+Conceptually, theory formalization is orthogonal to FAIR theory.
+FAIR Theory does not require theories to be formal, and formal theory can be represented in a way that is not FAIR.
+It is - in principle - possible to represent a collection of verbal statements as a FAIR Theory.
+While FAIR Theory is fully consistent with formal theory, it does not require theories to be formal.
 
@@
 
-## Formalizing the Empirical Cycle
-
-In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
-As several authors have taken inspiration from the work by De Groot,
-we compare our interpretation of the original theory to the interpretation of others.
-Originally, the theory has the following structure:
-
-```
-
-digraph {
-
-  observation;
-  induction;
-  deduction;
-  test;
-  evaluation;
-
-  observation -> induction;
-  induction -> deduction;
-  deduction -> test;
-  test -> evaluation;
-  evaluation -> observation;
-
-}
-```
-
-Subsequently, Wagenmakers and colleagues modified the theory by *"[adding the] Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*:
-
-```
-digraph {
-
-  subgraph cluster_discovery {
-    label="Discovery";
-    hypothesis [label="New hypothesis"];
-    prediction [label="New prediction"];
-  }
-  data  [label="Old knowledge and old data"];      
-  subgraph cluster_justification {
-    label="Justification";
-    test [label="Test on new data"];
-    evaluation;
-  }
-
-  data -> hypothesis [label="Speculate & explore"];
-  hypothesis -> prediction  [label="Deduce"];
-  prediction -> test  [label="Design new experiment"];
-  test -> evaluation  [label="Statistical analysis"];
-  evaluation -> data  [label="Knowledge accumulation"];
-
-}
-```
-
-Note, however, that there appear to be further changes:
-the phases of the cycle have been renamed,
-and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
-Moreover, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
-As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-
-Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:
-
-```
-digraph {
-
-  theory;
-  prediction;
-  test [label="inferential procedure"];
-  observation;
-
-  theory -> prediction [label="deduction"];
-  prediction -> test;
-  test -> observation;
-  observation -> theory [label="generalization"];
-
-}
-```
-
-In our representation,
-induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
-For example, the refutation of a hypothesized effect,
-or the serendipitous observation of some pattern in data,
-might be a reason to revise or construct theory.
-Induction, incidentally, also occurs within the link from prediction to testing:
-in the form of the inductive bias of methods used to perform the test,
-and auxiliary assumptions that must be made to address remaining theoretical ambiguities.
-
-<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced.
--->
-<!-- ``` -->
-
 ## Using FAIR Theory to Perform Causal Inference
 
-Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
-According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
-Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
-and could even be integrated into the analysis environment.
-
-In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
-We could use that to illustrate updating FAIR theory:
-
-https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45
-
-We can find more examples of causal inference with DAGs in these tutorials:
-
-https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/
-
-https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/
+Based on this example: https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/
 
@@
 
-# Discussion
-
-## Future Directions
-
-One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
-it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
-Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
-the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
-FAIR Theory can help address the measurement crisis:
-since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
-
-FAIR Theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers for knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
-
 \newpage

### 2024-12-17 manuscript.Rmd
resolve Andreas' comments

--- 2024-12-16 manuscript.Rmd
@@
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
-Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-psychological theories rarely produce hypotheses that are found to be true,
-and are often sufficiently vague, such that contradictory findings do not falsify the central claims.
-<!--AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
+Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+Psychological theories rarely produce hypotheses that are corroborated by evidence.
+Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
+as the theory's central claims evade falsification.<!--CJ: Resolved AB's comment?
+AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
+<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
 
 Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
-<!-- AB is it really a lack of clarity, ie.
-that everyone understands it?
-I think the core problem is a lack of precision, ie.
-accuracy in their predictions.
-Maybe a lot of theories are actually too clear but not precise; making them more precise also often makes them less clear, doesn't it?-->
-In other words, social scientific theories lack *formalization* .
+Two main concerns are that, first, social scientific theories lack precision compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
+ and clarity
+In other words, social scientific theories lack *formalization*,
+which means that they do not make very accurate predictions,
+and are thus hard to falsify and difficult to understand on their own,
+without either substantial interpretation or additional background knowledge.
 A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena by successively increasing the empirical content of a theory.
-In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
+<!-- by successively increasing the empirical content of a theory. -->
+<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
+<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
+This needs more work! -->
+.
 There are, however, indications that contemporary psychology falls short of this ideal.
@@
 One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like "theory", "model", and "framework" are used interchangeably.
-To curb this trend, we suggest using the keyword `"FAIR theory"` for all resources that constitute or reference a FAIR theory.
-<!-- this is problematic because many search engine will tear apart this term and interpret it as two keywords - I think we need a single word without space FAIRtheory maybe -->
+To curb this trend, we suggest using the keyword `"FAIRtheory"` for all resources that constitute or reference a FAIR theory (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines.
 This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
@@
 Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
-Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
-Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
+Post-hoc curation does not address the root cause of the lack of Findability, however.
+Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
 
@@
 Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
+This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
@@
 and Zenodo allows for access control of individual files/resources.
-An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics).
-In general, it makes sense to keep outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
 
@@
 
-A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
+A third impediment arises when theories have a "dependency on the author" (DOA).
 DOA occurs when a theory cannot be understood by independent scholars,
@@
 such as verbal statements in plain text,
-mathematical formulae, <!-- AB: US readers may expect formulas here ? not sure-->
+mathematical formulae,
 and statements expressed in some axiomatic system.
@@
 While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language <!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
+the same image represented in the DOT language for representing graphs does meet this ideal.
 
@@
 Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
-As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved), <!-- AB : maybe add a caveat that there may be country-specific legislation, company property rights, etc.
-and that this is of course no legal advice-->
+As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
 although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].

### 2024-12-19 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Felix Schönbrodt <felix@nicebread.de>

@@
 The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 These various practices ensure transparent and repeated testing of hypotheses.

### 2024-12-19 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Felix Schönbrodt <felix@nicebread.de>

@@
       - Writing – review & editing
-  - name: "Felix Schönbrodt"
+  - name: "Felix D. Schönbrodt"
     affiliation: "5"

### 2024-12-19 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Felix Schönbrodt <felix@nicebread.de>

@@
   - id            : "5"
-    institution   : "Other affiliations"
+    institution   : "Ludwig-Maximilians-Universität München, Germany"
 authornote: |

### 2024-12-19 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Felix Schönbrodt <felix@nicebread.de>

@@
 
-Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
 The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.

### 2024-12-19 manuscript.Rmd
update 2024-12-19 11:16:18.873357

--- 2024-12-19 manuscript.Rmd
@@
 ---
-title             : "FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
+title             : "FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
 shorttitle        : "FAIR THEORY"
@@
       - Writing – review & editing
-  - name: "Felix D. Schönbrodt"
+  - name: "Felix Schönbrodt"
     affiliation: "5"
@@
 a digital representation of a scientific theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR Theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
+By improving the efficiency of scholarly communication, FAIR theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
 
@@
 
-## The Need for FAIR Theory
+## The Need for FAIR theory
 
 The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 These various practices ensure transparent and repeated testing of hypotheses.
@@
 We apply the FAIR principles to scientific theories,
-introducing the concept of *FAIR Theory* to 
+introducing the concept of *FAIR theory* to 
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 There are, however, indications that contemporary psychology falls short of this ideal.
-Firstly, because deductive research is over-represented in the literature: 
+Firstly, because hypothesis-testing research is over-represented in the literature: 
 According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
@@
 thus advancing our principled understanding of the phenomena they describe.
-FAIR Theory does not require theories to be formal,
+FAIR theory does not require theories to be formal,
 and formal theory can be represented in a way that is not FAIR.
-It is therefore important for us to emphasize that FAIR Theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
-The guidelines introduced by FAIR Theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
+It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
+The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
 For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
@@
 then this formal theory is not FAIR.
-FAIR Theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
+FAIR theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
 
@@
 <!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
-In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
+In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,
 reflected in questions such as "who owns theory",
@@
 so copyright would pose few restrictions to Reusability in scholarly communication.
-Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
+Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR theory;
 it explicitly waives all rights and encourages reuse.
@@
 
-<!-- ## FAIR Theory and Recognition & Rewards -->
+<!-- ## FAIR theory and Recognition & Rewards -->
 
@@
 
-## Using FAIR Theory to Perform Causal Inference
+## Using FAIR theory to Perform Causal Inference
 
@@
 
-One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
+One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
 Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
@@
 the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
-FAIR Theory can help address the measurement crisis:
+FAIR theory can help address the measurement crisis:
 since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
 
-FAIR Theory incorporates theory into open science workflows,
+FAIR theory incorporates theory into open science workflows,
 facilitates scholarly communication about theories,

### 2024-12-19 manuscript.Rmd
update 2024-12-19 11:36:52.824189

@@
 and many of those definitions are consistent with the FAIR theory principles.
-This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward]. <!-- what is this ref exactly?! -->
-<!-- Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried]. -->
-<!-- This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-There is no principled difference, but a FAIR theory is the most general instance of a theory, whereas a model is highly specific - e.g., by assuming functional form and error families.
-We can make a distinction between statistical models - which make just enough assumptions to allow estimation of the remaining unknown parameters from data - and generative/computational models, which are completely parametrized to the point that a specific interpreter can use them to generate new data.
-To some extent, -->
+This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward].
 Meehl [-@meehlAppraisingAmendingTheories1990] provides guidance as to what kinds of "statements" such a theory might contain:
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
+
 Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
-From this definition, it becomes clear that FAIR models are a special case of FAIR theory.
-Yet, theory intentionally has a broader scope and may contain one or more models as specific instances.
-For example, a comprehensive theory of disease spread and pandemics may cover various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
+This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
+There is no principled difference;
+theories and models derived from it exist along a continuum of specificity, where
+a theory has a relatively broader scope and may contain one or more models as specific instances.
+For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
+From this theory, we could derive a more specific *statistical model* by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
+This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+Or, we could derive an even more specific *generative/computational model*, which is completely parametrized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Note that broadness and narrowness are relative terms,
+and one person's theory may be another person's model.
+This definition also implies that FAIR models are a special case of FAIR theory.
+
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
 The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

### 2024-12-19 manuscript.Rmd
update 2024-12-19 11:45:49.650496

@@
 Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
-<!-- Theories are made reusable by thorough documentation, -->
-<!-- appropriate licensing,  -->
-<!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
 In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,

### 2024-12-27 manuscript.Rmd
expand examples, draft vignette

@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
-Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprecht2020towards; @barker2022introducing], computational workflows [@wilkinson2024applying] or simulation studies [@amaro2024need].
+Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
@@
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-Or to adapt an analogy of @van2021theory, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.
+Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].
 
@@
 For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
-Conversely, a directed acyclic graph (DAG) is a type of formal theory,
+Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \@ref(fig:figec)),
 but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
-then this formal theory is not FAIR.
+it is not FAIR.
 FAIR theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
@@
 
-# Examples
+# Making a Theory FAIR
 
-## Formalizing the Empirical Cycle
+Open science infrastructure is an area of active development, and as such,
+the approach proposed here should not be considered definitive,
+but rather, as one proposal for a FAIR-compliant implementation of theory using infrastructure available at the time of writing.
 
-In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
-As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
-we compare our interpretation of the original theory to the interpretation of others.
-Originally, the theory has the following structure:
+<!-- <!-- The `worcs` R-package documentation contains a digital version of this tutorial, --> -->
+<!-- <!-- to allow for future updates. --> -->
+
+<!-- 1. Create project folder -->
+<!-- 1. Add theory file -->
+<!-- 1. Add LICENSE file -->
+<!-- 1. Add readme.md -->
+<!--     * Include project Title, Description, Usage of the theory file, How to Contribute, How to Cite the FAIR theory, Related Works (e.g., the paper that documents the theory) -->
+<!-- 1. Add .zenodo.json file to provide metadata that will allow Zenodo to index the repository as a FAIR theory -->
+<!-- 1. Initialize a Git repository in the project folder -->
+<!-- 1. Create Git remote repository (e.g., on GitHub) -->
+<!-- 1. Connect local to remote Git repository -->
+<!-- 1. Add the repository to Zenodo -->
+<!--     * Sync Zenodo to import recent GitHub repositories (may take some time before it shows) -->
+<!--     * Flip Switch -->
+<!-- 1. Publish a release on GitHub -->
+<!-- 1. *Optional:* Edit Zenodo metadata (e.g., to add formal cross-references to related works) https://help.zenodo.org/docs/deposit/manage-records/#edit -->
+
+<!-- https://developers.zenodo.org/#add-metadata-to-your-github-repository-release` -->
+
+## 1. Implementing the Theory
+
+Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
+we decided to use it as an example for this tutorial.
+Note that, while the empirical cycle is not explicitly referred to as a "theory" by De Groot, he derives it from "a theory of thinking" by Selz (REF).
+We can thus consider it a meta-theory of the process of theory construction.
+<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
+The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969:
+
+> *Phase 1:* 'Observation': collection and grouping of empirical materials; (tentative) formation of hypotheses.  
+*Phase 2:* 'Induction': formulation of hypotheses.  
+*Phase 3:* 'Deduction': derivation of specific consequences from the hypotheses, in the form of testable predictions.  
+*Phase 4:* 'Testing': of the hypotheses against new empirical materials, by way of checking whether or not the predictions are fulfilled.  
+*Phase 5:* 'Evaluation': of the outcome of the testing procedure with respect to the hypotheses or theories stated, as well as with a view to subsequent, continued or related, investigations.
+
+If we compare it to the levels of theory formalization [@guestHowComputationalModeling2021],
+De Groot's theory is either at the "theory" or "specification" level.
+It consists of a series of natural language statements.
+We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language:
 
 ```
-
 digraph {
@@
 ```
+
+This language describes the model as a directed graph.
+Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
+and the second half describes their proposed interrelations.
+This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
+
+We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
+
+## 2. Creating a Project Folder
+
+Create a new folder and copy the theory file from the previous step into it.
+To help meet the Interoperability and Reusability criteria,
+add two more files:
+A README.md file with instructions for future users of your theory,
+and a LICENSE file with the legal conditions for reuse.
+We recommend using the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
+
+### What's in a README?
+
+The readme should contain information to help people get started with using your FAIR theory.
+We suggest the following elements:
+
+* Title, prefaced with `# FAIR theory: The Theory's Name`
+* Description: A plain-text description of the theory and its scope
+* Interoperability: This section, also often labeled "Getting Started", "Instructions", or "How to Use" might be better labeled Interoperability from a FAIR perspective.
+Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
+* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
+* License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal constraints to reuse*.
+* Citing this work: Tell users how to cite the theory.
+* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
+Note that this is redundant with metadata entered in Zenodo (step 5).
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+The disadvantage is that you may have to maintain this section of the readme.
+A compromise might be to use this section to document one canonical reference for the theory that is unlikely to change going forward.
+For example, we referenced the original empirical cycle paper here:
+
+```
+This repository contains an implementation of the "empirical cycle",
+a model proposed by De Groot and Spiekerman (1969, p. 28)
+
+> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
+Foundations of inference and research in the behavioral sciences.
+De Gruyter Mouton. https://doi.org/10.1515/9783112313121
+```
+
+## 3. Version Control the Repository
+
+The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
+In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013, @vanlissaWORCSWorkflowOpen2021].
+To version control our project, we initiate a Git repository in the project folder.
+This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally,
+opening a terminal in the project folder,
+and running the command `git init`.
+We subsequently create a remote repository to host a copy of this local Git repository, which will in turn be archived.
+To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`.
+The default settings for the repository are fine;
+for the next steps of the tutorial it must be set to "Public".
+Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`.
+Next, in the terminal window, run the following lines of code sequentially:
+
+```
+git add .
+git commit -m "First commit"
+git remote add origin *[the repository URL you copied]*
+git branch -M main
+git push -u origin main
+```
+
+The local files should now be backed up to the Git remote repository;
+refresh the webpage to verify.
+
+## 4. Archive the Theory on Zenodo
+
+The process of archiving a GitHub repository on Zenodo is documented in a vignette in the `theorytools` R-package, so that it can be kept up-to-date.
+We present a brief summary of the instructions at the time of writing here.
+First, create a Zenodo account with your existing GitHub account.
+Then in Zenodo, go to the GitHub section under your account.
+Syncronize Zenodo to import all of your public GitHub repositories.
+Following the instructions on the page, activate Zenodo for your theory repository.
+Then, head to the repository's GitHub page, and create a new release.
+You have to choose a tag and release title;
+we suggest using semantic versioning for both, starting with version `0.1.0`,  unless you use another convention for versioning (which should be documented in README.md).
+After publishing the release,
+you should be able to see it in your Zenodo account.
+
+## 5. Entering Meta-Data
+
+By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
+To document our archive as a FAIR theory requires adding some extra information on Zenodo.
+Supplying the following information helps improve the Findability of a theory:
+
+- Set the *resource type* to `Model`
+- Verify that the *title* is prefaced with `FAIR theory:`
+- Add the *keyword* `fairtheory`
+- List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
+    + `Is documented by` can be used to reference a theory paper you wrote, in which you introduce this FAIR theory
+    + `Is derived from` could be used to reference a paper or book chapter that introduced an existing theory that was not previously made FAIR.
+We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
+
+## Comparing Implementations of a Theory
+
+As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
+we compare our interpretation of the original theory to the interpretation of others.
 
@@
 }
-```
 

### 2024-12-27 manuscript.tex
expand examples, draft vignette

@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 4906}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6763}
 \usepackage{lineno}
@@
 \hypersetup{
-  pdftitle={FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix Schönbrodt5},
+  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
   pdflang={en-EN},
@@
 
-\title{FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
-\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix Schönbrodt\textsuperscript{5}}
+\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
+\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix D. Schönbrodt\textsuperscript{5}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# aec0b761.
+This is a preprint paper, generated from Git Commit \# 70a3322a.
 
@@
 Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M.
-Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix Schönbrodt: Conceptualization, Writing -- review \& editing.
+Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D.
+Schönbrodt: Conceptualization, Writing -- review \& editing.
 
@@
 \affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Other affiliations}
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany}
 
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
-Since their inception, scholars have demonstrated their relevance for making other information artefacts more open.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable {[}REF{]} for both humans and computers.
+Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
+R.
+Wilkinson et al., 2024).
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
 To this end, we introduce ``FAIR theory'':
-a digital representation of theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR Theory has the potential to accelerate cumulative knowledge acquisition and advance social scientific research.
-
-\subsection{The Need for FAIR Theory}\label{the-need-for-fair-theory}
+a digital representation of a scientific theory, compliant with the FAIR principles.
+By improving the efficiency of scholarly communication, FAIR theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
+
+\subsection{The Need for FAIR theory}\label{the-need-for-fair-theory}
 
 The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
-Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
+Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
 These various practices ensure transparent and repeated testing of hypotheses.
-However, recent reviews show that most preregistered hypothesis tests are not supported (Scheel, Schijen, \& Lakens, 2021).
-Increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-psychological theories rarely produce hypotheses that are found to be true,
-and are often sufficiently vague to explain contradictory findings.
-
-Scholars have raised concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences (Szollosi \& Donkin, 2021).
-In other words, social scientific theories lack \emph{formalization} .
-A second concern is the lack of transparent and democratic scholarly communication about psychological theory.
-
-Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and -improvement have been largely overlooked.
+However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
+Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+Psychological theories rarely produce hypotheses that are corroborated by evidence.
+Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
+as the theory's central claims evade falsification.
+
+Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
+Two main concerns are that, first, social scientific theories lack precision compared to theories in the physical sciences (Szollosi \& Donkin, 2021).
+and clarity
+In other words, social scientific theories lack \emph{formalization},
+which means that they do not make very accurate predictions,
+and are thus hard to falsify and difficult to understand on their own,
+without either substantial interpretation or additional background knowledge.
+A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
+
+Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
 The present paper restores balance by applying, for the first time,
@@
 We apply the FAIR principles to scientific theories,
-introducing the concept of \emph{FAIR Theory} to
+introducing the concept of \emph{FAIR theory} to
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 In a progressive research program (Lakatos, 1971),
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
+.
 There are, however, indications that contemporary psychology falls short of this ideal.
-Firstly, because deductive research is over-represented in the literature:
-According to one estimate, 89.6\% of paper published in psychology tests hypotheses (Kühberger, Fritz, \& Scherndl, 2014).
-Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
-Only 15\% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis (McPhetres et al., 2021).
+Firstly, because hypothesis-testing research is over-represented in the literature:
+According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
+Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
+Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
 The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
+In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).
 
@@
 
-The present paper addresses the lack of open science methods for theory development and improvement by applying the FAIR principles to scientific theories.
-Merely publishing theory does not make it open;
+The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
+Merely publishing theory -- as in a classic research article -- does not make it open;
 to be open, theory should adhere to established open science standards.
-The FAIR principles, initially introduced for open research data,
-have since been applied to other information artifacts (Lamprecht et al., 2019).
 We apply the FAIR principles to digital representations of theory,
@@
 \emph{Interoperable} within the data analysis environment,
-and \emph{Reusable} in the practical and legal sense, so that they may be improved over time.
-Following Lamprecht and colleagues,
+and \emph{Reusable} in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
+Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
+Following the original proposal of Lamprecht and colleagues,
 we adapt the FAIR principles for theory, see Table \ref{tab:tabfair}.
@@
 
+Some have defined a model as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
+This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
+There is no principled difference;
+theories and models derived from it exist along a continuum of specificity, where
+a theory has a relatively broader scope and may contain one or more models as specific instances.
+For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
+From this theory, we could derive a more specific \emph{statistical model} by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
+This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+Or, we could derive an even more specific \emph{generative/computational model}, which is completely parametrized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Note that broadness and narrowness are relative terms,
+and one person's theory may be another person's model.
+This definition also implies that FAIR models are a special case of FAIR theory.
+
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
+The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+
 \subsubsection{The Role of Theory Formalization}\label{the-role-of-theory-formalization}
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' \emph{empirical content} {[}REF{]},
-making them easier to falsify,
+Greater formalization increases theories' \emph{empirical content} {[}REF{]} as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes that would otherwise get away with their vague formulation in verbal theories.
+For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented (Lewandowsky \& Farrell, 2010).
+More precise theories makes them easier to falsify,
 which necessitates revising them,
 thus advancing our principled understanding of the phenomena they describe.
-FAIR Theory does not require theories to be formal,
+FAIR theory does not require theories to be formal,
 and formal theory can be represented in a way that is not FAIR.
+It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down.
+The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
 For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
@@
 then this formal theory is not FAIR.
-FAIR Theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).
+FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).
 
@@
 One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like ``theory'', ``model'', and ``framework'' are used interchangeably.
-To curb this trend, we suggest using the keyword \texttt{"FAIR\ theory"} for all resources that constitute or reference a FAIR theory.
+To curb this trend, we suggest using the keyword \texttt{"FAIRtheory"} for all resources that constitute or reference a FAIR theory (separating the words \texttt{FAIR} and \texttt{theory} by a space or hyphen would lead them to be interpreted as separate tokens in many search engines.
 This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
@@
 Similarly, Borsboom and colleagues seek to establish a database of psychological theories {[}REF BORSBOOM{]}.
-Post-hoc curation does not address the root cause of the lack of Findability, however.
-Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.
+Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
+Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
 
@@
 Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
+This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via \texttt{https} or \texttt{git}).
@@
 
-Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (Wilkinson et al., 2016).
+Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
 The original interoperability principles can be rephrased somewhat to apply to theory.
@@
 
-As the original definition of interoperability was somewhat narrow (Wilkinson et al., 2016),
+As the original definition of interoperability was somewhat narrow (M. D. Wilkinson et al., 2016),
 the concept has recently been further refined in terms of facilitating ``successful communication between machines and between humans and machines'', where ``A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
@@
 Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data (de Groot, 1961).
-In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
+In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,
 reflected in questions such as ``who owns theory'',
@@
 
-\subsection{Some other stuff}\label{some-other-stuff}
-
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts, like computer code.
-Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
+\subsection{Additional considerations}\label{additional-considerations}
+
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts, like computer code (which we also have successfully applied in the domain of reproducible research findings, see XXX).
+Using version control systems, like git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
 Git also facilitates diffuse and adversarial collaboration,
@@
 
-Making Theories FAIR Accelerates Scientific Progress
+\subsubsection{Making Theories FAIR Accelerates Scientific Progress}\label{making-theories-fair-accelerates-scientific-progress}
+
 Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
@@
 
-\section{Examples}\label{examples}
-
-\subsection{Formalizing the Empirical Cycle}\label{formalizing-the-empirical-cycle}
-
-In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
-As several authors have taken inspiration from the work by De Groot,
-we compare our interpretation of the original theory to the interpretation of others.
-Originally, the theory has the following structure:
+\section{Making a Theory FAIR}\label{making-a-theory-fair}
+
+Open science infrastructure is an area of active development, and as such,
+the approach proposed here should not be considered definitive,
+but rather, as one proposal for a FAIR-compliant implementation of theory using infrastructure available at the time of writing.
+
+--\textgreater{}
+--\textgreater{}
+
+\subsection{1. Implementing the Theory}\label{implementing-the-theory}
+
+Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
+we decided to use it as an example for this tutorial.
+Note that, while the empirical cycle is not explicitly referred to as a ``theory'' by De Groot, he derives it from ``a theory of thinking'' by Selz (REF).
+We can thus consider it a meta-theory of the process of theory construction.
+The empirical cycle is described on page 28 of De Groot and Spiekerman (1969):
+
+\begin{quote}
+\emph{Phase 1:} `Observation': collection and grouping of empirical materials; (tentative) formation of hypotheses.\\
+\emph{Phase 2:} `Induction': formulation of hypotheses.\\
+\emph{Phase 3:} `Deduction': derivation of specific consequences from the hypotheses, in the form of testable predictions.\\
+\emph{Phase 4:} `Testing': of the hypotheses against new empirical materials, by way of checking whether or not the predictions are fulfilled.\\
+\emph{Phase 5:} `Evaluation': of the outcome of the testing procedure with respect to the hypotheses or theories stated, as well as with a view to subsequent, continued or related, investigations.
+\end{quote}
+
+If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
+De Groot's theory is either at the ``theory'' or ``specification'' level.
+It consists of a series of natural language statements.
+We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language:
 
 \begin{verbatim}
-
 digraph {
@@
 \end{verbatim}
+
+This language describes the model as a directed graph.
+Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
+and the second half describes their proposed interrelations.
+This follows the first two properties of good theory according to Meehl (Paul E. Meehl, 1990).
+
+We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
+
+\subsection{2. Creating a Project Folder}\label{creating-a-project-folder}
+
+Create a new folder and copy the theory file from the previous step into it.
+To help meet the Interoperability and Reusability criteria,
+add two more files:
+A README.md file with instructions for future users of your theory,
+and a LICENSE file with the legal conditions for reuse.
+We recommend using the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
+
+\subsubsection{What's in a README?}\label{whats-in-a-readme}
+
+The readme should contain information to help people get started with using your FAIR theory.
+We suggest the following elements:
+
+\begin{itemize}
+\tightlist
+\item
+  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
+\item
+  Description: A plain-text description of the theory and its scope
+\item
+  Interoperability: This section, also often labeled ``Getting Started'', ``Instructions'', or ``How to Use'' might be better labeled Interoperability from a FAIR perspective.
+Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
+\item
+  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
+\item
+  License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal constraints to reuse}.
+\item
+  Citing this work: Tell users how to cite the theory.
+\item
+  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
+Note that this is redundant with metadata entered in Zenodo (step 5).
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+The disadvantage is that you may have to maintain this section of the readme.
+A compromise might be to use this section to document one canonical reference for the theory that is unlikely to change going forward.
+For example, we referenced the original empirical cycle paper here:
+\end{itemize}
+
+\begin{verbatim}
+This repository contains an implementation of the "empirical cycle",
+a model proposed by De Groot and Spiekerman (1969, p. 28)
+
+> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
+Foundations of inference and research in the behavioral sciences.
+De Gruyter Mouton. https://doi.org/10.1515/9783112313121
+\end{verbatim}
+
+\subsection{3. Version Control the Repository}\label{version-control-the-repository}
+
+The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
+In particular, the practice of version control offers extensive benefits for scientific work Van Lissa et al. (2021).
+To version control our project, we initiate a Git repository in the project folder.
+This can be done by \href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{installing Git} locally,
+opening a terminal in the project folder,
+and running the command \texttt{git\ init}.
+We subsequently create a remote repository to host a copy of this local Git repository, which will in turn be archived.
+To \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{do this on GitHub}, create an account and press the button labeled \texttt{Create\ new\ repository}.
+The default settings for the repository are fine;
+for the next steps of the tutorial it must be set to ``Public''.
+Copy the repository's URL, which usually has the form \texttt{https://github.com/username/repository.git}.
+Next, in the terminal window, run the following lines of code sequentially:
+
+\begin{verbatim}
+git add .
+git commit -m "First commit"
+git remote add origin *[the repository URL you copied]*
+git branch -M main
+git push -u origin main
+\end{verbatim}
+
+The local files should now be backed up to the Git remote repository;
+refresh the webpage to verify.
+
+\subsection{4. Archive the Theory on Zenodo}\label{archive-the-theory-on-zenodo}
+
+The process of archiving a GitHub repository on Zenodo is documented in a vignette in the \texttt{theorytools} R-package, so that it can be kept up-to-date.
+We present a brief summary of the instructions at the time of writing here.
+First, create a Zenodo account with your existing GitHub account.
+Then in Zenodo, go to the GitHub section under your account.
+Syncronize Zenodo to import all of your public GitHub repositories.
+Following the instructions on the page, activate Zenodo for your theory repository.
+Then, head to the repository's GitHub page, and create a new release.
+You have to choose a tag and release title;
+we suggest using semantic versioning for both, starting with version \texttt{0.1.0}, unless you use another convention for versioning (which should be documented in README.md).
+After publishing the release,
+you should be able to see it in your Zenodo account.
+
+\subsection{5. Entering Meta-Data}\label{entering-meta-data}
+
+By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
+To document our archive as a FAIR theory requires adding some extra information on Zenodo.
+Supplying the following information helps improve the Findability of a theory:
+
+\begin{itemize}
+\tightlist
+\item
+  Set the \emph{resource type} to \texttt{Model}
+\item
+  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}
+\item
+  Add the \emph{keyword} \texttt{fairtheory}
+\item
+  List the DOIs/identifiers of \emph{related works}. Use the \texttt{Relation} field as appropriate. For example:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \texttt{Is\ documented\ by} can be used to reference a theory paper you wrote, in which you introduce this FAIR theory
+  \item
+    \texttt{Is\ derived\ from} could be used to reference a paper or book chapter that introduced an existing theory that was not previously made FAIR.
+We used \texttt{Is\ derived\ from} to reference De Groot and Spiekerman's empirical cycle.
+  \end{itemize}
+\end{itemize}
+
+\subsection{Comparing Implementations of a Theory}\label{comparing-implementations-of-a-theory}
+
+As several authors have taken inspiration from the work by De Groot (de Groot, 1961),\\
+we compare our interpretation of the original theory to the interpretation of others.
 
@@
 
-\begin{verbatim}
-digraph {
+```
+digraph \{
 
@@
 prediction;
-  test [label="inferential procedure"];
+test {[}label=``inferential procedure''{]};
 observation;
 
-  theory -> prediction [label="deduction"];
-  prediction -> test;
-  test -> observation;
-  observation -> theory [label="generalization"];
-
-}
-\end{verbatim}
+theory -\textgreater{} prediction {[}label=``deduction''{]};
+prediction -\textgreater{} test;
+test -\textgreater{} observation;
+observation -\textgreater{} theory {[}label=``generalization''{]};
+
+\}
 
@@
 
-\subsection{Using FAIR Theory to Perform Causal Inference}\label{using-fair-theory-to-perform-causal-inference}
+\subsection{Using FAIR theory to Perform Causal Inference}\label{using-fair-theory-to-perform-causal-inference}
 
@@
 
-One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
+One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
 Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
@@
 
+\bibitem[\citeproctext]{ref-degrootMethodologyFoundationsInference1969}
+De Groot, A. D., \& Spiekerman, J. A. A. (1969). \emph{Methodology: {Foundations} of inference and research in the behavioral sciences}. De Gruyter Mouton. \url{https://doi.org/10.1515/9783112313121}
+
 \bibitem[\citeproctext]{ref-dumas-malletPoorReplicationValidity2017}
@@
 Guest, O. (2024). What {Makes} a {Good Theory}, and {How Do We Make} a {Theory Good}? \emph{Computational Brain \& Behavior}. \url{https://doi.org/10.1007/s42113-023-00193-2}
+
+\bibitem[\citeproctext]{ref-guestHowComputationalModeling2021}
+Guest, O., \& Martin, A.
+E.
+(2021).
+How {Computational Modeling Can Force Theory Building} in {Psychological Science}.
+\emph{Perspectives on Psychological Science}, \emph{16}(4), 789--802.
+\url{https://doi.org/10.1177/1745691620970585}
 
@@
 
+\bibitem[\citeproctext]{ref-lewandowsky2010computational}
+Lewandowsky, S., \& Farrell, S. (2010). \emph{Computational modeling in cognition: {Principles} and practice}. Sage.
+
 \bibitem[\citeproctext]{ref-lewinPsychologyProcessGroup1943}
@@
 
+\bibitem[\citeproctext]{ref-taylor2022psychology}
+Taylor, S. (2022). The psychology of pandemics. \emph{Annual Review of Clinical Psychology}, \emph{18}(1), 581--609.
+
 \bibitem[\citeproctext]{ref-vanlissaUsingEndpointsCheck2023}
@@
 
+\bibitem[\citeproctext]{ref-vanrooijTheoryTestHow2021}
+van Rooij, I., \& Baggio, G.
+(2021).
+Theory {Before} the {Test}: {How} to {Build High-Verisimilitude Explanatory Theories} in {Psychological Science}.
+\emph{Perspectives on Psychological Science}, \emph{16}(4), 682--697.
+\url{https://doi.org/10.1177/1745691620970604}
+
 \bibitem[\citeproctext]{ref-vogtFAIR20Extending2024}
@@
 
+\bibitem[\citeproctext]{ref-wilkinson2024applying}
+Wilkinson, S.
+R., Aloqalaa, M., Belhajjame, K., Crusoe, M.
+R., Kinoshita, B.
+de P., Gadelha, L., et al.others.
+(2024).
+\emph{Applying the {FAIR} principles to computational workflows}.
+Retrieved from \url{https://arxiv.org/abs/2410.03490}
+
 \end{CSLReferences}

### 2024-12-27 add_fair_theory.R

@@ -1,0 +1,68 @@
+add_fair_theory <- function(path = ".",
+                            theory_file = NULL,
+                            add_license = "CC0"
+                            ){
+  # 1. Add theory file
+  has_theory_file <- !is.null(theory_file)
+  if(has_theory_file){
+    existing_theory_file <- file.exists(theory_file)
+  }
+  if(existing_theory_file) theory_file <- normalizePath(theory_file)
+  
+  # 1. Create project folder
+  oldwd <- getwd()
+  if(!dir.exists(path)){
+    dir.create(path)
+    setwd(path)
+  }
+  if(has_theory_file){
+    if(existing_theory_file){
+      file.copy(theory_file, file.path(getwd(), basename(theory_file)))
+    } else {
+      file.create(file.path(getwd(), theory_file))
+    }
+  }
+  # 1. Add LICENSE file
+  # Begin license
+  if(!add_license == "none"){
+    tryCatch({
+      dir.create(path, recursive = TRUE, showWarnings = FALSE)
+      
+      # copy 'resources' folder to path
+      license_dir = system.file('rstudio', 'templates', 'project', 'licenses', package = 'worcs', mustWork = TRUE)
+      license_file <- file.path(license_dir, paste0(add_license, ".txt"))
+      file.copy(license_file, file.path(path, "LICENSE"), copy.mode = FALSE)
+      col_message("Writing license file.", verbose = verbose)
+    }, error = function(e){
+      col_message("Writing license file.", success = FALSE)
+    })
+  }
+  # End license
+  # 1. Add readme.md
+  # * Include project Title, Description, Usage of the theory file, How to Contribute, How to Cite the FAIR theory, Related Works (e.g., the paper that documents the theory)
+  # 1. Add .zenodo.json file to provide metadata that will allow Zenodo to index the repository as a FAIR theory
+  # 1. Initialize a Git repository in the project folder
+  # 1. Create Git remote repository (e.g., on GitHub)
+  # 1. Connect local to remote Git repository
+  # 1. Add the repository to Zenodo
+  # * Sync Zenodo to import recent GitHub repositories (may take some time before it shows)
+  # * Flip Switch
+  # 1. Publish a release on GitHub
+  # 1. *Optional:* Edit Zenodo metadata (e.g., to add formal cross-references to related works) https://help.zenodo.org/docs/deposit/manage-records/#edit
+  #   
+    
[…]

### 2024-12-31 manuscript.Rmd
add steps to make theory FAIR

@@
 the approach proposed here should not be considered definitive,
-but rather, as one proposal for a FAIR-compliant implementation of theory using infrastructure available at the time of writing.
+but rather, as one proposal for a FAIR-compliant implementation of theory.
+At the time of writing (2024), the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
+Nevertheless, it is important to stress that alternatives to GitHub  (Gitlab, Bitbucket, etc.) and Zenodo (e.g., institutional repositories) exist.
+The principles described here
+(using version control and archiving major versions) could be implemented in a different workflow.
 
-<!-- <!-- The `worcs` R-package documentation contains a digital version of this tutorial, --> -->
-<!-- <!-- to allow for future updates. --> -->
+The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
+<!-- R-users can use the following code (see the package documentation for more information): -->
 
-<!-- 1. Create project folder -->
-<!-- 1. Add theory file -->
-<!-- 1. Add LICENSE file -->
-<!-- 1. Add readme.md -->
-<!--     * Include project Title, Description, Usage of the theory file, How to Contribute, How to Cite the FAIR theory, Related Works (e.g., the paper that documents the theory) -->
-<!-- 1. Add .zenodo.json file to provide metadata that will allow Zenodo to index the repository as a FAIR theory -->
-<!-- 1. Initialize a Git repository in the project folder -->
-<!-- 1. Create Git remote repository (e.g., on GitHub) -->
-<!-- 1. Connect local to remote Git repository -->
-<!-- 1. Add the repository to Zenodo -->
-<!--     * Sync Zenodo to import recent GitHub repositories (may take some time before it shows) -->
-<!--     * Flip Switch -->
-<!-- 1. Publish a release on GitHub -->
-<!-- 1. *Optional:* Edit Zenodo metadata (e.g., to add formal cross-references to related works) https://help.zenodo.org/docs/deposit/manage-records/#edit -->
-
-<!-- https://developers.zenodo.org/#add-metadata-to-your-github-repository-release` -->
+<!-- ``` -->
+<!-- install.packages("theorytools") -->
+<!-- library(theorytools) -->
+<!-- # Use worcs to check if GitHub permissions are set: -->
+<!-- library(worcs) -->
+<!-- check_github() -->
+<!-- # Create the theory repository: -->
+<!-- fair_theory(path = "directory/for/theory", -->
+<!--             title = "The Theory's Title", -->
+<!--             theory_file = "path_to_existing_theory.txt", -->
+<!--             remote_repo = "github_repo_name", -->
+<!--             add_license = "cc0") -->
+<!-- ``` -->
 
@@
 and a LICENSE file with the legal conditions for reuse.
-We recommend using the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
+We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
 
@@
 * Description: A plain-text description of the theory and its scope
-* Interoperability: This section, also often labeled "Getting Started", "Instructions", or "How to Use" might be better labeled Interoperability from a FAIR perspective.
-Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
+* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
+From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
+We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
 * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
-* License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal constraints to reuse*.
+* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
 * Citing this work: Tell users how to cite the theory.
+Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
+The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
 * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Note that this is redundant with metadata entered in Zenodo (step 5).
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-The disadvantage is that you may have to maintain this section of the readme.
-A compromise might be to use this section to document one canonical reference for the theory that is unlikely to change going forward.
+Again, this is redundant with metadata entered in Zenodo (step 5).
+We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
 For example, we referenced the original empirical cycle paper here:
@@
 This repository contains an implementation of the "empirical cycle",
-a model proposed by De Groot and Spiekerman (1969, p. 28)
+a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
 
@@
 To version control our project, we initiate a Git repository in the project folder.
-This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally,
-opening a terminal in the project folder,
-and running the command `git init`.
-We subsequently create a remote repository to host a copy of this local Git repository, which will in turn be archived.
-To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`.
-The default settings for the repository are fine;
-for the next steps of the tutorial it must be set to "Public".
-Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`.
-Next, in the terminal window, run the following lines of code sequentially:
+<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
+<!-- opening a terminal in the project folder, -->
+<!-- and running the command `git init`. -->
+We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
+<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`.
+-->
+<!-- The default settings for the repository are fine; -->
+<!-- for the next steps of the tutorial it  -->
+Note that the repository
+must be set to "Public" to take advantage of GitHub's Zenodo integration.
+<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
+<!-- Next, in the terminal window, run the following lines of code sequentially: -->
 
-```
-git add .
-git commit -m "First commit"
-git remote add origin *[the repository URL you copied]*
-git branch -M main
-git push -u origin main
-```
-
-The local files should now be backed up to the Git remote repository;
-refresh the webpage to verify.
+<!-- ``` -->
+<!-- git add . -->
+<!-- git commit -m "First commit" -->
+<!-- git remote add origin *[the repository URL you copied]* -->
+<!-- git branch -M main -->
+<!-- git push -u origin main -->
+<!-- ``` -->
+Push the local files to the Git remote repository, and keep them synchronized going forward.
 
@@
 Then in Zenodo, go to the GitHub section under your account.
-Syncronize Zenodo to import all of your public GitHub repositories.
 Following the instructions on the page, activate Zenodo for your theory repository.
-Then, head to the repository's GitHub page, and create a new release.
+Then, create a new release of the GitHub repository.
 You have to choose a tag and release title;
@@
 After publishing the release,
-you should be able to see it in your Zenodo account.
+you should be able to see the archived version in your Zenodo account,
+along with a DOI.
 
@@
 - Add the *keyword* `fairtheory`
+- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to increase its findability
 - List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
@@
 }
+```
 

### 2024-12-31 add_fair_theory.R

@@ -1,44 +1,76 @@
 add_fair_theory <- function(path = ".",
                             theory_file = NULL,
-                            add_license = "CC0"
+                            remote_repo = "https",
+                            add_license = c("cc0", "ccby", "gpl", "gpl3", "agpl", "agpl3", "apache", "apl2", "lgpl", "mit", "proprietary", "none")
                             ){
-  # 1. Add theory file
+  # Select first license
+  if(is.null(add_license)) add_license = "none"
+  add_license <- tryCatch(add_license[1], error = function(e){"none"})
+  
+  # 1. Create project folder
+  if(!dir.exists(path)){
+    dir.create(path)
+  }
+  
+  # 2. Initialize Git repo
+  gert::git_init(path = path)
+  # Connect to remote repo if possible
+  repo_url <- worcs:::parse_repo(remote_repo = remote_repo, verbose = FALSE)
+  valid_repo <- !is.null(repo_url)
+  if(valid_repo){
+    tryCatch({
+      Args_gert <- list(
+        name = "origin",
+        url = remote_repo,
+        repo = path
+      )
+      do.call(gert::git_remote_add, Args_gert)
+      col_message(paste0("Connected to remote repository at ", remote_repo), verbose = verbose)
+    }, error = function(e){
+      col_message("Could not connect to a remote 'GitHub' repository. You are working with a local 'Git' repository only.", success = FALSE, verbose = verbose)
+    })
+  } else {
+    col_message("No valid 'GitHub' address provided. You are working with a local 'Git' repository only.", success = FALSE)
+  }
+  
+  # 3. Add theory file
   has_theory_file <- !is.null(theory_file)
   if(has_theory_file){
     existing_theory_file <- file.exists(theory_file)
-  }
-  if(existing_theory_file) theory_file <- normalizePath(theory_file)
-  
-  # 1. Create project folder
-  oldwd <- getwd()
-  if(!dir.exists(path)){
-    dir.create(path)
-    setwd(path)
-  }
-  if(has_theory_file){
+    theory_file <- normalizePath(theory_file)
     if(existing_theory_file){
       file.copy(theory_file, file.path(getwd(), basename(theory_file)))
     } else {
-      file.create(file.path(getwd(), theory_file))
+      file.create(file.path(path, theory_file))
     }
   }
+  
   # 1. Add LICENSE file
-  # Begin license
   if(!add_license == "none"){
-    tryCatch({
[…]

### 2025-01-02 manuscript.Rmd
add svg code for DOT graphs

@@
 The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
-<!-- R-users can use the following code (see the package documentation for more information): -->
-
-<!-- ``` -->
-<!-- install.packages("theorytools") -->
-<!-- library(theorytools) -->
-<!-- # Use worcs to check if GitHub permissions are set: -->
-<!-- library(worcs) -->
-<!-- check_github() -->
-<!-- # Create the theory repository: -->
-<!-- fair_theory(path = "directory/for/theory", -->
-<!--             title = "The Theory's Title", -->
-<!--             theory_file = "path_to_existing_theory.txt", -->
-<!--             remote_repo = "github_repo_name", -->
-<!--             add_license = "cc0") -->
-<!-- ``` -->
 
@@
 }
+```
+
+```{r include = FALSE}
+library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
+ec_groot <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  observation [pos = "2,3!"];
+  induction [pos = "3,2!"];
+  deduction [pos = "3,1!"];
+  test [pos = "1,1!"];
+  evaluation [pos = "1,2!"];
+
+  observation -> induction;
+  induction -> deduction;
+  deduction -> test;
+  test -> evaluation;
+  evaluation -> observation;
+
+}')
+ec_wagenmakers <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  subgraph cluster_discovery {
+    label="Discovery";
+    induction [label="New hypothesis" pos = "3,2!"];
+    deduction [label="New prediction" pos = "3,1!"];
+  }
+  observation  [label="Old knowledge and old data" pos = "2,3!"];
+  subgraph cluster_justification {
+    label="Justification";
+    test [label="Test on new data" pos = "1,1!"];
+    evaluation [pos = "1,2!"];
+  }
+
+  observation -> induction [label="Speculate & explore" labeljust=r];
+  induction -> deduction  [label="Deduce"];
+  deduction -> test  [label="Design new experiment"];
+  test -> evaluation  [label="Statistical analysis"];
+  evaluation -> observation  [label="Knowledge accumulation"];
+
+}')
+rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
+rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "graph.svg")
 ```
@@
 
-## Comparing Implementations of a Theory
+## Automating these Steps
 
-As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
-we compare our interpretation of the original theory to the interpretation of others.
+R-users can use the `theorytools` package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):
 
-Subsequently, Wagenmakers and colleagues modified the theory by *"[adding the] Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*:
+```
+install.packages("theorytools")
+library(theorytools)
+# Use worcs to check if GitHub permissions are set:
+library(worcs)
+check_git()
+check_github()
+# Create the theory repository:
+fair_theory(path = "c:/theoryfolder/empirical_cycle",
+            title = "The Empirical Cycle",
+            theory_file = "empirical_cycle.dot",
+            remote_repo = "empirical_cycle",
+            add_license = "cc0")
+```
+
+Note that this function also automatically provides basic FAIR theory metadata to Zenodo.
+
+## Forking Different Implementations of a Theory
+
+De Groot's empirical cycle has inspired several authors,
+but closer examination reveals that not all of them have interpreted his work in the same way.
+For example, Wagenmakers and colleagues [-@wagenmakersCreativityVerificationCyclePsychological2018] write *"De Groot’s “empirical cycle,” shown here in Figure 6"* - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
+One advantage of FAIR theory is that we can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
+This illustrates an important advantage of FAIR theory:
+we can implement different versions of a theory, and study their relationships.
+In practice, this is achieved by ["forking the repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo),
+["cloning"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo) the forked repository to our local computer,
+making any changes we want, and then completing steps 4-5 of "Making a Theory FAIR".
+
+We attempted to implement Wagenmakers and colleagues' version as a DOT graph,
+to illustrate some clear deviations from the original.
+The authors deviate from De Groot in a minor way by renaming the five phases. 
+While this change was not made explicit (thus illustrating an advantage of FAIR theory), we assumed that the phases are identical, just with different labels.
+The authors do mention one explicit change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
+The DOT graph below illustrates this change:
 
@@
     label="Discovery";
-    hypothesis [label="New hypothesis"];
-    prediction [label="New prediction"];
+    induction [label="New hypothesis"];
+    deduction [label="New prediction"];
   }
-  data  [label="Old knowledge and old data"];      
+  observation  [label="Old knowledge and old data"];
   subgraph cluster_justification {
@@
 
-  data -> hypothesis [label="Speculate & explore"];
-  hypothesis -> prediction  [label="Deduce"];
-  prediction -> test  [label="Design new experiment"];
+  observation -> induction [label="Speculate & explore"];
+  induction -> deduction  [label="Deduce"];
+  deduction -> test  [label="Design new experiment"];
   test -> evaluation  [label="Statistical analysis"];
-  evaluation -> data  [label="Knowledge accumulation"];
+  evaluation -> observation  [label="Knowledge accumulation"];
 

### 2025-01-02 manuscript.tex
add svg code for DOT graphs

@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6763}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6897}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 70a3322a.
+This is a preprint paper, generated from Git Commit \# 9c5837f3.
 
@@
 For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
-Conversely, a directed acyclic graph (DAG) is a type of formal theory,
+Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \ref{fig:figec}),
 but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
-then this formal theory is not FAIR.
+it is not FAIR.
 FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).
@@
 the approach proposed here should not be considered definitive,
-but rather, as one proposal for a FAIR-compliant implementation of theory using infrastructure available at the time of writing.
-
---\textgreater{}
---\textgreater{}
+but rather, as one proposal for a FAIR-compliant implementation of theory.
+At the time of writing (2024), the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
+Nevertheless, it is important to stress that alternatives to GitHub (Gitlab, Bitbucket, etc.) and Zenodo (e.g., institutional repositories) exist.
+The principles described here
+(using version control and archiving major versions) could be implemented in a different workflow.
+
+The process described here can be largely automated in R using the \texttt{theorytools} package; see the package vignette on FAIR theory, \texttt{vignette("fairtheory",\ package\ =\ "theorytools")}.
 
@@
 and a LICENSE file with the legal conditions for reuse.
-We recommend using the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
+We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
 
@@
 \item
-  Interoperability: This section, also often labeled ``Getting Started'', ``Instructions'', or ``How to Use'' might be better labeled Interoperability from a FAIR perspective.
-Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
+  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''.
+From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''.
+We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
 \item
@@
 \item
-  License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal constraints to reuse}.
+  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
 \item
   Citing this work: Tell users how to cite the theory.
+Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
+The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
 \item
   Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Note that this is redundant with metadata entered in Zenodo (step 5).
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-The disadvantage is that you may have to maintain this section of the readme.
-A compromise might be to use this section to document one canonical reference for the theory that is unlikely to change going forward.
+Again, this is redundant with metadata entered in Zenodo (step 5).
+We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
 For example, we referenced the original empirical cycle paper here:
@@
 To version control our project, we initiate a Git repository in the project folder.
-This can be done by \href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{installing Git} locally,
-opening a terminal in the project folder,
-and running the command \texttt{git\ init}.
-We subsequently create a remote repository to host a copy of this local Git repository, which will in turn be archived.
-To \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{do this on GitHub}, create an account and press the button labeled \texttt{Create\ new\ repository}.
-The default settings for the repository are fine;
-for the next steps of the tutorial it must be set to ``Public''.
-Copy the repository's URL, which usually has the form \texttt{https://github.com/username/repository.git}.
-Next, in the terminal window, run the following lines of code sequentially:
-
-\begin{verbatim}
-git add .
-git commit -m "First commit"
-git remote add origin *[the repository URL you copied]*
-git branch -M main
-git push -u origin main
-\end{verbatim}
-
-The local files should now be backed up to the Git remote repository;
-refresh the webpage to verify.
+We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
+Note that the repository
+must be set to ``Public'' to take advantage of GitHub's Zenodo integration.
+
+Push the local files to the Git remote repository, and keep them synchronized going forward.
 
@@
   Add the \emph{keyword} \texttt{fairtheory}
+\item
+  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to increase its findability
 \item
@@
 \end{itemize}
+
+\subsection{Automating these Steps}\label{automating-these-steps}
+
+R-users can use the \texttt{theorytools} package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):
+
+\begin{verbatim}
+install.packages("theorytools")
+library(theorytools)
+# Use worcs to check if GitHub permissions are set:
+library(worcs)
+check_git()
+check_github()
+# Create the theory repository:
+fair_theory(path = "c:/theoryfolder/empirical_cycle",
+            title = "The Empirical Cycle",
+            theory_file = "empirical_cycle.dot",
+            remote_repo = "empirical_cycle",
+            add_license = "cc0")
+\end{verbatim}
+
+Note that this function also automatically provides basic FAIR theory metadata to Zenodo.
 
@@
 
-```
-digraph \{
+\begin{verbatim}
+digraph {
 
@@
   prediction;
-test {[}label=``inferential procedure''{]};
+  test [label="inferential procedure"];
   observation;
 
-theory -\textgreater{} prediction {[}label=``deduction''{]};
-prediction -\textgreater{} test;
-test -\textgreater{} observation;
-observation -\textgreater{} theory {[}label=``generalization''{]};
-
-\}
+  theory -> prediction [label="deduction"];
+  prediction -> test;
+  test -> observation;
+  observation -> theory [label="generalization"];
+
+}
+\end{verbatim}
 

### 2025-01-03 manuscript.Rmd
full draft

@@
 classoption       : "man"
-output            : papaja::apa6_pdf
+output            :
+  papaja::apa6_pdf: default
+  md_document: default
 ---
@@
 }')
+ec_lissa <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  theory [pos = "2,3!"];
+  prediction [pos = "3,2!"];
+  data [pos = "1,1!"];
+  test [pos = "2,1!"];
+  results [pos = "1,2!"];
+
+  theory -> prediction [label="deduction"];
+  prediction -> test [label = "implement inferential procedure"];
+  data -> results;
+  test -> results;
+  results -> theory [label="interpretation and generalization"];
+
+}')
 rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
-rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "graph.svg")
+rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.svg")
+rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
 ```
@@
 De Groot's empirical cycle has inspired several authors,
-but closer examination reveals that not all of them have interpreted his work in the same way.
+but not all of them have interpreted his work the same.
 For example, Wagenmakers and colleagues [-@wagenmakersCreativityVerificationCyclePsychological2018] write *"De Groot’s “empirical cycle,” shown here in Figure 6"* - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
-One advantage of FAIR theory is that we can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-This illustrates an important advantage of FAIR theory:
-we can implement different versions of a theory, and study their relationships.
-In practice, this is achieved by ["forking the repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo),
+An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
+We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
+This is achieved by ["forking the repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo),
 ["cloning"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo) the forked repository to our local computer,
@@
 
-We attempted to implement Wagenmakers and colleagues' version as a DOT graph,
+We have implemented Wagenmakers and colleagues' version as a DOT graph
 to illustrate some clear deviations from the original.
-The authors deviate from De Groot in a minor way by renaming the five phases. 
-While this change was not made explicit (thus illustrating an advantage of FAIR theory), we assumed that the phases are identical, just with different labels.
-The authors do mention one explicit change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
-The DOT graph below illustrates this change:
+First, the phases of the cycle have been renamed.
+While this change was not described in the paper,
+we assumed that the labels are meant to illustrate the phases, not substantially change the ontology.
+We represent this change by adding labels to the original DOT graph.
+Note, however, that the labels suggest a focus on empirical psychology that was absent in the original formulation, which was more general.
+Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
+As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
+Second, the authors mention an explicit change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
+The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
 
@@
 
-<!--AB TODO: Also add images of the graph syntax - not everyone will be able to read this in syntax and see the graph in their minds-->
-
-Note, however, that there appear to be further changes:
-the phases of the cycle have been renamed,
-and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
-Moreover, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
-As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-
-Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:
+The first author was inspired by De Groot too,
+but they conceive of the empirical cycle in yet another way.
+First, notice that the nodes in De Groot's formulation mostly refer to processes.
+This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
+In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
+Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
+<!-- In our representation, -->
+<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
+<!-- For example, the refutation of a hypothesized effect, -->
+<!-- or the serendipitous observation of some pattern in data, -->
+<!-- might be a reason to revise or construct theory. -->
+<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
+<!-- in the form of the inductive bias of methods used to perform the test, -->
+<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
+<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced.
+-->
+<!-- ``` -->
+For example, many have argued that observation is value-laden, and as such, involves induction.
+The derivation of hypotheses from theory is also not purely deductive,
+as auxiliary assumptions are often made (which are, again, an inductive process).
+Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [REF Peikert].
+With these alterations, we implement the empirical cycle as follows:
 
@@
   prediction;
-  test [label="inferential procedure"];
-  observation;
+  data;
+  test;
+  results;
 
   theory -> prediction [label="deduction"];
-  prediction -> test;
-  test -> observation;
-  observation -> theory [label="generalization"];
+  prediction -> test [label = "implement inferential procedure"];
+  data -> results;
+  test -> results [label = "apply to data"];
+  results -> theory [label="interpretation and generalization"];
 
@@
 ```
-
-In our representation,
-induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
-For example, the refutation of a hypothesized effect,
-or the serendipitous observation of some pattern in data,
-might be a reason to revise or construct theory.
-Induction, incidentally, also occurs within the link from prediction to testing:
-in the form of the inductive bias of methods used to perform the test,
-and auxiliary assumptions that must be made to address remaining theoretical ambiguities.
-
-<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced.
--->
-<!-- ``` -->
 
@@
 
-* Theory is the vehicle of cumulative knowledge acquisition
-* According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights.
+<!-- * Theory is the vehicle of cumulative knowledge acquisition -->
+<!-- * According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights.
 When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena.
-* At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory.
-* One source of potential improvements of theory methodology that has not been previously considered is computer science.
-* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
-* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
-* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
-* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
-* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
+-->
+<!-- * At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory. -->
+<!-- * One source of potential improvements of theory methodology that has not been previously considered is computer science. -->
+<!-- * The process of "iteratively improving" digital objects - in this case, computer code - is well understood. -->
+<!-- * Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well. -->
+<!-- * This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
+-->
+<!-- * We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories -->
+<!-- * FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
 Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.
-* FAIR theory allows for algorithmic derivation of hypotheses implied by the theory.
-* FAIR theory enables integration testing: researchers can build a "test suite" of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
-* To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.
+-->
+<!-- * FAIR theory allows for algorithmic derivation of hypotheses implied by the theory. -->
+<!-- * FAIR theory enables integration testing: researchers can build a "test suite" of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
+-->
+<!-- * To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.
+-->
 
 # Discussion
+
+The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
+Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
+These limitations impede cumulative knowledge production in our field,
+leading to an accumulation of "one-shot" empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
+We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
+We demonstrated how to create, version-control, and archive theories as digital information artifacts.
+We introduced the `theorytools` R-package to partly automate these processes, reducing barrier of entry for researchers,
+and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
+
+Making theory FAIR allows researchers to more easily find a relevant framework;
+access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
+Whereas the idea of theory can be quite nebulous to empirical social scientists,
+FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
+Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
+While FAIR theory does not directly resolve the problem of strategic ambiguity, it does provide a framework within which scholars can increase the precision and formalization of theories.
+FAIR principles also facilitates new ways of collaboration,
+leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
+<!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->
+
+## Strengths
+
+One important strength of FAIR theory is that it provides much-needed open science methods for the underemphasized inductive phase of the empirical cycle.
+Most extant open science methods focus on increased rigor in testing, but provide little guidance as to what to do with the newly collected empirical evidence.
+By providing much-needed open science methods for theory construction,
+FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the "open empirical cycle" [REF Hoijtink].
+
+Our approach aligns closely with contemporary developments in open science,
+such as modular publishing, interdisciplinarity, meta-research, and team science.
+The advantage of modular publishing is that authors can be credited for theory development.
+Given the current emphasis on empirical papers [REF], theoretical papers can be hard to publish.
+FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
+
+Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [REF Oisin paper].
+Meta-research benefit from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
+In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
+clarifying any remaining areas of contention or misunderstanding. 
+Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
+This facilitates collaboration across geographical boundaries,
+and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.
+
+Finally, FAIR theory plays an important role in science communication, because theory synthesizes contemporary scientific understanding about a phenomenon.
+Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
+By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
+
+## Limitations
+
+One important limitation of the present work is that,
+while we build on well-established information architecture like Zenodo,
+it is unlikely that the proposed metadata standard is definitive.
+Community adoption can reveal areas of further improvement.
+Furthermore, at the time of writing, dedicated indexing systems for FAIR theory are non-existent.
+Using the Zenodo search function and submitting theories to the "FAIR Theory Community" on Zenodo can help overcome this limitation in the short term.
+
+Another limitation is the learning curve associated with tools like Git and Zenodo.
+The `theorytools` R-package mitigates this limitation by automating key steps in the process.
+Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
+One barrier to adoption of FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
+Education might help address this limitation; with this in mind,
+we are developing open educational materials on theory development.
+
+One limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [REF] and lack of theory formalization [REF].
+However, our work does establish a framework within which theories can be further formalized.
+The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
+Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [REF] and lack of standardized ontologies for psychological constructs [REF].
+However, our work here provides a template for addressing such challenges,
+and any advancements in the areas of measurement and ontology will serve to amplify the value of FAIR theories, particularly when such resources are cross-referenced in the metadata (e.g., on Zenodo).
 
@@
 
+## Conclusion
+
+FAIR theory is a major step forwards towards more transparent, collaborative, and efficient theory construction.
+It provides much-needed open science methods for the inductive phase of the empirical cycle,
+closing a critical gap in the scientific process.
+FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
+This paves the way for more theory-driven scholarship,
+and accelerates cumulative knowledge acquisition in the social sciences and beyond.
+
 \newpage

### 2025-01-03 manuscript.tex
full draft

@@
   {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
+    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
     {}{\InnerPatchFailed}%
@@
 {}{\PatchFailed}
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6897}
+\makeatother
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6844}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 9c5837f3.
+This is a preprint paper, generated from Git Commit \# ec81348c.
 
@@
 
-\subsection{Comparing Implementations of a Theory}\label{comparing-implementations-of-a-theory}
-
-As several authors have taken inspiration from the work by De Groot (de Groot, 1961),\\
-we compare our interpretation of the original theory to the interpretation of others.
-
-Subsequently, Wagenmakers and colleagues modified the theory by \emph{``{[}adding the{]} Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}:
+\subsection{Forking Different Implementations of a Theory}\label{forking-different-implementations-of-a-theory}
+
+De Groot's empirical cycle has inspired several authors,
+but not all of them have interpreted his work the same.
+For example, Wagenmakers and colleagues (\textbf{wagenmakersCreativityVerificationCyclePsychological2018?}) write \emph{``De Groot's ``empirical cycle,'' shown here in Figure 6''} - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
+An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
+We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
+This is achieved by \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``forking the repository''},
+\href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``cloning''} the forked repository to our local computer,
+making any changes we want, and then completing steps 4-5 of ``Making a Theory FAIR''.
+
+We have implemented Wagenmakers and colleagues' version as a DOT graph
+to illustrate some clear deviations from the original.
+First, the phases of the cycle have been renamed.
+While this change was not described in the paper,
+we assumed that the labels are meant to illustrate the phases, not substantially change the ontology.
+We represent this change by adding labels to the original DOT graph.
+Note, however, that the labels suggest a focus on empirical psychology that was absent in the original formulation, which was more general.
+Furthermore, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
+As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
+Second, the authors mention an explicit change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}.
+The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
 
@@
 
-Note, however, that there appear to be further changes:
-the phases of the cycle have been renamed,
-and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
-Moreover, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
-As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-
-Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:
+The first author was inspired by De Groot too,
+but they conceive of the empirical cycle in yet another way.
+First, notice that the nodes in De Groot's formulation mostly refer to processes.
+This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
+In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
+Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
+For example, many have argued that observation is value-laden, and as such, involves induction.
+The derivation of hypotheses from theory is also not purely deductive,
+as auxiliary assumptions are often made (which are, again, an inductive process).
+Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well {[}REF Peikert{]}.
+With these alterations, we implement the empirical cycle as follows:
 
@@
 \end{verbatim}
-
-In our representation,
-induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
-For example, the refutation of a hypothesized effect,
-or the serendipitous observation of some pattern in data,
-might be a reason to revise or construct theory.
-Induction, incidentally, also occurs within the link from prediction to testing:
-in the form of the inductive bias of methods used to perform the test,
-and auxiliary assumptions that must be made to address remaining theoretical ambiguities.
 
@@
 \url{https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/}
-
-\begin{itemize}
-\tightlist
-\item
-  Theory is the vehicle of cumulative knowledge acquisition
-\item
-  According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights.
-When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena.
-\item
-  At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory.
-\item
-  One source of potential improvements of theory methodology that has not been previously considered is computer science.
-\item
-  The process of ``iteratively improving'' digital objects - in this case, computer code - is well understood.
-\item
-  Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
-\item
-  This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
-\item
-  We introduce ``FAIR theory'', a digital research artifact to represent formal social scientific theories
-\item
-  FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
-Version control also enables diffuse collaboration in theory development, as other researchers can submit ``pull requests'' to suggest modifications of a theory, or can ``fork'' existing theories to create a spin-off from an existing theory.
-\item
-  FAIR theory allows for algorithmic derivation of hypotheses implied by the theory.
-\item
-  FAIR theory enables integration testing: researchers can build a ``test suite'' of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
-\item
-  To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.
-\end{itemize}
 

### 2025-01-04 dk_example/example.tex
start adding the dk example

\documentclass[a4paper,11pt]{article}

% load packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{mathtools}
\usepackage{bbm}

% define own colors and use colored text
\usepackage{xcolor}

\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}

%%% standard additional commands
%epsilon
\newcommand*{\oldepsilon}{\epsilon}%
\renewcommand*{\epsilon}{\varepsilon}%

% Number Sets
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\1}{\mathbf{1}}

% Statistics
\newcommand{\Var}{\mathbb{V}}
\newcommand{\Exp}{\mathbb{E}}

% Matrices
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}

% operators
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\di}{d}

\begin{document}

\section{Ontology}
The constructs used are
\begin{itemize}
 \item performance
 \item performance accessment
 \item skill
 \item confidence
\end{itemize}

\section{Mathematical Formulation}

\section{(Statistical) Models}

\section{Open Questions / Ambiguities}

\end{document}

### 2025-01-06 dk_example/example.tex
add short textual description
add to dk example

@@
 
+\section{Description}
+The Dunning-Kruger effect is stated in terms of skill and overconfidence to show that measurement error can cause significant bias in the relationship between performance and overestimation.
+Instrumental variable methods can be used to correct for this bias.
+
 \section{Ontology}
@@
 \begin{itemize}
- \item performance
- \item performance accessment
- \item skill
- \item confidence
+ \item performance as a test score
+ \item performance estimation as the difference between the expected and the actual test score
+ \item skill as the ability to perform well on a given test
+ \item overconfidence as the difference between self-assessed and actual skill
+ \item measurement error as luck on a test
+\end{itemize}
+
+\section{Relationships}
+
+\subsection{Original DK Paper}
+
+\begin{figure}[ht]
+  \vspace{1cm}
+  \begin{center}
+  \input{figures/original_dk.tex}
+  \end{center}
+  %\caption{Subtypes of `AbstractSem`.}
+  \label{fig:typetree}
+\end{figure}
+
+With $-$ signifying a negative association and $\simeq$ meaning measured by.
+
+\begin{itemize}
+ \item low skilled are more overconfident in assessing their skill than high skilled
 \end{itemize}
@@
 
+\subsection{Correction}
+There are different ways to correct for the bias introduced by measurement error:
+\begin{itemize}
+ \item use different test performances to calculate skill and overconfidence, but a small test-retest correlation can bias the estimate towrds zero then (Krueger and Mueller, 2002)
+\end{itemize}
+
 \section{Open Questions / Ambiguities}

### 2025-01-06 dk_example/figures/original_dk.tex
add to dk example

\begin{tikzpicture}
        [manifest/.style={rectangle,draw=black!80,semithick,minimum width=1cm,inner sep=4pt,text centered},
        latent/.style={circle,draw=black!80,thick,inner sep=2,text centered},
        latent_wide/.style={circle,draw=black!80,thick,inner sep=4,text centered},
        from/.style={<-,>=stealth',semithick},
        text height=1.75ex, text depth=.5ex]

\begin{scope}

% types
\node[manifest] at (0,0) (skill) {\texttt{skill}};

\node[manifest] at (6, 0) (overconfidence) {\texttt{overconfidence}};

\node[manifest] at (0, -2) (performance) {\texttt{performance}};

\node[manifest] at (6,-2) (estimate) {\texttt{overestimate performance}};

%arrows
\draw[-,>=stealth',semithick] (skill) -- node [fill=white] {$-$} (overconfidence);
\draw[-,>=stealth',semithick] (performance) -- node [fill=white] {$-$} (estimate);
\draw[-,>=stealth',semithick] (skill) -- node [fill=white] {$\simeq$} (performance);
\draw[-,>=stealth',semithick] (overconfidence) -- node [fill=white] {$\simeq$} (estimate);

\end{scope}

\end{tikzpicture}

### 2025-01-07 manuscript.Rmd
add a proposal for semver

@@
 This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
-The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
-This can be achieved by using Git for version control and Zenodo for archiving.
+However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
+The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
+In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
+Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
+We propose the following adaptation of semantic versioning for theories:
+
+```
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
+MINOR version when you expand the set of empierical statements in a backward compatible manner
+PATCH version when you make backward compatible bug fixes, or clarifications
+```
+
+Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
 
@@
 Then, create a new release of the GitHub repository.
-You have to choose a tag and release title;
-we suggest using semantic versioning for both, starting with version `0.1.0`,  unless you use another convention for versioning (which should be documented in README.md).
+Choose a tag and release title using our adapted semantic versioning, starting with version 1.0.0, if you intend to share your theory with the broader scientific community.
 After publishing the release,

### 2025-01-08 manuscript.Rmd
Update

--- 2025-01-03 manuscript.Rmd
@@
 This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-There is no principled difference;
-theories and models derived from it exist along a continuum of specificity, where
-a theory has a relatively broader scope and may contain one or more models as specific instances.
+While there may be philosophical differences,
+these differences are largely irrelevant for the purpose of making theories more FAIR.
+When organizing knowledge, we think it is helpful to view theories and models derived from them as existing along a continuum of specificity,
+from broad to narrow, from general to specific,
+where a theory has a relatively broader scope and may contain one or more models as specific instances.
+<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
+Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
+-->
 For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.

### 2025-01-08 manuscript.tex
Update

@@
 \makeatother
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 6844}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 7856}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# ec81348c.
+This is a preprint paper, generated from Git Commit \# a125ad91.
 
@@
 
+The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
+Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
+These limitations impede cumulative knowledge production in our field,
+leading to an accumulation of ``one-shot'' empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
+We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
+We demonstrated how to create, version-control, and archive theories as digital information artifacts.
+We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing barrier of entry for researchers,
+and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
+
+Making theory FAIR allows researchers to more easily find a relevant framework;
+access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
+Whereas the idea of theory can be quite nebulous to empirical social scientists,
+FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
+Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
+While FAIR theory does not directly resolve the problem of strategic ambiguity, it does provide a framework within which scholars can increase the precision and formalization of theories.
+FAIR principles also facilitates new ways of collaboration,
+leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
+
+\subsection{Strengths}\label{strengths}
+
+One important strength of FAIR theory is that it provides much-needed open science methods for the underemphasized inductive phase of the empirical cycle.
+Most extant open science methods focus on increased rigor in testing, but provide little guidance as to what to do with the newly collected empirical evidence.
+By providing much-needed open science methods for theory construction,
+FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the ``open empirical cycle'' {[}REF Hoijtink{]}.
+
+Our approach aligns closely with contemporary developments in open science,
+such as modular publishing, interdisciplinarity, meta-research, and team science.
+The advantage of modular publishing is that authors can be credited for theory development.
+Given the current emphasis on empirical papers {[}REF{]}, theoretical papers can be hard to publish.
+FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
+
+Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
+Meta-research benefit from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
+In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
+clarifying any remaining areas of contention or misunderstanding.
+Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
+This facilitates collaboration across geographical boundaries,
+and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.
+
+Finally, FAIR theory plays an important role in science communication, because theory synthesizes contemporary scientific understanding about a phenomenon.
+Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
+By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
+
+\subsection{Limitations}\label{limitations}
+
+One important limitation of the present work is that,
+while we build on well-established information architecture like Zenodo,
+it is unlikely that the proposed metadata standard is definitive.
+Community adoption can reveal areas of further improvement.
+Furthermore, at the time of writing, dedicated indexing systems for FAIR theory are non-existent.
+Using the Zenodo search function and submitting theories to the ``FAIR Theory Community'' on Zenodo can help overcome this limitation in the short term.
+
+Another limitation is the learning curve associated with tools like Git and Zenodo.
+The \texttt{theorytools} R-package mitigates this limitation by automating key steps in the process.
+Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
+One barrier to adoption of FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
+Education might help address this limitation; with this in mind,
+we are developing open educational materials on theory development.
+
+One limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity {[}REF{]} and lack of theory formalization {[}REF{]}.
+However, our work does establish a framework within which theories can be further formalized.
+The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
+Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis {[}REF{]} and lack of standardized ontologies for psychological constructs {[}REF{]}.
+However, our work here provides a template for addressing such challenges,
+and any advancements in the areas of measurement and ontology will serve to amplify the value of FAIR theories, particularly when such resources are cross-referenced in the metadata (e.g., on Zenodo).
+
 \subsection{Future Directions}\label{future-directions}
@@
 opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
+
+\subsection{Conclusion}\label{conclusion}
+
+FAIR theory is a major step forwards towards more transparent, collaborative, and efficient theory construction.
+It provides much-needed open science methods for the inductive phase of the empirical cycle,
+closing a critical gap in the scientific process.
+FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
+This paves the way for more theory-driven scholarship,
+and accelerates cumulative knowledge acquisition in the social sciences and beyond.
 

### 2025-01-08 manuscript.Rmd
update

@@
 This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
-However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
-The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
-In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
-Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
-We propose the following adaptation of semantic versioning for theories:
-
-```
-Given a version number MAJOR.MINOR.PATCH, increment the:
-
-MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
-MINOR version when you expand the set of empierical statements in a backward compatible manner
-PATCH version when you make backward compatible bug fixes, or clarifications
-```
-
-Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
+The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
+This can be achieved by using Git for version control and Zenodo for archiving.
+Git tracks the internal provenance of a theory repository; Zenodo is used to archive relations to external references (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
 
@@
 ### Making Theories FAIR Accelerates Scientific Progress
+
 Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
@@
 We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
+
+## 6. Making Changes
+
+<!-- CJ: Aaron, I moved your part here because I don't think it relates to interoperability. But perhaps there's another place where it fits better? -->
+Aside from technical solutions to However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
+The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
+In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
+Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
+We propose the following adaptation of semantic versioning for theories:
+
+```
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
+MINOR version when you expand the set of empierical statements in a backward compatible manner
+PATCH version when you make backward compatible bug fixes, or clarifications
+```
+
+Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
 

### 2025-01-09 manuscript.tex
meeting notes and minor changes

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 % definitions for citeproc citations
@@
 \makeatother
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 7856}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8100}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# a125ad91.
+This is a preprint paper, generated from Git Commit \# eed54ad2.
 
@@
 This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-There is no principled difference;
-theories and models derived from it exist along a continuum of specificity, where
-a theory has a relatively broader scope and may contain one or more models as specific instances.
+While there may be philosophical differences,
+these differences are largely irrelevant for the purpose of making theories more FAIR.
+When organizing knowledge, we think it is helpful to view theories and models derived from them as existing along a continuum of specificity,
+from broad to narrow, from general to specific,
+where a theory has a relatively broader scope and may contain one or more models as specific instances.
 For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
@@
 This can be achieved by using Git for version control and Zenodo for archiving.
+Git tracks the internal provenance of a theory repository; Zenodo is used to archive relations to external references (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
 
@@
 Then, create a new release of the GitHub repository.
-You have to choose a tag and release title;
-we suggest using semantic versioning for both, starting with version \texttt{0.1.0}, unless you use another convention for versioning (which should be documented in README.md).
+Choose a tag and release title using our adapted semantic versioning, starting with version 1.0.0, if you intend to share your theory with the broader scientific community.
 After publishing the release,
@@
 \end{itemize}
+
+\subsection{6. Making Changes}\label{making-changes}
+
+Aside from technical solutions to However, ``{[}i{]}n systems with many dependencies, releasing new {[}\ldots{]} versions can quickly become a nightmare.''
+The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
+In software development this problem is colloquially called ``dependency hell'' and usually addressed via semantic versioning.
+Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
+We propose the following adaptation of semantic versioning for theories:
+
+\begin{verbatim}
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
+MINOR version when you expand the set of empierical statements in a backward compatible manner
+PATCH version when you make backward compatible bug fixes, or clarifications
+\end{verbatim}
+
+Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
 
@@
 R., Aloqalaa, M., Belhajjame, K., Crusoe, M.
-R., Kinoshita, B.
-de P., Gadelha, L., et al.others.
+R., Paula Kinoshita, B.
+de, Gadelha, L., \ldots{} Goble, C.
 (2024).
-\emph{Applying the {FAIR} principles to computational workflows}.
+\emph{Applying the FAIR principles to computational workflows}.
 Retrieved from \url{https://arxiv.org/abs/2410.03490}

### 2025-01-12 manuscript.Rmd
update 2025-01-12 12:29:47.77178

@@
 To this end, we introduce "FAIR theory":
-a digital representation of a scientific theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
-
-<!-- QUOTES
-
-Kurt Lewin: Nothing is as practical as a good theory
-
-"Call Leon": When finding unexpected results in cognitive dissonance research, students were told to call Leon Festinger.
-He could explain why the study did not work, e.g.
-because you used the wrong font for the questionnaire.
-
-END QUOTES -->
+a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
+compliant with the FAIR principles.
+FAIR theory has the potential to improve the efficiency of scholarly communication and
+accelerate cumulative knowledge acquisition.
 
@@
 Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
-as the theory's central claims evade falsification.<!--CJ: Resolved AB's comment?
-AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
+as the theory's central claims evade falsification.
 <!-- CJ: Maybe we should give an example here. Does anyone know one? -->
@@
 
-<!-- First, we reflect on the differences between theory and data. <!-- CJ: Should we? --> -->
-<!-- Second, we adapt the FAIR principles for theory, respecting these differences. -->
-<!-- Third, we discuss the current state and future directions of psychological theory with respect to each criterion. -->
-<!-- Fourth, we present worked examples of theory made FAIR. -->
-<!-- We conclude with a summary and future directions for FAIR theory. -->
-
 ## Making Theory FAIR
@@
 The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
-Merely publishing theory -- as in a classic research article -- does not make it open;
+Merely publishing theory in a research article does not make it open;
 to be open, theory should adhere to established open science standards.
-<!--AB  I deleted this sentence because it was already in the introduction -->
 We apply the FAIR principles to digital representations of theory,
@@
 
+### Modular Publishing
+
+FAIR theory is an example of modular publishing [@kirczModularityNextForm1998].
+At present, the primary unit of social scientific communication is the academic paper.
+A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
+Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
+with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
+Data sharing is a good example of a modular publishing practice that is  widely adopted and increasingly required by funding agencies, journals, and universities.
+Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
+which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
+To maintain a persistent record of scholarly communication,
+Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
+which is used by many academic publishers.
+Finally, the [DataCite Metadata Schema]() offers a standard way to document the nature of relationships between information artifacts.
+<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
+For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
+and cross-reference the published paper with `relationType: IsSupplementTo`.
+Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.
+
+### Version Control
+
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts.
+Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
+Git tracks line-by-line changes to text-based files,
+and maintains a complete history of those changes.
+It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013].
+Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
+The present paper considers advantages of Git for FAIR theory.
+Git enables explicitly comparing versions of a file (or: theory),
+incorporating changes by different authors,
+and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
+This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
+
+# The FAIR Principles
+
 ## Findability
@@
 grounding their work in established theoretical foundations.
-Making theories Findable also increases the impact and reuse potential of theories across disciplines,
+It further increases the impact and reuse potential of theories across disciplines,
 either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
@@
 Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
-Finally, these metadata should be registered or indexed in a searchable registry (F4).
-This final criterion is less straightforward.
-Ideally, FAIR theories should be indexed in search engines used by academics, like Google Scholar.
-At present, however, these search engines are designed to index traditional print publications.
-The *data paper* solves this problem for research data;
-the idea is that scholars publish a paper (or even preprint) as documentation for the data resource [REF McGIllivray on data papers].
-The data paper is indexed by search engines, and in turn points to the relevant information artifact.
-The same solution could be applied to theories - but it seems superfluous to generate papers whose only purpose is to redirect to a specific resource.
-Another solution is to manually index FAIR theories,
-for example by adding them to one's Google Scholar profile,
-or entering them in PURE.
-<!-- CJ: Can we guarantee automatically indexing FAIR theory somehow? -->
-
-At present, theories have poor findability, which impedes cumulative knowledge acquisition.
-One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like "theory", "model", and "framework" are used interchangeably.
-To curb this trend, we suggest using the keyword `"FAIRtheory"` for all resources that constitute or reference a FAIR theory (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines.
-This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
-Another factor contributing to the present lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
-A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
-This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
-similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
-As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
+
+Finally, metadata should be registered or indexed in a searchable registry (F4).
+Zenodo and GitHub are both searchable.
+Standardized metadata further enhance Findability in these repositories.
+The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
+<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
+Furthermore, a standard keyword can be used; we suggest using the keyword `"fairtheory"` for all resources that constitute or reference a FAIR theory.
+<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines. -->
+
+Findability is substantially amplified if intended users of a resource know where to search for it.
+This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
+Regrettably, most academic search engines are designed to index traditional print publications, not other information artifacts.
+Since the status quo is to publish theories in papers,
+the FAIR requirements are met if scholars continue to do so,
+and additionally publish theories as separate information artifacts.
+The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
+<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
+<!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
+<!-- The same solution could be applied to theories. -->
+In the longer term, it may not be necessary to write a paper for each theory.
+If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
+Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
 
@@
 
-The Accessibility criteria serve to *regulate* access, not to maximize it.
-These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
+The Accessibility principles pertain to *regulating* access, not only maximizing it.
+They apply to theory with minor changes.
 Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
@@
 and Zenodo allows for access control of individual files/resources.
-An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics).
-In general, it makes sense to keep outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
+<!-- An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics). -->
+In general, it makes sense to retain outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
 
@@
 This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
-The current prevalence of strategic ambiguity renders psychological theory difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
+At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
 It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
@@
 Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
-The original interoperability principles can be rephrased somewhat to apply to theory.
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
-The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
+The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
 Instead, FAIR theory should, ad minimum,
-be instantiated as 
-as a type of data that is human- and machine-readable with as few interpretative steps as possible,
+be instantiated in a human- and machine-readable datatype,
 as previously recommended [@vanlissaWORCSWorkflowOpen2021].
@@
 Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
-This is essentially a call to establish standardized ontologies,
-which are themselves a type of theory [@meehlAppraisingAmendingTheories1990]. 
+Aside from the aforementioned Datacite metadata schema [@datacitemetadataworkinggroupDataCiteMetadataSchema2024],
+in the context of theory, this highlights the importance of establishing standardized ontologies.
 Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by linking the DOI of those nested theories [@ContributingCitationsReferences].
+This can be achieved by cross-referencing the DOI of those nested theories [@ContributingCitationsReferences].
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
-This can be achieved by using Git for version control and Zenodo for archiving.
-Git tracks the internal provenance of a theory repository; Zenodo is used to archive relations to external references (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
-
-As the original definition of interoperability was somewhat narrow [@wilkinsonFAIRGuidingPrinciples2016a],
-the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
-This definition invites the question: *interoperable for what?*
-Suitable answers for FAIR theory may be:
-this theory is X-interoperable for deriving testable hypotheses,
+This is achieved by using Git for version control and Zenodo for archiving.
+Git tracks the internal provenance of a theory repository; Zenodo is used to cross-reference external relationships (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
+
+Recent work points out that interoperability is not an all-or-nothing property.
+The concept of X-interoperability was introduced to answer the question: *interoperable for what?*
+X-interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
+This revised definition makes it possible to outline a theory's affordances in terms of X-interoperability.
+For example, a FAIR theory may be X-interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
 or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-This revised definition implies that theories have specific properties that incur affordances in terms of X-interoperability;
-for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form).
+If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
+we see how each of these properties incurs certain affordances in terms of X-interoperability (Table \@ref(tab:tabmeehl)).
 
@@
 
-If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
-As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved), <!-- AB : maybe add a caveat that there may be country-specific legislation, company property rights, etc.
-and that this is of course no legal advice-->
-although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].
-Criterion R1.2 is met by version control with Git and archival on Zenodo.
-Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
+If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
+Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+It should further have detailed provenance (R1.2), 
+which is achieved through version control with Git and archival on Zenodo.
+Finally, the (meta)data which meets domain-relevant community standards (R1.3).
+The Datacite metadata schema offers an initial template in this regard,
+and this paper takes one step towards establishing more fine grained community standards for FAIR theory.
 
@@
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
-This norm impedes scientific progress.
-Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
-In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,
-reflected in questions such as "who owns theory",
+As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologieGrondslagenVan1961].
+In FAIR theory workshops, we similarly notice reluctance to reusing and adapting existing theories.
+Students ask questions such as "who owns theory",
 and "who determines how a theory may be reused or changed"?
@@
 
-Licensing theories for reuse provides an unambiguous answer to such questions.
+Licensing theories for reuse unambiguously answers these questions.
+With the caveat that legislation may vary across contexts and jurisdictions, the following should not be interpreted as legal advice.
 In determining what license is appropriate for theory,
-it is important to consider that copyright law limits authors' rights based on the *idea-expression dichotomy* [@bently2010copyright],
-which holds that copyright explicitly does not
+a key consideration is that copyright law protects authors' rights according to the *idea-expression dichotomy* [@bently2010copyright].
+It explicitly does not
 *"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
-Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
-It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
-then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
-Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
-so copyright would pose few restrictions to Reusability in scholarly communication.
-Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR theory;
-it explicitly waives all rights and encourages reuse.
-In principle, CC0 does not require attribution.
-Nevertheless, is essential that scholars do comprehensively cite theory,
-including prior work that new theories are based on,
-even in absence of legal obligations to do so,
-to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].
-
-## Additional considerations
-
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code (which we also have successfully applied in the domain of reproducible research findings, see XXX).
-<!--AB cite WORCS and Aarons tutorial on how to do the same thing again..-->
-Using version control systems, like git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
-Git also facilitates diffuse and adversarial collaboration,
-as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
-In sum, version control using Git enables systematic, collaborative, and transparent theory development,
-enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
-
-Even if scholars wish to diverge substantially from prior theory,
-explicitly referring back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
-
-From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
-
-### Making Theories FAIR Accelerates Scientific Progress
-
-Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
-By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
-This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
-
-<!-- ## FAIR theory and Recognition & Rewards -->
-
-<!-- FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory. -->
-<!-- In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging. -->
+Copyright thus extends to creative works expressing a theory (e.g., writing, visual illustrations),
+but not to the underlying theoretical idea.
+It thus seems that theories expressed in prose or depicted visually - in other words, that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+<!-- If copyright limits Reusability and does not cover ideas in their purest form (like formal theories), -->
+<!-- then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection. -->
+Another important consideration is that academic research is covered under "fair use" exemptions to copyright.
+<!-- so copyright would pose few restrictions to Reusability in scholarly communication. -->
+Given that copyright does not cover ideas in their purest form (like formal theories), and that academic use is likely exempted from its protection,
+in many cases, it may be sensible to explicitly license theories in a way that encourages Reusability.
+The CC0 (no rights reserved) license does this by explicitly waiving all rights and encouraging reuse.
+
+Aside from legal conditions for reuse, there are also social considerations.
+For example, while a CC0 license does not require attribution,
+it is nonetheless essential that scholars comprehensively cite theory and related works to comply with the norms of open scholarship [@aalbersbergMakingScienceTransparent2018].
+Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
+It is also possible to create or adopt a "Code of Conduct" which prescribes
 
@@
 but rather, as one proposal for a FAIR-compliant implementation of theory.
-At the time of writing (2024), the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
-Nevertheless, it is important to stress that alternatives to GitHub  (Gitlab, Bitbucket, etc.) and Zenodo (e.g., institutional repositories) exist.
-The principles described here
-(using version control and archiving major versions) could be implemented in a different workflow.
-
+The guiding principle of our implementation is to align and build upon existing successful open science infrastructures to the maximum possible extent.
+At the time of writing (2024), 
+the value of using Git for version control of academic research is well-established,
+and the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
+Zenodo and GitHub are also integrated with the Open Science Framework (OSF),
+a popular platform in psychology.
+Creating a front page on the OSF increases the visibility of a FAIR theory,
+while the integration with Zenodo and GitHub removes the need for uploading and maintaining the same information on multiple platforms.
+While we make use of specific open science infrastructures, it is important to stress that our workflow illustrates general principles which can also be implemented using other open science infrastructures.
 The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
@@
 Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
-While FAIR theory does not directly resolve the problem of strategic ambiguity, it does provide a framework within which scholars can increase the precision and formalization of theories.
+While FAIR theory does not directly reduce ambiguity,
+it provides a framework within which scholars can iteratively increase precision and formalization.
 FAIR principles also facilitates new ways of collaboration,
@@
 <!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->
+
+## How to Incentivize FAIR Theory Development
+
+FAIR theory requires a departure from contemporary practice.
+Several factors can expedite such a culture change.
+One key factor is the *recognition and rewards* movement:
+practices for evaluating scientific output are evolving, with initiatives like the [*Declaration on Research Assessment* (DORA)](https://sfdora.org/read/) and [Coalition for Advancing Research Assessment](https://coara.eu/)
+promoting the use of more diverse and meaningful metrics beyond journal impact factors.
+Modular publishing capitalizes on these changing metrics,
+and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory [@kirczModularityNextForm1998].
+Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
+to expedite its effective reuse and iterative improvement.
+A second factor is to lower barriers for the adoption of FAIR theory by building upon existing widely adopted open science infrastructures.
+For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
+Barriers of entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
+Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
+These materials allow teachers to incorporate theory development into their curriculum without investing substantial time into course development,
+thus educating the next generation to make use of and contribute to FAIR theory.
+Finally, community building plays an important role;
+the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
+Authors can also share their FAIR theories with other early adopters by submitting them to the "FAIR Theory Community" on Zenodo.
 

### 2025-01-13 manuscript.Rmd
suggest a new definition of theory

@@
 and many of those definitions are consistent with the FAIR theory principles.
-This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data [bogen/woodward].
+@scientifictheories describe different schools of thought when it comes to scientific theories: the syntactic, the semantic, and the pragmatic view.
+The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch.
+2], where domain refers to the scientific field (such as psychology or physics), with each domain having it's own theoretical terms (such as "intelligence" in psychology).
 Meehl [-@meehlAppraisingAmendingTheories1990] provides guidance as to what kinds of "statements" such a theory might contain:
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
-
-Some have defined a model as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
-This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-While there may be philosophical differences,
-these differences are largely irrelevant for the purpose of making theories more FAIR.
-When organizing knowledge, we think it is helpful to view theories and models derived from them as existing along a continuum of specificity,
-from broad to narrow, from general to specific,
-where a theory has a relatively broader scope and may contain one or more models as specific instances.
-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
-Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
--->
-For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
-From this theory, we could derive a more specific *statistical model* by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
+The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematical tools.
+It shifts the focus from theories as collections of sentences to mathematical models.
+The term model is not uniquely defined within the literature, one rather general definition would be a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
+-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations be
+tween concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
+The pragmatic view argues that there might not be one structure or definition of scientific theories, but instead, different scientific domains might very well define theory differently.
+It also argues that nonformal aspects (e.g. commonly used analogies) and practice (e.g. experimental designs) can be an important part of scientific theories.
+Since the primary purpose of FAIR theories should be to help advancing science, we believe that it is best left to the scientists to decide which parts of theory, models, or other aspects should belong to a specific FAIR theory, depending on whether certain information is assumed to be useful to the their scientific (sub)community.
+As a FAIR theory evolves, the nature of information tracked might also change.
+For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
+From this theory, more specific *statistical/mathematical models* could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
 This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
-Or, we could derive an even more specific *generative/computational model*, which is completely parametrized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
-Note that broadness and narrowness are relative terms,
-and one person's theory may be another person's model.
-This definition also implies that FAIR models are a special case of FAIR theory.
+Then, an even more specific *generative/computational model* could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.
+longitudinal designs observing change over time), measurement tools (e.g.
+different questionnaires used to assess the same construct), or study subjects (e.g.
+specific strains of rats).
 

### 2025-01-14 manuscript.tex
update 2025-01-14 07:02:28.379281

@@
 \makeatother
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8100}
+\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8368}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# eed54ad2.
+This is a preprint paper, generated from Git Commit \# cca0d45c.
 
@@
 To this end, we introduce ``FAIR theory'':
-a digital representation of a scientific theory, compliant with the FAIR principles.
-By improving the efficiency of scholarly communication, FAIR theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.
+a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
+compliant with the FAIR principles.
+FAIR theory has the potential to improve the efficiency of scholarly communication and
+accelerate cumulative knowledge acquisition.
 
@@
 
---\textgreater{}
-
 \subsection{Making Theory FAIR}\label{making-theory-fair}
@@
 The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
-Merely publishing theory -- as in a classic research article -- does not make it open;
+Merely publishing theory in a research article does not make it open;
 to be open, theory should adhere to established open science standards.
@@
 
+\subsubsection{Modular Publishing}\label{modular-publishing}
+
+FAIR theory is an example of modular publishing (\textbf{kirczModularityNextForm1998?}).
+At present, the primary unit of social scientific communication is the academic paper.
+A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
+Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
+with adequate metadata that is indexed in standardized repositories (\textbf{vandesompelRethinkingScholarlyCommunication2004?}).
+Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
+Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
+which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
+To maintain a persistent record of scholarly communication,
+Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
+which is used by many academic publishers.
+Finally, the \href{}{DataCite Metadata Schema} offers a standard way to document the nature of relationships between information artifacts.
+For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
+and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
+Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.
+
+\subsubsection{Version Control}\label{version-control}
+
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts.
+Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
+Git tracks line-by-line changes to text-based files,
+and maintains a complete history of those changes.
+It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
+Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
+The present paper considers advantages of Git for FAIR theory.
+Git enables explicitly comparing versions of a file (or: theory),
+incorporating changes by different authors,
+and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
+This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
+
+\section{The FAIR Principles}\label{the-fair-principles}
+
 \subsection{Findability}\label{findability}
@@
 Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
-Finally, these metadata should be registered or indexed in a searchable registry (F4).
-This final criterion is less straightforward.
-Ideally, FAIR theories should be indexed in search engines used by academics, like Google Scholar.
-At present, however, these search engines are designed to index traditional print publications.
-The \emph{data paper} solves this problem for research data;
-the idea is that scholars publish a paper (or even preprint) as documentation for the data resource {[}REF McGIllivray on data papers{]}.
-The data paper is indexed by search engines, and in turn points to the relevant information artifact.
-The same solution could be applied to theories - but it seems superfluous to generate papers whose only purpose is to redirect to a specific resource.
-Another solution is to manually index FAIR theories,
-for example by adding them to one's Google Scholar profile,
-or entering them in PURE.
-
-At present, theories have poor findability, which impedes cumulative knowledge acquisition.
-One factor contributing to theories' lack of Findability is the lack of standardized metadata, or even a standardized keyword to signal the presence of theory within a paper - terms like ``theory'', ``model'', and ``framework'' are used interchangeably.
-To curb this trend, we suggest using the keyword \texttt{"FAIRtheory"} for all resources that constitute or reference a FAIR theory (separating the words \texttt{FAIR} and \texttt{theory} by a space or hyphen would lead them to be interpreted as separate tokens in many search engines.
-This would allow theoretical resources to be systematically indexed, tagged, and made searchable.
-Another factor contributing to the present lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
-A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
-This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
-similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
-As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.
+
+Finally, metadata should be registered or indexed in a searchable registry (F4).
+Zenodo and GitHub are both searchable.
+Standardized metadata further enhance Findability in these repositories.
+The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}).
+Furthermore, a standard keyword can be used; we suggest using the keyword \texttt{"fairtheory"} for all resources that constitute or reference a FAIR theory.
+
+Findability is substantially amplified if intended users of a resource know where to search for it.
+This is a known problem in relation to research data and software (\textbf{katzSpecialIssueSoftware2024?}).
+Regrettably, most academic search engines are designed to index traditional print publications, not other information artifacts.
+Since the status quo is to publish theories in papers,
+the FAIR requirements are met if scholars continue to do so,
+and additionally publish theories as separate information artifacts.
+The \texttt{"fairtheory"} keyword can also be used to signal the presence of theory within a paper.
+In the longer term, it may not be necessary to write a paper for each theory.
+If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
+Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
 
@@
 
-The Accessibility criteria serve to \emph{regulate} access, not to maximize it.
-These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
+The Accessibility principles pertain to \emph{regulating} access, not only maximizing it.
+They apply to theory with minor changes.
 Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
@@
 and Zenodo allows for access control of individual files/resources.
-An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics).
-In general, it makes sense to keep outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
+In general, it makes sense to retain outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
 
@@
 This property of good theories has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
-The current prevalence of strategic ambiguity renders psychological theory difficult to understand (Frankenhuis et al., 2023).
+At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
 It is important to acknowledge the \emph{indeterminacy of translation} (Quine, 1970):
@@
 Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
-The original interoperability principles can be rephrased somewhat to apply to theory.
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
-The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
+The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
 Instead, FAIR theory should, ad minimum,
-be instantiated as
-as a type of data that is human- and machine-readable with as few interpretative steps as possible,
+be instantiated in a human- and machine-readable datatype,
 as previously recommended (Van Lissa et al., 2021).
@@
 Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
-This is essentially a call to establish standardized ontologies,
-which are themselves a type of theory (Paul E. Meehl, 1990).
+Aside from the aforementioned Datacite metadata schema (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}),
+in the context of theory, this highlights the importance of establishing standardized ontologies.
 Thirdly, theory (meta)data should include qualified references to other (meta)data, including previous versions of the theory (I3).
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by linking the DOI of those nested theories ({``Contributing {Citations} and {References},''} n.d.).
+This can be achieved by cross-referencing the DOI of those nested theories ({``Contributing {Citations} and {References},''} n.d.).
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
-This can be achieved by using Git for version control and Zenodo for archiving.
-Git tracks the internal provenance of a theory repository; Zenodo is used to archive relations to external references (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
-
-As the original definition of interoperability was somewhat narrow (M. D. Wilkinson et al., 2016),
-the concept has recently been further refined in terms of facilitating ``successful communication between machines and between humans and machines'', where ``A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
-This definition invites the question: \emph{interoperable for what?}
-Suitable answers for FAIR theory may be:
-this theory is X-interoperable for deriving testable hypotheses,
+This is achieved by using Git for version control and Zenodo for archiving.
+Git tracks the internal provenance of a theory repository; Zenodo is used to cross-reference external relationships (e.g., papers that influenced the theory, previous theories that inspired it, models based upon the theory).
+
+Recent work points out that interoperability is not an all-or-nothing property.
+The concept of X-interoperability was introduced to answer the question: \emph{interoperable for what?}
+X-interoperability is defined as facilitating ``successful communication between machines and between humans and machines {[}, where{]} A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
+This revised definition makes it possible to outline a theory's affordances in terms of X-interoperability.
+For example, a FAIR theory may be X-interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
 or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
-This revised definition implies that theories have specific properties that incur affordances in terms of X-interoperability;
-for example, Table \ref{tab:tabmeehl} illustrates the affordances of Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form).
+If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
+we see how each of these properties incurs certain affordances in terms of X-interoperability (Table \ref{tab:tabmeehl}).
 
@@
 to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
-or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert, Van Lissa, \& Brandmaier, 2021),
+or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert et al., 2021),
 or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
@@
 
-If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
-As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
-although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements {[}REF TOP guidelines{]}.
-Criterion R1.2 is met by version control with Git and archival on Zenodo.
-Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
+If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
+Applied to FAIR theory, reusability requires that theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+It should further have detailed provenance (R1.2),
+which is achieved through version control with Git and archival on Zenodo.
+Finally, the (meta)data which meets domain-relevant community standards (R1.3).
+The Datacite metadata schema offers an initial template in this regard,
+and this paper takes one step towards establishing more fine grained community standards for FAIR theory.
 
@@
 \emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
-This norm impedes scientific progress.
-Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data (de Groot, 1961).
-In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,
-reflected in questions such as ``who owns theory'',
+As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (de Groot, 1961).
+In FAIR theory workshops, we similarly notice reluctance to reusing and adapting existing theories.
+Students ask questions such as ``who owns theory'',
 and ``who determines how a theory may be reused or changed''?
@@
 
-Licensing theories for reuse provides an unambiguous answer to such questions.
+Licensing theories for reuse unambiguously answers these questions.
+With the caveat that legislation may vary across contexts and jurisdictions, the following should not be interpreted as legal advice.
 In determining what license is appropriate for theory,
-it is important to consider that copyright law limits authors' rights based on the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010),
-which holds that copyright explicitly does not
+a key consideration is that copyright law protects authors' rights according to the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010).
+It explicitly does not
 \emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''}.
-Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
-It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
-then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
-Furthermore, even if copyright would apply, academic research is covered under ``fair use'' exemptions,
-so copyright would pose few restrictions to Reusability in scholarly communication.
-Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR theory;
-it explicitly waives all rights and encourages reuse.
-In principle, CC0 does not require attribution.
-Nevertheless, is essential that scholars do comprehensively cite theory,
-including prior work that new theories are based on,
-even in absence of legal obligations to do so,
-to meet the definition of Reusability (R1.2, Table \ref{tab:tabfair} and to comply with other definitions of open scholarship (Aalbersberg et al., 2018).
-
-\subsection{Additional considerations}\label{additional-considerations}
-
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts, like computer code (which we also have successfully applied in the domain of reproducible research findings, see XXX).
-Using version control systems, like git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
-Git also facilitates diffuse and adversarial collaboration,
-as independent researchers can create independent versions of existing theories through ``forking'', or suggest modifications to existing theories via ``pull requests''.
-In sum, version control using Git enables systematic, collaborative, and transparent theory development,
-enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence (Van Lissa, 2023).
-
-Even if scholars wish to diverge substantially from prior theory,
-explicitly referring back to it enables clear comparison of the differences (Ram, 2013).
-
-From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.
-
-\subsubsection{Making Theories FAIR Accelerates Scientific Progress}\label{making-theories-fair-accelerates-scientific-progress}
-
-Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized.
-By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge.
-This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.
+Copyright thus extends to creative works expressing a theory (e.g., writing, visual illustrations),
+but not to the underlying theoretical idea.
+It thus seems that theories expressed in prose or depicted visually - in other words, that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
+Another important consideration is that academic research is covered under ``fair use'' exemptions to copyright.
+Given that copyright does not cover ideas in their purest form (like formal theories), and that academic use is likely exempted from its protection,
+in many cases, it may be sensible to explicitly license theories in a way that encourages Reusability.
+The CC0 (no rights reserved) license does this by explicitly waiving all rights and encouraging reuse.
+
+Aside from legal conditions for reuse, there are also social considerations.
+For example, while a CC0 license does not require attribution,
+it is nonetheless essential that scholars comprehensively cite theory and related works to comply with the norms of open scholarship (Aalbersberg et al., 2018).
+Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
+It is also possible to create or adopt a ``Code of Conduct'' which prescribes
 
@@
 but rather, as one proposal for a FAIR-compliant implementation of theory.
-At the time of writing (2024), the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
-Nevertheless, it is important to stress that alternatives to GitHub (Gitlab, Bitbucket, etc.) and Zenodo (e.g., institutional repositories) exist.
-The principles described here
-(using version control and archiving major versions) could be implemented in a different workflow.
-
+The guiding principle of our implementation is to align and build upon existing successful open science infrastructures to the maximum possible extent.
+At the time of writing (2024),
+the value of using Git for version control of academic research is well-established,
+and the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
+Zenodo and GitHub are also integrated with the Open Science Framework (OSF),
+a popular platform in psychology.
+Creating a front page on the OSF increases the visibility of a FAIR theory,
+while the integration with Zenodo and GitHub removes the need for uploading and maintaining the same information on multiple platforms.
+While we make use of specific open science infrastructures, it is important to stress that our workflow illustrates general principles which can also be implemented using other open science infrastructures.
 The process described here can be largely automated in R using the \texttt{theorytools} package; see the package vignette on FAIR theory, \texttt{vignette("fairtheory",\ package\ =\ "theorytools")}.
@@
 Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
-While FAIR theory does not directly resolve the problem of strategic ambiguity, it does provide a framework within which scholars can increase the precision and formalization of theories.
+While FAIR theory does not directly reduce ambiguity,
+it provides a framework within which scholars can iteratively increase precision and formalization.
 FAIR principles also facilitates new ways of collaboration,
 leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
+
+\subsection{How to Incentivize FAIR Theory Development}\label{how-to-incentivize-fair-theory-development}
+
+FAIR theory requires a departure from contemporary practice.
+Several factors can expedite such a culture change.
+One key factor is the \emph{recognition and rewards} movement:
+practices for evaluating scientific output are evolving, with initiatives like the \href{https://sfdora.org/read/}{\emph{Declaration on Research Assessment} (DORA)} and \href{https://coara.eu/}{Coalition for Advancing Research Assessment}
+promoting the use of more diverse and meaningful metrics beyond journal impact factors.
+Modular publishing capitalizes on these changing metrics,
+and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory (\textbf{kirczModularityNextForm1998?}).
+Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
+to expedite its effective reuse and iterative improvement.
+A second factor is to lower barriers for the adoption of FAIR theory by building upon existing widely adopted open science infrastructures.
+For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
+Barriers of entry can also be lowered by simplifying workflows, which is the goal of the \texttt{theorytools} R-package.
+Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
+These materials allow teachers to incorporate theory development into their curriculum without investing substantial time into course development,
+thus educating the next generation to make use of and contribute to FAIR theory.
+Finally, community building plays an important role;
+the international network of open science communities, reproducibility networks, and other similar initiatives provide platforms for disseminating FAIR theories and related methodological innovations.
+Authors can also share their FAIR theories with other early adopters by submitting them to the ``FAIR Theory Community'' on Zenodo.
 

### 2025-01-14 manuscript.Rmd
minor tweak to Max' rewrite

@@
 
-There are different definitions of theory,
-and many of those definitions are consistent with the FAIR theory principles.
-@scientifictheories describe different schools of thought when it comes to scientific theories: the syntactic, the semantic, and the pragmatic view.
-The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch.
-2], where domain refers to the scientific field (such as psychology or physics), with each domain having it's own theoretical terms (such as "intelligence" in psychology).
-Meehl [-@meehlAppraisingAmendingTheories1990] provides guidance as to what kinds of "statements" such a theory might contain:
+## What is Theory?
+
+<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g.
+also such things as scientific practice (experimental design, materials)).
+-->
+<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
+<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added).
+-->
+Definitions of theory abound and are the subject of extensive scholarly debate.
+Given that a pluriformity of definitions are consistent with FAIR theory principles,
+our paper is not aligned with one particular definition.
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
+The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
+acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
+We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
 statements about the types of entities postulated (i.e., ontology),
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
-The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematical tools.
+
+The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
 It shifts the focus from theories as collections of sentences to mathematical models.
-The term model is not uniquely defined within the literature, one rather general definition would be a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
+The term "model" is not uniquely defined within the literature;
+it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
+This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
 -<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations be
 tween concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
-The pragmatic view argues that there might not be one structure or definition of scientific theories, but instead, different scientific domains might very well define theory differently.
-It also argues that nonformal aspects (e.g. commonly used analogies) and practice (e.g. experimental designs) can be an important part of scientific theories.
-Since the primary purpose of FAIR theories should be to help advancing science, we believe that it is best left to the scientists to decide which parts of theory, models, or other aspects should belong to a specific FAIR theory, depending on whether certain information is assumed to be useful to the their scientific (sub)community.
-As a FAIR theory evolves, the nature of information tracked might also change.
+
+The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
+It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.
+
+Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
+the method is not contained to any one particular definition.
+It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
+As the practice of FAIRification becomes more embedded,
+we expect that it will become increasingly clear what kind of information is useful.
+As a particular FAIR theory evolves, the nature of information tracked will likely change.
 For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.

### 2025-01-14 manuscript.Rmd
Update manuscript.Rmd
Co-authored-by: Maximilian Ernst <34346372+Maximilian-Stefan-Ernst@users.noreply.github.com>

@@
 This demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.
 

### 2025-01-14 manuscript.Rmd
continue writing

--- 2025-01-14 manuscript.Rmd
@@
   Test test.
-keywords          : "meta theory, theory formation, cumulative science, formal models"
+keywords          : "fairtheory, meta theory, theory formation, cumulative science, formal models"
 wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
@@
 Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-Two main concerns are that, first, social scientific theories lack precision compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
- and clarity
-In other words, social scientific theories lack *formalization*,
-which means that they do not make very accurate predictions,
-and are thus hard to falsify and difficult to understand on their own,
+One main concern is that social scientific theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+When theories do not make very accurate predictions,
+they are hard to falsify and difficult to understand on their own,
 without either substantial interpretation or additional background knowledge.
-A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
+A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
 
@@
 we decided to use it as an example for this tutorial.
-Note that, while the empirical cycle is not explicitly referred to as a "theory" by De Groot, he derives it from "a theory of thinking" by Selz (REF).
-We can thus consider it a meta-theory of the process of theory construction.
+The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
+The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
+Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
+We can thus consider it a meta-theory of theory construction.
 <!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
-The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969:
+The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:
 
@@
 If we compare it to the levels of theory formalization [@guestHowComputationalModeling2021],
-De Groot's theory is either at the "theory" or "specification" level.
-It consists of a series of natural language statements.
+it is defined at either the "theory" or "specification" level.
 We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language:
@@
 
-This language describes the model as a directed graph.
+This implementation describes the model as a directed graph.
 Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,

### 2025-01-14 dk_example/example.tex
empty lines
add to dk example

@@
 
-With $-$ signifying a negative association and $\simeq$ meaning measured by.
+With $-$ signifying a negative association and $\simeq$ signifying measured by.
 
@@
 \section{Mathematical Formulation}
+Define random variables
+\begin{itemize}
+ \item $s^*$ denoting skill
+ \item $\epsilon$ denoting measurement error, with $\Exp[\epsilon] = 0$, $\epsilon$ independent of all other random variables included in the model
+ \item $s^*_s$ denoting self-assessed skill
+\end{itemize}
 
+\noindent Then we define performance $p$ as
+\begin{equation} \label{p}
+  p \coloneq s^* + \epsilon
+\end{equation}
+and overconfidence $oc^*$ as
+\begin{equation} \label{oc}
+  oc^* \coloneq s^*_s-s^*
+\end{equation}
+and expected performance $p_e$ as
+\begin{equation} \label{ep}
+  p_e \coloneq s^* + oc^*
+\end{equation}
+Overconfidence $oc^*$ is measured by overestimation $oe$ defined as
+\begin{equation}
+  oe \coloneq p_e - p
+\end{equation}
+
+\noindent From eq. \ref{oc} and \ref{ep} it follows that $p_e = s^*_s$ and further from eq. \ref{ep} and \ref{p} we see
+\begin{align}
+  oe &= p_e - p \\
+  &= (s^* + oc^*) - (s^* + \epsilon) \\
+  &= oc^* - \epsilon
+\end{align}
 \section{(Statistical) Models}
@@
 \begin{itemize}
- \item use different test performances to calculate skill and overconfidence, but a small test-retest correlation can bias the estimate towrds zero then (Krueger and Mueller, 2002)
+ \item use different test performances to calculate skill and overconfidence, but a small test-retest correlation can bias the estimate towards zero then (Krueger and Mueller, 2002)
 \end{itemize}
+
+\section{Common Experimental Paradigm}
+\begin{itemize}
+ \item participants take a test
+ \item participants guess their test performance either before or after the test
+ \item bottom quartile performers vastly overestimate their performance, while top quartile performers access their performance more accurately
+\end{itemize}
+
+There are many replications for different populations and different tasks.
 
@@
 
+\section{Notes}
+\begin{itemize}
+ \item difference in metacognitive skills drives the relationship between skill and overconfidence
+\end{itemize}
+
 \end{document}

### 2025-01-15 Zenodo.Rmd

---
title: "Make FAIR Theory citable using GitHub and Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Make FAIR Theory citable using GitHub and Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "vignettes.bib"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>"
)
```

# Make FAIR Theory citable using GitHub and Zenodo

This tutorial is derived from Module 5, Task 2 of @jon_tennant_2018_1434288.

Estimated time to complete: 45-60 minutes.

<!-- fig.width=600, fig.height=861,  -->

```{r fig.align='center', out.width = "70%", fig.cap="Conceptual workflow for this task."}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/Task2.png?raw=true")
```

<!-- <p align="center"> -->
<!--   <img src="https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/Task2.png?raw=true" alt="Task 2 workflow" width="600" height="861" style="margin-right: 30px; margin-left: 10px;" onmouseover="this.width='1200'; this.height='1722'" onmouseout="this.width='600'; this.height='861'"> -->
<!-- </p> -->

## Foreword

Although the integration of GitHub and Zenodo makes it really easier to work with these tools nowadays (January 2019), it is important to stress that there are alternatives to GitHub (Gitlab, Bitbucket, ...) and alternatives to Zenodo (Other repositories might be more suited to your community, you might ask your colleagues). For instance, one can work with Gitlab and manually upload each new versions to your university repository, getting a DOI. The principles (working with a version control system online, and archiving major versions 
in a repository which provides a persistent unique identifier) can be applied in different workflow.

## Set up a GitHub repository

Find out how to set up a GitHub repository using the [GitHub documentation](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/),
or by reading this paper on reproducible research workflows [@vanlissaWORCSWorkflowOpen2021].

**Important**: Here's a checklist of steps to complete:

- [ ] The repository contains a theory file
- [ ] The repository contains a LICENSE file (we recommend CC0)
- [ ] The repository contains a README file; we recommend the following sections:
    * Title, prefaced with `# FAIR theory: The Theory's Name`*
    * Description: A plain-text description of the theory and its scope
    * Interoperability: This section, also often labeled "Getting Started", "Instructions", or "How to Use" might be better labeled Interoperability from a FAIR perspective. Following the discussion in the respective section, we propose focusing on X-interoperability, and telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would provide instructions here on how to plot a DOT graph.
    * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
    * License: As a complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal constraints to reuse*.
    * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
    * Citing this work: Tell users how to cite the theory.
- [ ] The repository settings must be set to 'public', not private, so Zenodo can discover it.

<!-- ## Choose your GitHub repository -->

<!-- Once on your GitHub project listings page at [github.com](https://github.com) head to the 'Repositories' tab. Select which repository you would like to archive, and open it up. -->

<!-- <br/> -->

## Login to Zenodo

Head over to [zenodo.org](https://zenodo.org). Zenodo is a platform where you can permanently archive your code and other project elements. Zenodo does this by assigning projects a **Digital Object Identifier** (DOI), which also helps to make the work more citable. This is different to GitHub, which acts as a place where the actual work on a project takes place, rather than long-term archiving of it. At GitHub, content can be modified, deleted, rewritten, and irreversibly changed, which makes it a bit concerning to be used for longer lasting referencing purposes. Zenodo offers more security and permanence for research outputs.

<!-- S -->

```{r fig.align='center', out.width = "70%", fig.cap="Sign up for Zenodo"}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/zenodo.png?raw=true")
```

If you already have a Zenodo account, this is easy.
If not, follow the steps to create one — you can **login using your GitHub account** or ORCID profile to make things simpler, as Zenodo has a built in integration for it. This might be easier than creating yet another research account and profile.

## Authorise GitHub to connect with Zenodo

On the Zenodo website authorise it to connect to your GitHub account in the '[Using GitHub](https://zenodo.org/account/settings/github/)' section. Here, Zenodo will redirect you to GitHub to ask for permissions to use '[webhooks](https://developer.github.com/webhooks/)' on your repositories. You want to authorise Zenodo here with the permissions it needs to form those links.

```{r fig.align='center', out.width = "70%", fig.cap="Authorize Zenodo to connect with GitHub"}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/zenodo_github.png?raw=true")
```

## Select the repository to archive

If you have got this far, this means that Zenodo is now authorised to configure the repository webhooks that it needs to archive the repository and issue it a DOI. To do this, on the Zenodo website navigate to the [GitHub repository listing page](https://zenodo.org/account/settings/github/) and simply "flip the switch" next to your repository.

```{r fig.align='center', out.width = "70%", fig.cap="Enable individual GitHub repositories to be archived in Zenodo"}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/enabled_repos.png?raw=true")
```

## *Optional:* Check repository settings

If you were successful, you have now set up a new webhook between Zenodo and your repository.

Optionally, you can verify this. In GitHub, click on the settings for your repository, and the Webhooks tab on the left hand side menu. This should display the new Zenodo webhook configured to Zenodo. Note, it may take a little time for the webhook listing to show up.

```{r fig.align='center', out.width = "70%", fig.cap="Check that webhooks are enabled for your GitHub repository."}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/webhooks.png?raw=true")
```

## Create a new release

To archive a  repository on Zenodo, you must create a new release. This is done in the 'releases' tab on the GitHub repository front page (top center).

The first time you archive a repository is known as the 'first release'. All subsequent releases are automatically archived.

```{r fig.align='center', out.width = "70%", fig.cap="Check that the repository first release was successful."}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/first_release.png?raw=true")
```

To archive your repository, click 'Create a new release' back in Zenodo. In the form, you have to choose a tag and release title.
Unless you use another convention for versioning (which should be documented in README.md),
we suggest using semantic versioning for both, starting with version `0.1.0`.
Do not use the label `v0.1.0`, as this has caused compatibility issues with Zenodo in the past.
Fill out this value in both fields (tag and title).

```{r fig.align='center', out.width = "70%", fig.cap="Create a new release."}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/create_release.png?raw=true")
```

Finally, click 'publish release', and your archive will be published and versioned on GitHub.

To verify that your release was archived on Zenodo and assigned a DOI,
you need to visit the [Uploads](https://zenodo.org/deposit) tab.

```{r fig.align='center', out.width = "70%", fig.cap="Check the new release has been uploaded."}
knitr::include_graphics("https://github.com/OpenScienceMOOC/Module-5-Open-Research-Software-and-Open-Source/blob/master/content_development/images/upload_release.png?raw=true")
```

## Entering Meta-Data

By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
To document our archive as a FAIR theory requires adding some extra information on Zenodo.
On Zenodo click the [Upload](https://zenodo.org/deposit) tab in the main menu, where you should find your newly uploaded repository.

```{r fig.align='center', out.width = "70%", fig.cap="Click the orange Edit button."}
knitr::include_graphics("zenodo_edit.png")
```

Click the orange `Edit` button, and supply the following information:

- [ ] Resource type: Set this to `Model`
- [ ] Title: Verify that the title is prefaced with `FAIR theory:`
- [ ] Keywords and subjects: Add the keyword `fairtheory`
- [ ] Related works: List the DOIs/identifiers of related (print) works. Use the `Relation` field as appropriate. For example:
    + `Is documented by` a theory paper you wrote, in which you introduce this FAIR theory
    + `Is derived from` an existing theory, which was published in print (paper, book chapter) but not made FAIR

To save these changes, click 'Publish'.

> **Pro-tip**: Check the `Citation` field on the Zenodo page, and copy-paste it into the README file of your GitHub repo to make cross-linking even easier (or refer users to the Zenodo page to find the citation, which obviates the need to manually update this information). Click the DOI bage in the `Details` field to get instructions on how to add a clear highlighted DOI badge to your GitHub repository, for users to see and make use of your DOI:

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1323437.svg)](https://doi.org/10.5281/zenodo.1323437)

The GitHub/Zenodo integration will assign one "mother-DOI" to the project, as well as a unique DOI to each version/release of the FAIR theory. This enables users to refer to and cite specific versions of the theory. The list of authors for the citation is automatically determined by the GitHub user account names used by the repository - this can be edited on Zenodo, as explained above. DOIs used in Zenodo are registered through the [DataCite](https://www.datacite.org/) service.

**CONGRATULATIONS!!**

Your FAIR theory is now archived in Zenodo, and with a DOI that can be versioned to reflect updates to the repository version through time. You should be able to see details of this on the GitHub Zenodo page for your repository. This also means that your archived projects can get picked up by other indexing services and search engines that use DOIs too.

Providing a long-term archive and a DOI for your work is required for others to be able to properly cite it, as this provides basic citation metadata. For Open Science, it is important to be able to comprehensively cite the resources that you use in your research, including theory, and this workflow enables that to happen, in line with best practices. Making theory FAIR also helps elevate the standard of theory to that of the standard of other research outputs, like papers and software.

> **Pro-tip**: Is your research funded by an EU grant? Now you can directly connect your FAIR theory to your grant by updating the grant section of the metadata on the project's Zenodo record. This massively helps to increase its discoverability!

## Checklist for citing your project

So now you have a sustainably archived GitHub repository in Zenodo that is ready to be re-used and cited! Before continuing, make sure that you have:

- [ ] Linked your GitHub project to Zenodo. If you see a complete copy of your GitHub repository in Zenodo then things are working.
- [ ] Zenodo and GitHub integrated setup works nicely. For example have all the author names, and correct project title come across to Zenodo. If not, or if authors just have nicknames you can edit these details in Zenodo.
- [ ] Project has a first release, with a DOI. You should have a DOI displayed on your projects Zenodo page. This first DOI is called the 'concept DOI' and is the master DOI linking to all subsequent release DOIs. Copy this DOI link and embed it in your GitHub projects README page. You're done!

### Additional resources

[Making your code citable](https://guides.github.com/activities/citable-code/) - GitHub Guides.

# References

### 2025-01-15 manuscript.Rmd

@@
 One main concern is that social scientific theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
-When theories do not make very accurate predictions,
+When theories do not make precise predictions,
 they are hard to falsify and difficult to understand on their own,
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' *empirical content* [REF] as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes  that would otherwise get away with their vague formulation in verbal theories.
-For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented [@lewandowsky2010computational].
-More precise theories makes them easier to falsify,
-which necessitates revising them,
-thus advancing our principled understanding of the phenomena they describe.
-FAIR theory does not require theories to be formal,
-and formal theory can be represented in a way that is not FAIR.
-It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
-The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
-For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
-Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \@ref(fig:figec)),
-but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
-it is not FAIR.
-FAIR theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).
+Greater formalization increases theories' *empirical content* [REF] because it involves expressing ideas as precise statements.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
+Without committing to specific implementations a-priori,
+the theory becomes hard to test.
+Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
+More precise theories are easier to falsify,
+which necessitates specific revisions and advances our principled understanding of the phenomena they describe.
+
+To some extent, FAIRness and formalization are orthogonal.
+<!-- : FAIR theory does not require theories to be formal, -->
+<!-- and formal theory can be represented in a way that is not FAIR. -->
+<!-- We stress that  -->
+FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
+rather, it increases the precision with which they are communicated.
+The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
+For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
+Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
+FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
 
@@
 
+### Semantic Versioning
+
+Aside from technical solutions, version control is a social process as well.
+On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
+For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
+In computer science, these challenges are navigated by assigning version numbers.
+Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to information artifacts.
+Whereas version control tracks changes,
+semantic versioning communicates what those changes mean to users of the theory.
+We propose the following adaptation of semantic versioning for theories:
+
+```
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make backwards incompatible changes, i.e., the theory now contains empirical statements that are at odds with a previous version
+MINOR version when you expand the set of empirical statements in a backward compatible manner (i.e., the previous version is subsumed within the new version)
+PATCH version when you make backward compatible bug fixes,
+cosmetic changes, fix spelling errors, or add clarifications
+```
+
+Semantic versioning guides the social process of theory development, communicating how much a theory is changing over tiem.
+
 # The FAIR Principles
@@
 in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
-In a similar way, it would become possible to compare all theories of a specific phenomenon,
+In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
 or to study structural properties of theories.
@@
 Furthermore, a standard keyword can be used; we suggest using the keyword `"fairtheory"` for all resources that constitute or reference a FAIR theory.
-<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines. -->
+<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->
 
@@
 and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.
+In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
+and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
+will advance its Accessibility.
 
@@
 mathematical formulae,
-and statements expressed in some axiomatic system.
+and statements expressed in some formal language.
 Examples of the latter include pseudo-code,
@@
 While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language <!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
+the same image represented in the DOT language [@DOTLanguage]<!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not require attribution,
-it is nonetheless essential that scholars comprehensively cite theory and related works to comply with the norms of open scholarship [@aalbersbergMakingScienceTransparent2018].
+For example, while a CC0 license does not mandate attribution,
+it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation [@aalbersbergMakingScienceTransparent2018].
 Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
@@
 The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
-In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013, @vanlissaWORCSWorkflowOpen2021].
+In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
 To version control our project, we initiate a Git repository in the project folder.
@@
 
-- Set the *resource type* to `Model`
-- Verify that the *title* is prefaced with `FAIR theory:`
-- Add the *keyword* `fairtheory`
-- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to increase its findability
+- Set the *resource type* to `Model`; this ensures proper archival in Zenodo
+- Verify that the *title* is prefaced with `FAIR theory:`; this allows sentient readers to recognize the work as a FAIR theory
+- Add the *keyword* `fairtheory`; this aids search engine indexation
+- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building
 - List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
@@
 <!-- CJ: Aaron, I moved your part here because I don't think it relates to interoperability. But perhaps there's another place where it fits better? -->
-Aside from technical solutions to However, "[i]n systems with many dependencies, releasing new [...] versions can quickly become a nightmare."
-The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
-In software development this problem is colloquially called "dependency hell" and usually addressed via semantic versioning.
-Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
-We propose the following adaptation of semantic versioning for theories:
-
-```
-Given a version number MAJOR.MINOR.PATCH, increment the:
-
-MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
-MINOR version when you expand the set of empierical statements in a backward compatible manner
-PATCH version when you make backward compatible bug fixes, or clarifications
-```
-
-Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
 

### 2025-01-15 manuscript.Rmd
update 2025-01-15 10:49:17.858478

@@
 a philosophical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
+research ideally follows a cyclical process with two phases (Figure \@ref(fig:figecs)).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 
-```{r}
+```{r eval = FALSE}
 library(tidySEM)
@@
 
-```{r figec, fig.cap="A take on the empirical cycle by De Groot"}
+```{r figec, eval = FALSE, fig.cap="A take on the empirical cycle by De Groot"}
 knitr::include_graphics("empirical_cycle.pdf")
@@
 
-MAJOR version when you make backwards incompatible changes, i.e., the theory now contains empirical statements that are at odds with a previous version
-MINOR version when you expand the set of empirical statements in a backward compatible manner (i.e., the previous version is subsumed within the new version)
+MAJOR version when you make backwards incompatible changes, i.e.,
+      the theory now contains empirical statements that are at odds
+      with a previous version
+MINOR version when you expand the set of empirical statements in a
+      backward compatible manner (i.e., the previous version is
+      subsumed within the new version)
 PATCH version when you make backward compatible bug fixes,
@@
 
-Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
-we decided to use it as an example for this tutorial.
-The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
-The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
-Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
-We can thus consider it a meta-theory of theory construction.
+We will use the empirical cycle as a running example for this tutorial.
+The resulting FAIR version of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
+The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969,
+<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
+<!-- We can thus consider it a meta-theory of theory construction. -->
 <!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
-The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:
+and consists of a series of natural language statements:
 
@@
 rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
+rsvg::rsvg_pdf(svg = "ec_groot.svg",
+               file = "ec_groot.pdf")
+rsvg::rsvg_pdf(svg = "ec_wagenmakers.svg",
+               file = "ec_wagenmakers.pdf")
+rsvg::rsvg_pdf(svg = "ec_lissa.svg",
+               file = "ec_lissa.pdf")
+```
+
+```{r figecs, figures-side, fig.show="hold", out.width="33%", fig.cap="The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)"}
+knitr::include_graphics("ec_groot.pdf")
+knitr::include_graphics("ec_wagenmakers.pdf")
+knitr::include_graphics("ec_lissa.pdf")
 ```
@@
 
-We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
-
-## 2. Creating a Project Folder
-
-Create a new folder and copy the theory file from the previous step into it.
+## 2. Creating the Project Folder
+
+We can write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`, and save it to a new folder.
 To help meet the Interoperability and Reusability criteria,
-add two more files:
-A README.md file with instructions for future users of your theory,
-and a LICENSE file with the legal conditions for reuse.
+add two more files.
+Include a `README.md` file with instructions for future users of your theory (see: [What to Include in a README?]())
+and a `LICENSE` file with the legal conditions for reuse.
 We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
-
-### What's in a README?
-
-The readme should contain information to help people get started with using your FAIR theory.
-We suggest the following elements:
-
-* Title, prefaced with `# FAIR theory: The Theory's Name`
-* Description: A plain-text description of the theory and its scope
-* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
-From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
-We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
-* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
-* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
-* Citing this work: Tell users how to cite the theory.
-Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
-The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
-* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Again, this is redundant with metadata entered in Zenodo (step 5).
-We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
-For example, we referenced the original empirical cycle paper here:
-
-```
-This repository contains an implementation of the "empirical cycle",
-a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
-
-> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
-Foundations of inference and research in the behavioral sciences.
-De Gruyter Mouton. https://doi.org/10.1515/9783112313121
-```
 
@@
 <!-- Next, in the terminal window, run the following lines of code sequentially: -->
-
 <!-- ``` -->

### 2025-01-17 manuscript.Rmd
Update manuscript.Rmd
Added myself as author

@@
       - Writing – review & editing
+  - name: "Noah N.N. van Dongen"
+    affiliation: "6"
+    role:
+      - Writing – review & editing
 affiliation:
@@
     institution   : "Ludwig-Maximilians-Universität München, Germany"
+  - id            : "6"
+    institution   : "University of Amsterdam, the Netherlands"
 authornote: |

### 2025-01-22 manuscript.tex
Minor edits by Felix

@@
   {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
+    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
     {}{\InnerPatchFailed}%
@@
 {}{\PatchFailed}
-\makeatother
-\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8368}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8200}
 \usepackage{lineno}
@@
   pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
   pdflang={en-EN},
-  pdfkeywords={meta theory, theory formation, cumulative science, formal models},
+  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
   hidelinks,
@@
 \title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
-\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix D. Schönbrodt\textsuperscript{5}}
+\author{Caspar J.
+Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M.
+Brandmaier\textsuperscript{2,3,4}, Felix D.
+Schönbrodt\textsuperscript{5}, \& Noah N.N.
+van Dongen\textsuperscript{6}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# cca0d45c.
+This is a preprint paper, generated from Git Commit \# 09ba7dc6.
 
@@
 Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D.
-Schönbrodt: Conceptualization, Writing -- review \& editing.
+Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N.
+van Dongen: Writing -- review \& editing.
 
@@
 \affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany}
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands}
 
@@
 Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-Two main concerns are that, first, social scientific theories lack precision compared to theories in the physical sciences (Szollosi \& Donkin, 2021).
-and clarity
-In other words, social scientific theories lack \emph{formalization},
-which means that they do not make very accurate predictions,
-and are thus hard to falsify and difficult to understand on their own,
+One main concern is that social scientific theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
+When theories do not make precise predictions,
+they are hard to falsify and difficult to understand on their own,
 without either substantial interpretation or additional background knowledge.
-A second concern is the lack of transparent and participative scholarly communication about psychological theory and their development over time.
+A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
 
@@
 a philosophical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
+research ideally follows a cyclical process with two phases (Figure \ref{fig:figecs}).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
-
-\begin{figure}
-\centering
-\includegraphics{empirical_cycle.pdf}
-\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
-\end{figure}
 
@@
 The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
+In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
@@
 I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
-I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
+I2 & (Meta)data use vocabularies that follow FAIR principles & Theory and its associated metadata\ \ use vocabularies that follow FAIR principles & Rephrased\\
 I2S.1 & - &  & \\
 I2S.2 & - &  & \\
-I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
+I3 & (Meta)data include qualified references to other (meta)data & Theory and its associated metadata includes qualified references to other theories and their metadata, including previous versions of the theory & Extended\\
 I4S & - &  & \\
 R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
-R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
-R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
-R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
+R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory and its associated metadata are released with a clear and accessible license & Rephrased\\
+R1.2 & (Meta)data are associated with detailed provenance & Theory and its associated metadata are associated with detailed provenance & Rephrased\\
+R1.3 & (Meta)data meet domain-relevant community standards & Theory and its associated metadata and documentation meet domain-relevant community standards & Rephrased\\
 \bottomrule
@@
 
-There are different definitions of theory,
-and many of those definitions are consistent with the FAIR theory principles.
-This paper defines theory as an integrated set of statements that explain phenomena consistently evidenced by patterns in data {[}bogen/woodward{]}.
-Meehl (1990) provides guidance as to what kinds of ``statements'' such a theory might contain:
+\subsection{What is Theory?}\label{what-is-theory}
+
+Definitions of theory abound and are the subject of extensive scholarly debate.
+Given that a pluriformity of definitions are consistent with FAIR theory principles,
+our paper is not aligned with one particular definition.
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
+The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
+acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
+We recognize the syntactic view in Meehl's hierarchy of ever-more specific ``statements'' a theory might contain (1990):
 statements about the types of entities postulated (i.e., ontology),
@@
 
-Some have defined a model as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
-This invites the question: if a FAIR theory is a specific instantiation of theory, how does FAIR theory differ from a model?
-While there may be philosophical differences,
-these differences are largely irrelevant for the purpose of making theories more FAIR.
-When organizing knowledge, we think it is helpful to view theories and models derived from them as existing along a continuum of specificity,
-from broad to narrow, from general to specific,
-where a theory has a relatively broader scope and may contain one or more models as specific instances.
-For example, following Meehl, we could envision a theory that merely specifies how specific constructs are causally connected.
-From this theory, we could derive a more specific \emph{statistical model} by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
+The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
+It shifts the focus from theories as collections of sentences to mathematical models.
+The term ``model'' is not uniquely defined within the literature;
+it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
+This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
+-
+
+The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
+It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.
+
+Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
+the method is not contained to any one particular definition.
+It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
+As the practice of FAIRification becomes more embedded,
+we expect that it will become increasingly clear what kind of information is useful.
+As a particular FAIR theory evolves, the nature of information tracked will likely change.
+For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are defined and causally connected.
+From this theory, more specific \emph{statistical/mathematical models} could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
 This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
-Or, we could derive an even more specific \emph{generative/computational model}, which is completely parametrized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
-Note that broadness and narrowness are relative terms,
-and one person's theory may be another person's model.
-This definition also implies that FAIR models are a special case of FAIR theory.
-
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
-The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).
+
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., TODO), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety; Taylor, 2022).
+While the latter psychological factors might be narratively described as vague directional causal statements,
+the theory may also encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' \emph{empirical content} {[}REF{]} as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes that would otherwise get away with their vague formulation in verbal theories.
-For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented (Lewandowsky \& Farrell, 2010).
-More precise theories makes them easier to falsify,
-which necessitates revising them,
-thus advancing our principled understanding of the phenomena they describe.
-FAIR theory does not require theories to be formal,
-and formal theory can be represented in a way that is not FAIR.
-It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down.
-The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
-For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
-Conversely, a directed acyclic graph (DAG) is a type of formal theory (for example, see the empirical cycle in Figure \ref{fig:figec}),
-but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
-it is not FAIR.
-FAIR theory is thus consistent with, but does not require, formal theory (also see \emph{Accessibility}).
+Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it involves expressing ideas as precise statements.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
+Without committing to specific implementations a-priori,
+the theory becomes hard to test.
+Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
+More precise theories are easier to falsify,
+which necessitates specific revisions and advances our principled understanding of the phenomena they describe.
+
+To some extent, FAIRness and formalization are orthogonal.
+FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
+rather, it increases the precision with which they are communicated.
+The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
+For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
+Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
+FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).
 
@@
 
+\subsubsection{Semantic Versioning}\label{semantic-versioning}
+
+Aside from technical solutions, version control is a social process as well.
+On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
+For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
+In computer science, these challenges are navigated by assigning version numbers.
+Specifically, \emph{semantic versioning} comprises a simple set of rules for assigning version numbers to information artifacts.
+Whereas version control tracks changes,
+semantic versioning communicates what those changes mean to users of the theory.
+We propose the following adaptation of semantic versioning for theories:
+
+\begin{verbatim}
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make backwards incompatible changes, i.e.,
+      the theory now contains empirical statements that are at odds
+      with a previous version
+MINOR version when you expand the set of empirical statements in a
+      backward compatible manner (i.e., the previous version is
+      subsumed within the new version)
+PATCH version when you make backward compatible bug fixes,
+      cosmetic changes, fix spelling errors, or add clarifications
+\end{verbatim}
+
+Semantic versioning guides the social process of theory development, communicating how much a theory is changing over tiem.
+
 \section{The FAIR Principles}\label{the-fair-principles}
@@
 or through analogical modeling.
-In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
 For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
@@
 The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
-First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+First, this requires assigning a globally unique and persistent identifier (such as a DOI) to each theory (F1).
 Of the many services that provide DOIs for scientific information artefacts,
@@
 and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, embedding within shared contexts, and explication of assumptions, will advance its Accessibility.
+In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
+and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
+will advance its Accessibility (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
 
@@
 While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language for representing graphs does meet this ideal.
+the same image represented in the DOT language (\textbf{DOTLanguage?}) for representing graphs does meet this ideal.
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not require attribution,
-it is nonetheless essential that scholars comprehensively cite theory and related works to comply with the norms of open scholarship (Aalbersberg et al., 2018).
+For example, while a CC0 license does not mandate attribution from a legal perspective,
+the norms of good scientific practice require that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
 Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a ``Code of Conduct'' which prescribes
+It is also possible to create or adopt a ``Code of Conduct'' which prescribes how others can contribute to or reuse a theory.
 
@@
 
-Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
-we decided to use it as an example for this tutorial.
-Note that, while the empirical cycle is not explicitly referred to as a ``theory'' by De Groot, he derives it from ``a theory of thinking'' by Selz (REF).
-We can thus consider it a meta-theory of the process of theory construction.
-The empirical cycle is described on page 28 of De Groot and Spiekerman (1969):
+We will use the empirical cycle as a running example for this tutorial.
+The resulting FAIR version of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
+The empirical cycle is described on page 28 of De Groot and Spiekerman (1969),
+and consists of a series of natural language statements:
 
@@
 If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
-De Groot's theory is either at the ``theory'' or ``specification'' level.
-It consists of a series of natural language statements.
-We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language:
+it is defined at either the ``theory'' or ``specification'' level.
+We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \ref{tab:tabfair}):
 
@@
 
-This language describes the model as a directed graph.
+\begin{figure}
+\includegraphics[width=0.33\linewidth]{ec_groot} \includegraphics[width=0.33\linewidth]{ec_wagenmakers} \includegraphics[width=0.33\linewidth]{ec_lissa} \caption{The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)}(\#fig:figecs, figures-side)
+\end{figure}
+
+This implementation describes the model as a directed graph.
 Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
@@
 
-We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
-
-\subsection{2. Creating a Project Folder}\label{creating-a-project-folder}
-
-Create a new folder and copy the theory file from the previous step into it.
+\subsection{2. Creating the Project Folder}\label{creating-the-project-folder}
+
+We can write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}, and save it to a new folder.
 To help meet the Interoperability and Reusability criteria,
-add two more files:
-A README.md file with instructions for future users of your theory,
-and a LICENSE file with the legal conditions for reuse.
+add two more files.
+Include a \texttt{README.md} file with instructions for future users of your theory (see: \href{}{What to Include in a README?})
+and a \texttt{LICENSE} file with the legal conditions for reuse.
 We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
 
-\subsubsection{What's in a README?}\label{whats-in-a-readme}
-
-The readme should contain information to help people get started with using your FAIR theory.
-We suggest the following elements:
-
-\begin{itemize}
-\tightlist
-\item
-  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
-\item
-  Description: A plain-text description of the theory and its scope
-\item
-  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''.
-From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''.
-We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
-\item
-  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
-\item
-  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
-\item
-  Citing this work: Tell users how to cite the theory.
-Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
-The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
-\item
-  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Again, this is redundant with metadata entered in Zenodo (step 5).
-We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
-For example, we referenced the original empirical cycle paper here:
-\end{itemize}
-
-\begin{verbatim}
-This repository contains an implementation of the "empirical cycle",
-a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
-
-> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
-Foundations of inference and research in the behavioral sciences.
-De Gruyter Mouton. https://doi.org/10.1515/9783112313121
-\end{verbatim}
-
 \subsection{3. Version Control the Repository}\label{version-control-the-repository}
@@
 The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
-In particular, the practice of version control offers extensive benefits for scientific work Van Lissa et al. (2021).
+In particular, the practice of version control offers extensive benefits for scientific work (Ram, 2013; Van Lissa et al., 2021).
 To version control our project, we initiate a Git repository in the project folder.
@@
 must be set to ``Public'' to take advantage of GitHub's Zenodo integration.
-
 Push the local files to the Git remote repository, and keep them synchronized going forward.
@@
 \item
-  Set the \emph{resource type} to \texttt{Model}
+  Set the \emph{resource type} to \texttt{Model}; this ensures proper archival in Zenodo
 \item
-  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}
+  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}; this allows sentient readers to recognize the work as a FAIR theory
 \item
-  Add the \emph{keyword} \texttt{fairtheory}
+  Add the \emph{keyword} \texttt{fairtheory}; this aids search engine indexation
 \item
-  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to increase its findability
+  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to contribute to community building
 \item
@@
 \subsection{6. Making Changes}\label{making-changes}
-
-Aside from technical solutions to However, ``{[}i{]}n systems with many dependencies, releasing new {[}\ldots{]} versions can quickly become a nightmare.''
-The tension lies in balancing regular versioned updates with maintaining interconnected theories: changes risk breaking links between theories, but assuming every change does so leads to an overly sparse network.
-In software development this problem is colloquially called ``dependency hell'' and usually addressed via semantic versioning.
-Traditional version control only tracks changes but semantic versioning tracks what those changes mean to other users of the theory by using a specific format of version numbers.
-We propose the following adaptation of semantic versioning for theories:
-
-\begin{verbatim}
-Given a version number MAJOR.MINOR.PATCH, increment the:
-
-MAJOR version when you make incompatible changes, i.e., change the meaning or derived empierical statements
-MINOR version when you expand the set of empierical statements in a backward compatible manner
-PATCH version when you make backward compatible bug fixes, or clarifications
-\end{verbatim}
-
-Semantic versioning and clearly documented interdependencies allow for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
 
@@
 
+\bibitem[\citeproctext]{ref-langeChecklist2025}
+Lange, J., Freyer, N., Musfeld, P., Schönbrodt, F.
+D., \& Leising, D.
+(2025).
+\emph{A checklist for incentivizing and facilitating good theory building}.
+OSF Preprints.
+\url{https://doi.org/10.31219/osf.io/7qvfz}
+
 \bibitem[\citeproctext]{ref-lavelleWhenCrisisBecomes2021}
@@
 
+\bibitem[\citeproctext]{ref-scientifictheories}
+Winther, R.
+G.
+(2021).
+The structure of scientific theories.
+In E.
+N.
+Zalta (Ed.), \emph{The {Stanford} encyclopedia of philosophy} ({S}pring 2021).
+\url{https://plato.stanford.edu/archives/spr2021/entries/structure-scientific-theories/}; Metaphysics Research Lab, Stanford University.
+
 \end{CSLReferences}

### 2025-01-22 manuscript.tex
Revert "Minor edits by Felix"
This reverts commit f95eaa65161f175c7d4ad9485f354356b897c441.

@@
 {}{\PatchFailed}
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8200}
+\makeatother
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8163}
 \usepackage{lineno}
@@
   pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
   pdflang={en-EN},
@@
 \title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
-\author{Caspar J.
-Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M.
-Brandmaier\textsuperscript{2,3,4}, Felix D.
-Schönbrodt\textsuperscript{5}, \& Noah N.N.
-van Dongen\textsuperscript{6}}
+\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix D. Schönbrodt\textsuperscript{5}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# 09ba7dc6.
+This is a preprint paper, generated from Git Commit \# 070e7292.
 
@@
 Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D.
-Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N.
-van Dongen: Writing -- review \& editing.
+Schönbrodt: Conceptualization, Writing -- review \& editing.
 
@@
 \affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands}
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany}
 
@@
 The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
+In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
@@
 I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
-I2 & (Meta)data use vocabularies that follow FAIR principles & Theory and its associated metadata\ \ use vocabularies that follow FAIR principles & Rephrased\\
+I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
 I2S.1 & - &  & \\
 I2S.2 & - &  & \\
-I3 & (Meta)data include qualified references to other (meta)data & Theory and its associated metadata includes qualified references to other theories and their metadata, including previous versions of the theory & Extended\\
+I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
 I4S & - &  & \\
 R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
-R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory and its associated metadata are released with a clear and accessible license & Rephrased\\
-R1.2 & (Meta)data are associated with detailed provenance & Theory and its associated metadata are associated with detailed provenance & Rephrased\\
-R1.3 & (Meta)data meet domain-relevant community standards & Theory and its associated metadata and documentation meet domain-relevant community standards & Rephrased\\
+R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
+R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
+R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
 \bottomrule
@@
 As a particular FAIR theory evolves, the nature of information tracked will likely change.
-For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are defined and causally connected.
+For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
 From this theory, more specific \emph{statistical/mathematical models} could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
@@
 
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., TODO), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety; Taylor, 2022).
-While the latter psychological factors might be narratively described as vague directional causal statements,
-the theory may also encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
+The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 
@@
 or through analogical modeling.
-In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
 For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
@@
 The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
-First, this requires assigning a globally unique and persistent identifier (such as a DOI) to each theory (F1).
+First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
 Of the many services that provide DOIs for scientific information artefacts,
@@
 and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
+will advance its Accessibility.
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not mandate attribution from a legal perspective,
-the norms of good scientific practice require that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
+For example, while a CC0 license does not mandate attribution,
+it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation (Aalbersberg et al., 2018).
 Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a ``Code of Conduct'' which prescribes how others can contribute to or reuse a theory.
+It is also possible to create or adopt a ``Code of Conduct'' which prescribes
 
@@
 it is defined at either the ``theory'' or ``specification'' level.
-We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \ref{tab:tabfair}):
+We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language:
 
@@
 
-\bibitem[\citeproctext]{ref-langeChecklist2025}
-Lange, J., Freyer, N., Musfeld, P., Schönbrodt, F.
-D., \& Leising, D.
-(2025).
-\emph{A checklist for incentivizing and facilitating good theory building}.
-OSF Preprints.
-\url{https://doi.org/10.31219/osf.io/7qvfz}
-
 \bibitem[\citeproctext]{ref-lavelleWhenCrisisBecomes2021}

### 2025-01-22 manuscript.Rmd
Revert "Revert "Minor edits by Felix""
This reverts commit 70cbfdba75719d8881eb289222024ddfea80cf62.

@@
 The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
+In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020;@norouziCapturingCausalClaims2024].
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
@@
 As a particular FAIR theory evolves, the nature of information tracked will likely change.
-For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
+For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are defined and causally connected.
 From this theory, more specific *statistical/mathematical models* could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
@@
 
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
-The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., TODO), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems [e.g., anxiety\; Taylor, -@taylor2022psychology].
+While the latter psychological factors might be narratively described as vague directional causal statements, 
+the theory may also encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 
@@
 The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
-First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+First, this requires assigning a globally unique and persistent identifier (such as a DOI) to each theory (F1).
 Of the many services that provide DOIs for scientific information artefacts,
@@
 and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility.
+will advance its Accessibility [@langeChecklist2025].
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not mandate attribution,
-it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation [@aalbersbergMakingScienceTransparent2018].
+For example, while a CC0 license does not mandate attribution from a legal perspective,
+the norms of good scientific practice require that scholars comprehensively cite theory and related works [@aalbersbergMakingScienceTransparent2018].
 Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a "Code of Conduct" which prescribes
+It is also possible to create or adopt a "Code of Conduct" which prescribes how others can contribute to or reuse a theory.
 
@@
 it is defined at either the "theory" or "specification" level.
-We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language:
+We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \@ref(tab:tabfair)):
 

### 2025-01-27 manuscript.Rmd
Update manuscript.Rmd
first update

--- 2025-01-17 manuscript.Rmd
@@
 The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to the widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 These various practices ensure transparent and repeated testing of hypotheses.
@@
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-Psychological theories rarely produce hypotheses that are corroborated by evidence.
-Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
+Psychological theories rarely provide hypotheses that are corroborated by evidence.
+Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
 as the theory's central claims evade falsification.
@@
 The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
-The present paper restores balance by applying, for the first time,
+<!-- NvD: In philosophy of science, induction is considered to be problematic (logically invalid).
+It is more appropriate to talk about abduction when it comes to theory development.
+I am fine with sticking to induction if that is what psychologists "know".
+In that case, I suggest adding a clarifying footnote on abduction vs induction.
+-->
+The present paper restores balance by applying, for the first time, <!-- NvD: Are we sure that this is the first time? Do we need to make this claim? -->
 open science principles to psychological theory.
@@
 According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
-a philosophical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \@ref(fig:figecs)).
+a meta-theoretical model of cumulative knowledge acquisition, <!-- NvD: De Groot was not a philosopher and his model is not commonly known/accepted in philosophy of science -->
+research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 
-```{r eval = FALSE}
+```{r}
 library(tidySEM)
@@
 
-```{r figec, eval = FALSE, fig.cap="A take on the empirical cycle by De Groot"}
+```{r figec, fig.cap="A take on the empirical cycle by De Groot"}
 knitr::include_graphics("empirical_cycle.pdf")
@@
 
-MAJOR version when you make backwards incompatible changes, i.e.,
-      the theory now contains empirical statements that are at odds
-      with a previous version
-MINOR version when you expand the set of empirical statements in a
-      backward compatible manner (i.e., the previous version is
-      subsumed within the new version)
+MAJOR version when you make backwards incompatible changes, i.e., the theory now contains empirical statements that are at odds with a previous version
+MINOR version when you expand the set of empirical statements in a backward compatible manner (i.e., the previous version is subsumed within the new version)
 PATCH version when you make backward compatible bug fixes,
@@
 
-We will use the empirical cycle as a running example for this tutorial.
-The resulting FAIR version of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
-The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969,
-<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
-<!-- We can thus consider it a meta-theory of theory construction. -->
+Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
+we decided to use it as an example for this tutorial.
+The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
+The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
+Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
+We can thus consider it a meta-theory of theory construction.
 <!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
-and consists of a series of natural language statements:
+The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:
 
@@
 rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
-rsvg::rsvg_pdf(svg = "ec_groot.svg",
-               file = "ec_groot.pdf")
-rsvg::rsvg_pdf(svg = "ec_wagenmakers.svg",
-               file = "ec_wagenmakers.pdf")
-rsvg::rsvg_pdf(svg = "ec_lissa.svg",
-               file = "ec_lissa.pdf")
-```
-
-```{r figecs, figures-side, fig.show="hold", out.width="33%", fig.cap="The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)"}
-knitr::include_graphics("ec_groot.pdf")
-knitr::include_graphics("ec_wagenmakers.pdf")
-knitr::include_graphics("ec_lissa.pdf")
 ```
@@
 
-## 2. Creating the Project Folder
-
-We can write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`, and save it to a new folder.
+We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
+
+## 2. Creating a Project Folder
+
+Create a new folder and copy the theory file from the previous step into it.
 To help meet the Interoperability and Reusability criteria,
-add two more files.
-Include a `README.md` file with instructions for future users of your theory (see: [What to Include in a README?]())
-and a `LICENSE` file with the legal conditions for reuse.
+add two more files:
+A README.md file with instructions for future users of your theory,
+and a LICENSE file with the legal conditions for reuse.
 We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
+
+### What's in a README?
+
+The readme should contain information to help people get started with using your FAIR theory.
+We suggest the following elements:
+
+* Title, prefaced with `# FAIR theory: The Theory's Name`
+* Description: A plain-text description of the theory and its scope
+* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
+From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
+We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
+* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
+* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
+* Citing this work: Tell users how to cite the theory.
+Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
+The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
+* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
+Again, this is redundant with metadata entered in Zenodo (step 5).
+We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
+For example, we referenced the original empirical cycle paper here:
+
+```
+This repository contains an implementation of the "empirical cycle",
+a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
+
+> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
+Foundations of inference and research in the behavioral sciences.
+De Gruyter Mouton. https://doi.org/10.1515/9783112313121
+```
 
@@
 <!-- Next, in the terminal window, run the following lines of code sequentially: -->
+
 <!-- ``` -->

### 2025-01-29 manuscript.Rmd

@@
 <!-- CJ: Maybe we should give an example here. Does anyone know one? -->
+<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->
 
@@
 One main concern is that social scientific theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning).
+However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains.
+E, M, and C stand form something is E = MC<sup>2</sup>).
+-->
 When theories do not make precise predictions,
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
+<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
 this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
@@
 This needs more work! -->
+<!-- NvD: SOmething like?
+"In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories.
+-->
 .
@@
 According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+In addition, the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
 Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
-The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
+The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
+<!-- NvD: (most of) The other papers also don't have a true derivation chain. It is rare to see a logical derivation, analytical proof, or simulation that connects hypothesis to theory.  -->
 In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
 Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
+<!-- NvD: I am not sure what is being said here.
+In the statistical sense, people are testing the null hypothesis.
+However, that is (almost always) not the hypothesis they are interested in nor the one they mention the introduction of their papers.
+Also, I thought that gigerenzer's paper was about the ritualistic use of NHST, not the origin of the hypotheses that were being tested.
+-->
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
+<!-- NvD: I don't think that statement completely covers it.
+How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.).
+However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena.
+-->
 Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].
@@
 The majority of papers ostensibly test hypotheses,
-but these are rarely derived from theory,
-and test results do not routinely contribute to the improvement of existing theories.
+but these are rarely connected, let alone derived, from theory,
+and test results do not routinely reference back to theories, contributing to their improvement or rejection.
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
@@
 -->
+<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives.
+We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated.
+But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory.
+We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.
+-->
 Definitions of theory abound and are the subject of extensive scholarly debate.
@@
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
--<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations be
-tween concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->
+-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
+Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
+-->
 
@@
 As the practice of FAIRification becomes more embedded,
-we expect that it will become increasingly clear what kind of information is useful.
-As a particular FAIR theory evolves, the nature of information tracked will likely change.
+we expect that it will become increasingly clear what kind and form of information is useful.
+As a particular FAIR theory evolves, the nature of the information tracked will likely change.
 For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
-From this theory, more specific *statistical/mathematical models* could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
-This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+From this theory, more precise *statistical/mathematical models* could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
+This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
 Then, an even more specific *generative/computational model* could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
-Greater formalization increases theories' *empirical content* [REF] because it involves expressing ideas as precise statements.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
+<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.
+-->
+Greater formalization increases theories' *empirical content* [REF]<!-- NvD: We could add Popper 1956 "Logic of Scientific Discovery" here.
+--> because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
 Without committing to specific implementations a-priori,
@@
 
+Although we are in favor of the advancement towards formal theories, 
+this might not always be desirable or feasible. 
+Fortunately, formalization is not required to make theories FAIR.
 To some extent, FAIRness and formalization are orthogonal.
@@
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
-rather, it increases the precision with which they are communicated.
-The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
+rather, it increases the fidelity and ease with which they are communicated.
+The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
 For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
@@
 
-FAIR theory is an example of modular publishing [@kirczModularityNextForm1998].
+We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
 At present, the primary unit of social scientific communication is the academic paper.
@@
 
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts.
+We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
 Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
@@
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
-The present paper considers advantages of Git for FAIR theory.
+The present paper considers the advantages of Git for FAIR theory.
 Git enables explicitly comparing versions of a file (or: theory),
@@
 Aside from technical solutions, version control is a social process as well.
-On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
+On the one hand, regular updates can improve theories - but on the other hand, it risks breaking the compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
 For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.
+<!-- NvD: Maybe we can add an  example (e.g., the theory that explains ego-depletion and the failed replication of the phenomenon). -->
 In computer science, these challenges are navigated by assigning version numbers.

### 2025-01-29 manuscript.Rmd
update 2025-01-29 13:15:59.904362

@@
 The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to the widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 These various practices ensure transparent and repeated testing of hypotheses.
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
-<!-- NvD: In philosophy of science, induction is considered to be problematic (logically invalid).
-It is more appropriate to talk about abduction when it comes to theory development.
-I am fine with sticking to induction if that is what psychologists "know".
-In that case, I suggest adding a clarifying footnote on abduction vs induction.
--->
-The present paper restores balance by applying, for the first time, <!-- NvD: Are we sure that this is the first time? Do we need to make this claim? -->
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
+However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881).
+Defining what makes a theory "best" requires making auxiliary assumptions.
+For present purposes, induction and abduction are interchangeable terms.].
+<!-- Peirce: CP 8.233 Cross-Ref:†† -->
+<!-- 233. This is not essentially different from induction. Only it is most -->
+<!-- usually an induction from instances which are not discrete and numerable. I now -->
+<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
+<!-- line of procedure, or at least not to distinguish it sharply. -->
+The present paper restores balance by applying, for the first time,
 open science principles to psychological theory.
@@
 According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
-In addition, the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
 Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
@@
 Aside from technical solutions, version control is a social process as well.
-On the one hand, regular updates can improve theories - but on the other hand, it risks breaking the compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
+On the one hand, regular updates can improve theories - but on the other hand, it risks breaking compatibility between theories and hypotheses derived from them, or compatibility between one theory and others that depend upon it.
 For example, if we construct a theory to explain a specific phenomenon, and we cross-reference an existing theory comprising an ontology for our field - that dependency is broken if the ontology is later updated and our phenomenon of interest is removed.

### 2025-01-29 manuscript.tex
update 2025-01-29 13:15:59.904362

@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8163}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8688}
 \usepackage{lineno}
@@
   pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
   pdflang={en-EN},
@@
 
-This is a preprint paper, generated from Git Commit \# 070e7292.
+This is a preprint paper, generated from Git Commit \# c910eaa6.
 
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' {[}REF Peirce{]}.
+However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4481).
+Defining what makes a theory ``best'' requires making auxiliary assumptions.
+For present purposes, induction and abduction are interchangeable terms.}.
 The present paper restores balance by applying, for the first time,
@@
 According to the \emph{empirical cycle} (de Groot, 1961),
-a philosophical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \ref{fig:figecs}).
+a meta-theoretical model of cumulative knowledge acquisition,
+research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
+
+\begin{figure}
+\centering
+\includegraphics{empirical_cycle.pdf}
+\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
+\end{figure}
 
@@
 According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
+In addition, the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
 Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
-The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
+The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
 In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
@@
 The majority of papers ostensibly test hypotheses,
-but these are rarely derived from theory,
-and test results do not routinely contribute to the improvement of existing theories.
+but these are rarely connected, let alone derived, from theory,
+and test results do not routinely reference back to theories, contributing to their improvement or rejection.
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
@@
 As the practice of FAIRification becomes more embedded,
-we expect that it will become increasingly clear what kind of information is useful.
-As a particular FAIR theory evolves, the nature of information tracked will likely change.
+we expect that it will become increasingly clear what kind and form of information is useful.
+As a particular FAIR theory evolves, the nature of the information tracked will likely change.
 For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
-From this theory, more specific \emph{statistical/mathematical models} could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
-This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+From this theory, more precise \emph{statistical/mathematical models} could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
+This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
 Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it involves expressing ideas as precise statements.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
+Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
 Without committing to specific implementations a-priori,
@@
 
+Although we are in favor of the advancement towards formal theories,
+this might not always be desirable or feasible.
+Fortunately, formalization is not required to make theories FAIR.
 To some extent, FAIRness and formalization are orthogonal.
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
-rather, it increases the precision with which they are communicated.
-The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
+rather, it increases the fidelity and ease with which they are communicated.
+The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
 For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
@@
 
-FAIR theory is an example of modular publishing (\textbf{kirczModularityNextForm1998?}).
+We propose FAIR theory as an instantiation of modular publishing (\textbf{kirczModularityNextForm1998?}).
 At present, the primary unit of social scientific communication is the academic paper.
@@
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
-The present paper considers advantages of Git for FAIR theory.
+The present paper considers the advantages of Git for FAIR theory.
 Git enables explicitly comparing versions of a file (or: theory),
@@
 
-We will use the empirical cycle as a running example for this tutorial.
-The resulting FAIR version of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
-The empirical cycle is described on page 28 of De Groot and Spiekerman (1969),
-and consists of a series of natural language statements:
+Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
+we decided to use it as an example for this tutorial.
+The resulting FAIR theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
+The empirical cycle is described on page 28 of De Groot and Spiekerman (1969).
+Note that, while De Groot does not explicitly refer to the empirical cycle as a ``theory'', he derives it from ``a theory of thinking''.
+We can thus consider it a meta-theory of theory construction.
+The ``empirical cycle'' theory presented by De Groot consists of a series of natural language statements:
 
@@
 
-\begin{figure}
-\includegraphics[width=0.33\linewidth]{ec_groot} \includegraphics[width=0.33\linewidth]{ec_wagenmakers} \includegraphics[width=0.33\linewidth]{ec_lissa} \caption{The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)}(\#fig:figecs, figures-side)
-\end{figure}
-
 This implementation describes the model as a directed graph.
@@
 
-\subsection{2. Creating the Project Folder}\label{creating-the-project-folder}
-
-We can write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}, and save it to a new folder.
+We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
+
+\subsection{2. Creating a Project Folder}\label{creating-a-project-folder}
+
+Create a new folder and copy the theory file from the previous step into it.
 To help meet the Interoperability and Reusability criteria,
-add two more files.
-Include a \texttt{README.md} file with instructions for future users of your theory (see: \href{}{What to Include in a README?})
-and a \texttt{LICENSE} file with the legal conditions for reuse.
+add two more files:
+A README.md file with instructions for future users of your theory,
+and a LICENSE file with the legal conditions for reuse.
 We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
+
+\subsubsection{What's in a README?}\label{whats-in-a-readme}
+
+The readme should contain information to help people get started with using your FAIR theory.
+We suggest the following elements:
+
+\begin{itemize}
+\tightlist
+\item
+  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
+\item
+  Description: A plain-text description of the theory and its scope
+\item
+  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''.
+From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''.
+We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
+For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
+\item
+  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
+\item
+  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
+\item
+  Citing this work: Tell users how to cite the theory.
+Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
+The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
+The advantage is that documenting related works in the README makes it more readily accessible to users.
+We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
+\item
+  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
+Again, this is redundant with metadata entered in Zenodo (step 5).
+We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
+For example, we referenced the original empirical cycle paper here:
+\end{itemize}
+
+\begin{verbatim}
+This repository contains an implementation of the "empirical cycle",
+a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
+
+> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
+Foundations of inference and research in the behavioral sciences.
+De Gruyter Mouton. https://doi.org/10.1515/9783112313121
+\end{verbatim}
 
@@
 must be set to ``Public'' to take advantage of GitHub's Zenodo integration.
+
 Push the local files to the Git remote repository, and keep them synchronized going forward.

### 2025-01-31 manuscript.Rmd
Update manuscript.Rmd
Noah's suggestions for the rest

@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
-However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881).
-Defining what makes a theory "best" requires making auxiliary assumptions.
-For present purposes, induction and abduction are interchangeable terms.].
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+This is a more general interpretation than enumerative induction where a general conclusion (e.g., all swans are white) is drawn from a limited set of observations (e.g., the *n* observed white swans).
+Another term is *abduction*, which is used to describe the logical process of proposing explanatory propositions that make the observations unsurprising or "follow as a matter of course" [@peirceCollectedPapersCharles1960].
+We decided to use the term induction, because this is what most psychologists are familiar with.
+We also note that Peirce acknowledged that abduction "is not essentially different from induction" (p.4881).
+Thus, for present purposes, induction and abduction are interchangeable terms.].
 <!-- Peirce: CP 8.233 Cross-Ref:†† -->
@@
 
-Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
+Making theories Findable would allow researchers to easily identify relevant theories <!-- NvD: for hypothesis derivation --> <!-- NvD: "to inform their hypotheses" reads a bit like they already have hypotheses and go for a "post-hoc search" to justify them -->,
 grounding their work in established theoretical foundations.
@@
 In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
-For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+For example, predator-prey models have inspired theories of <!-- NvD: intelligence (van der maas et al.
+2006 https://doi.org/10.1037/0033-295x.113.4.842)-->, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
 Findability also enables meta-research on theories, 
@@
 Zenodo and GitHub are both searchable.
-Standardized metadata further enhance Findability in these repositories.
+Standardized metadata further improve Findability in these repositories.
 The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
@@
 and post many examples on their website [@grayHowMapTheory2017].
-Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
+Similarly, <!-- NvD: van Dongen and colleagues are working on a database of models and formalized theories  PREPRINT will be uploaded soon, I hope. -->
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
@@
 Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-If theories are accessible, researchers can reuse and refine them,
-thus accelerating cumulative knowledge acquisition.
-Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+If theories are not accessible, researchers cannot reuse and refine them. <!-- NvD: the original was logically unsound. Accessibility is necessary for reuse, but not sufficient, right?) --> 
+Thus, accessibility can accelerate cumulative knowledge acquisition.
+Making theories accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
 While isolated empirical findings can appear fragmented and contradictory [@dumas-malletPoorReplicationValidity2017],
-theories offer a top-down, big picture representation of the phenomena studied in a field.
+theories offer a top-down, big-picture representation of the phenomena studied in a field.
 In other words, theories are an important instrument in science communication.
@@
 and Zenodo allows for access control of individual files/resources.
-<!-- An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics). -->
+<!-- An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by the oxygen theory). -->
 In general, it makes sense to retain outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.
@@
 accessibility also requires clear and explicit communication.
-This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024].
-At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
+<!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
+This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. <!-- NvD: I suggest keeping this -->
+At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].<!-- NvD: I suggest keeping this -->
 It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
@@
 From a pragmatic perspective, these considerations indicate that,
-when striving to make theory accessible,
-it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions),
-while acknowledging that accessibility exists on a spectrum,
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
+when striving to make theory accessible, <!-- NvD: I suggest keeping this -->
+it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), <!-- NvD: I suggest keeping this -->
+while acknowledging that accessibility exists on a spectrum, <!-- NvD: I suggest keeping this -->
+and that it is impossible to eliminate all ambiguity.<!-- NvD: I suggest keeping this -->
+Rather, it may benefit scientific discourse to anticipate misunderstanding,<!-- NvD: I suggest keeping this -->
+and use it to drive further explication of theory.<!-- NvD: I suggest keeping this -->
 In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
-and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility.
+and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)<!-- NvD: I suggest keeping this -->
+will advance its Accessibility.<!-- NvD: I suggest keeping this -->
 
@@
 With regard to the state of interoperability in contemporary psychology, 
-Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
+Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that theories ought to be highly X-interoperable in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
@@
 If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
 It should further have detailed provenance (R1.2), 
@@
 The Datacite metadata schema offers an initial template in this regard,
-and this paper takes one step towards establishing more fine grained community standards for FAIR theory.
+and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
 
@@
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologieGrondslagenVan1961].
-In FAIR theory workshops, we similarly notice reluctance to reusing and adapting existing theories.
-Students ask questions such as "who owns theory",
-and "who determines how a theory may be reused or changed"?
-These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
+In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
+Students ask questions such as "Who owns a theory?",
+and "Who determines how a theory may be reused or changed?".
+These questions imply a norm against modifying theory without consent from the author.
+<!-- NvD: I suggest removing the rest of the sentence (as a result of an earlier suggestion).
+-- > reminiscent of the aforementioned problem of dependency on the author.
 
@@
 
-The first author was inspired by De Groot too,
-but they conceive of the empirical cycle in yet another way.
+The first author was inspired by De Groot as well,
+but they specify the empirical cycle differently.
 First, notice that the nodes in De Groot's formulation mostly refer to processes.
@@
 explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
-Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
+Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
 and could even be integrated into the analysis environment.
@@
 We demonstrated how to create, version-control, and archive theories as digital information artifacts.
-We introduced the `theorytools` R-package to partly automate these processes, reducing barrier of entry for researchers,
+We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
 and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
@@
 it provides a framework within which scholars can iteratively increase precision and formalization.
-FAIR principles also facilitates new ways of collaboration,
+FAIR principles also facilitate new ways of collaboration,
 leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
@@
 to expedite its effective reuse and iterative improvement.
-A second factor is to lower barriers for the adoption of FAIR theory by building upon existing widely adopted open science infrastructures.
+A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
 For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
-Barriers of entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
+Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
 Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
-These materials allow teachers to incorporate theory development into their curriculum without investing substantial time into course development,
+These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
 thus educating the next generation to make use of and contribute to FAIR theory.
@@
 Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [REF Oisin paper].
-Meta-research benefit from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
+Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
 In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
@@
 Another limitation is the learning curve associated with tools like Git and Zenodo.
-The `theorytools` R-package mitigates this limitation by automating key steps in the process.
-Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
-One barrier to adoption of FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
+The `theorytools` R-package mitigates this limitation to some extent by automating key steps in the process. <!-- NvD: However, some skill with R is still required. -->
+Moreover, the initial investment in time can be offset by long-term productivity gains and an increased impact of FAIR theory.
+One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
 Education might help address this limitation; with this in mind,
-we are developing open educational materials on theory development.
-
-One limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [REF] and lack of theory formalization [REF].
-However, our work does establish a framework within which theories can be further formalized.
+and we are developing open educational materials on theory development.
+
+A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [REF] and lack of theory formalization [REF].
+However, our work does establish a framework that allows for and promotes the formalization of theories. 
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [REF] and lack of standardized ontologies for psychological constructs [REF].
-However, our work here provides a template for addressing such challenges,
+Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [REF] and lack of standardized ontologies for psychological constructs [REF].<!-- NvD: Adding this might go a bit too far, we are also not solving world hunger and poverty ;-) -->
+However, our work here provides a template for addressing such challenges, <!-- NvD: I am not sure if we are offering a template to address these challenges.
+If we do, we should make this explicit.
+-->
 and any advancements in the areas of measurement and ontology will serve to amplify the value of FAIR theories, particularly when such resources are cross-referenced in the metadata (e.g., on Zenodo).
@@
 One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
+Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) --> <!-- NvD: I agree.
+Also, if we mention it here, we should definitely not mention it under limitations ->
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
@@
 making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
+FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories.
 More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers for knowledge exchange with the global scholarly community,
+removes barriers to knowledge exchange with the global scholarly community,
 opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
@@
 
-FAIR theory is a major step forwards towards more transparent, collaborative, and efficient theory construction.
+FAIR theory is a major step forward. A step towards more transparent, collaborative, and efficient theory construction.
 It provides much-needed open science methods for the inductive phase of the empirical cycle,

### 2025-02-03 manuscript.Rmd
update 2025-02-03 09:44:07.710923

--- 2025-01-29 manuscript.Rmd
@@
 In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
-Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
+Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
 <!-- In our representation, -->
@@
 <!-- ``` -->
-For example, many have argued that observation is value-laden, and as such, involves induction.
-The derivation of hypotheses from theory is also not purely deductive,
-as auxiliary assumptions are often made (which are, again, an inductive process).
-Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [REF Peikert].
+For example, many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
+Deriving hypotheses from theory is also not purely deductive,
+as auxiliary assumptions are often made (which, again, involves induction).
+Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
 With these alterations, we implement the empirical cycle as follows:
@@
 
-Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
+Some have argued that *causal explanations* are a property of good theory [@meehlAppraisingAmendingTheories1990].
 According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
+explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995].
 Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,

### 2025-02-03 manuscript.Rmd

--- 2025-01-31 manuscript.Rmd
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
-This is a more general interpretation than enumerative induction where a general conclusion (e.g., all swans are white) is drawn from a limited set of observations (e.g., the *n* observed white swans).
-Another term is *abduction*, which is used to describe the logical process of proposing explanatory propositions that make the observations unsurprising or "follow as a matter of course" [@peirceCollectedPapersCharles1960].
-We decided to use the term induction, because this is what most psychologists are familiar with.
-We also note that Peirce acknowledged that abduction "is not essentially different from induction" (p.4881).
-Thus, for present purposes, induction and abduction are interchangeable terms.].
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
+However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881).
+Defining what makes a theory "best" requires making auxiliary assumptions.
+For present purposes, induction and abduction are interchangeable terms.].
 <!-- Peirce: CP 8.233 Cross-Ref:†† -->
@@
 
-Making theories Findable would allow researchers to easily identify relevant theories <!-- NvD: for hypothesis derivation --> <!-- NvD: "to inform their hypotheses" reads a bit like they already have hypotheses and go for a "post-hoc search" to justify them -->,
-grounding their work in established theoretical foundations.
+Making theories Findable would allow researchers to easily identify relevant theories
+and ground their hypotheses in established theoretical foundations.
 It further increases the impact and reuse potential of theories across disciplines,
@@
 In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
-For example, predator-prey models have inspired theories of <!-- NvD: intelligence (van der maas et al.
-2006 https://doi.org/10.1037/0033-295x.113.4.842)-->, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+For example, predator-prey models have inspired theories of intelligence [@vandermaasDynamicalModelGeneral2006], and the Eysenck model of atomic magnetism has inspired a network theory of depression [@cramerMajorDepressionComplex2016].
 Findability also enables meta-research on theories, 
@@
 and post many examples on their website [@grayHowMapTheory2017].
-Similarly, <!-- NvD: van Dongen and colleagues are working on a database of models and formalized theories  PREPRINT will be uploaded soon, I hope. -->
+Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
@@
 
-Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
+Some have argued that *causal explanations* are a property of good theory [@meehlAppraisingAmendingTheories1990].
 According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
+explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995].
 Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
@@
 it provides a framework within which scholars can iteratively increase precision and formalization.
-FAIR principles also facilitate new ways of collaboration,
+The FAIR principles also facilitate new ways of collaboration,
 leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
@@
 Another limitation is the learning curve associated with tools like Git and Zenodo.
-The `theorytools` R-package mitigates this limitation to some extent by automating key steps in the process. <!-- NvD: However, some skill with R is still required. -->
-Moreover, the initial investment in time can be offset by long-term productivity gains and an increased impact of FAIR theory.
+The `theorytools` R-package mitigates this limitation by automating key steps in the process.
+Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
 One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
 Education might help address this limitation; with this in mind,
-and we are developing open educational materials on theory development.
-
-A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [REF] and lack of theory formalization [REF].
+we are developing open educational materials on theory development.
+
+A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [@frankenhuisStrategicAmbiguitySocial2023] and lack of theory formalization [@robinaughInvisibleHandsFine2021].
 However, our work does establish a framework that allows for and promotes the formalization of theories. 
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [REF] and lack of standardized ontologies for psychological constructs [REF].<!-- NvD: Adding this might go a bit too far, we are also not solving world hunger and poverty ;-) -->
+Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [@bringmannBackBasicsImportance2022] and lack of standardized ontologies for psychological constructs [@boscoMetaBUSVehicleFacilitating2017].
 However, our work here provides a template for addressing such challenges, <!-- NvD: I am not sure if we are offering a template to address these challenges.
@@
 
-FAIR theory is a major step forward. A step towards more transparent, collaborative, and efficient theory construction.
+FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
 It provides much-needed open science methods for the inductive phase of the empirical cycle,

### 2025-02-03 manuscript.Rmd
do something with felix's comments

@@
 
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
+such as adherence to infection prevention protocols (e.g., social distancing),
+pandemic-related behavior (e.g., panic buying),
+and pandemic-related distress
+[@taylor2022psychology].
 The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
@@
 or through analogical modeling.
-In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
 For example, predator-prey models have inspired theories of intelligence [@vandermaasDynamicalModelGeneral2006], and the Eysenck model of atomic magnetism has inspired a network theory of depression [@cramerMajorDepressionComplex2016].
@@
 The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
-First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
 Of the many services that provide DOIs for scientific information artefacts,
@@
 <!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
-This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. <!-- NvD: I suggest keeping this -->
-At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].<!-- NvD: I suggest keeping this -->
+This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
+At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
 It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
@@
 From a pragmatic perspective, these considerations indicate that,
-when striving to make theory accessible, <!-- NvD: I suggest keeping this -->
-it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), <!-- NvD: I suggest keeping this -->
-while acknowledging that accessibility exists on a spectrum, <!-- NvD: I suggest keeping this -->
-and that it is impossible to eliminate all ambiguity.<!-- NvD: I suggest keeping this -->
-Rather, it may benefit scientific discourse to anticipate misunderstanding,<!-- NvD: I suggest keeping this -->
-and use it to drive further explication of theory.<!-- NvD: I suggest keeping this -->
+when striving to make theory accessible, 
+it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), 
+while acknowledging that accessibility exists on a spectrum, 
+and that it is impossible to eliminate all ambiguity.
+Rather, it may benefit scientific discourse to anticipate misunderstanding,
+and use it to drive further explication of theory.
 In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
-and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)<!-- NvD: I suggest keeping this -->
-will advance its Accessibility.<!-- NvD: I suggest keeping this -->
+and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
+will advance its Accessibility [@langeChecklist2025].
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not mandate attribution,
-it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation [@aalbersbergMakingScienceTransparent2018].
+For example, while a CC0 license does not legally mandate attribution,
+the norms of good scientific practice require that scholars comprehensively cite theory and related works [@aalbersbergMakingScienceTransparent2018].
 Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a "Code of Conduct" which prescribes
+It is also possible to create or adopt a "Code of Conduct" which prescribes behavioral norms for contributors and users of a theory.
 
@@
 
-Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
-we decided to use it as an example for this tutorial.
-The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
-The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
-Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
-We can thus consider it a meta-theory of theory construction.
+We will use the *empirical cycle* as a running example for this tutorial.
+The empirical cycle, described on page 28 of @degrootMethodologyFoundationsInference1969,
+is a kind of meta-theory of theory construction.
+The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
+<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
+<!-- We can thus consider it a meta-theory of theory construction. -->
 <!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
-The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:
+The original theory consists of a series of natural language statements:
 
@@
 This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
-
 We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
@@
 and a LICENSE file with the legal conditions for reuse.
+For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html)
 We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
 
-### What's in a README?
-
-The readme should contain information to help people get started with using your FAIR theory.
-We suggest the following elements:
-
-* Title, prefaced with `# FAIR theory: The Theory's Name`
-* Description: A plain-text description of the theory and its scope
-* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
+<!-- ### What's in a README? -->
+
+<!-- The readme should contain information to help people get started with using your FAIR theory. -->
+<!-- We suggest the following elements: -->
+
+<!-- * Title, prefaced with `# FAIR theory: The Theory's Name` -->
+<!-- * Description: A plain-text description of the theory and its scope -->
+<!-- * Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
 From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
@@
 For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
-* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
-* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
-* Citing this work: Tell users how to cite the theory.
+-->
+<!-- * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*. -->
+<!-- * License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*. -->
+<!-- * Citing this work: Tell users how to cite the theory.
 Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
@@
 We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
-* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
+-->
+<!-- * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
 Again, this is redundant with metadata entered in Zenodo (step 5).
 We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
-For example, we referenced the original empirical cycle paper here:
-
-```
-This repository contains an implementation of the "empirical cycle",
-a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
-
-> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
-Foundations of inference and research in the behavioral sciences.
-De Gruyter Mouton. https://doi.org/10.1515/9783112313121
-```
+For example, we referenced the original empirical cycle paper here: -->
+
+<!-- ``` -->
+<!-- This repository contains an implementation of the "empirical cycle", -->
+<!-- a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works. -->
+
+<!-- > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: -->
+<!-- Foundations of inference and research in the behavioral sciences. -->
+<!-- De Gruyter Mouton. https://doi.org/10.1515/9783112313121 -->
+<!-- ``` -->
 

### 2025-02-04 manuscript.Rmd
add abstract

@@
 
-abstract: |
-  Test test.
+abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
 keywords          : "fairtheory, meta theory, theory formation, cumulative science, formal models"

### 2025-02-14 manuscript.Rmd
add title
should be theory specification
add example of ambiguity

@@
 ---
-title             : "FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
+title             : "To be FAIR: Theory Specification Needs an Update"
 shorttitle        : "FAIR THEORY"
@@
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
-Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-Psychological theories rarely provide hypotheses that are corroborated by evidence.
-Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
-as the theory's central claims evade falsification.
+Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental: Psychological theories rarely provide hypotheses that are corroborated by evidence.
+Furthermore, theories are often so steeped in ambiguity that they can accommodate mutually inconsistent findings, as the theory’s central claims evade falsification.
+A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Since its introduction, the very definition of intrinsic and extrinsic motivation has shifted.
+Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
+Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
+Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
+Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
+
 <!-- CJ: Maybe we should give an example here. Does anyone know one? -->

### 2025-02-17 manuscript.Rmd
reduce changes, shorten
Update manuscript.Rmd
revert this change; it doesn't make a meaningful difference but it's less concise

@@
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
-Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental: Psychological theories rarely provide hypotheses that are corroborated by evidence.
-Furthermore, theories are often so steeped in ambiguity that they can accommodate mutually inconsistent findings, as the theory’s central claims evade falsification.
+Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+Psychological theories rarely provide hypotheses that are corroborated by evidence.
+Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
+as the theory's central claims evade falsification.
 A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
-Since its introduction, the very definition of intrinsic and extrinsic motivation has shifted.
 Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].

### 2025-02-19 manuscript.Rmd
Rephrase reference to paradigms

--- 2025-02-04 manuscript.Rmd
@@
 Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
-The Kuhnian notion of "normal science", conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding [@kuhnStructureScientificRevolutions2009].
-From a pragmatic perspective, these considerations indicate that,
-when striving to make theory accessible, 
-it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), 
-while acknowledging that accessibility exists on a spectrum, 
+The notion of "normal science" describes a phase in which a scientific community operates within the context of a shared paradigm,
+creating the background knowledge needed for mutual understanding and productive scientific work [@kuhnStructureScientificRevolutions2009].
+In reality, however, paradigms are rarely clear-cut, and shared background knowledge can come from different sources, or be problematically absent.
+From a pragmatic perspective,
+this underscores the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
+while acknowledging that Accessibility exists on a spectrum, 
 and that it is impossible to eliminate all ambiguity.

### 2025-02-19 manuscript.Rmd
rephrase paradigm section

@@
 creating the background knowledge needed for mutual understanding and productive scientific work [@kuhnStructureScientificRevolutions2009].
-In reality, however, paradigms are rarely clear-cut, and shared background knowledge can come from different sources, or be problematically absent.
+In reality, however, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
+including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
 From a pragmatic perspective,

### 2025-02-19 manuscript.tex
rephrase paradigm section

@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8688}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8186}
 \usepackage{lineno}
@@
 \hypersetup{
-  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
+  pdftitle={To be FAIR: Theory Specification Needs an Update},
   pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
@@
 
-\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
+\title{To be FAIR: Theory Specification Needs an Update}
 \author{Caspar J.
@@
 
-This is a preprint paper, generated from Git Commit \# c910eaa6.
+This is a preprint paper, generated from Git Commit \# c8b13da7.
 
@@
 \abstract{%
-Test test.
+Many meta‐scientific initiatives have focused on enhancing the rigor of theory testing, yet the systematic specification of theories remains underdeveloped.
+We present FAIR theory, a new approach that transforms scientific theories into digital, evolving artifacts---developed collaboratively---to bolster the robustness and interconnectedness of scientific knowledge.
+By representing theories as formal, machine‐readable objects (e.g., mathematical models, computer code, nomological networks, or explicit operational definitions) and tracking their evolution through version control and semantic versioning, FAIR theory provides a structured, cumulative framework for theory development.
+We propose that each theory is assigned a universally unique identifier and enriched with detailed metadata, ensuring it is Findable, Accessible, Interoperable, and Reusable.
+This digital infrastructure addresses a critical gap in scholarly practice by introducing the collaborative tools and standards for theory specification that have already revolutionized methodology, data sharing, and publishing.
+Ultimately, FAIR theory aims to transform how scientific ideas are communicated, evaluated, and iteratively refined, thereby accelerating cumulative knowledge acquisition across disciplines.
 }
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' {[}REF Peirce{]}.
-However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4481).
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' (\textbf{peirceCollectedPapersCharles1960?}).
+However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4881).
 Defining what makes a theory ``best'' requires making auxiliary assumptions.
@@
 According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
-In addition, the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
+Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
 Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
@@
 
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) (Taylor, 2022).
+As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
+such as adherence to infection prevention protocols (e.g., social distancing),
+pandemic-related behavior (e.g., panic buying),
+and pandemic-related distress
+(Taylor, 2022).
 The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
@@
 
-Making theories Findable would allow researchers to easily identify relevant theories to inform their hypotheses,
-grounding their work in established theoretical foundations.
+Making theories Findable would allow researchers to easily identify relevant theories
+and ground their hypotheses in established theoretical foundations.
 It further increases the impact and reuse potential of theories across disciplines,
@@
 or through analogical modeling.
-In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
-For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.
+In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
+For example, predator-prey models have inspired theories of intelligence (\textbf{vandermaasDynamicalModelGeneral2006?}), and the Eysenck model of atomic magnetism has inspired a network theory of depression (\textbf{cramerMajorDepressionComplex2016?}).
 Findability also enables meta-research on theories,
@@
 The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
-First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
+First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
 Of the many services that provide DOIs for scientific information artefacts,
@@
 Zenodo and GitHub are both searchable.
-Standardized metadata further enhance Findability in these repositories.
+Standardized metadata further improve Findability in these repositories.
 The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}).
@@
 and post many examples on their website (Gray, 2017).
-Similarly, Borsboom and colleagues seek to establish a database of psychological theories {[}REF BORSBOOM{]}.
+Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
@@
 Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-If theories are accessible, researchers can reuse and refine them,
-thus accelerating cumulative knowledge acquisition.
-Making theories accessible also allows stakeholders (e.g., practitioners, policy makers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+If theories are not accessible, researchers cannot reuse and refine them.
+Thus, accessibility can accelerate cumulative knowledge acquisition.
+Making theories accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
 While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
-theories offer a top-down, big picture representation of the phenomena studied in a field.
+theories offer a top-down, big-picture representation of the phenomena studied in a field.
 In other words, theories are an important instrument in science communication.
@@
 and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility.
+will advance its Accessibility (\textbf{langeChecklist2025?}).
 
@@
 With regard to the state of interoperability in contemporary psychology,
-Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that ought to be highly X-interoperable in psychological researchers' day-to-day work.
+Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that theories ought to be highly X-interoperable in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
@@
 If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+Applied to FAIR theory, reusability requires that each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
 It should further have detailed provenance (R1.2),
@@
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (de Groot, 1961).
-In FAIR theory workshops, we similarly notice reluctance to reusing and adapting existing theories.
-Students ask questions such as ``who owns theory'',
-and ``who determines how a theory may be reused or changed''?
-These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
+In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
+Students ask questions such as ``Who owns a theory?'',
+and ``Who determines how a theory may be reused or changed?''.
+These questions imply a norm against modifying theory without consent from the author.
+\textless!-- NvD: I suggest removing the rest of the sentence (as a result of an earlier suggestion).
+-- \textgreater{} reminiscent of the aforementioned problem of dependency on the author.
 
@@
 Aside from legal conditions for reuse, there are also social considerations.
-For example, while a CC0 license does not mandate attribution,
-it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation (Aalbersberg et al., 2018).
+For example, while a CC0 license does not legally mandate attribution,
+the norms of good scientific practice require that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
 Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a ``Code of Conduct'' which prescribes
+It is also possible to create or adopt a ``Code of Conduct'' which prescribes behavioral norms for contributors and users of a theory.
 
@@
 
-Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
-we decided to use it as an example for this tutorial.
-The resulting FAIR theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
-The empirical cycle is described on page 28 of De Groot and Spiekerman (1969).
-Note that, while De Groot does not explicitly refer to the empirical cycle as a ``theory'', he derives it from ``a theory of thinking''.
-We can thus consider it a meta-theory of theory construction.
-The ``empirical cycle'' theory presented by De Groot consists of a series of natural language statements:
+We will use the \emph{empirical cycle} as a running example for this tutorial.
+The empirical cycle, described on page 28 of De Groot and Spiekerman (1969),
+is a kind of meta-theory of theory construction.
+The resulting FAIR implementation of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
+The original theory consists of a series of natural language statements:
 
@@
 it is defined at either the ``theory'' or ``specification'' level.
-We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language:
+We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \ref{tab:tabfair}):
 
@@
 This follows the first two properties of good theory according to Meehl (Paul E. Meehl, 1990).
-
 We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
@@
 and a LICENSE file with the legal conditions for reuse.
+For guidance on writing the README file, see \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{this vignette}
 We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
-
-\subsubsection{What's in a README?}\label{whats-in-a-readme}
-
-The readme should contain information to help people get started with using your FAIR theory.
-We suggest the following elements:
-
-\begin{itemize}
-\tightlist
-\item
-  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
-\item
-  Description: A plain-text description of the theory and its scope
-\item
-  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''.
-From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''.
-We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
-\item
-  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
-\item
-  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
-\item
-  Citing this work: Tell users how to cite the theory.
-Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
-The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
-\item
-  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Again, this is redundant with metadata entered in Zenodo (step 5).
-We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
-For example, we referenced the original empirical cycle paper here:
-\end{itemize}
-
-\begin{verbatim}
-This repository contains an implementation of the "empirical cycle",
-a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.
-
-> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
-Foundations of inference and research in the behavioral sciences.
-De Gruyter Mouton. https://doi.org/10.1515/9783112313121
-\end{verbatim}
 
@@
 In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
-Second, De Groot's strict distinction between processes of observation, induction, and deduction is not widely supported by philosophy of science.
-For example, many have argued that observation is value-laden, and as such, involves induction.
-The derivation of hypotheses from theory is also not purely deductive,
-as auxiliary assumptions are often made (which are, again, an inductive process).
-Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well {[}REF Peikert{]}.
+Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
+For example, many have argued that observation is theory-laden, and as such, involves induction (\textbf{brewerTheoryLadennessObservationTheoryLadenness2001?}).
+Deriving hypotheses from theory is also not purely deductive,
+as auxiliary assumptions are often made (which, again, involves induction).
+Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (\textbf{peikertTransparencyOpenScience2023?}).
 With these alterations, we implement the empirical cycle as follows:
@@
 
-Some have argued that \emph{causal explanations} are a property of good theory {[}REF Meehl, etc?{]}.
+Some have argued that \emph{causal explanations} are a property of good theory (Paul E. Meehl, 1990).
 According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
-Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
+explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data (\textbf{pearlCausalDiagramsEmpirical1995?}).
+Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
 and could even be integrated into the analysis environment.
@@
 We demonstrated how to create, version-control, and archive theories as digital information artifacts.
-We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing barrier of entry for researchers,
+We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing the barrier of entry for researchers,
 and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
@@
 it provides a framework within which scholars can iteratively increase precision and formalization.
-FAIR principles also facilitates new ways of collaboration,
+The FAIR principles also facilitate new ways of collaboration,
 leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
@@
 to expedite its effective reuse and iterative improvement.
-A second factor is to lower barriers for the adoption of FAIR theory by building upon existing widely adopted open science infrastructures.
+A second factor is to lower barriers to adopting FAIR theory by building upon existing widely adopted open science infrastructures.
 For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
-Barriers of entry can also be lowered by simplifying workflows, which is the goal of the \texttt{theorytools} R-package.
+Barriers to entry can also be lowered by simplifying workflows, which is the goal of the \texttt{theorytools} R-package.
 Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
-These materials allow teachers to incorporate theory development into their curriculum without investing substantial time into course development,
+These materials allow teachers to incorporate theory development into their curriculum without investing substantial time on course development,
 thus educating the next generation to make use of and contribute to FAIR theory.
@@
 Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
-Meta-research benefit from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
+Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
 In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
@@
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
-One barrier to adoption of FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
+One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
 Education might help address this limitation; with this in mind,
@@
 
-One limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity {[}REF{]} and lack of theory formalization {[}REF{]}.
-However, our work does establish a framework within which theories can be further formalized.
+A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity (Frankenhuis et al., 2023) and lack of theory formalization (Robinaugh et al., 2021).
+However, our work does establish a framework that allows for and promotes the formalization of theories.
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis {[}REF{]} and lack of standardized ontologies for psychological constructs {[}REF{]}.
+Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis (\textbf{bringmannBackBasicsImportance2022?}) and lack of standardized ontologies for psychological constructs (Bosco et al., 2017).
 However, our work here provides a template for addressing such challenges,
@@
 One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
+Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'': \textless!-- NvD: I agree.
+Also, if we mention it here, we should definitely not mention it under limitations -\textgreater{}
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties {[}REF{]}.
@@
 
-FAIR theory is a major step forwards towards more transparent, collaborative, and efficient theory construction.
+FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
 It provides much-needed open science methods for the inductive phase of the empirical cycle,

### 2025-02-19 manuscript.Rmd
move conclusion to end of accessibility section

@@
 including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
-From a pragmatic perspective,
-this underscores the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
-while acknowledging that Accessibility exists on a spectrum, 
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
-and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility [@langeChecklist2025].
 
@@
 DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.
+
+From a pragmatic perspective,
+these considerations underscore the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
+while acknowledging that Accessibility exists on a spectrum, 
+and that it is impossible to eliminate all ambiguity.
+Rather, it may benefit scientific discourse to anticipate misunderstanding,
+and use it to drive further explication of theory.
+In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
+and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
+will advance its Accessibility [@langeChecklist2025].
 

### 2025-02-19 manuscript.Rmd

@@
 as the theory's central claims evade falsification.
+A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
+Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
+Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
+Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
+
 <!-- CJ: Maybe we should give an example here. Does anyone know one? -->

### 2025-02-19 manuscript.Rmd
Add examples for relationship types (see #32)

@@
 and cross-reference the published paper with `relationType: IsSupplementTo`.
-Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.
+Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
+Other types are useful for making relationships between multiple theory objects explicit: If an existing narrative theory is FAIRified without much own contribution, the FAIR theory would cross-reference the paper with the narrative theory as `relationType: IsDerivedFrom`; if substantial own contributions were necessary, `relationType:IsNewVersionOf` could be used.
+If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
+<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->
 

### 2025-02-19 manuscript.Rmd
update 2025-02-19 10:53:20.811478

--- 2025-02-19 manuscript.Rmd
@@
 It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
-which holds that every communicative utterance has multiple alternative translations, with no *objective* means of choosing the correct one.
+which holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+has multiple alternative translations, with no *objective* means of choosing the correct one.
 It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.

### 2025-02-20 manuscript.Rmd
closes #53

@@
 -->
-Greater formalization increases theories' *empirical content* [REF]<!-- NvD: We could add Popper 1956 "Logic of Scientific Discovery" here.
---> because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
+Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
@@
 the theory becomes hard to test.
+Compared to theories expressed in natural language,
+formal theories facilitate inconsistency.
 Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
@@
 Although we are in favor of the advancement towards formal theories, 
-this might not always be desirable or feasible. 
-Fortunately, formalization is not required to make theories FAIR.
-To some extent, FAIRness and formalization are orthogonal.
-<!-- : FAIR theory does not require theories to be formal, -->
-<!-- and formal theory can be represented in a way that is not FAIR. -->
-<!-- We stress that  -->
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
 rather, it increases the fidelity and ease with which they are communicated.
+Thus, FAIR theory does not require theories to be formal,
+and formal theory can be represented in a way that is not FAIR.
 The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
@@
 and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
-and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility [@langeChecklist2025].
+In sum, while it is impossible to communicate a theory completely unambiguously,
+scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+A theory's Accessibility is increased by
+reducing dependencies on (implicit) background knowledge,
+explication of assumptions,
+formalization,
+and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [@langeChecklist2025].
 

### 2025-02-20 manuscript.Rmd
closes #51

@@
 Finally, metadata should be registered or indexed in a searchable registry (F4).
-Zenodo and GitHub are both searchable.
-Standardized metadata further improve Findability in these repositories.
+It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
+only few are specifically designed for FAIR-compliant archival.
+Zenodo stands out in this respect, and it is in turn [indexed by Google Scholar](https://support.zenodo.org/help/en-gb/29-indexing/61-is-zenodo-indexed-by-google-scholar), Google Dataset Search, and other services.
+Thus, while it is fine (and recommended) to use GitHub for version control,
+the version of record should be archived in a FAIR repository like Zenodo.
+Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
 The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
 <!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
-Furthermore, a standard keyword can be used; we suggest using the keyword `"fairtheory"` for all resources that constitute or reference a FAIR theory.
+Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
 <!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->
@@
 
-One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) --> <!-- NvD: I agree.
-Also, if we mention it here, we should definitely not mention it under limitations ->
+One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
+Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
@@
 FAIR theory can help address the measurement crisis:
-since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
-
-FAIR theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers to knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
+since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.
+
+<!-- FAIR theory incorporates theory into open science workflows, -->
+<!-- facilitates scholarly communication about theories, -->
+<!-- making it easier to share theories with less opportunity for ambiguity and misunderstanding. -->
+<!-- FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories. -->
+<!-- More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition, -->
+<!-- removes barriers to knowledge exchange with the global scholarly community, -->
+<!-- opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration. -->
 

### 2025-02-20 manuscript.Rmd
update 2025-02-20 12:48:57.507521
closes #37

@@
 wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
-bibliography      : "theory-specification.bib"
+bibliography      : "fair_theory.bib"
 floatsintext      : yes
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
-However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881).
-Defining what makes a theory "best" requires making auxiliary assumptions.
-For present purposes, induction and abduction are interchangeable terms.].
+The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
+However, Peirce later acknowledged that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
+As defining what makes a theory "best" requires auxiliary assumptions, we consider abduction to be a special case of induction.
+For present purposes, however, the terms are interchangeable.].
 <!-- Peirce: CP 8.233 Cross-Ref:†† -->
@@
 only few are specifically designed for FAIR-compliant archival.
-Zenodo stands out in this respect, and it is in turn [indexed by Google Scholar](https://support.zenodo.org/help/en-gb/29-indexing/61-is-zenodo-indexed-by-google-scholar), Google Dataset Search, and other services.
+Zenodo stands out in this respect.
 Thus, while it is fine (and recommended) to use GitHub for version control,
@@
 formalization,
-and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [@langeChecklist2025].
+and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [REF langeChecklist2025].
 

### 2025-02-21 manuscript.Rmd
update 2025-02-21 07:54:50.647059

--- 2025-02-20 manuscript.Rmd
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and reusable [REF] for both humans and computers.
 Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
@@
 We apply the FAIR principles to scientific theories,
-introducing the concept of *FAIR theory* to 
+introducing *FAIR theory* to 
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 
-The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
+The present paper introduces open science methods for theory specification.
 Merely publishing theory in a research article does not make it open;
 to be open, theory should adhere to established open science standards.
-We apply the FAIR principles to digital representations of theory,
-introducing a FAIR metadata format to make theories *Findable* via a DOI,
+We do so by implementing theories as information artifacts compliant with the FAIR principles.
+The resulting FAIR theories are archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo). 
+FAIR theories are *Findable* via a DOI or search,
 *Accessible* in a machine- and human-readable filetype,
 *Interoperable* within the data analysis environment,
-and *Reusable* in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
+and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
 Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
@@
 
-There have been notable efforts to improve theories' findability through post-hoc curation.
+There have been notable efforts to improve theories' Findability through post-hoc curation.
 For example, Gray and colleagues introduced a format for representing theories,
@@
 
-Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-If theories are not accessible, researchers cannot reuse and refine them. <!-- NvD: the original was logically unsound. Accessibility is necessary for reuse, but not sufficient, right?) --> 
-Thus, accessibility can accelerate cumulative knowledge acquisition.
-Making theories accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+Transparent scholarly communication about theory requires that theories are Accessible to all researchers and other stakeholders.
+If theories are not Accessible, researchers cannot reuse and refine them. <!-- NvD: the original was logically unsound. Accessibility is necessary for reuse, but not sufficient, right?) --> 
+Thus, Accessibility can accelerate cumulative knowledge acquisition.
+Making theories Accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
 While isolated empirical findings can appear fragmented and contradictory [@dumas-malletPoorReplicationValidity2017],
@@
 They apply to theory with minor changes.
-Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
+Firstly, theory and its associated metadata should be Accessible by their identifier using a standardized communications protocol (A1).
 This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
-Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
+Secondly (A2), theory metadata should be Accessible, even when the theory is no longer available,
 which is also achieved via long-term storage (e.g., on Zenodo).
@@
 
-At present, there are several impediments to theories' accessibility.
+At present, there are several impediments to theories' Accessibility.
 To the extent that theories are still contained within papers,
 paywalls erected by commercial publishers constitute a barrier.
-Open Access publishing thus increases the accessibility of all academic output, including theory.
+Open Access publishing thus increases the Accessibility of all academic output, including theory.
 A second impediment is more indirect:
 While open access publishing increases practical access to theories,
-accessibility also requires clear and explicit communication.
+Accessibility also requires clear and explicit communication.
 <!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
@@
 This demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible in terms of both availability and interpretability.
 
@@
 
-<!-- The data usually need to be integrated with other data. In addition, the data need to interoperate with applications or workflows for analysis, storage, and processing. -->
+<!-- The data usually need to be integrated with other data. In addition, the data need to Interoperate with applications or workflows for analysis, storage, and processing. -->
 
@@
 Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
-Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
+Firstly, theory and its associated metadata should use a formal, Accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
 The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
@@
 and Gray's theory maps [@grayHowMapTheory2017].
-While a theory represented as a bitmap image is not very interoperable,
+While a theory represented as a bitmap image is not very Interoperable,
 the same image represented in the DOT language [@DOTLanguage]<!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
@@
 
-Recent work points out that interoperability is not an all-or-nothing property.
-The concept of X-interoperability was introduced to answer the question: *interoperable for what?*
-X-interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
-This revised definition makes it possible to outline a theory's affordances in terms of X-interoperability.
-For example, a FAIR theory may be X-interoperable for deriving testable hypotheses,
+Recent work points out that Interoperability is not an all-or-nothing property.
+The concept of X-Interoperability was introduced to answer the question: *Interoperable for what?*
+X-Interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-Interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
+This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
+For example, a FAIR theory may be X-Interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
@@
 If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
-we see how each of these properties incurs certain affordances in terms of X-interoperability (Table \@ref(tab:tabmeehl)).
+we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \@ref(tab:tabmeehl)).
 
@@
   "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
-  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
+  "X-Interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
 ) |>
@@
 
-With regard to the state of interoperability in contemporary psychology, 
-Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that theories ought to be highly X-interoperable in psychological researchers' day-to-day work.
+With regard to the state of Interoperability in contemporary psychology, 
+Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that theories ought to be highly X-Interoperable in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
-The examples of X-interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
-For example, interoperable theory could be used
+The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
+For example, Interoperable theory could be used
 to select control variables for causal inference [@cinelliCrashCourseGood2022],
@@
 or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
-Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
+Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
 For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
@@
 If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+Applied to FAIR theory, reusability requires that  each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
 It should further have detailed provenance (R1.2), 
@@
 From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
-We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how.
+We propose explicitly addressing the theory's X-Interoperability, telling users exactly what they can use the theory for, and how.
 For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
@@
 
-<!-- CJ: Aaron, I moved your part here because I don't think it relates to interoperability. But perhaps there's another place where it fits better? -->
+<!-- CJ: Aaron, I moved your part here because I don't think it relates to Interoperability. But perhaps there's another place where it fits better? -->
 
@@
 The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
-Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
+Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
 These limitations impede cumulative knowledge production in our field,
@@
 
-Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [REF Oisin paper].
+Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [REF Oisin paper].
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.

### 2025-02-21 manuscript.Rmd
closes #1

@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and reusable [REF] for both humans and computers.
-Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
+Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
@@
 accelerate cumulative knowledge acquisition.
+We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.
 
@@
 
-The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
+The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
 Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
@@
 
-Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One main concern is that social scientific theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
+One main concern is that theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
 <!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning).
@@
 
-The present paper introduces open science methods for theory specification.
+The present paper introduces open science methods for theory specification and archival.
 Merely publishing theory in a research article does not make it open;
 to be open, theory should adhere to established open science standards.
-We do so by implementing theories as information artifacts compliant with the FAIR principles.
-The resulting FAIR theories are archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo). 
+We do so by implementing theories as information artifacts,
+archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
 FAIR theories are *Findable* via a DOI or search,
@@
 and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
-Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
 <!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
+<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
 Following the original proposal of Lamprecht and colleagues,
@@
 We reflect on the necessary changes (which are minor),
-as well as on the current state and future of FAIR theory in the social sciences.
+as well as on the current state and future of FAIR theory in psychology.
 The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
@@
 The term "model" is not uniquely defined within the literature;
-it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
+it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
@@
 the method is not contained to any one particular definition.
+
 It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
@@
 Compared to theories expressed in natural language,
-formal theories facilitate inconsistency.
+formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
 Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
@@
 
-Although we are in favor of the advancement towards formal theories, 
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
@@
 and formal theory can be represented in a way that is not FAIR.
-The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
-For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
-Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
+The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
+They can be applied to theories representated in natural language,
+as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
+Thus, for example, "grounded theory", derived from qualitative research,
+can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
+Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
@@
 We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
-At present, the primary unit of social scientific communication is the academic paper.
+In most fields, the primary unit of scientific communication is the academic paper.
 A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
@@
 which is used by many academic publishers.
-Finally, the [DataCite Metadata Schema]() offers a standard way to document the nature of relationships between information artifacts.
+Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
 <!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
@@
 
-The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
+The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
 Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
@@
 access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
-Whereas the idea of theory can be quite nebulous to empirical social scientists,
+Whereas the idea of theory can be quite nebulous to empirical psychologists,
 FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
@@
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [@bringmannBackBasicsImportance2022] and lack of standardized ontologies for psychological constructs [@boscoMetaBUSVehicleFacilitating2017].
+Another limitation of scope is that FAIR theory does not resolve other related issues in psychology, such as the measurement crisis [@bringmannBackBasicsImportance2022] and lack of standardized ontologies for psychological constructs [@boscoMetaBUSVehicleFacilitating2017].
 However, our work here provides a template for addressing such challenges, <!-- NvD: I am not sure if we are offering a template to address these challenges.
@@
 Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
-it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
-Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
+it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [@bringmannBackBasicsImportance2022].
+Additionally, the "jingle-jangle" fallacy is prevalent in psychology:
 the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
@@
 This paves the way for more theory-driven scholarship,
-and accelerates cumulative knowledge acquisition in the social sciences and beyond.
+and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
 

### 2025-02-21 manuscript.Rmd
Make it clear that using GitHub and Zenodo actually meets ALL fair principles. closes #29

@@
 
-Open science infrastructure is an area of active development, and as such,
-the approach proposed here should not be considered definitive,
-but rather, as one proposal for a FAIR-compliant implementation of theory.
-The guiding principle of our implementation is to align and build upon existing successful open science infrastructures to the maximum possible extent.
+To concretize the FAIR principles,
+we propose an applied workflow for making theory FAIR.
+The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
 At the time of writing (2024),
-the value of using Git for version control of academic research is well-established,
-and the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
-Zenodo and GitHub are also integrated with the Open Science Framework (OSF),
+the value of using Git for version control of academic research is well-established [@ramGitCanFacilitate2013],
+and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* of the FAIR principles (Table \@ref(tab:tabfair)).
+Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
 a popular platform in psychology.
-Creating a front page on the OSF increases the visibility of a FAIR theory,
-while the integration with Zenodo and GitHub removes the need for uploading and maintaining the same information on multiple platforms.
-While we make use of specific open science infrastructures, it is important to stress that our workflow illustrates general principles which can also be implemented using other open science infrastructures.
-The process described here can be largely automated in R using the `theorytools` package; see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
+Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory amongst users of that platform,
+while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
+Note that open science infrastructure is an area of active development, and as such,
+the approach proposed here is subject to change.
+Nevertheless, it illustrates general principles which can also be implemented using other open science infrastructures.
+Our workflow can be largely automated in R using the `theorytools` package.
+For the most recent version of our proposed workflow, see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
 

### 2025-02-21 manuscript.Rmd
update 2025-02-21 11:33:31.377054

@@
 The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
-Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
+Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
 This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
@@
 archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
-FAIR theories are *Findable* via a DOI or search,
-*Accessible* in a machine- and human-readable filetype,
-*Interoperable* within the data analysis environment,
-and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
+Metadata are "data about the data", they provide information about the nature and content of an information artifact.
+Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
+FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
+*Accessible* in a machine- and human-readable filetype;
+*Interoperable* for specific purposed, for example, within the data analysis environment;
+and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
 <!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
@@
 First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
-Of the many services that provide DOIs for scientific information artefacts,
+Of the many services that provide DOIs for scientific information artifacts,
 Zenodo and the Open Science Framework are commonly used in psychology.
@@
 Zenodo stands out in this respect.
-Thus, while it is fine (and recommended) to use GitHub for version control,
+Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
 the version of record should be archived in a FAIR repository like Zenodo.
@@
 This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
-Regrettably, most academic search engines are designed to index traditional print publications, not other information artifacts.
+Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
 Since the status quo is to publish theories in papers,
@@
 Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
-Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
+Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
 Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
+Both approaches can be complementary, however.
+for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
+Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
+although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places and prevent inconsistencies from arising.
 
@@
 
-Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
+Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
 Firstly, theory and its associated metadata should use a formal, Accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
+[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/DGPSabstract.md) extracted from Zenodo.
 
@@
 
-The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
+The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved.
 In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
@@
 
-Note that this function also automatically provides basic FAIR theory metadata to Zenodo.
+Note that this function also automatically submits basic FAIR theory metadata to Zenodo.
 

### 2025-02-21 manuscript.Rmd
closes #30

@@
 and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
-[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/DGPSabstract.md) extracted from Zenodo.
+[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) extracted from Zenodo.
 

### 2025-02-23 manuscript.Rmd
complete draft for final comments

@@
 footnotelist      : no
-classoption       : "man"
+classoption       : "man, noextraspace"
 output            :
@@
   md_document: default
+header-includes:
+  - \usepackage{longtable, setspace}
+  - \raggedbottom
 ---
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+The equally critical inductive processeses of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
 Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
@@
 According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
-a meta-theoretical model of cumulative knowledge acquisition, <!-- NvD: De Groot was not a philosopher and his model is not commonly known/accepted in philosophy of science -->
-research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
+a meta-theoretical model of cumulative knowledge acquisition,
+research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs), panel (a).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
+
+```{r include = FALSE}
+library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
+ec_groot <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  observation [pos = "2,3!"];
+  induction [pos = "3,2!"];
+  deduction [pos = "3,1!"];
+  test [pos = "1,1!"];
+  evaluation [pos = "1,2!"];
+
+  observation -> induction;
+  induction -> deduction;
+  deduction -> test;
+  test -> evaluation;
+  evaluation -> observation;
+
+}')
+ec_wagenmakers <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  subgraph cluster_discovery {
+    label="Discovery";
+    induction [label="New hypothesis" pos = "3,2!"];
+    deduction [label="New prediction" pos = "3,1!"];
+  }
+  observation  [label="Old knowledge and old data" pos = "2,3!"];
+  subgraph cluster_justification {
+    label="Justification";
+    test [label="Test on new data" pos = "1,1!"];
+    evaluation [pos = "1,2!"];
+  }
+
+  observation -> induction [label="Speculate & explore" labeljust=r];
+  induction -> deduction  [label="Deduce"];
+  deduction -> test  [label="Design new experiment"];
+  test -> evaluation  [label="Statistical analysis"];
+  evaluation -> observation  [label="Knowledge accumulation"];
+
+}')
+ec_lissa <- grViz(diagram = 
+'digraph neato { node [fontsize = 6]; 
+    edge [fontsize = 6];
+    overlap = false; fontsize = 7;
+  graph [layout = neato]
+
+  theory [pos = "2,3!"];
+  prediction [pos = "3,2!"];
+  data [pos = "1,1!"];
+  test [pos = "2,1!"];
+  results [pos = "1,2!"];
+
+  theory -> prediction [label="deduction"];
+  prediction -> test [label = "implement inferential procedure"];
+  data -> results;
+  test -> results [label = "evaluate reproducible code"];
+  results -> theory [label="interpretation and generalization"];
+
+}')
+rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
+rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.svg")
+rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
+rsvg_pdf(charToRaw(export_svg(ec_groot)), "ec_groot.pdf")
+rsvg_pdf(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.pdf")
+rsvg_pdf(charToRaw(export_svg(ec_lissa)), "ec_lissa.pdf")
+
+rsvg_png(charToRaw(export_svg(ec_groot)), "ec_groot.png")
+rsvg_png(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.png")
+rsvg_png(charToRaw(export_svg(ec_lissa)), "ec_lissa.png")
+```
+
+```{r, eval = FALSE, out.width = "32%", fig.cap='Three implementations of the "empirical cycle".', fig.subcap=c('De Groot, 1969', 'Wagenmakers et al., 2018', 'Van Lissa, 2025')}
+knitr::include_graphics("ec_groot.pdf")
+knitr::include_graphics("ec_wagenmakers.pdf")
+knitr::include_graphics("ec_lissa.pdf")
+```
+
+```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle".'}
+knitr::include_graphics("ec_visio.pdf")
+```
 
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
 <!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
 <!-- by successively increasing the empirical content of a theory. -->
@@
 -->
-.
 There are, however, indications that contemporary psychology falls short of this ideal.
@@
 The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
-<!-- NvD: (most of) The other papers also don't have a true derivation chain. It is rare to see a logical derivation, analytical proof, or simulation that connects hypothesis to theory.  -->
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
-Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
-<!-- NvD: I am not sure what is being said here.
-In the statistical sense, people are testing the null hypothesis.
-However, that is (almost always) not the hypothesis they are interested in nor the one they mention the introduction of their papers.
-Also, I thought that gigerenzer's paper was about the ritualistic use of NHST, not the origin of the hypotheses that were being tested.
--->
+Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
+in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
+Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
+but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
@@
 The majority of papers ostensibly test hypotheses,
-but these are rarely connected, let alone derived, from theory,
-and test results do not routinely reference back to theories, contributing to their improvement or rejection.
+but these are rarely connected to, let alone derived from, theory.
+Moreover, test results do not routinely contribute to the improvement or rejection of theories.
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
@@
 Following the original proposal of Lamprecht and colleagues,
-we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
+we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
 We reflect on the necessary changes (which are minor),
@@
 
-```{r tabfair}
+```{r tabfair, eval = FALSE}
 tab_fair <- read.csv("fair_principles.csv", stringsAsFactors = FALSE)[, c("Criterion", "Original", "Theory", "Our.action")]
@@
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-our paper is not aligned with one particular definition.
+the method is not contained to any one particular definition.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
@@
 statements about the functional form of those connections,
-and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023, @guestWhatMakesGood2024].
+and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
 
@@
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
--<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
+<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
 Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
 -->
-
 The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
 It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.
-
-Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
-the method is not contained to any one particular definition.
 
@@
 we expect that it will become increasingly clear what kind and form of information is useful.
-As a particular FAIR theory evolves, the nature of the information tracked will likely change.
-For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
-From this theory, more precise *statistical/mathematical models* could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
-This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
-Then, an even more specific *generative/computational model* could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
-Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.
+As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
+For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
+After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
+Over time, more precise *statistical/mathematical models* could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
+This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+Going even further, a *generative/computational model* could be specified,
+which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
+Examples include experimental designs (e.g.
 longitudinal designs observing change over time), measurement tools (e.g.
@@
 Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
-Other types are useful for making relationships between multiple theory objects explicit: If an existing narrative theory is FAIRified without much own contribution, the FAIR theory would cross-reference the paper with the narrative theory as `relationType: IsDerivedFrom`; if substantial own contributions were necessary, `relationType:IsNewVersionOf` could be used.
+Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
+the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
+If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
 If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
@@
 We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
-Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
+Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
 Git tracks line-by-line changes to text-based files,
@@
 Whereas version control tracks changes,
-semantic versioning communicates what those changes mean to users of the theory.
+semantic versioning communicates what those changes mean to users of the theory,
+guides the social process of theory development, and signals how much a theory has been changed.
 We propose the following adaptation of semantic versioning for theories:
@@
 
-MAJOR version when you make backwards incompatible changes, i.e., the theory now contains empirical statements that are at odds with a previous version
-MINOR version when you expand the set of empirical statements in a backward compatible manner (i.e., the previous version is subsumed within the new version)
-PATCH version when you make backward compatible bug fixes,
-cosmetic changes, fix spelling errors, or add clarifications
-```
-
-Semantic versioning guides the social process of theory development, communicating how much a theory is changing over tiem.
+MAJOR version when you make backwards incompatible changes, i.e., the
+      theory now contains empirical statements that are at odds with
+      a previous version
+MINOR version when you expand the set of empirical statements in a
+      backward compatible manner (i.e., the previous version is
+      subsumed within the new version)
+PATCH version when you make backward compatible bug fixes, cosmetic
+      changes, fix spelling errors, or add clarifications
+```
 
@@
 
-The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
+The four Findability criteria are applicable to theory with only minor adjustments, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
 First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
@@
 ontology [@guyonMeasurementOntologyEpistemology2018],
-or catalog of psychological phenomena [REF Noah Denny].
+or catalog of psychological phenomena.
 Metadata should also include identifiers for all the versions of the theory it describes	(F3);
@@
 <!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
-Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
+Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a specific) FAIR theory.
 <!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->
@@
 Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
-although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places and prevent inconsistencies from arising.
+although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.
 
@@
 
-The Accessibility principles pertain to *regulating* access, not only maximizing it.
+The Accessibility principles are sometimes misunderstood as pertaining to *maximizing* access; rather, they should be understood as *regulating* it.
 They apply to theory with minor changes.
 Firstly, theory and its associated metadata should be Accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
+This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as Git), and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
@@
 To the extent that theories are still contained within papers,
-paywalls erected by commercial publishers constitute a barrier.
-Open Access publishing thus increases the Accessibility of all academic output, including theory.
+paywalls erected by publishers constitute a barrier.
+Open Access publishing increases the Accessibility of all academic output, including theory.
 A second impediment is more indirect:
@@
 creating the background knowledge needed for mutual understanding and productive scientific work [@kuhnStructureScientificRevolutions2009].
-In reality, however, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
+In real life, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
 including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
 
-A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
+A third impediment arises when theories have, what we call, a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
 DOA occurs when a theory cannot be understood by independent scholars,
-thus requiring the original author for interpretation and clarification.
-We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
-as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
+requiring the original author to provide interpretation and clarification.
+<!-- We have heard DOA referred to apocryphally as the "ask Leon" phenomenon, -->
+<!-- as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results. -->
 DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
@@
 because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
-The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
-This demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible in terms of both availability and interpretability.
-
-From a pragmatic perspective,
-these considerations underscore the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
-while acknowledging that Accessibility exists on a spectrum, 
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
+DOA is inherently problematic, as illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
+This example demonstrates that original authors are not the ultimate authority on their theories.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interpretability.
+
 In sum, while it is impossible to communicate a theory completely unambiguously,
 scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+<!-- Accessibility requires theoreticians to explicit as possible about assumptions and ontological definitions to eliminate unnecessary ambiguity. -->
+<!-- At the same time, it is important to recognize that there is an upper bound on interpretability, which means that it is impossible to eliminate all ambiguity. -->
+It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
+to invite others to fill in the blanks and motivate ever further explication of theory.
 A theory's Accessibility is increased by
@@
 formalization,
-and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [REF langeChecklist2025].
+and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs [REF LANGE CHECKLIST].
 
@@
 Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
-Firstly, theory and its associated metadata should use a formal, Accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
+Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
 The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
@@
 be instantiated in a human- and machine-readable datatype,
-as previously recommended [@vanlissaWORCSWorkflowOpen2021].
+as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
 Depending on the level of formalization of the theory,
@@
 While a theory represented as a bitmap image is not very Interoperable,
-the same image represented in the DOT language [@DOTLanguage]<!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.
+the same image represented in the DOT language [@DOTLanguage2024]
+for representing graphs does meet this ideal.
 
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by cross-referencing the DOI of those nested theories [@ContributingCitationsReferences].
+This can be achieved by cross-referencing the DOI of those nested theories [@ContributingCitationsReferences2024].
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
@@
 
-With regard to the state of Interoperability in contemporary psychology, 
-Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] implies that theories ought to be highly X-Interoperable in psychological researchers' day-to-day work.
+With regard to the state of Interoperability in psychology, 
+Kurt Lewin's -@lewinPsychologyProcessGroup1943 adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
@@
 
-If we consider the current state of Reusability in psychological theory, there appears to be a norm *against* theory reuse:
+If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
@@
 and "Who determines how a theory may be reused or changed?".
-These questions imply a norm against modifying theory without consent from the author.
-<!-- NvD: I suggest removing the rest of the sentence (as a result of an earlier suggestion).
--- > reminiscent of the aforementioned problem of dependency on the author.
-
-Licensing theories for reuse unambiguously answers these questions.
-With the caveat that legislation may vary across contexts and jurisdictions, the following should not be interpreted as legal advice.
-In determining what license is appropriate for theory,
-a key consideration is that copyright law protects authors' rights according to the *idea-expression dichotomy* [@bently2010copyright].
+These questions imply a norm against modifying theory without its author's consent, reminiscent of the aforementioned problem of dependency on the author.
+
+Licensing theories for reuse unambiguously answers these questions,
+with the caveats that legislation may vary across contexts and jurisdictions, and that this paper does not constitute legal advice.
+Two considerations are important when determining what license is appropriate for theory.
+A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy [@bently2010copyright].
 It explicitly does not
 *"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
-Copyright thus extends to creative works expressing a theory (e.g., writing, visual illustrations),
+Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
 but not to the underlying theoretical idea.
@@
 <!-- then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection. -->
-Another important consideration is that academic research is covered under "fair use" exemptions to copyright.
+A second consideration is that academic research is covered under "fair use" exemptions to copyright.
 <!-- so copyright would pose few restrictions to Reusability in scholarly communication. -->
-Given that copyright does not cover ideas in their purest form (like formal theories), and that academic use is likely exempted from its protection,
-in many cases, it may be sensible to explicitly license theories in a way that encourages Reusability.
-The CC0 (no rights reserved) license does this by explicitly waiving all rights and encouraging reuse.
+Given these two considerations - that copyright does not protect ideas in their purest form and that academic use offers exemptions to copyright - it may be counterproductive and possibly misleading to adopt a license that assumes copyright protection to theories.
+For psychological theories without commercial aspects, we suggest using a licence that explicitly waives copyright and encourages Reusability,
+such as CC0 (no rights reserved).
 
@@
 For example, while a CC0 license does not legally mandate attribution,
-the norms of good scientific practice require that scholars comprehensively cite theory and related works [@aalbersbergMakingScienceTransparent2018].
-Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a "Code of Conduct" which prescribes behavioral norms for contributors and users of a theory.
+the norms of good scientific practice mandate that scholars comprehensively cite theory and related works [@aalbersbergMakingScienceTransparent2018].
+Particularly when FAIRifying an existing theory, failing to credit its author amounts to scientific malpractice.
+Another instrument for guiding the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
+A final suggestion is to create or adopt a "Code of Conduct" which prescribes behavioral norms for contributors and users of a theory [@ehmkeContributorCovenantCode2014].
 
@@
 the value of using Git for version control of academic research is well-established [@ramGitCanFacilitate2013],
-and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* of the FAIR principles (Table \@ref(tab:tabfair)).
+and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* of the FAIR principles ([Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
 Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
@@
 Note that open science infrastructure is an area of active development, and as such,
-the approach proposed here is subject to change.
-Nevertheless, it illustrates general principles which can also be implemented using other open science infrastructures.
+the approach proposed here might change.
 Our workflow can be largely automated in R using the `theorytools` package.
-For the most recent version of our proposed workflow, see the package vignette on FAIR theory, `vignette("fairtheory", package = "theorytools")`.
+To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
+It can be accessed by running `vignette("fairtheory", package = "theorytools")` in R.
+We present a brief summary of the instructions at the time of writing here,
+to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
 
@@
 The empirical cycle, described on page 28 of @degrootMethodologyFoundationsInference1969,
-is a kind of meta-theory of theory construction.
+is a meta-theory of theory construction.
 The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
@@
 it is defined at either the "theory" or "specification" level.
-We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \@ref(tab:tabfair)):
+We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
+The implementation below describes the model as a directed graph.
+Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
+and the second half describes their proposed interrelations.
+This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
+We can save this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
 
@@
 ```
-
-```{r include = FALSE}
-library(DiagrammeR)
-library(DiagrammeRsvg)
-library(rsvg)
-ec_groot <- grViz(diagram = 
-'digraph neato { node [fontsize = 6]; 
-    edge [fontsize = 6];
-    overlap = false; fontsize = 7;
-  graph [layout = neato]
-
-  observation [pos = "2,3!"];
-  induction [pos = "3,2!"];
-  deduction [pos = "3,1!"];
-  test [pos = "1,1!"];
-  evaluation [pos = "1,2!"];
-
-  observation -> induction;
-  induction -> deduction;
-  deduction -> test;
-  test -> evaluation;
-  evaluation -> observation;
-
-}')
-ec_wagenmakers <- grViz(diagram = 
-'digraph neato { node [fontsize = 6]; 
-    edge [fontsize = 6];
-    overlap = false; fontsize = 7;
-  graph [layout = neato]
-
-  subgraph cluster_discovery {
-    label="Discovery";
-    induction [label="New hypothesis" pos = "3,2!"];
-    deduction [label="New prediction" pos = "3,1!"];
-  }
-  observation  [label="Old knowledge and old data" pos = "2,3!"];
-  subgraph cluster_justification {
-    label="Justification";
-    test [label="Test on new data" pos = "1,1!"];
-    evaluation [pos = "1,2!"];
-  }
-
-  observation -> induction [label="Speculate & explore" labeljust=r];
-  induction -> deduction  [label="Deduce"];
-  deduction -> test  [label="Design new experiment"];
-  test -> evaluation  [label="Statistical analysis"];
-  evaluation -> observation  [label="Knowledge accumulation"];
-
-}')
-ec_lissa <- grViz(diagram = 
-'digraph neato { node [fontsize = 6]; 
-    edge [fontsize = 6];
-    overlap = false; fontsize = 7;
-  graph [layout = neato]
-
-  theory [pos = "2,3!"];
-  prediction [pos = "3,2!"];
-  data [pos = "1,1!"];
-  test [pos = "2,1!"];
-  results [pos = "1,2!"];
-
-  theory -> prediction [label="deduction"];
-  prediction -> test [label = "implement inferential procedure"];
-  data -> results;
-  test -> results;
-  results -> theory [label="interpretation and generalization"];
-
-}')
-rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
-rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.svg")
-rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
-```
-
-This implementation describes the model as a directed graph.
-Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
-and the second half describes their proposed interrelations.
-This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
-We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
 
@@
 must be set to "Public" to take advantage of GitHub's Zenodo integration.
-<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
-<!-- Next, in the terminal window, run the following lines of code sequentially: -->
-
-<!-- ``` -->
-<!-- git add . -->
-<!-- git commit -m "First commit" -->
-<!-- git remote add origin *[the repository URL you copied]* -->
-<!-- git branch -M main -->
-<!-- git push -u origin main -->
-<!-- ``` -->
 Push the local files to the Git remote repository, and keep them synchronized going forward.
@@
 
-The process of archiving a GitHub repository on Zenodo is documented in a vignette in the `theorytools` R-package, so that it can be kept up-to-date.
-We present a brief summary of the instructions at the time of writing here.
 First, create a Zenodo account with your existing GitHub account.
@@
 To document our archive as a FAIR theory requires adding some extra information on Zenodo.
-Supplying the following information helps improve the Findability of a theory:
+Supplying the information below helps improve the Findability of a theory.
+See here for [an example of the resulting FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) archived in Zenodo.
 
@@
 
-## 6. Making Changes
-
+<!-- ## 6. Making Changes -->
 <!-- CJ: Aaron, I moved your part here because I don't think it relates to Interoperability. But perhaps there's another place where it fits better? -->
@@
 
-Note that this function also automatically submits basic FAIR theory metadata to Zenodo.
-
-## Forking Different Implementations of a Theory
+## Updating a Theory
 
@@
 but not all of them have interpreted his work the same.
-For example, Wagenmakers and colleagues [-@wagenmakersCreativityVerificationCyclePsychological2018] write *"De Groot’s “empirical cycle,” shown here in Figure 6"* - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
+For example, Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle diverges substantially from De Groot's description.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-This is achieved by ["forking the repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo),
-["cloning"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo) the forked repository to our local computer,
-making any changes we want, and then completing steps 4-5 of "Making a Theory FAIR".
-
-We have implemented Wagenmakers and colleagues' version as a DOT graph
+Wagenmakers and colleagues' version can also be implemented as a DOT graph
 to illustrate some clear deviations from the original.
-First, the phases of the cycle have been renamed.
-While this change was not described in the paper,
-we assumed that the labels are meant to illustrate the phases, not substantially change the ontology.
-We represent this change by adding labels to the original DOT graph.
-Note, however, that the labels suggest a focus on empirical psychology that was absent in the original formulation, which was more general.
+Notice that, first, the phases of the cycle have been renamed.
+This change was not described in the paper.
+If we assume that the labels are meant to illustrate the phases,
+not substantially change the ontology,
+then we can represent it by adding labels to the original DOT graph.
+The labels do suggest a focus on empirical psychology not present in De Groot's version.
 Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
 As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-Second, the authors mention an explicit change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
+The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
 The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
@@
 The first author was inspired by De Groot as well,
-but they specify the empirical cycle differently.
+but again specified the empirical cycle differently.
 First, notice that the nodes in De Groot's formulation mostly refer to processes.
 This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
-In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
+In the implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
 Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
@@
 <!-- ``` -->
-For example, many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
+Many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
 Deriving hypotheses from theory is also not purely deductive,
@@
 
-## Using FAIR theory to Perform Causal Inference
-
-Some have argued that *causal explanations* are a property of good theory [@meehlAppraisingAmendingTheories1990].
-According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995].
-Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
-and could even be integrated into the analysis environment.
-
-In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
-We could use that to illustrate updating FAIR theory:
-
-https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45
-
-We can find more examples of causal inference with DAGs in these tutorials:
-
-https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/
-
-https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/
+The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
+If you start with an existing GitHub repository, and wish to make some changes to it,
+this is commonly done by [creating a "branch"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository).
+A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
+Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
+Figure \@ref(fig:figbranch) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
+A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
+If you wish to develop a new version of someone else's FAIR theory,
+it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
+This creates a copy of their repository onto your GitHub account.
+Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
+Figure \@ref(fig:figcompare) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+
+```{r figbranch, fig.cap="Branches can contain different versions of a theory.", out.width = "50%"}
+knitr::include_graphics("git_branches.png")
+```
+
+```{r figcompare, fig.cap="GitHub's compare functionality.", out.width = "50%"}
+knitr::include_graphics("git_compare.png")
+```
+
+## Further Uses of FAIR Theory
+
+As uses of FAIR theory are best illustrated using tutorial examples,
+the `theorytools` package contains several vignettes that showcase specific applications.
+At the time of writing, the package includes a [vignette on theory specification](),
+that is to say, moving from a theory represented as prose to a FAIR theory,
+based on the many-theorists project by Glöckner and Fiedler (in preparation).
+Another vignette illustrates the use of FAIR theory for causal inference [@pearlCausalDiagramsEmpirical1995].
+Furthermore, one vignette illustrates the use of FAIR theory for [conducting a simulation study]().
+More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
+<!-- Some have argued that good theories involve *causal explanations* [@meehlAppraisingAmendingTheories1990]. -->
+<!-- Explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995]. -->
+<!-- Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference, -->
+<!-- and could even be integrated into the analysis environment. -->
 
@@
 The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
-Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
+Psychological theories often fall short of the FAIR principles:
+they are hard to find and access, have no practical uses in scholars' daily workflows beyond providing context for a literature review,
+and are more likely to be forgotten or replaced than reused.
 These limitations impede cumulative knowledge production in our field,
@@
 We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
-and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
+and creating a FAIR resource for theory construction tools and documentation that can be updated as best practices continue to develop.
 
 Making theory FAIR allows researchers to more easily find a relevant framework;
-access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
+access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
 Whereas the idea of theory can be quite nebulous to empirical psychologists,
@@
 By providing much-needed open science methods for theory construction,
-FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the "open empirical cycle" [REF Hoijtink].
+FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the "open empirical cycle" [@hoijtinkOpenEmpiricalCycle2023].
 
@@
 The advantage of modular publishing is that authors can be credited for theory development.
-Given the current emphasis on empirical papers [REF], theoretical papers can be hard to publish.
+Given the current emphasis on empirical papers [@mcphetresDecadeTheoryReflected2021], theoretical papers can be hard to publish.
 FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
 
-Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [REF Oisin paper].
+Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [@haslbeckModelingPsychopathologyData2021].
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
 In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
-clarifying any remaining areas of contention or misunderstanding. 
+clarifying any remaining areas of contention or misunderstanding [@vanlissaVisionTeamScience2024]. 
 Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
@@
 
-Finally, FAIR theory plays an important role in science communication, because theory synthesizes contemporary scientific understanding about a phenomenon.
+Finally, FAIR theory holds substantial potential for science communication.
+While the general public is ill-equipped to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon,
+theories encapsulate the bigger picture of contemporary scientific understanding.
+For example, while few people read empirical studies on attachment,
+attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
 Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
@@
 Another limitation is the learning curve associated with tools like Git and Zenodo.
-The `theorytools` R-package mitigates this limitation by automating key steps in the process.
+The `theorytools` R-package mitigates this limitation for R-users by automating key steps in the process.
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
@@
 Education might help address this limitation; with this in mind,
-we are developing open educational materials on theory development.
+we have shared several open educational materials on theory development in the "FAIR Theory Community" on Zenodo, and we encourage others to reuse these and share their materials.
 
@@
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [@bringmannBackBasicsImportance2022].
-Additionally, the "jingle-jangle" fallacy is prevalent in psychology:
+The "jingle-jangle" fallacy is common in psychology:
 the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.

### 2025-02-23 manuscript.tex
complete draft for final comments

@@
 \documentclass[
-  man,floatsintext]{apa6}
+  man, noextraspace,floatsintext]{apa6}
 \usepackage{amsmath,amssymb}
@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8186}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8881}
 \usepackage{lineno}
@@
 \usepackage{csquotes}
+\usepackage{longtable, setspace}
+\raggedbottom
 \ifLuaTeX
@@
 
-This is a preprint paper, generated from Git Commit \# c8b13da7.
+This is a preprint paper, generated from Git Commit \# ecb21ba7.
 
@@
 \abstract{%
-Many meta‐scientific initiatives have focused on enhancing the rigor of theory testing, yet the systematic specification of theories remains underdeveloped.
-We present FAIR theory, a new approach that transforms scientific theories into digital, evolving artifacts---developed collaboratively---to bolster the robustness and interconnectedness of scientific knowledge.
-By representing theories as formal, machine‐readable objects (e.g., mathematical models, computer code, nomological networks, or explicit operational definitions) and tracking their evolution through version control and semantic versioning, FAIR theory provides a structured, cumulative framework for theory development.
-We propose that each theory is assigned a universally unique identifier and enriched with detailed metadata, ensuring it is Findable, Accessible, Interoperable, and Reusable.
-This digital infrastructure addresses a critical gap in scholarly practice by introducing the collaborative tools and standards for theory specification that have already revolutionized methodology, data sharing, and publishing.
-Ultimately, FAIR theory aims to transform how scientific ideas are communicated, evaluated, and iteratively refined, thereby accelerating cumulative knowledge acquisition across disciplines.
+Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped.
+To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts.
+FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved.
+This paper adapts the FAIR priciples for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
+We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing.
+By applying the principles and infrastructure that have already revolutionized sharing of data and publications to theory, we establish a sustainable, transparent, and collaborative approach to theory development.
+FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond.
+FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
 }
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable {[}REF{]} for both humans and computers.
-Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M.
+D.
+Wilkinson et al., 2016a) for both humans and computers.
+Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
 R.
@@
 
-The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
+The so-called ``replication crisis'' has prompted extensive reforms in psychology and other scientific fields (Lavelle, 2021; Scheel, 2022).
 Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
@@
 as the theory's central claims evade falsification.
-
-Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-One main concern is that social scientific theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
+A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
+Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
+The profound difference between these definitions becomes clear when considering the act of changing a child's dirty diaper.
+Under the original definition, few would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience
+Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.
+
+Scholars have been raising concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
+One main concern is that theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
 When theories do not make precise predictions,
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' (\textbf{peirceCollectedPapersCharles1960?}).
-However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4881).
-Defining what makes a theory ``best'' requires making auxiliary assumptions.
-For present purposes, induction and abduction are interchangeable terms.}.
+The equally critical inductive processeses of theory construction and improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories.
+Others have used the term ``abduction'', coined by Peirce to describe ``inference to the best explanation''.
+However, Peirce later acknowledged that abduction ``is not essentially different from induction'' (Peirce, 1960, p.
+4881).
+As defining what makes a theory ``best'' requires auxiliary assumptions, we consider abduction to be a special case of induction.
+For present purposes, however, the terms are interchangeable.}.
 The present paper restores balance by applying, for the first time,
@@
 We apply the FAIR principles to scientific theories,
-introducing the concept of \emph{FAIR theory} to
+introducing \emph{FAIR theory} to
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
+research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}, panel (a).
 In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
@@
 \centering
-\includegraphics{empirical_cycle.pdf}
-\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
+\includegraphics{ec_visio.pdf}
+\caption{\label{fig:figecs}Three implementations of the ``empirical cycle''.}
 \end{figure}
@@
 In a progressive research program (Lakatos, 1971),
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
-.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
 There are, however, indications that contemporary psychology falls short of this ideal.
@@
 The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
-In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
-Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
+Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
+in which case it would be straightforward and important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
+Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
+but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
@@
 The majority of papers ostensibly test hypotheses,
-but these are rarely connected, let alone derived, from theory,
-and test results do not routinely reference back to theories, contributing to their improvement or rejection.
+but these are rarely connected to, let alone derived from, theory.
+Moreover, test results do not routinely contribute to the improvement or rejection of theories.
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Paul E.
-Meehl, 1978).
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
 
@@
 
-The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
+The present paper introduces open science methods for theory specification and archival.
 Merely publishing theory in a research article does not make it open;
 to be open, theory should adhere to established open science standards.
-We apply the FAIR principles to digital representations of theory,
-introducing a FAIR metadata format to make theories \emph{Findable} via a DOI,
-\emph{Accessible} in a machine- and human-readable filetype,
-\emph{Interoperable} within the data analysis environment,
-and \emph{Reusable} in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
-Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
+We do so by implementing theories as information artifacts,
+archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+Metadata are ``data about the data'', they provide information about the nature and content of an information artifact.
+Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
+FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
+\emph{Accessible} in a machine- and human-readable filetype;
+\emph{Interoperable} for specific purposed, for example, within the data analysis environment;
+and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
 Following the original proposal of Lamprecht and colleagues,
-we adapt the FAIR principles for theory, see Table \ref{tab:tabfair}.
+we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
 We reflect on the necessary changes (which are minor),
-as well as on the current state and future of FAIR theory in the social sciences.
+as well as on the current state and future of FAIR theory in psychology.
 The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
@@
 
-\begin{lltable}
-
-\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
-\caption{\label{tab:tabfair}}\\
-\toprule
-Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
-\midrule
-\endfirsthead
-\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
-\toprule
-Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
-\midrule
-\endhead
-F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory (meta)data has a global, unique and persistent identifier & Rephrased\\
-F2 & Data are described with rich metadata & Theory is described with rich metadata & Rephrased\\
-F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & Rephrased and extended\\
-F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & Rephrased, needs work\\
-A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & Rephrased\\
-A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Remain the same\\
-A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Remain the same, but less relevant\\
-A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Rephrased, but less relevant\\
-I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
-I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
-I2S.1 & - &  & \\
-I2S.2 & - &  & \\
-I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
-I4S & - &  & \\
-R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
-R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
-R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
-R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
-\bottomrule
-\end{longtable}
-
-\end{lltable}
-
 \subsection{What is Theory?}\label{what-is-theory}
@@
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-our paper is not aligned with one particular definition.
+the method is not contained to any one particular definition.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
@@
 statements about the functional form of those connections,
-and statements about their specific numerical values Guest (2024).
+and statements about their specific numerical values (cf. Frankenhuis, Panchanathan, \& Smaldino, 2023; Guest, 2024).
 
@@
 The term ``model'' is not uniquely defined within the literature;
-it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
+it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020).
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
--
-
 The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
@@
 
-Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
-the method is not contained to any one particular definition.
 It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
@@
 we expect that it will become increasingly clear what kind and form of information is useful.
-As a particular FAIR theory evolves, the nature of the information tracked will likely change.
-For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
-From this theory, more precise \emph{statistical/mathematical models} could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
-This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
-Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
-Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).
+As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
+For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
+After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
+Over time, more precise \emph{statistical/mathematical models} could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
+This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
+Going even further, a \emph{generative/computational model} could be specified,
+which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
+Examples include experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).
 
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
+Greater formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
@@
 the theory becomes hard to test.
+Compared to theories expressed in natural language,
+formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
 Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
@@
 
-Although we are in favor of the advancement towards formal theories,
-this might not always be desirable or feasible.
-Fortunately, formalization is not required to make theories FAIR.
-To some extent, FAIRness and formalization are orthogonal.
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
 rather, it increases the fidelity and ease with which they are communicated.
-The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
-For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
-Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
+Thus, FAIR theory does not require theories to be formal,
+and formal theory can be represented in a way that is not FAIR.
+The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
+They can be applied to theories representated in natural language,
+as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
+Thus, for example, ``grounded theory'', derived from qualitative research,
+can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
+Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).
@@
 
-We propose FAIR theory as an instantiation of modular publishing (\textbf{kirczModularityNextForm1998?}).
-At present, the primary unit of social scientific communication is the academic paper.
+We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
+In most fields, the primary unit of scientific communication is the academic paper.
 A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
 Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
-with adequate metadata that is indexed in standardized repositories (\textbf{vandesompelRethinkingScholarlyCommunication2004?}).
+with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
 Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
@@
 which is used by many academic publishers.
-Finally, the \href{}{DataCite Metadata Schema} offers a standard way to document the nature of relationships between information artifacts.
+Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
 For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
 and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
-Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.
+Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}.
+Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
+the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
+If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
+If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.
 
@@
 
-MAJOR version when you make backwards incompatible changes, i.e., the theory now contains empirical statements that are at odds with a previous version
-MINOR version when you expand the set of empirical statements in a backward compatible manner (i.e., the previous version is subsumed within the new version)
-PATCH version when you make backward compatible bug fixes,
-cosmetic changes, fix spelling errors, or add clarifications
+MAJOR version when you make backwards incompatible changes, i.e., the
+      theory now contains empirical statements that are at odds with
+      a previous version
+MINOR version when you expand the set of empirical statements in a
+      backward compatible manner (i.e., the previous version is
+      subsumed within the new version)
+PATCH version when you make backward compatible bug fixes, cosmetic
+      changes, fix spelling errors, or add clarifications
 \end{verbatim}
-
-Semantic versioning guides the social process of theory development, communicating how much a theory is changing over tiem.
 
@@
 In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
-For example, predator-prey models have inspired theories of intelligence (\textbf{vandermaasDynamicalModelGeneral2006?}), and the Eysenck model of atomic magnetism has inspired a network theory of depression (\textbf{cramerMajorDepressionComplex2016?}).
+For example, predator-prey models have inspired theories of intelligence (Van Der Maas et al., 2006), and the Eysenck model of atomic magnetism has inspired a network theory of depression (Cramer et al., 2016).
 Findability also enables meta-research on theories,
@@
 
-The four Findability criteria are applicable to theory with only minor adjustments, see Table \ref{tab:tabfair}.
+The four Findability criteria are applicable to theory with only minor adjustments, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
 First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
-Of the many services that provide DOIs for scientific information artefacts,
+Of the many services that provide DOIs for scientific information artifacts,
 Zenodo and the Open Science Framework are commonly used in psychology.
@@
 ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
-or catalog of psychological phenomena {[}REF Noah Denny{]}.
+or catalog of psychological phenomena.
 Metadata should also include identifiers for all the versions of the theory it describes (F3);
@@
 Finally, metadata should be registered or indexed in a searchable registry (F4).
-Zenodo and GitHub are both searchable.
-Standardized metadata further improve Findability in these repositories.
-The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}).
-Furthermore, a standard keyword can be used; we suggest using the keyword \texttt{"fairtheory"} for all resources that constitute or reference a FAIR theory.
+It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
+only few are specifically designed for FAIR-compliant archival.
+Zenodo stands out in this respect.
+Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work (Ram, 2013),
+the version of record should be archived in a FAIR repository like Zenodo.
+Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
+The DataCite Metadata Schema provides a controlled vocabulary for research output, and the \texttt{resource\_type:\ model} matches the description of FAIR theory (DataCite Metadata Working Group, 2024).
+Furthermore, we suggest using the keyword \texttt{"fairtheory"} for all resources that constitute or reference (a specific) FAIR theory.
 
 Findability is substantially amplified if intended users of a resource know where to search for it.
-This is a known problem in relation to research data and software (\textbf{katzSpecialIssueSoftware2024?}).
-Regrettably, most academic search engines are designed to index traditional print publications, not other information artifacts.
+This is a known problem in relation to research data and software (Katz \& Chue Hong, 2024).
+Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
 Since the status quo is to publish theories in papers,
@@
 Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
-Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
+Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
 Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
+Both approaches can be complementary, however.
+for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
+Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
+although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.
 
@@
 
-Transparent scholarly communication about theory requires that theories are accessible to all researchers and other stakeholders.
-If theories are not accessible, researchers cannot reuse and refine them.
-Thus, accessibility can accelerate cumulative knowledge acquisition.
-Making theories accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
+Transparent scholarly communication about theory requires that theories are Accessible to all researchers and other stakeholders.
+If theories are not Accessible, researchers cannot reuse and refine them.
+Thus, Accessibility can accelerate cumulative knowledge acquisition.
+Making theories Accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
 While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
@@
 
-The Accessibility principles pertain to \emph{regulating} access, not only maximizing it.
+The Accessibility principles are sometimes misunderstood as pertaining to \emph{maximizing} access; rather, they should be understood as \emph{regulating} it.
 They apply to theory with minor changes.
-Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
-This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
+Firstly, theory and its associated metadata should be Accessible by their identifier using a standardized communications protocol (A1).
+This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as Git), and archiving that repository on Zenodo for long-term storage.
 The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via \texttt{https} or \texttt{git}).
-Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
+Secondly (A2), theory metadata should be Accessible, even when the theory is no longer available,
 which is also achieved via long-term storage (e.g., on Zenodo).
@@
 
-At present, there are several impediments to theories' accessibility.
+At present, there are several impediments to theories' Accessibility.
 To the extent that theories are still contained within papers,
-paywalls erected by commercial publishers constitute a barrier.
-Open Access publishing thus increases the accessibility of all academic output, including theory.
+paywalls erected by publishers constitute a barrier.
+Open Access publishing increases the Accessibility of all academic output, including theory.
 A second impediment is more indirect:
 While open access publishing increases practical access to theories,
-accessibility also requires clear and explicit communication.
+Accessibility also requires clear and explicit communication.
 This property of good theories has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
@@
 It is important to acknowledge the \emph{indeterminacy of translation} (Quine, 1970):
-which holds that every communicative utterance has multiple alternative translations, with no \emph{objective} means of choosing the correct one.
+which holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+has multiple alternative translations, with no \emph{objective} means of choosing the correct one.
 It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
@@
 Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
-The Kuhnian notion of ``normal science'', conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding (Kuhn, 2009).
-From a pragmatic perspective, these considerations indicate that,
-when striving to make theory accessible,
-it is important to be as explicit as possible (e.g., about assumptions and ontological definitions),
-while acknowledging that accessibility exists on a spectrum,
-and that it is impossible to eliminate all ambiguity.
-Rather, it may benefit scientific discourse to anticipate misunderstanding,
-and use it to drive further explication of theory.
-In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
-and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
-will advance its Accessibility (\textbf{langeChecklist2025?}).
-
-A third impediment arises when theories have a ``dependency on the author'' (DOA).
+The notion of ``normal science'' describes a phase in which a scientific community operates within the context of a shared paradigm,
+creating the background knowledge needed for mutual understanding and productive scientific work (Kuhn, 2009).
+In real life, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
+including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
+
+A third impediment arises when theories have, what we call, a ``dependency on the author'' (DOA).
 DOA occurs when a theory cannot be understood by independent scholars,
-thus requiring the original author for interpretation and clarification.
-We have heard DOA referred to apocryphally as the ``ask Leon'' phenomenon,
-as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
+requiring the original author to provide interpretation and clarification.
 DOA relates to the discourse on ``Great Man Theorizing'' (Guest, 2024) because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
@@
 because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target (Szollosi \& Donkin, 2021).
-The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
-This demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible in terms of both availability and interpretability.
+DOA is inherently problematic, as illustrated by cases where third parties identify logical inconsistencies within a theory (e.g., Kissner, 2008).
+This example demonstrates that original authors are not the ultimate authority on their theories.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interpretability.
+
+In sum, while it is impossible to communicate a theory completely unambiguously,
+scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
+to invite others to fill in the blanks and motivate ever further explication of theory.
+A theory's Accessibility is increased by
+reducing dependencies on (implicit) background knowledge,
+explication of assumptions,
+formalization,
+and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs {[}REF LANGE CHECKLIST{]}.
 
@@
 
-Interoperability pertains to the property of information artefacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
+Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016b).
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 be instantiated in a human- and machine-readable datatype,
-as previously recommended (Van Lissa et al., 2021).
+as should all information artifacts created while performing scholarly work (Van Lissa et al., 2021).
 Depending on the level of formalization of the theory,
@@
 and Gray's theory maps (Gray, 2017).
-While a theory represented as a bitmap image is not very interoperable,
-the same image represented in the DOT language (\textbf{DOTLanguage?}) for representing graphs does meet this ideal.
+While a theory represented as a bitmap image is not very Interoperable,
+the same image represented in the DOT language ({``{DOT Language},''} 2024)
+for representing graphs does meet this ideal.
 
 Secondly, theory (meta)data should use vocabularies that follow FAIR principles (I2).
-Aside from the aforementioned Datacite metadata schema (\textbf{datacitemetadataworkinggroupDataCiteMetadataSchema2024?}),
+Aside from the aforementioned Datacite metadata schema (DataCite Metadata Working Group, 2024),
 in the context of theory, this highlights the importance of establishing standardized ontologies.
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by cross-referencing the DOI of those nested theories ({``Contributing {Citations} and {References},''} n.d.).
+This can be achieved by cross-referencing the DOI of those nested theories ({``Contributing {Citations} and {References},''} 2024).
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
@@
 
-Recent work points out that interoperability is not an all-or-nothing property.
-The concept of X-interoperability was introduced to answer the question: \emph{interoperable for what?}
-X-interoperability is defined as facilitating ``successful communication between machines and between humans and machines {[}, where{]} A and B are considered X-interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
-This revised definition makes it possible to outline a theory's affordances in terms of X-interoperability.
-For example, a FAIR theory may be X-interoperable for deriving testable hypotheses,
+Recent work points out that Interoperability is not an all-or-nothing property.
+The concept of X-Interoperability was introduced to answer the question: \emph{Interoperable for what?}
+X-Interoperability is defined as facilitating ``successful communication between machines and between humans and machines {[}, where{]} A and B are considered X-Interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
+This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
+For example, a FAIR theory may be X-Interoperable for deriving testable hypotheses,
 or for the purpose of selecting relevant control variables,
@@
 If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
-we see how each of these properties incurs certain affordances in terms of X-interoperability (Table \ref{tab:tabmeehl}).
+we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \ref{tab:tabmeehl}).
 
@@
 \toprule
-Property & \multicolumn{1}{c}{X-interoperability}\\
+Property & \multicolumn{1}{c}{X-Interoperability}\\
 \midrule
@@
 
-With regard to the state of interoperability in contemporary psychology,
-Kurt Lewin's adage ``there's nothing as practical as a good theory'' (Lewin, 1943) implies that theories ought to be highly X-interoperable in psychological researchers' day-to-day work.
+With regard to the state of Interoperability in psychology,
+Kurt Lewin's -Lewin (1943) adage ``there's nothing as practical as a good theory'' paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
-The examples of X-interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
-For example, interoperable theory could be used
+The examples of X-Interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
+For example, Interoperable theory could be used
 to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
@@
 or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
-Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
+Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
 For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris, Silk, Steinberg, Myers, \& Robinson, 2007).
@@
 If we take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
-Applied to FAIR theory, reusability requires that each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1).
+Applied to FAIR theory, reusability requires that each theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and Accessible license for reuse (R1.1).
 It should further have detailed provenance (R1.2),
@@
 and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
-
-If we consider the current state of Reusability in psychological theory, there appears to be a norm \emph{against} theory reuse:
+\href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{This is an example of FAIR metadata} extracted from Zenodo.
+
+If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
 \emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
@@
 and ``Who determines how a theory may be reused or changed?''.
-These questions imply a norm against modifying theory without consent from the author.
-\textless!-- NvD: I suggest removing the rest of the sentence (as a result of an earlier suggestion).
--- \textgreater{} reminiscent of the aforementioned problem of dependency on the author.
-
-Licensing theories for reuse unambiguously answers these questions.
-With the caveat that legislation may vary across contexts and jurisdictions, the following should not be interpreted as legal advice.
-In determining what license is appropriate for theory,
-a key consideration is that copyright law protects authors' rights according to the \emph{idea-expression dichotomy} (Bently, Davis, \& Ginsburg, 2010).
+These questions imply a norm against modifying theory without its author's consent, reminiscent of the aforementioned problem of dependency on the author.
+
+Licensing theories for reuse unambiguously answers these questions,
+with the caveats that legislation may vary across contexts and jurisdictions, and that this paper does not constitute legal advice.
+Two considerations are important when determining what license is appropriate for theory.
+A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy (Bently, Davis, \& Ginsburg, 2010).
 It explicitly does not
 \emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''}.
-Copyright thus extends to creative works expressing a theory (e.g., writing, visual illustrations),
+Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
 but not to the underlying theoretical idea.
 It thus seems that theories expressed in prose or depicted visually - in other words, that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
-Another important consideration is that academic research is covered under ``fair use'' exemptions to copyright.
-Given that copyright does not cover ideas in their purest form (like formal theories), and that academic use is likely exempted from its protection,
-in many cases, it may be sensible to explicitly license theories in a way that encourages Reusability.
-The CC0 (no rights reserved) license does this by explicitly waiving all rights and encouraging reuse.
+A second consideration is that academic research is covered under ``fair use'' exemptions to copyright.
+Given these two considerations - that copyright does not protect ideas in their purest form and that academic use offers exemptions to copyright - it may be counterproductive and possibly misleading to adopt a license that assumes copyright protection to theories.
+For psychological theories without commercial aspects, we suggest using a licence that explicitly waives copyright and encourages Reusability,
+such as CC0 (no rights reserved).
 
@@
 For example, while a CC0 license does not legally mandate attribution,
-the norms of good scientific practice require that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
-Another way to guide the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
-It is also possible to create or adopt a ``Code of Conduct'' which prescribes behavioral norms for contributors and users of a theory.
+the norms of good scientific practice mandate that scholars comprehensively cite theory and related works (Aalbersberg et al., 2018).
+Particularly when FAIRifying an existing theory, failing to credit its author amounts to scientific malpractice.
+Another instrument for guiding the social process of (diffuse) collaboration is to include a ``README'' file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
+A final suggestion is to create or adopt a ``Code of Conduct'' which prescribes behavioral norms for contributors and users of a theory (Ehmke, 2014).
 
@@
 
-Open science infrastructure is an area of active development, and as such,
-the approach proposed here should not be considered definitive,
-but rather, as one proposal for a FAIR-compliant implementation of theory.
-The guiding principle of our implementation is to align and build upon existing successful open science infrastructures to the maximum possible extent.
+To concretize the FAIR principles,
+we propose an applied workflow for making theory FAIR.
+The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
 At the time of writing (2024),
-the value of using Git for version control of academic research is well-established,
-and the integration of GitHub and Zenodo makes for a particularly user-friendly approach.
-Zenodo and GitHub are also integrated with the Open Science Framework (OSF),
+the value of using Git for version control of academic research is well-established (Ram, 2013),
+and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets \emph{all} of the FAIR principles (\href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
+Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
 a popular platform in psychology.
-Creating a front page on the OSF increases the visibility of a FAIR theory,
-while the integration with Zenodo and GitHub removes the need for uploading and maintaining the same information on multiple platforms.
-While we make use of specific open science infrastructures, it is important to stress that our workflow illustrates general principles which can also be implemented using other open science infrastructures.
-The process described here can be largely automated in R using the \texttt{theorytools} package; see the package vignette on FAIR theory, \texttt{vignette("fairtheory",\ package\ =\ "theorytools")}.
+Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory amongst users of that platform,
+while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
+Note that open science infrastructure is an area of active development, and as such,
+the approach proposed here might change.
+Our workflow can be largely automated in R using the \texttt{theorytools} package.
+To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
+It can be accessed by running \texttt{vignette("fairtheory",\ package\ =\ "theorytools")} in R.
+We present a brief summary of the instructions at the time of writing here,
+to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
 
@@
 The empirical cycle, described on page 28 of De Groot and Spiekerman (1969),
-is a kind of meta-theory of theory construction.
+is a meta-theory of theory construction.
 The resulting FAIR implementation of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
@@
 it is defined at either the ``theory'' or ``specification'' level.
-We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of Table \ref{tab:tabfair}):
+We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
+The implementation below describes the model as a directed graph.
+Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
+and the second half describes their proposed interrelations.
+This follows the first two properties of good theory according to Meehl (Meehl, 1990).
+We can save this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
 
@@
 \end{verbatim}
-
-This implementation describes the model as a directed graph.
-Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
-and the second half describes their proposed interrelations.
-This follows the first two properties of good theory according to Meehl (Paul E. Meehl, 1990).
-We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
 
@@
 
-The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
+The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved.
 In particular, the practice of version control offers extensive benefits for scientific work (Ram, 2013; Van Lissa et al., 2021).
@@
 
-The process of archiving a GitHub repository on Zenodo is documented in a vignette in the \texttt{theorytools} R-package, so that it can be kept up-to-date.
-We present a brief summary of the instructions at the time of writing here.
 First, create a Zenodo account with your existing GitHub account.
@@
 To document our archive as a FAIR theory requires adding some extra information on Zenodo.
-Supplying the following information helps improve the Findability of a theory:
+Supplying the information below helps improve the Findability of a theory.
+See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the resulting FAIR metadata} archived in Zenodo.
 
@@
 \end{itemize}
-
-\subsection{6. Making Changes}\label{making-changes}
 
@@
 
-Note that this function also automatically provides basic FAIR theory metadata to Zenodo.
-
-\subsection{Forking Different Implementations of a Theory}\label{forking-different-implementations-of-a-theory}
+\subsection{Updating a Theory}\label{updating-a-theory}
 
@@
 but not all of them have interpreted his work the same.
-For example, Wagenmakers and colleagues (\textbf{wagenmakersCreativityVerificationCyclePsychological2018?}) write \emph{``De Groot's ``empirical cycle,'' shown here in Figure 6''} - but Figure 6 diverges substantially from De Groot's description, and from our implementation of it.
+For example, Wagenmakers and colleagues' (2018) interpretation of the empirical cycle diverges substantially from De Groot's description.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-This is achieved by \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``forking the repository''},
-\href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``cloning''} the forked repository to our local computer,
-making any changes we want, and then completing steps 4-5 of ``Making a Theory FAIR''.
-
-We have implemented Wagenmakers and colleagues' version as a DOT graph
+Wagenmakers and colleagues' version can also be implemented as a DOT graph
 to illustrate some clear deviations from the original.
-First, the phases of the cycle have been renamed.
-While this change was not described in the paper,
-we assumed that the labels are meant to illustrate the phases, not substantially change the ontology.
-We represent this change by adding labels to the original DOT graph.
-Note, however, that the labels suggest a focus on empirical psychology that was absent in the original formulation, which was more general.
+Notice that, first, the phases of the cycle have been renamed.
+This change was not described in the paper.
+If we assume that the labels are meant to illustrate the phases,
+not substantially change the ontology,
+then we can represent it by adding labels to the original DOT graph.
+The labels do suggest a focus on empirical psychology not present in De Groot's version.
 Furthermore, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
 As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-Second, the authors mention an explicit change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}.
+The authors explicitly mention a second change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}.
 The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
@@
 The first author was inspired by De Groot as well,
-but they specify the empirical cycle differently.
+but again specified the empirical cycle differently.
 First, notice that the nodes in De Groot's formulation mostly refer to processes.
 This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
-In our implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
+In the implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
 Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
-For example, many have argued that observation is theory-laden, and as such, involves induction (\textbf{brewerTheoryLadennessObservationTheoryLadenness2001?}).
+Many have argued that observation is theory-laden, and as such, involves induction (Brewer \& Lambert, 2001).
 Deriving hypotheses from theory is also not purely deductive,
 as auxiliary assumptions are often made (which, again, involves induction).
-Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (\textbf{peikertTransparencyOpenScience2023?}).
+Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (Peikert, 2023).
 With these alterations, we implement the empirical cycle as follows:
@@
 
-\subsection{Using FAIR theory to Perform Causal Inference}\label{using-fair-theory-to-perform-causal-inference}
-
-Some have argued that \emph{causal explanations} are a property of good theory (Paul E. Meehl, 1990).
-According to Pearl and colleagues,
-explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data (\textbf{pearlCausalDiagramsEmpirical1995?}).
-Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference,
-and could even be integrated into the analysis environment.
-
-In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG.
-We could use that to illustrate updating FAIR theory:
-
-\url{https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45}
-
-We can find more examples of causal inference with DAGs in these tutorials:
-
-\url{https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/}
-
-\url{https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/}
+The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
+If you start with an existing GitHub repository, and wish to make some changes to it,
+this is commonly done by \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository}{creating a ``branch''}.
+A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
+Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
+Figure \ref{fig:figbranch} shows how our example \texttt{empirical\_cycle} repository contains branches with Wagenmakers' and Van Lissa's implementations.
+A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
+If you wish to develop a new version of someone else's FAIR theory,
+it is possible to \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``fork their repository''}.
+This creates a copy of their repository onto your GitHub account.
+Both branches and forks can be compared and merged via ``pull requests'', which are essentially a request to incorporate the changes you have made.
+Figure \ref{fig:figcompare} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+
+\begin{figure}
+\includegraphics[width=0.5\linewidth]{git_branches} \caption{Branches can contain different versions of a theory.}\label{fig:figbranch}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\linewidth]{git_compare} \caption{GitHub's compare functionality.}\label{fig:figcompare}
+\end{figure}
+
+\subsection{Further Uses of FAIR Theory}\label{further-uses-of-fair-theory}
+
+As uses of FAIR theory are best illustrated using tutorial examples,
+the \texttt{theorytools} package contains several vignettes that showcase specific applications.
+At the time of writing, the package includes a \href{}{vignette on theory specification},
+that is to say, moving from a theory represented as prose to a FAIR theory,
+based on the many-theorists project by Glöckner and Fiedler (in preparation).
+Another vignette illustrates the use of FAIR theory for causal inference (Pearl, 1995).
+Furthermore, one vignette illustrates the use of FAIR theory for \href{}{conducting a simulation study}.
+More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
 
@@
 
-The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
-Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited interoperability, and are rarely reused.
+The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
+Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
 These limitations impede cumulative knowledge production in our field,
@@
 Modular publishing capitalizes on these changing metrics,
-and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory (\textbf{kirczModularityNextForm1998?}).
+and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory (Kircz, 1998).
 Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
@@
 
-Interdisciplinarity benefits from FAIR theory's accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
+Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
@@
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis (\textbf{bringmannBackBasicsImportance2022?}) and lack of standardized ontologies for psychological constructs (Bosco et al., 2017).
+Another limitation of scope is that FAIR theory does not resolve other related issues in psychology, such as the measurement crisis (Bringmann, Elmer, \& Eronen, 2022) and lack of standardized ontologies for psychological constructs (Bosco et al., 2017).
 However, our work here provides a template for addressing such challenges,
@@
 
-One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'': \textless!-- NvD: I agree.
-Also, if we mention it here, we should definitely not mention it under limitations -\textgreater{}
-it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties {[}REF{]}.
-Additionally, the ``jingle-jangle'' fallacy is prevalent in the social sciences:
+One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
+Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
+it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties (Bringmann et al., 2022).
+Additionally, the ``jingle-jangle'' fallacy is prevalent in psychology:
 the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
 FAIR theory can help address the measurement crisis:
-since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
-
-FAIR theory incorporates theory into open science workflows,
-facilitates scholarly communication about theories,
-making it easier to share theories with less opportunity for ambiguity and misunderstanding.
-FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories.
-More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
-removes barriers to knowledge exchange with the global scholarly community,
-opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
+since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.
 
@@
 
+\bibitem[\citeproctext]{ref-brewerTheoryLadennessObservationTheoryLadenness2001}
+Brewer, W.
+F., \& Lambert, B.
+L.
+(2001).
+The {Theory-Ladenness} of {Observation} and the {Theory-Ladenness} of the {Rest} of the {Scientific Process}.
+\emph{Philosophy of Science}, \emph{68}(S3), S176--S186.
+\url{https://doi.org/10.1086/392907}
+
+\bibitem[\citeproctext]{ref-bringmannBackBasicsImportance2022}
+Bringmann, L.
+F., Elmer, T., \& Eronen, M.
+I.
+(2022).
+Back to {Basics}: {The Importance} of {Conceptual Clarification} in {Psychological Science}.
+\emph{Current Directions in Psychological Science}, 09637214221096485.
+\url{https://doi.org/10.1177/09637214221096485}
+
 \bibitem[\citeproctext]{ref-cinelliCrashCourseGood2022}
@@
 
-\bibitem[\citeproctext]{ref-ContributingCitationsReferences}
-Contributing {Citations} and {References}. (n.d.). Retrieved December 5, 2024, from \url{https://support.datacite.org/docs/data-citation}
+\bibitem[\citeproctext]{ref-ContributingCitationsReferences2024}
+Contributing {Citations} and {References}. (2024). Retrieved December 5, 2024, from \url{https://support.datacite.org/docs/data-citation}
+
+\bibitem[\citeproctext]{ref-cramerMajorDepressionComplex2016}
+Cramer, A.
+O.
+J., Borkulo, C.
+D.
+van, Giltay, E.
+J., Maas, H.
+L.
+J.
+van der, Kendler, K.
+S., Scheffer, M., \& Borsboom, D.
+(2016).
+Major {Depression} as a {Complex Dynamic System}.
+\emph{PLOS ONE}, \emph{11}(12), e0167490.
+\url{https://doi.org/10.1371/journal.pone.0167490}
+
+\bibitem[\citeproctext]{ref-datacitemetadataworkinggroupDataCiteMetadataSchema2024}
+DataCite Metadata Working Group.
+(2024).
+\emph{{DataCite Metadata Schema Documentation} for the {Publication} and {Citation} of {Research Data} and {Other Research Outputs} v4.6}.
+\url{https://doi.org/10.14454/MZV1-5B55}
 
@@
 
+\bibitem[\citeproctext]{ref-deciEffectsExternallyMediated1971}
+Deci, E. L. (1971). Effects of externally mediated rewards on intrinsic motivation. \emph{Journal of Personality and Social Psychology}, \emph{18}(1), 105--115. \url{https://doi.org/10.1037/h0030644}
+
+\bibitem[\citeproctext]{ref-DOTLanguage2024}
+{DOT Language}. (2024). Retrieved December 5, 2024, from \url{https://graphviz.org/doc/info/lang.html}
+
 \bibitem[\citeproctext]{ref-dumas-malletPoorReplicationValidity2017}
@@
 \url{https://doi.org/10.1371/journal.pone.0172650}
+
+\bibitem[\citeproctext]{ref-ehmkeContributorCovenantCode2014}
+Ehmke, C. (2014). Contributor {Covenant}: {A Code} of {Conduct} for {Open Source} and {Other Digital Commons Communities}. Retrieved April 3, 2024, from \url{https://www.contributor-covenant.org/}
 
@@
 \url{https://doi.org/10.1177/0959354318761606}
+
+\bibitem[\citeproctext]{ref-katzSpecialIssueSoftware2024}
+Katz, D. S., \& Chue Hong, N. P. (2024). Special issue on software citation, indexing, and discoverability. \emph{PeerJ Computer Science}, \emph{10}, e1951. \url{https://doi.org/10.7717/peerj-cs.1951}
+
+\bibitem[\citeproctext]{ref-kirczModularityNextForm1998}
+Kircz, J. G. (1998). Modularity: The next form of scientific information presentation? \emph{Journal of Documentation}, \emph{54}(2), 210--235. \url{https://doi.org/10.1108/EUM0000000007185}
 
@@
 \bibitem[\citeproctext]{ref-meehlTheoreticalRisksTabular1978}
-Meehl, Paul E.
+Meehl, P.
+E.
 (1978).
@@
 \bibitem[\citeproctext]{ref-meehlAppraisingAmendingTheories1990}
-Meehl, Paul E.
+Meehl, P.
+E.
 (1990).
@@
 
+\bibitem[\citeproctext]{ref-pearlCausalDiagramsEmpirical1995}
+Pearl, J. (1995). Causal {Diagrams} for {Empirical Research}. \emph{Biometrika}, \emph{82}(4), 669--688. \url{https://doi.org/10.2307/2337329}
+
+\bibitem[\citeproctext]{ref-peikertTransparencyOpenScience2023}
+Peikert, A. (2023). \emph{Towards {Transparency} and {Open Science}} (doctoralThesis, Humboldt-Universität zu Berlin). Humboldt-Universität zu Berlin. \url{https://doi.org/10.18452/27056}
+
 \bibitem[\citeproctext]{ref-peikertReproducibleResearchTutorial2021}
@@
 \url{https://doi.org/10.3390/psych3040053}
+
+\bibitem[\citeproctext]{ref-peirceCollectedPapersCharles1960}
+Peirce, C. (1960). \emph{Collected {Papers Of Charles Peirce}}. Retrieved from \url{http://archive.org/details/collected-papers-of-charles-peirce}
+
+\bibitem[\citeproctext]{ref-popperLogicScientificDiscovery2002}
+Popper, K. R. (2002). \emph{The logic of scientific discovery}. Retrieved from \url{http://public.ebookcentral.proquest.com/choice/publicfullrecord.aspx?p=254228}
 
@@
 
+\bibitem[\citeproctext]{ref-ryanSelfdeterminationTheoryFacilitation2000}
+Ryan, R.
+M., \& Deci, E.
+L.
+(2000).
+Self-determination theory and the facilitation of intrinsic motivation, social development, and well-being.
+\emph{American Psychologist}, \emph{55}(1), 68--78.
+\url{https://doi.org/10.1037/0003-066X.55.1.68}
+
 \bibitem[\citeproctext]{ref-scheelWhyMostPsychological2022}
@@
 Taylor, S. (2022). The psychology of pandemics. \emph{Annual Review of Clinical Psychology}, \emph{18}(1), 581--609.
+
+\bibitem[\citeproctext]{ref-vandesompelRethinkingScholarlyCommunication2004}
+Van De Sompel, H., Payette, S., Erickson, J., Lagoze, C., \& Warner, S.
+(2004).
+Rethinking {Scholarly Communication}: {Building} the {System} that {Scholars Deserve}.
+\emph{D-Lib Magazine}, \emph{10}(9).
+\url{https://doi.org/10.1045/september2004-vandesompel}
+
+\bibitem[\citeproctext]{ref-vandermaasDynamicalModelGeneral2006}
+Van Der Maas, H.
+L.
+J., Dolan, C.
+V., Grasman, R.
+P.
+P.
+P., Wicherts, J.
+M., Huizenga, H.
+M., \& Raijmakers, M.
+E.
+J.
+(2006).
+A dynamical model of general intelligence: {The} positive manifold of intelligence by mutualism.
+\emph{Psychological Review}, \emph{113}(4), 842--861.
+\url{https://doi.org/10.1037/0033-295X.113.4.842}
 
@@
 
+\bibitem[\citeproctext]{ref-wagenmakersCreativityVerificationCyclePsychological2018}
+Wagenmakers, E.-J., Dutilh, G., \& Sarafoglou, A.
+(2018).
+The {Creativity-Verification Cycle} in {Psychological Science}: {New Methods} to {Combat Old Idols}.
+\emph{Perspectives on Psychological Science}, \emph{13}(4), 418--427.
+\url{https://doi.org/10.1177/1745691618771357}
+
+\bibitem[\citeproctext]{ref-wilkinsonFAIRGuidingPrinciples2016}
+Wilkinson, M.
+D., Dumontier, M., Aalbersberg, Ij.
+J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
+(2016a).
+The {FAIR Guiding Principles} for scientific data management and stewardship.
+\emph{Scientific Data}, \emph{3}(1, 1), 1--9.
+\url{https://doi.org/10.1038/sdata.2016.18}
+
 \bibitem[\citeproctext]{ref-wilkinsonFAIRGuidingPrinciples2016a}
@@
 J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
-(2016).
+(2016b).
 The {FAIR Guiding Principles} for scientific data management and stewardship.
@@
 R., Aloqalaa, M., Belhajjame, K., Crusoe, M.
-R., Paula Kinoshita, B.
-de, Gadelha, L., \ldots{} Goble, C.
+R., Kinoshita, B.
+de P., Gadelha, L., et al.others.
 (2024).
-\emph{Applying the FAIR principles to computational workflows}.
+\emph{Applying the {FAIR} principles to computational workflows}.
 Retrieved from \url{https://arxiv.org/abs/2410.03490}
@@
 N.
-Zalta (Ed.), \emph{The {Stanford} encyclopedia of philosophy} ({S}pring 2021).
-\url{https://plato.stanford.edu/archives/spr2021/entries/structure-scientific-theories/}; Metaphysics Research Lab, Stanford University.
+Zalta (Ed.), \emph{The {Stanford} encyclopedia of philosophy} (Spring 2021).
+Metaphysics Research Lab, Stanford University.
+Retrieved from \url{https://plato.stanford.edu/archives/spr2021/entries/structure-scientific-theories/}
 

### 2025-02-23 synthesize_data.R

@@ -1,0 +1,119 @@
+# txt <- readClipboard()
+c("Variable\t(Score Range) Age n M SD ADV1 INV1 SEF1 GPA1 GPA2 GPA3 CON1 CON2 CON3 EMO1 EMO2 EMO3 POP1 POP2 POP3", 
+  "Immigrant (0 = no, 1 = yes)\t13 IMM1 1057 0.50 0.50 .27\t2.40\t2.15\t2.42\t2.41\t2.39\t2.14\t2.17\t2.02\t2.01\t2.01\t2.07\t2.13\t2.11\t2.04", 
+  "Adversity (0–4) 13 ADV1 1017 0.83 0.80\t2.21\t2.15\t2.29\t2.24\t2.24\t2.07\t2.04 .04 .02 .01 .04\t2.08\t2.03\t2.01", 
+  "Involvement (1–5) 13 INV1 889 3.38 1.15 .17 .54 .48 .47 .27 .13 .03 .05\t2.01 .00 .16 .12 .07", 
+  "Self-efficacy (1–7) 13 SEF1 967 5.43 0.87 .33 .26 .22 .16 .15\t2.01 .19 .18 .19 .13 .08 .09", 
+  "Academic (1–20) 13 GPA1 843 13.7 3.02 .89 .84 .39 .29 .19 .09 .06 .03 .34 .21 .15", 
+  "Academic (1–20) 14 GPA2 748 13.3 2.97 .89 .34 .35 .18 .09 .00\t2.01 .26 .18 .11", 
+  "Academic (1–20) 15 GPA3 620 13.4 3.02 .31 .38 .22 .09 .00\t2.02 .22 .12 .08", 
+  "Conduct (1–5) 13 CON1 1026 4.59 0.72 .43 .32\t2.01 .00\t2.06 .13 .06\t2.01", 
+  "Conduct (1–5) 14 CON2 546 4.45 0.76 .44 .01 .00\t2.06\t2.05\t2.02\t2.04", 
+  "Conduct (1–5) 15 CON3 522 4.37 0.76\t2.07\t2.08\t2.06 .00\t2.01\t2.03", 
+  "Emotional (0–2) 13 EMO1 965 1.45 0.46 .49 .37 .06 .06 .06", 
+  "Emotional (0–2) 14 EMO2 730 1.54 0.44 .59 .10 .11 .12", "Emotional (0–2) 15 EMO3 589 1.46 0.51 .05 .08 .04", 
[…]

### 2025-02-25 manuscript.Rmd
update 2025-02-25 08:07:30.666758
Update manuscript.Rmd

@@
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
+This paper argues that the FAIR principles can also promote effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
@@
 Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
-These various practices ensure transparent and repeated testing of hypotheses.
+These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
@@
 The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
-Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
-Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
+Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
+Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
 
@@
 
-```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle".'}
+```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologieGrondslagenVan1961].'}
 knitr::include_graphics("ec_visio.pdf")

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

@@
 Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
+in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
 Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
@@
 *Accessible* in a machine- and human-readable filetype;
-*Interoperable* for specific purposed, for example, within the data analysis environment;
+*Interoperable* for specific purposes, for example, within the data analysis environment;
 and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
@@
 -->
-Definitions of theory abound and are the subject of extensive scholarly debate.
+Definitions of theory are abundant, and are the subject of extensive scholarly debate.
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-the method is not contained to any one particular definition.
-Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
+our suggested approach is not limited to any one particular definition.
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
 The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
 acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
+We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
 statements about the types of entities postulated (i.e., ontology),
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
-
 The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
@@
 Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
-Examples include experimental designs (e.g.
-longitudinal designs observing change over time), measurement tools (e.g.
-different questionnaires used to assess the same construct), or study subjects (e.g.
-specific strains of rats).
+Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).
 
@@
 -->
-Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
+More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should be observed and what not if the theory were true.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
 Without committing to specific implementations a-priori,
@@
 rather, it increases the fidelity and ease with which they are communicated.
-Thus, FAIR theory does not require theories to be formal,
-and formal theory can be represented in a way that is not FAIR.
+Thus, FAIR theory does not require theories to be formal.
+At the same time, not every formal theory is FAIR.
 The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

@@
 can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
-Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
+Conversely, a formal theory is not FAIR if it is confined to a publication in a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
@@
 
-```
-Given a version number MAJOR.MINOR.PATCH, increment the:
-
-MAJOR version when you make backwards incompatible changes, i.e., the
+Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:
+
+- MAJOR version when you commit backwards incompatible changes, i.e., the
       theory now contains empirical statements that are at odds with
-      a previous version
-MINOR version when you expand the set of empirical statements in a
+      a previous version of the theory
+- MINOR version when you expand the set of empirical statements in a
       backward compatible manner (i.e., the previous version is
       subsumed within the new version)
-PATCH version when you make backward compatible bug fixes, cosmetic
+- PATCH version when you make backward compatible bug fixes, cosmetic
       changes, fix spelling errors, or add clarifications
-```
 

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

@@
 With regard to the state of Interoperability in psychology, 
-Kurt Lewin's -@lewinPsychologyProcessGroup1943 adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
+Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

@@
 Note that open science infrastructure is an area of active development, and as such,
-the approach proposed here might change.
-Our workflow can be largely automated in R using the `theorytools` package.
+the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
+Our suggested workflow can be largely automated in R using the `theorytools` package but note that our workflow is not contingent upon researchers using R at all.
 To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
@@
 - Add the *keyword* `fairtheory`; this aids search engine indexation
-- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building
+- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
 - List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
@@
 ```
+The first two lines install and load the `theorytools` package.
+Lines 4--6 use the `worcs` package to check whether git for version control is installed and a connection to github can be established to publish the theory.
+Line 8 calls the main entry function `fair_theory` for a local repository in a specified local `path` with a given `title` and the aforementioned dot `theory_file` containing the theory, uploading it to the user's remote github repository with name given in `remote_repo` and a given license specified in `add_license`.
 

### 2025-02-25 manuscript.Rmd
Update manuscript.Rmd

@@
 FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
-This paves the way for more theory-driven scholarship,
-and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
+This paves the way for more theory-driven scholarship
+and acceleration of cumulative knowledge acquisition in psychology, the social sciences, and beyond.
 

### 2025-02-26 manuscript.Rmd
add comments and changes

--- 2025-02-25 manuscript.Rmd
@@
       - Writing – review & editing
+  - name: "Maximilian S. Ernst"
+    affiliation: "2,7"
+    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
+      - Writing – review & editing
+      - Investigation
 affiliation:
@@
     institution   : "University of Amsterdam, the Netherlands"
+  - id            : "7"
+    institution   : "Max Planck School of Cognition, Leipzig, Germany"
 authornote: |
@@
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
-This paper argues that the FAIR principles can also promote effective and transparent scholarly communication about theory.
+This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
@@
 accelerate cumulative knowledge acquisition.
-We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.
+We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.
 
@@
 Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
-These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
-However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
+These various practices ensure transparent and repeated testing of hypotheses.
+However, recent reviews show that most preregistered hypotheses tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
@@
 as the theory's central claims evade falsification.
-A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
-Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
-Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
-The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
-Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
-Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
 
@@
 Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One main concern is that theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+One main concern is that theories lack or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
 <!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning).
@@
 -->
-When theories do not make precise predictions,
-they are hard to falsify and difficult to understand on their own,
-without either substantial interpretation or additional background knowledge.
+When theories are difficult to understand on their own,
+without either substantial interpretation or additional background knowledge,
+it becomes difficult do derive precise predictions,
+and therefore hard to falsify the theory. 
 A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
@@
 
-```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologieGrondslagenVan1961].'}
+```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle".'}
 knitr::include_graphics("ec_visio.pdf")
@@
 Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
+in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
 Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
@@
 *Accessible* in a machine- and human-readable filetype;
-*Interoperable* for specific purposes, for example, within the data analysis environment;
+*Interoperable* for specific purposed, for example, within the data analysis environment;
 and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
@@
 -->
-Definitions of theory are abundant, and are the subject of extensive scholarly debate.
+Definitions of theory abound and are the subject of extensive scholarly debate.
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-our suggested approach is not limited to any one particular definition.
+the method is not contained to any one particular definition.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
@@
 acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
+We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
 statements about the types of entities postulated (i.e., ontology),
@@
 and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
+
 The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
@@
 Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
-Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).
+Examples include experimental designs (e.g.
+longitudinal designs observing change over time), measurement tools (e.g.
+different questionnaires used to assess the same construct), or study subjects (e.g.
+specific strains of rats).
 
@@
 -->
-More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should be observed and what not if the theory were true.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
+Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
 Without committing to specific implementations a-priori,
@@
 rather, it increases the fidelity and ease with which they are communicated.
-Thus, FAIR theory does not require theories to be formal.
-At the same time, not every formal theory is FAIR.
+Thus, FAIR theory does not require theories to be formal,
+and formal theory can be represented in a way that is not FAIR.
 The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
@@
 can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
-Conversely, a formal theory is not FAIR if it is confined to a publication in a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
+Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
@@
 
-Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:
-
-- MAJOR version when you commit backwards incompatible changes, i.e., the
+```
+Given a version number MAJOR.MINOR.PATCH, increment the:
+
+MAJOR version when you make backwards incompatible changes, i.e., the
       theory now contains empirical statements that are at odds with
-      a previous version of the theory
-- MINOR version when you expand the set of empirical statements in a
+      a previous version
+MINOR version when you expand the set of empirical statements in a
       backward compatible manner (i.e., the previous version is
       subsumed within the new version)
-- PATCH version when you make backward compatible bug fixes, cosmetic
+PATCH version when you make backward compatible bug fixes, cosmetic
       changes, fix spelling errors, or add clarifications
+```
 
@@
 In the longer term, it may not be necessary to write a paper for each theory.
-If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
+If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
 Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
@@
 Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
+<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
@@
 With regard to the state of Interoperability in psychology, 
-Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
+Kurt Lewin's -@lewinPsychologyProcessGroup1943 adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
@@
 We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
+- Optionally, add *References* to related works in plain text. For example, here we can provide the full citation of De Groot and Spiekerman:
+    > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: Foundations of inference and research in the behavioral sciences. De Gruyter Mouton. https://doi.org/10.1515/9783112313121
 
@@
 it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
+<!-- I think it would be nice if links are colored (blue), so people can see in the PDF that there is something to click. -->
 This creates a copy of their repository onto your GitHub account.
@@
 This paves the way for more theory-driven scholarship,
-and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
+accelerating cumulative knowledge acquisition in psychology, other social sciences, and beyond.
 

### 2025-02-26 manuscript.Rmd
Update manuscript.Rmd

@@
     affiliation: "2,7"
-    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
+    role:
       - Writing – review & editing

### 2025-02-26 dk_example/example.tex
add to dk aexample
finalize DK example draft

@@
 \usepackage{bbm}
+\usepackage{float}
 
@@
 
-\section{Relationships}
+\section{Relationships between constructs}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
   \vspace{1cm}
@@
 
-With $-$ signifying a negative association and $\simeq$ signifying measured by.
+With $-$ signifying a negative association and $\simeq$ signifying ``measured by''.
 
@@
 \end{itemize}
+
+\subsection{Formulation using measurement error}
+
+\begin{figure}[H]
+  \vspace{1cm}
+  \begin{center}
+  \input{figures/error_dk.tex}
+  \end{center}
+  \caption{A caption}
+  \label{fig:typetree}
+\end{figure}
+
+With $-$ signifying a negative association, $\simeq$ signifying ``measured by'' and $\coloneq$ signifying ``defined as''.
 
@@
 \noindent From eq. \ref{oc} and \ref{ep} it follows that $p_e = s^*_s$ and further from eq. \ref{ep} and \ref{p} we see
-\begin{align}
+\begin{align} \label{dd}
   oe &= p_e - p \\
@@
 \end{align}
+
 \section{(Statistical) Models}
+Using a linear regression model, the Dunning-Kruger effect can be stated as
+\begin{equation}
+ oc^* = \alpha + \beta_1 s^* + u
+\end{equation}
+with $\beta_1 < 0$.
+Substituting the obervable variables and rearranging according to eq. \ref{p} and \ref{dd}:
+\begin{equation}
+  oe = \alpha + \beta_1 p + u - \epsilon(1 + \beta_1)
+\end{equation}
 
@@
 \begin{itemize}
- \item use different test performances to calculate skill and overconfidence, but a small test-retest correlation can bias the estimate towards zero then (Krueger and Mueller, 2002)
+ \item Bias correction: use a bias correction formula that takes into account the correlation between performance and the error term
+ \item IV approach: measure performance on a second test ($p_2$) and compute $\beta_1 = \frac{\cov(oe, p_2)}{\cov(p, p_2)}$.
 \end{itemize}
@@
 \section{Open Questions / Ambiguities}
+The paper states they assume performance expectations are measured before the test - the derivations differ substantially if they are measured after the test.
 

### 2025-02-26 dk_example/figures/error_dk.tex
finalize DK example draft

--- 2025-01-06 dk_example/figures/original_dk.tex
@@
 
-\node[manifest] at (6,-2) (estimate) {\texttt{overestimate performance}};
+\node[manifest] at (6,-2) (estimate) {\texttt{overestimation}};
+
+\node[manifest] at (0, -6) (skillerror) {\texttt{skill + error}};
+
+\node[manifest] at (6, -4) (pminusp) {\texttt{expected performance - performance}};
+
+\node[manifest] at (6, -6) (ocerror) {\texttt{overconfidence - error}};
 
@@
 \draw[-,>=stealth',semithick] (overconfidence) -- node [fill=white] {$\simeq$} (estimate);
+\draw[-,>=stealth',semithick] (skillerror) -- node [fill=white] {$-$} (ocerror);
+\draw[-,>=stealth',semithick] (performance) -- node [fill=white] {$\coloneq$} (skillerror);
+\draw[-,>=stealth',semithick] (estimate) -- node [fill=white] {$\coloneq$} (pminusp);
+\draw[-,>=stealth',semithick] (pminusp) -- node [fill=white] {$=$ (derived below)} (ocerror);
 

### 2025-02-27 manuscript.Rmd

@@
 classoption       : "man, noextraspace"
+urlcolor          : blue
 output            :
@@
 These various practices ensure transparent and repeated testing of hypotheses.
-However, recent reviews show that most preregistered hypotheses tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
+However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
@@
 Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One main concern is that theories lack or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
+One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
 <!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->

### 2025-02-27 manuscript.Rmd

@@
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
+This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
@@
 as the theory's central claims evade falsification.
+A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
+Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
+Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
+Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
 
@@
 -->
-Definitions of theory abound and are the subject of extensive scholarly debate.
+Definitions of theory are abundant, and are the subject of extensive scholarly debate.
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-the method is not contained to any one particular definition.
+our suggested approach is not limited to any one particular definition.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
@@
 acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's hierarchy of ever-more specific "statements" a theory might contain [-@meehlAppraisingAmendingTheories1990]:
+We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
 statements about the types of entities postulated (i.e., ontology),
@@
 -->
-Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
+More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed if the theory were true.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
 Without committing to specific implementations a-priori,
@@
 rather, it increases the fidelity and ease with which they are communicated.
-Thus, FAIR theory does not require theories to be formal,
-and formal theory can be represented in a way that is not FAIR.
+Thus, FAIR theory does not require theories to be formal.
+At the same time, formal theories are not automatically FAIR.
 The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
@@
 can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
-Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
+Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
@@
 FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
-This paves the way for more theory-driven scholarship,
-accelerating cumulative knowledge acquisition in psychology, other social sciences, and beyond.
+This paves the way for more theory-driven scholarship
+and accelerates cumulative knowledge acquisition in psychology, the social sciences, and beyond.
 

### 2025-03-02 manuscript.Rmd
Move paragraph about science communication to future directions, #65

--- 2025-02-26 manuscript.Rmd
@@
       - Writing – review & editing
-  - name: "Maximilian S. Ernst"
-    affiliation: "2,7"
-    role:
-      - Writing – review & editing
-      - Investigation
 affiliation:
@@
     institution   : "University of Amsterdam, the Netherlands"
-  - id            : "7"
-    institution   : "Max Planck School of Cognition, Leipzig, Germany"
 authornote: |
@@
 accelerate cumulative knowledge acquisition.
-We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.
+We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.
 
@@
 as the theory's central claims evade falsification.
+A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
+Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
+Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
+Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
 
@@
 Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One main concern is that theories lack or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
-<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
+One main concern is that theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
 <!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning).
@@
 -->
-When theories are difficult to understand on their own,
-without either substantial interpretation or additional background knowledge,
-it becomes difficult do derive precise predictions,
-and therefore hard to falsify the theory. 
+When theories do not make precise predictions,
+they are hard to falsify and difficult to understand on their own,
+without either substantial interpretation or additional background knowledge.
 A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
@@
 the method is not contained to any one particular definition.
-Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories]
 The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
@@
 In the longer term, it may not be necessary to write a paper for each theory.
-If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
+If Zenodo becomes more recognized as centralized repository for information artifacts, researchers may begin to search there more regularly.
 Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
@@
 Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
-<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
@@
 it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
-<!-- I think it would be nice if links are colored (blue), so people can see in the PDF that there is something to click. -->
 This creates a copy of their repository onto your GitHub account.
@@
 
-Finally, FAIR theory holds substantial potential for science communication.
-While the general public is ill-equipped to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon,
-theories encapsulate the bigger picture of contemporary scientific understanding.
-For example, while few people read empirical studies on attachment,
-attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
-Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
-By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
-
 ## Limitations
@@
 
+Another future direction for FAIR theory is as an instrument of science communication.
+Practitioners and the general public are rarely able to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon.
+Theories are more accessible, because they encapsulate the bigger picture of contemporary scientific understanding.
+For example, while few people read empirical studies on attachment,
+attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
+Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
+By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
+
 ## Conclusion
@@
 This paves the way for more theory-driven scholarship,
-accelerating cumulative knowledge acquisition in psychology, other social sciences, and beyond.
+and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
 

### 2025-03-02 manuscript.Rmd
Add lange reference, closes #63

@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs [REF LANGE CHECKLIST].
+and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs [@langeChecklistIncentivizingFacilitating2025a].
 

### 2025-03-02 manuscript.Rmd
Footnote Quine, closes #62

@@
 At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
-It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
-which holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
-has multiple alternative translations, with no *objective* means of choosing the correct one.
-It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
+It is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+has multiple alternative translations, with no objective means of choosing the correct one.],
+which implies that it is not possible to remove all ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
 This places a theoretical upper bound on theories' ability to be understood.

### 2025-03-02 manuscript.Rmd
cite zenodo closes #59

@@
 Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
-which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
+which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
 To maintain a persistent record of scholarly communication,

### 2025-03-02 manuscript.Rmd
remove vertical dividers, closes #58

@@
 ```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologieGrondslagenVan1961].'}
+# rsvg::rsvg_pdf("empirical_cycle1.svg", "ec_visio.pdf")
 knitr::include_graphics("ec_visio.pdf")

### 2025-03-02 manuscript.Rmd
I think this is a bit redundant, but here goes.. closes #66

@@
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
-The present paper considers the advantages of Git for FAIR theory.
-Git enables explicitly comparing versions of a file (or: theory),
+Git provides a useful framework for developing FAIR theory,
+because it enables explicitly comparing versions of a file (or: theory),
 incorporating changes by different authors,
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
+Note, however, that the function of Git(Hub) is *not* to archive a FAIR theory.
+FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
+Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.
 

### 2025-03-02 manuscript.Rmd
update 2025-03-02 14:23:45.942159

@@
 
-## Updating a Theory
+## Changing a Theory
 
@@
 but again specified the empirical cycle differently.
-First, notice that the nodes in De Groot's formulation mostly refer to processes.
-This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
-In the implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
-Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
+First, in De Groot's formulation, each stage describes a process.
+This invites the question of what the concrete outcomes of these processes are.
+In other words: what actually changes when going through the cycle, except the scholar's mind.
+To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs), panel (c).
+Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
 <!-- In our representation, -->
@@
 <!-- ``` -->
-Many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
-Deriving hypotheses from theory is also not purely deductive,
-as auxiliary assumptions are often made (which, again, involves induction).
+For example, if observation is theory-laden, it involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
+Deriving hypotheses from theory is not purely deductive either,
+as auxiliary assumptions are often made, which involves induction (e.g., assuming equal variances across groups because groups often have equal variances).
 Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
-With these alterations, we implement the empirical cycle as follows:
+These alterations result in the following implementation of the empirical cycle:
 

### 2025-03-02 manuscript.tex
update 2025-03-02 14:23:45.942159

@@
 \PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8881}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 9205}
 \usepackage{lineno}
@@
   pdftitle={To be FAIR: Theory Specification Needs an Update},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, Noah N.N. van Dongen6, \& Maximilian S. Ernst2,7},
   pdflang={en-EN},
   pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
-  hidelinks,
+  colorlinks=true,
+  linkcolor={Maroon},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
   pdfcreator={LaTeX via pandoc}}
@@
 Brandmaier\textsuperscript{2,3,4}, Felix D.
-Schönbrodt\textsuperscript{5}, \& Noah N.N.
-van Dongen\textsuperscript{6}}
+Schönbrodt\textsuperscript{5}, Noah N.N.
+van Dongen\textsuperscript{6}, \& Maximilian S.
+Ernst\textsuperscript{2,7}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# ecb21ba7.
+This is a preprint paper, generated from Git Commit \# b77adbc3.
 
@@
 Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N.
-van Dongen: Writing -- review \& editing.
+van Dongen: Writing -- review \& editing; Maximilian S.
+Ernst: Writing -- review \& editing, Investigation.
 
@@
 \affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands}
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands\\\textsuperscript{7} Max Planck School of Cognition, Leipzig, Germany}
 
@@
 To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts.
-FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved.
+FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved through collaborative efforts.
 This paper adapts the FAIR priciples for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
@@
 Wilkinson et al., 2024).
-This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
+This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
 To this end, we introduce ``FAIR theory'':
@@
 Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
-These various practices ensure transparent and repeated testing of hypotheses.
+These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
@@
 The profound difference between these definitions becomes clear when considering the act of changing a child's dirty diaper.
-Under the original definition, few would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience
-Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.
+Under the original definition, few caregivers would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience.
+Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.
 
 Scholars have been raising concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
-One main concern is that theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
-When theories do not make precise predictions,
-they are hard to falsify and difficult to understand on their own,
-without either substantial interpretation or additional background knowledge.
+One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
+When theories are difficult to understand on their own,
+without either substantial interpretation or additional background knowledge,
+it becomes difficult do derive precise predictions,
+and therefore hard to falsify the theory.
 A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
@@
 \includegraphics{ec_visio.pdf}
-\caption{\label{fig:figecs}Three implementations of the ``empirical cycle''.}
+\caption{\label{fig:figecs}Three implementations of the ``empirical cycle'' (de Groot, 1961).}
 \end{figure}
@@
 Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
+in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
 Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
@@
 \emph{Accessible} in a machine- and human-readable filetype;
-\emph{Interoperable} for specific purposed, for example, within the data analysis environment;
+\emph{Interoperable} for specific purposes, for example, within the data analysis environment;
 and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
@@
 
-Definitions of theory abound and are the subject of extensive scholarly debate.
+Definitions of theory are abundant, and are the subject of extensive scholarly debate.
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-the method is not contained to any one particular definition.
-Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
+our suggested approach is not limited to any one particular definition.
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021).
 The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
 acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's hierarchy of ever-more specific ``statements'' a theory might contain (1990):
+We recognize the syntactic view in Meehl's (1990) hierarchy of ever-more specific ``statements'' a theory might contain:
 statements about the types of entities postulated (i.e., ontology),
@@
 and statements about their specific numerical values (cf. Frankenhuis, Panchanathan, \& Smaldino, 2023; Guest, 2024).
-
 The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
@@
 Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
-Examples include experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).
+Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).
 
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-Greater formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
+More rigorous formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as precise statements, clearly demarcating what should (not) be observed if the theory were true.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
 Without committing to specific implementations a-priori,
@@
 can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
-Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
+Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
 FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).
@@
 Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
-which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
+which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
 To maintain a persistent record of scholarly communication,
@@
 
-\begin{verbatim}
-Given a version number MAJOR.MINOR.PATCH, increment the:
-
-MAJOR version when you make backwards incompatible changes, i.e., the
+Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:
+
+\begin{itemize}
+\tightlist
+\item
+  MAJOR version when you commit backwards incompatible changes, i.e., the
   theory now contains empirical statements that are at odds with
-      a previous version
+  a previous version of the theory
+\item
   MINOR version when you expand the set of empirical statements in a
@@
   subsumed within the new version)
+\item
   PATCH version when you make backward compatible bug fixes, cosmetic
   changes, fix spelling errors, or add clarifications
-\end{verbatim}
+\end{itemize}
 
@@
 At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
-It is important to acknowledge the \emph{indeterminacy of translation} (Quine, 1970):
-which holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
-has multiple alternative translations, with no \emph{objective} means of choosing the correct one.
-It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
+It is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+  has multiple alternative translations, with no objective means of choosing the correct one.},
+which implies that it is not possible to remove all ambiguity when communicating an idea (Quine, 1970).
 This places a theoretical upper bound on theories' ability to be understood.
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs {[}REF LANGE CHECKLIST{]}.
+and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
 
@@
 With regard to the state of Interoperability in psychology,
-Kurt Lewin's -Lewin (1943) adage ``there's nothing as practical as a good theory'' paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
+Kurt Lewin's (1943) adage ``there's nothing as practical as a good theory'' paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
 But, as we argued, this is not the case.
@@
 Note that open science infrastructure is an area of active development, and as such,
-the approach proposed here might change.
-Our workflow can be largely automated in R using the \texttt{theorytools} package.
+the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
+Our suggested workflow can be largely automated in R using the \texttt{theorytools} package but note that our workflow is not contingent upon researchers using R at all.
 To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
@@
 \item
-  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to contribute to community building
+  Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
 \item
@@
   \end{itemize}
+\item
+  Optionally, add \emph{References} to related works in plain text. For example, here we can provide the full citation of De Groot and Spiekerman:
+  \textgreater{} De Groot, A.
+D., \& Spiekerman, J.
+A.
+A.
+(1969).
+Methodology: Foundations of inference and research in the behavioral sciences.
+De Gruyter Mouton.
+\url{https://doi.org/10.1515/9783112313121}
 \end{itemize}
@@
 \end{verbatim}
+
+The first two lines install and load the \texttt{theorytools} package.
+Lines 4--6 use the \texttt{worcs} package to check whether git for version control is installed and a connection to github can be established to publish the theory.
+Line 8 calls the main entry function \texttt{fair\_theory} for a local repository in a specified local \texttt{path} with a given \texttt{title} and the aforementioned dot \texttt{theory\_file} containing the theory, uploading it to the user's remote github repository with name given in \texttt{remote\_repo} and a given license specified in \texttt{add\_license}.
 
@@
 We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing the barrier of entry for researchers,
-and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.
+and creating a FAIR resource for theory construction tools and documentation that can be updated as best practices continue to develop.
 
 Making theory FAIR allows researchers to more easily find a relevant framework;
-access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
+access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
 Whereas the idea of theory can be quite nebulous to empirical psychologists,
@@
 By providing much-needed open science methods for theory construction,
-FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the ``open empirical cycle'' {[}REF Hoijtink{]}.
+FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the ``open empirical cycle'' (Hoijtink et al., 2023).
 
@@
 The advantage of modular publishing is that authors can be credited for theory development.
-Given the current emphasis on empirical papers {[}REF{]}, theoretical papers can be hard to publish.
+Given the current emphasis on empirical papers (McPhetres et al., 2021), theoretical papers can be hard to publish.
 FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
 
-Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling {[}REF Oisin paper{]}.
+Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck, Ryan, Robinaugh, Waldorp, \& Borsboom, 2021).
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
 In terms of team science, FAIR theory facilitates collaboration by ensuring that all contributors have access to the same information and
-clarifying any remaining areas of contention or misunderstanding.
+clarifying any remaining areas of contention or misunderstanding (Van Lissa et al., 2024).
 Version control provides a framework to resolve parallel developments from multiple collaborators in a non-destructive manner.
@@
 and adversarial collaboration, where others strive to falsify a theory or identify its inconsistencies, and democratizes collaboration with as-of-yet unknown collaborators via platforms like GitHub, where researchers outside one's network can identify issues or suggest improvements to theories.
-
-Finally, FAIR theory plays an important role in science communication, because theory synthesizes contemporary scientific understanding about a phenomenon.
-Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
-By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
 
@@
 Another limitation is the learning curve associated with tools like Git and Zenodo.
-The \texttt{theorytools} R-package mitigates this limitation by automating key steps in the process.
+The \texttt{theorytools} R-package mitigates this limitation for R-users by automating key steps in the process.
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
@@
 Education might help address this limitation; with this in mind,
-we are developing open educational materials on theory development.
+we have shared several open educational materials on theory development in the ``FAIR Theory Community'' on Zenodo, and we encourage others to reuse these and share their materials.
 
@@
 it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties (Bringmann et al., 2022).
-Additionally, the ``jingle-jangle'' fallacy is prevalent in psychology:
+The ``jingle-jangle'' fallacy is common in psychology:
 the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
@@
 since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.
+
+Another future direction for FAIR theory is as an instrument of science communication.
+Practitioners and the general public are rarely able to read and derive actionable insight from large quantities of empirical papers about a particular phenomenon.
+Theories are more accessible, because they encapsulate the bigger picture of contemporary scientific understanding.
+For example, while few people read empirical studies on attachment,
+attachment theory plays a prominent role in popular scientific books about parenting and romantic relationships.
+Theory bridges the gap between academic research and practitioners by summarizing actionable insights, relieving practitioners from the need to sift through extensive empirical literature.
+By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
 
@@
 FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
-This paves the way for more theory-driven scholarship,
-and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
+This paves the way for more theory-driven scholarship
+and accelerates cumulative knowledge acquisition in psychology, the social sciences, and beyond.
 
@@
 
+\bibitem[\citeproctext]{ref-zenodo}
+European Organization For Nuclear Research, \& OpenAIRE. (2013). \emph{Zenodo}. CERN. \url{https://doi.org/10.25495/7GXK-RD71}
+
 \bibitem[\citeproctext]{ref-frankenhuisStrategicAmbiguitySocial2023}
@@
 \url{https://doi.org/10.1177/0959354318761606}
+
+\bibitem[\citeproctext]{ref-haslbeckModelingPsychopathologyData2021}
+Haslbeck, J.
+M.
+B., Ryan, O., Robinaugh, D.
+J., Waldorp, L.
+J., \& Borsboom, D.
+(2021).
+Modeling psychopathology: {From} data models to formal theories.
+\emph{Psychological Methods}.
+\url{https://doi.org/10.1037/met0000303}
+
+\bibitem[\citeproctext]{ref-hoijtinkOpenEmpiricalCycle2023}
+Hoijtink, H., Bruin, J.
+de, Duken, S.
+B., Flores, J., Frankenhuis, W., \& Lissa, C.
+J.
+van.
+(2023).
+\emph{The {Open Empirical Cycle} for {Hypothesis Evaluation} in {Psychology}}.
+\url{https://doi.org/10.31234/osf.io/wsxbh}
 
@@
 \url{https://doi.org/10.3233/DS-190026}
+
+\bibitem[\citeproctext]{ref-langeChecklistIncentivizingFacilitating2025a}
+Lange, J., Freyer, N., Musfeld, P., Schönbrodt, F., \& Leising, D. (2025, January 10). A checklist for incentivizing and facilitating good theory building. \url{https://doi.org/10.31219/osf.io/7qvfz}
 
@@
 
+\bibitem[\citeproctext]{ref-vanlissaVisionTeamScience2024}
+Van Lissa, C.
+J., Keymolen, E., Hoek, S.
+van den, Klingner, A., Schurman, L., \& Hunnik, M.
+van.
+(2024).
+\emph{Towards a {Vision} for {Team Science} at {Tilburg University}}.
+\url{https://doi.org/10.31234/osf.io/jsbuv}
+
 \bibitem[\citeproctext]{ref-vanrooijTheoryTestHow2021}

### 2025-03-02 manuscript.Rmd
use other example for SDT closes #56

@@
 Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
-The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
-Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
-Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
+Consider, for instance, a research scientist working in a highly competitive laboratory.
+Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
+Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness.
+This broader definition enables SDT to absorb a wide range of contradictory evidence—external circumstances that once would have undermined intrinsic motivation can now be reframed as fulfilling deeper psychological needs.
 

### 2025-03-03 manuscript.Rmd
Fix some typos

--- 2025-03-02 manuscript.Rmd
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processeses of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+The equally critical inductive processes of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
 Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
@@
 We do so by implementing theories as information artifacts,
-archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
 Metadata are "data about the data", they provide information about the nature and content of an information artifact.
@@
 Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
-FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
+FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).
 
@@
 Finally, metadata should be registered or indexed in a searchable registry (F4).
-It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
+It is important to note that, while many archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
 only few are specifically designed for FAIR-compliant archival.
@@
 Both approaches can be complementary, however.
-for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
+For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
 Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
@@
 the same image represented in the DOT language [@DOTLanguage2024]
-for representing graphs does meet this ideal.
+for representing graphs does meet this ideal (an example of such a DOT representation is given below).
 
@@
 We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
-The implementation below describes the model as a directed graph.
+The implementation below describes the model as a directed graph (see also Figure \@ref(fig:figecs)a).
 Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
@@
 and a LICENSE file with the legal conditions for reuse.
-For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html)
+For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html).
 We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

### 2025-03-03 manuscript.Rmd
No content change; just to open a PR …
No content change; add empty row for comment

@@
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+
 Psychological theories rarely provide hypotheses that are corroborated by evidence.

### 2025-03-05 manuscript.Rmd
Update manuscript.Rmd
Added a few comments about some small things

--- 2025-03-03 manuscript.Rmd
@@
 For present purposes, however, the terms are interchangeable.].
+
+<!-- NvD: I cannot accept this footnote as an adequate characterization.
+Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more).
+See for instance Brian Haig's 2005 paper.
+Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions".
+At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree.
+I am fine with the last sentence though ;-) My suggestion is to write the following: -->
+
+<!-- NvD: We use induction to describe inferences from specific observations to general theories.
+In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations.
+We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable.
+-->
+
 <!-- Peirce: CP 8.233 Cross-Ref:†† -->
@@
 This example demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interpretability.
+DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interoperability.
+<!-- NvD: Above we mention that "The Accessibility principles are sometimes misunderstood as pertaining to *maximizing* access; rather, they should be understood as *regulating* it." This feels a bit at odds.
+I suspect that we are not talking about the same thing here, this could be interpreted as a contradiction.
+-->
 
@@
 
+<!-- NvD: we use the term 'macro-theory' here for the first time. What does it mean? Do we need it? -->
 <!-- The data usually need to be integrated with other data. In addition, the data need to Interoperate with applications or workflows for analysis, storage, and processing. -->

### 2025-03-05 manuscript.Rmd
stash changes

--- 2025-03-03 manuscript.Rmd
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+The equally critical inductive processeses of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
 Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
@@
 We do so by implementing theories as information artifacts,
-archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
 Metadata are "data about the data", they provide information about the nature and content of an information artifact.
@@
 Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
-FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).
+FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).
 
@@
 Finally, metadata should be registered or indexed in a searchable registry (F4).
-It is important to note that, while many archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
+It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
 only few are specifically designed for FAIR-compliant archival.
@@
 Both approaches can be complementary, however.
-For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
+for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
 Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
@@
 the same image represented in the DOT language [@DOTLanguage2024]
-for representing graphs does meet this ideal (an example of such a DOT representation is given below).
+for representing graphs does meet this ideal.
 
@@
 We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
-The implementation below describes the model as a directed graph (see also Figure \@ref(fig:figecs)a).
+The implementation below describes the model as a directed graph.
 Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
@@
 and a LICENSE file with the legal conditions for reuse.
-For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html).
+For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html)
 We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
@@
 ```
+
+##
 

### 2025-03-05 manuscript.Rmd
edits and meeting notes

--- 2025-03-05 manuscript.Rmd
@@
       - Writing – review & editing
+  - name: "Maximilian S. Ernst"
+    affiliation: "2,4"
+    role:
+      - Writing – review & editing
+      - Investigation
+  - name: "Noah N.N. van Dongen"
+    affiliation: "5"
+    role:
+      - Writing – review & editing
+  - name: "Felix D. Schönbrodt"
+    affiliation: "6"
+    role:
+      - Conceptualization
+      - Writing – review & editing
   - name: "Andreas M. Brandmaier"
-    affiliation: "2,3,4"
+    affiliation: "2,3,7"
     role:
@@
       - Writing – review & editing
-  - name: "Felix D. Schönbrodt"
-    affiliation: "5"
-    role:
-      - Conceptualization
-      - Writing – review & editing
-  - name: "Noah N.N. van Dongen"
-    affiliation: "6"
-    role:
-      - Writing – review & editing
-  - name: "Maximilian S. Ernst"
-    affiliation: "2,7"
-    role:
-      - Writing – review & editing
-      - Investigation
 affiliation:
@@
   - id            : "4"
+    institution   : "Max Planck School of Cognition, Leipzig, Germany"
+  - id            : "5"
+    institution   : "University of Amsterdam, the Netherlands"
+  - id            : "6"
+    institution   : "Ludwig-Maximilians-Universität München, Germany"
+  - id            : "7"
     institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
-  - id            : "5"
-    institution   : "Ludwig-Maximilians-Universität München, Germany"
-  - id            : "6"
-    institution   : "University of Amsterdam, the Netherlands"
-  - id            : "7"
-    institution   : "Max Planck School of Cognition, Leipzig, Germany"
 authornote: |
   This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.
+The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
+See the CRediT authorship contributions and GitHub repository for specific author contributions.
 
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories.
+The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories.
 Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
-However, Peirce later acknowledged that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
-As defining what makes a theory "best" requires auxiliary assumptions, we consider abduction to be a special case of induction.
 For present purposes, however, the terms are interchangeable.].
-
+<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
+Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
+NvD dissents: -->
 <!-- NvD: I cannot accept this footnote as an adequate characterization.
@@
 I am fine with the last sentence though ;-) My suggestion is to write the following: -->
-
 <!-- NvD: We use induction to describe inferences from specific observations to general theories.
@@
 -->
-
 <!-- Peirce: CP 8.233 Cross-Ref:†† -->
@@
 -->
-More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed if the theory were true.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
+More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
 Without committing to specific implementations a-priori,
@@
 
-The Accessibility principles are sometimes misunderstood as pertaining to *maximizing* access; rather, they should be understood as *regulating* it.
-They apply to theory with minor changes.
+The Accessibility principles 
+<!-- are sometimes misunderstood as pertaining to *maximizing* access; rather, they should be understood as *regulating* it [@landiFAIROpenPossible2020]. -->
+<!-- They  -->
+apply to theory with minor changes.
 Firstly, theory and its associated metadata should be Accessible by their identifier using a standardized communications protocol (A1).
@@
 At present, there are several impediments to theories' Accessibility.
-To the extent that theories are still contained within papers,
-paywalls erected by publishers constitute a barrier.
-Open Access publishing increases the Accessibility of all academic output, including theory.
-A second impediment is more indirect:
-While open access publishing increases practical access to theories,
-Accessibility also requires clear and explicit communication.
-<!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
-This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
+One impediment is that, 
+when theories are published in paywalled journal articles,
+they might not be practically Accessible to all,
+even if they are in principle Accessible to those who can pay the fee.
+Open Access publishing increases the practical Accessibility of all academic output, including theory.
+
+A second impediment to Accessibility is more indirect
+and pertains to a theory's intelligibility to those with practical Access.
+It has been proposed that good theories have the property of "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
 At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
-It is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
+Shared background knowledge can come from paradigms held by members of a scientific community [@kuhnStructureScientificRevolutions2009],
+from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
+Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
+At the same time, it is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
 has multiple alternative translations, with no objective means of choosing the correct one.],
-which implies that it is not possible to remove all ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
+which implies that it is not possible to remove *all* ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
 This places a theoretical upper bound on theories' ability to be understood.
 
-Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
-The notion of "normal science" describes a phase in which a scientific community operates within the context of a shared paradigm,
-creating the background knowledge needed for mutual understanding and productive scientific work [@kuhnStructureScientificRevolutions2009].
-In real life, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
-including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
-
-A third impediment arises when theories have, what we call, a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
+A third impediment arises when theories have, what we call, a "dependency on the author" (DOA).
 DOA occurs when a theory cannot be understood by independent scholars,
@@
 This example demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interoperability.
+DOA thus unduly impedes scientific progress.
+<!-- DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interoperability. -->
 <!-- NvD: Above we mention that "The Accessibility principles are sometimes misunderstood as pertaining to *maximizing* access; rather, they should be understood as *regulating* it." This feels a bit at odds.
@@
 
-In sum, while it is impossible to communicate a theory completely unambiguously,
-scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
-<!-- Accessibility requires theoreticians to explicit as possible about assumptions and ontological definitions to eliminate unnecessary ambiguity. -->
-<!-- At the same time, it is important to recognize that there is an upper bound on interpretability, which means that it is impossible to eliminate all ambiguity. -->
+In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
+It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
+Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
 It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs [@langeChecklistIncentivizingFacilitating2025a].
-
-<!-- NvD: we use the term 'macro-theory' here for the first time. What does it mean? Do we need it? -->
+and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs [@langeChecklistIncentivizingFacilitating2025a].
 <!-- The data usually need to be integrated with other data. In addition, the data need to Interoperate with applications or workflows for analysis, storage, and processing. -->

### 2025-03-05 manuscript.Rmd
Final tweaks

@@
 The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
-See the CRediT authorship contributions and GitHub repository for specific author contributions.
 
@@
 
-The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
-Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
+Concern that the abundance of non-replicable findings 
+was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
 These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
@@
 Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
-The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
-Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
-Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.
-
-<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
-<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->
-
-Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
+The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
+Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
+Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
+
+Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
 One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
@@
 -->
-When theories are difficult to understand on their own,
-without either substantial interpretation or additional background knowledge,
+When theories are difficult to understand without either subjective interpretation or additional background knowledge,
 it becomes difficult do derive precise predictions,
@@
 A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
-
-Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
 The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories.
-Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation".
-For present purposes, however, the terms are interchangeable.].
+Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
+For present purposes, the terms are interchangeable.].
 <!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
@@
 <!-- line of procedure, or at least not to distinguish it sharply. -->
-The present paper restores balance by applying, for the first time,
+The present paper addresses this knowledge gap by applying, for the first time,
 open science principles to psychological theory.
-We apply the FAIR principles to scientific theories,
-introducing *FAIR theory* to 
+We apply introduce *FAIR theory* to 
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
-Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
+Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
 but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
@@
 -->
-Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].
+<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
+Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].
 
@@
 They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
+The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.
 
@@
 
-The present paper introduces open science methods for theory specification and archival.
-Merely publishing theory in a research article does not make it open;
-to be open, theory should adhere to established open science standards.
-We do so by implementing theories as information artifacts,
-archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
-Metadata are "data about the data", they provide information about the nature and content of an information artifact.
-Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
+Merely publishing theory in a journal article does not make it open;
+to be open, theory should adhere to established open science standards for specification and archival.
+We propose to implement theories as information artifacts,
+and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+Metadata are "data about the data". 
+They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
 FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
@@
 At the same time, formal theories are not automatically FAIR.
-The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
-They can be applied to theories representated in natural language,
+The FAIR principles
+<!-- pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility. -->
+<!-- They  -->
+can be applied to theories representated in natural language,
 as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
@@
 A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
-Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
+Modular publishing is the practice of making each of these resources available as independent citable *information artifacts*,
 with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
-Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
+Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
 Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
@@
 and maintains a complete history of those changes.
-It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013].
+It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
-Note, however, that the function of Git(Hub) is *not* to archive a FAIR theory.
-FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
-Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.
+Note that archival of the version of record is *not* a function of Git(Hub).
+While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.
 
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website [@grayHowMapTheory2017].
-Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
-<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
+and posted many examples on their website <www.theorymaps.org> [@grayHowMapTheory2017].
+Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
@@
 
-Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
+Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016].
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 At the time of writing (2024),
-the value of using Git for version control of academic research is well-established [@ramGitCanFacilitate2013],
-and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* of the FAIR principles ([Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
+the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* FAIR principles.
 Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
@@
 the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
-Our suggested workflow can be largely automated in R using the `theorytools` package but note that our workflow is not contingent upon researchers using R at all.
-To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
-It can be accessed by running `vignette("fairtheory", package = "theorytools")` in R.
-We present a brief summary of the instructions at the time of writing here,
+The following conceptual workflow does not require the use of R, but most steps can be automated in R using the `theorytools` package.
+The package also includes a living document with the [latest version of the workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
+We present a brief summary of the workflow at the time of writing here,
 to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
@@
 
-If we compare it to the levels of theory formalization [@guestHowComputationalModeling2021],
+If we compare this to the levels of theory formalization [@guestHowComputationalModeling2021],
 it is defined at either the "theory" or "specification" level.
@@
 
-The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved.
-In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
+<!-- The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved. -->
+<!-- In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021]. -->
 To version control our project, we initiate a Git repository in the project folder.
@@
 
-By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
-To document our archive as a FAIR theory requires adding some extra information on Zenodo.
-Supplying the information below helps improve the Findability of a theory.
-See here for [an example of the resulting FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) archived in Zenodo.
+To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
+See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle, as archived in Zenodo.
 
@@
 - Optionally, add *References* to related works in plain text. For example, here we can provide the full citation of De Groot and Spiekerman:
+
 > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: Foundations of inference and research in the behavioral sciences. De Gruyter Mouton. https://doi.org/10.1515/9783112313121
-
-<!-- ## 6. Making Changes -->
-<!-- CJ: Aaron, I moved your part here because I don't think it relates to Interoperability. But perhaps there's another place where it fits better? -->
 
@@
 
-De Groot's empirical cycle has inspired several authors,
-but not all of them have interpreted his work the same.
-For example, Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle diverges substantially from De Groot's description.
+Several authors have reinterpreted De Groot's empirical cycle.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-Wagenmakers and colleagues' version can also be implemented as a DOT graph
-to illustrate some clear deviations from the original.
-Notice that, first, the phases of the cycle have been renamed.
+For example, the DOT graph below implements Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle.
+First, notice that the phases of the cycle have been renamed.
 This change was not described in the paper.
@@
 then we can represent it by adding labels to the original DOT graph.
-The labels do suggest a focus on empirical psychology not present in De Groot's version.
+These labels suggest a focus on empirical psychology that was not present in De Groot's version.
 Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
@@
 This invites the question of what the concrete outcomes of these processes are.
-In other words: what actually changes when going through the cycle, except the scholar's mind.
+In other words: what actually changes when going through the cycle, except the scholar's mind?
 To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs), panel (c).
@@
 <!-- ``` -->
-For example, if observation is theory-laden, it involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
+For example, observation is theory-laden, and thus involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
 Deriving hypotheses from theory is not purely deductive either,

### 2025-03-05 manuscript.tex
Final tweaks

@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 9205}
+\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 9279}
 \usepackage{lineno}
@@
   pdftitle={To be FAIR: Theory Specification Needs an Update},
-  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, Noah N.N. van Dongen6, \& Maximilian S. Ernst2,7},
+  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Maximilian S. Ernst2,4, Noah N.N. van Dongen5, Felix D. Schönbrodt6, \& Andreas M. Brandmaier2,3,7},
   pdflang={en-EN},
@@
 \author{Caspar J.
-Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M.
-Brandmaier\textsuperscript{2,3,4}, Felix D.
-Schönbrodt\textsuperscript{5}, Noah N.N.
-van Dongen\textsuperscript{6}, \& Maximilian S.
-Ernst\textsuperscript{2,7}}
+Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Maximilian S.
+Ernst\textsuperscript{2,4}, Noah N.N.
+van Dongen\textsuperscript{5}, Felix D.
+Schönbrodt\textsuperscript{6}, \& Andreas M.
+Brandmaier\textsuperscript{2,3,7}}
 \date{}
@@
 
-This is a preprint paper, generated from Git Commit \# b77adbc3.
+This is a preprint paper, generated from Git Commit \# 3cbbaf56.
+The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
+See the CRediT authorship contributions and GitHub repository for specific author contributions.
 
@@
 Caspar J.
-Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M.
-Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D.
-Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N.
-van Dongen: Writing -- review \& editing; Maximilian S.
-Ernst: Writing -- review \& editing, Investigation.
+Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S.
+Ernst: Writing -- review \& editing, Investigation; Noah N.N.
+van Dongen: Writing -- review \& editing; Felix D.
+Schönbrodt: Conceptualization, Writing -- review \& editing; Andreas M.
+Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing.
 
@@
 \affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Department of Psychology, MSB Medical School Berlin, Berlin, Germany\\\textsuperscript{5} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{6} University of Amsterdam, the Netherlands\\\textsuperscript{7} Max Planck School of Cognition, Leipzig, Germany}
+Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Max Planck School of Cognition, Leipzig, Germany\\\textsuperscript{5} University of Amsterdam, the Netherlands\\\textsuperscript{6} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{7} Department of Psychology, MSB Medical School Berlin, Berlin, Germany}
 
@@
 FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved through collaborative efforts.
-This paper adapts the FAIR priciples for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
+This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
 We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing.
@@
 Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processeses of theory construction and improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories.
+The equally critical inductive processes of theory construction and improvement have been largely overlooked\footnote{Here, we use ``induction'' to refer to inferences from specific observations to general theories.
 Others have used the term ``abduction'', coined by Peirce to describe ``inference to the best explanation''.
-However, Peirce later acknowledged that abduction ``is not essentially different from induction'' (Peirce, 1960, p.
-4881).
-As defining what makes a theory ``best'' requires auxiliary assumptions, we consider abduction to be a special case of induction.
 For present purposes, however, the terms are interchangeable.}.
@@
 We do so by implementing theories as information artifacts,
-archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
 Metadata are ``data about the data'', they provide information about the nature and content of an information artifact.
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-More rigorous formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as precise statements, clearly demarcating what should (not) be observed if the theory were true.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for its particular clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
+More rigorous formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
 Without committing to specific implementations a-priori,
@@
 Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
-FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).
+FAIR theory is thus consistent with, but does not require, formalization (also see the section on \emph{Accessibility} below).
 
@@
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
-The present paper considers the advantages of Git for FAIR theory.
-Git enables explicitly comparing versions of a file (or: theory),
+Git provides a useful framework for developing FAIR theory,
+because it enables explicitly comparing versions of a file (or: theory),
 incorporating changes by different authors,
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
+Note, however, that the function of Git(Hub) is \emph{not} to archive a FAIR theory.
+FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
+Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.
 
@@
 Both approaches can be complementary, however.
-for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
+For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
 Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
@@
 
-The Accessibility principles are sometimes misunderstood as pertaining to \emph{maximizing} access; rather, they should be understood as \emph{regulating} it.
-They apply to theory with minor changes.
+The Accessibility principles
+apply to theory with minor changes.
 Firstly, theory and its associated metadata should be Accessible by their identifier using a standardized communications protocol (A1).
@@
 At present, there are several impediments to theories' Accessibility.
-To the extent that theories are still contained within papers,
-paywalls erected by publishers constitute a barrier.
-Open Access publishing increases the Accessibility of all academic output, including theory.
-A second impediment is more indirect:
-While open access publishing increases practical access to theories,
-Accessibility also requires clear and explicit communication.
-This property of good theories has been dubbed ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
+One impediment is that,
+when theories are published in paywalled journal articles,
+they might not be practically Accessible to all,
+even if they are in principle Accessible to those who can pay the fee.
+Open Access publishing increases the practical Accessibility of all academic output, including theory.
+
+A second impediment to Accessibility is more indirect
+and pertains to a theory's intelligibility to those with practical Access.
+It has been proposed that good theories have the property of ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
 At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
-It is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
+Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
+Shared background knowledge can come from paradigms held by members of a scientific community (Kuhn, 2009),
+from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
+Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
+At the same time, it is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
   has multiple alternative translations, with no objective means of choosing the correct one.},
-which implies that it is not possible to remove all ambiguity when communicating an idea (Quine, 1970).
+which implies that it is not possible to remove \emph{all} ambiguity when communicating an idea (Quine, 1970).
 This places a theoretical upper bound on theories' ability to be understood.
-
-Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
-The notion of ``normal science'' describes a phase in which a scientific community operates within the context of a shared paradigm,
-creating the background knowledge needed for mutual understanding and productive scientific work (Kuhn, 2009).
-In real life, paradigms are rarely clear-cut, and shared background knowledge can come from different sources,
-including education, cultures of communication, and availability of particular instruments for observation, measurement, and analysis - or it can be problematically absent.
 
@@
 This example demonstrates that original authors are not the ultimate authority on their theories.
-DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability and interpretability.
-
-In sum, while it is impossible to communicate a theory completely unambiguously,
-scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+DOA thus unduly impedes scientific progress.
+
+In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
+It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
+Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
 It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, macro-theories, measurement instruments, experimental designs (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
+and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
 
@@
 the same image represented in the DOT language ({``{DOT Language},''} 2024)
-for representing graphs does meet this ideal.
+for representing graphs does meet this ideal (an example of such a DOT representation is given below).
 
@@
 We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
-The implementation below describes the model as a directed graph.
+The implementation below describes the model as a directed graph (see also Figure \ref{fig:figecs}a).
 Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
@@
 and a LICENSE file with the legal conditions for reuse.
-For guidance on writing the README file, see \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{this vignette}
+For guidance on writing the README file, see \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{this vignette}.
 We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
@@
 
-\subsection{Updating a Theory}\label{updating-a-theory}
+\subsection{Changing a Theory}\label{changing-a-theory}
 
@@
 but again specified the empirical cycle differently.
-First, notice that the nodes in De Groot's formulation mostly refer to processes.
-This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
-In the implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
-Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
-Many have argued that observation is theory-laden, and as such, involves induction (Brewer \& Lambert, 2001).
-Deriving hypotheses from theory is also not purely deductive,
-as auxiliary assumptions are often made (which, again, involves induction).
+First, in De Groot's formulation, each stage describes a process.
+This invites the question of what the concrete outcomes of these processes are.
+In other words: what actually changes when going through the cycle, except the scholar's mind.
+To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}, panel (c).
+Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
+For example, if observation is theory-laden, it involves induction (Brewer \& Lambert, 2001).
+Deriving hypotheses from theory is not purely deductive either,
+as auxiliary assumptions are often made, which involves induction (e.g., assuming equal variances across groups because groups often have equal variances).
 Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (Peikert, 2023).
-With these alterations, we implement the empirical cycle as follows:
+These alterations result in the following implementation of the empirical cycle:
 
@@
 \end{figure}
+
+\subsection{}\label{section}
 
@@
 
-\bibitem[\citeproctext]{ref-peirceCollectedPapersCharles1960}
-Peirce, C. (1960). \emph{Collected {Papers Of Charles Peirce}}. Retrieved from \url{http://archive.org/details/collected-papers-of-charles-peirce}
-
 \bibitem[\citeproctext]{ref-popperLogicScientificDiscovery2002}

### 2025-03-05 manuscript.Rmd
update 2025-03-05 13:36:35.725556

@@
 
-##
-
 ## Further Uses of FAIR Theory
@@
 the `theorytools` package contains several vignettes that showcase specific applications.
-At the time of writing, the package includes a [vignette on theory specification](),
-that is to say, moving from a theory represented as prose to a FAIR theory,
-based on the many-theorists project by Glöckner and Fiedler (in preparation).
-Another vignette illustrates the use of FAIR theory for causal inference [@pearlCausalDiagramsEmpirical1995].
-Furthermore, one vignette illustrates the use of FAIR theory for [conducting a simulation study]().
+At the time of writing, the package includes a
+vignette [introducing augmented Directed Acyclic Graphs (aDAGs)](https://cjvanlissa.github.io/theorytools/articles/augmented_dags.html) as a format for theory specification that meets the requirements of good psychological theory.
+This vignette also shows the `theorytools` package's functionality for generating scripts for simulation studies, based on the information about distributions and functional form encoded in the aDAG.
+Another vignette describes how to take Self-Determination Theory [@langeSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
+Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning_krueger.html) [@feldEstimatingRelationshipSkill2017].
+Another vignette illustrates the use of FAIR theory for [covariate selection and causal inference](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].
 More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.

### 2025-03-05 manuscript.Rmd
consistent discrimination of vague and ambiguous (striking empierical content)

@@
 Psychological theories rarely provide hypotheses that are corroborated by evidence.
-Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
+Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
 as the theory's central claims evade falsification.
@@
 -->
-When theories are difficult to understand without either subjective interpretation or additional background knowledge,
+When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
 it becomes difficult do derive precise predictions,
@@
 -->
-More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
@@
 Compared to theories expressed in natural language,
-formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
+formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
 Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
@@
 Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
-While FAIR theory does not directly reduce ambiguity,
+While FAIR theory does not directly reduce vagueness,
 it provides a framework within which scholars can iteratively increase precision and formalization.

### 2025-03-06 manuscript.Rmd

@@
 Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+<!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
+Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
+Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness.
+-->
 The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
@@
 Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
+The expanded definition thus enables SDT to absorb potentially falsifying evidence.
 

### 2025-03-09 manuscript.Rmd
submitted version

@@
 abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
-keywords          : "fairtheory, meta theory, theory formation, cumulative science, formal models"
+keywords          : "fairtheory, meta science, theory formation, open science, formal models"
 wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
@@
 header-includes:
-  - \usepackage{longtable, setspace}
+  - \usepackage{longtable, setspace, subfig}
   - \raggedbottom
@@
 was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
-These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
+These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
@@
 Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories.
-Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
-For present purposes, the terms are interchangeable.].
+The equally critical processes of theory construction and improvement have been largely overlooked.
 <!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
@@
 a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs), panel (a).
-In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
+research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs)a.
+In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data.
+In the other half of the cycle (the "Context of Discovery"),
+patterns observed in data are generalized to theoretical principles, Figure \@ref(fig:figecs)b.
 In this model, theories are the vehicle of scientists' understanding of phenomena.
@@
 ```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologieGrondslagenVan1961].'}
-# rsvg::rsvg_pdf("empirical_cycle1.svg", "ec_visio.pdf")
+# rsvg::rsvg_pdf("empirical_cycle2.svg", "ec_visio.pdf")
 knitr::include_graphics("ec_visio.pdf")
@@
 We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
-- Optionally, add *References* to related works in plain text. For example, here we can provide the full citation of De Groot and Spiekerman:
-
-> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: Foundations of inference and research in the behavioral sciences. De Gruyter Mouton. https://doi.org/10.1515/9783112313121
+- Optionally, add *References* to related works in plain text. For example, we cite De Groot and Spiekerman in this field.
 
@@
 In other words: what actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs), panel (c).
+To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
 Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
@@
 <!-- ``` -->
-For example, observation is theory-laden, and thus involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
-Deriving hypotheses from theory is not purely deductive either,
-as auxiliary assumptions are often made, which involves induction (e.g., assuming equal variances across groups because groups often have equal variances).
+Theory testing, as takes place in the "context of justification", can be said to involve mostly deductive reasoning.
+Theory development and amendment, as takes place in the "context of discovery", involves primarily inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations.
+Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
+For present purposes, the terms are interchangeable.].
+<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
+Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
+NvD dissents: -->
+<!-- NvD: I cannot accept this footnote as an adequate characterization.
+Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more).
+See for instance Brian Haig's 2005 paper.
+Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions".
+At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree.
+I am fine with the last sentence though ;-) My suggestion is to write the following: -->
+<!-- NvD: We use induction to describe inferences from specific observations to general theories.
+In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations.
+We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable.
+-->
+<!-- Peirce: CP 8.233 Cross-Ref:†† -->
+<!-- 233. This is not essentially different from induction. Only it is most -->
+<!-- usually an induction from instances which are not discrete and numerable. I now -->
+<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
+<!-- line of procedure, or at least not to distinguish it sharply. -->
+However, deriving hypotheses from theory is not purely deductive
+as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
+A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
+Similarly, if we consider the claim that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
 Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
@@
 Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
-Figure \@ref(fig:figbranch) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
+Figure \@ref(fig:figgithub-1) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
 A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
@@
 Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
-Figure \@ref(fig:figcompare) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
-
-```{r figbranch, fig.cap="Branches can contain different versions of a theory.", out.width = "50%"}
+Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+
+```{r figgithub, out.width = "50%", fig.show='hold', fig.cap='FAIR Theories on GitHub', fig.subcap=c('Branches contain different versions of a theory', 'These versions can be compared'), fig.ncol = 2}
+#par(mfrow=c(1,2))
 knitr::include_graphics("git_branches.png")
-```
-
-```{r figcompare, fig.cap="GitHub's compare functionality.", out.width = "50%"}
+knitr::include_graphics("git_compare_short.png")
+```
+
+```{r figbranch, eval = FALSE, fig.cap="Branches can contain different versions of a theory.", out.width = "50%"}
+knitr::include_graphics("git_branches.png")
+```
+
+```{r figcompare, eval = FALSE, fig.cap="GitHub's compare functionality.", out.width = "50%"}
 knitr::include_graphics("git_compare.png")
@@
 vignette [introducing augmented Directed Acyclic Graphs (aDAGs)](https://cjvanlissa.github.io/theorytools/articles/augmented_dags.html) as a format for theory specification that meets the requirements of good psychological theory.
-This vignette also shows the `theorytools` package's functionality for generating scripts for simulation studies, based on the information about distributions and functional form encoded in the aDAG.
+These aDAGs are X-interoperable for plotting (using `dagitty` and `tidySEM`),
+for automatically selecting control variables, and for simulating data (using `theorytools`).
 Another vignette describes how to take Self-Determination Theory [@langeSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
-Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning_krueger.html) [@feldEstimatingRelationshipSkill2017].
+Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].
 Another vignette illustrates the use of FAIR theory for [covariate selection and causal inference](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].

### 2025-03-09 manuscript.tex
submitted version

@@
 \makeatother
-\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 9279}
+\keywords{fairtheory, meta science, theory formation, open science, formal models\newline\indent Word count: 9211}
 \usepackage{lineno}
@@
 \usepackage{csquotes}
-\usepackage{longtable, setspace}
+\usepackage{longtable, setspace, subfig}
 \raggedbottom
@@
   pdflang={en-EN},
-  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
+  pdfkeywords={fairtheory, meta science, theory formation, open science, formal models},
   colorlinks=true,
@@
 
-This is a preprint paper, generated from Git Commit \# 3cbbaf56.
+This is a preprint paper, generated from Git Commit \# 985c9433.
 The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
-See the CRediT authorship contributions and GitHub repository for specific author contributions.
 
@@
 To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts.
-FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved through collaborative efforts.
+FAIR theories are Findable in well-established archives, Accessible in practical terms and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable so that they can be iteratively improved through collaborative efforts.
 This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
@@
 FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond.
-FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
+FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition
 }
@@
 D.
-Wilkinson et al., 2016a) for both humans and computers.
+Wilkinson et al., 2016) for both humans and computers.
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
@@
 
-The so-called ``replication crisis'' has prompted extensive reforms in psychology and other scientific fields (Lavelle, 2021; Scheel, 2022).
-Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
-These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
+The so-called ``replication crisis'' has prompted extensive scientific reforms (Lavelle, 2021; Scheel, 2022).
+Concern that the abundance of non-replicable findings
+was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
+These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
+
 Psychological theories rarely provide hypotheses that are corroborated by evidence.
-Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
+Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
 as the theory's central claims evade falsification.
@@
 Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
-The profound difference between these definitions becomes clear when considering the act of changing a child's dirty diaper.
-Under the original definition, few caregivers would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience.
-Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.
-
-Scholars have been raising concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
+The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
+Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
+Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
+The expanded definition thus enables SDT to absorb potentially falsifying evidence.
+
+Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
 One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
-When theories are difficult to understand on their own,
-without either substantial interpretation or additional background knowledge,
+When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
 it becomes difficult do derive precise predictions,
@@
 A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
-
-Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
-The equally critical inductive processes of theory construction and improvement have been largely overlooked\footnote{Here, we use ``induction'' to refer to inferences from specific observations to general theories.
-Others have used the term ``abduction'', coined by Peirce to describe ``inference to the best explanation''.
-For present purposes, however, the terms are interchangeable.}.
-The present paper restores balance by applying, for the first time,
+Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
+The equally critical processes of theory construction and improvement have been largely overlooked.
+The present paper addresses this knowledge gap by applying, for the first time,
 open science principles to psychological theory.
-We apply the FAIR principles to scientific theories,
-introducing \emph{FAIR theory} to
+We apply introduce \emph{FAIR theory} to
 facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
@@
 a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}, panel (a).
-In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
+research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}a.
+In one half of the cycle, labeled the ``Context of Justification'' by Wagenmakers and colleagues, hypotheses derived from theory are tested on data.
+In the other half of the cycle (the ``Context of Discovery''),
+patterns observed in data are generalized to theoretical principles, Figure \ref{fig:figecs}b.
 In this model, theories are the vehicle of scientists' understanding of phenomena.
@@
 in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
-Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
+Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
 but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).
+Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).
 
@@
 They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
+The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.
 
@@
 
-The present paper introduces open science methods for theory specification and archival.
-Merely publishing theory in a research article does not make it open;
-to be open, theory should adhere to established open science standards.
-We do so by implementing theories as information artifacts,
-archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
-Metadata are ``data about the data'', they provide information about the nature and content of an information artifact.
-Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
+Merely publishing theory in a journal article does not make it open;
+to be open, theory should adhere to established open science standards for specification and archival.
+We propose to implement theories as information artifacts,
+and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+Metadata are ``data about the data''.
+They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
 FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
@@
 Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-More rigorous formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+More rigorous formalization increases theories' falsifiability (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
@@
 At the same time, formal theories are not automatically FAIR.
-The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
-They can be applied to theories representated in natural language,
+The FAIR principles
+can be applied to theories representated in natural language,
 as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
@@
 A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
-Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
+Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
 with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
-Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
+Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
 Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
@@
 and maintains a complete history of those changes.
-It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
+It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
 Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
-Note, however, that the function of Git(Hub) is \emph{not} to archive a FAIR theory.
-FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
-Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.
+Note that archival of the version of record is \emph{not} a function of Git(Hub).
+While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.
 
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and post many examples on their website (Gray, 2017).
-Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
+and posted many examples on their website \textless www.theorymaps.org\textgreater{} (Gray, 2017).
+Similarly, \href{http://psychomodels.org/}{PsychoModels} seeks to inventorize theories and models in psychology (van Dongen \& Volz, 2025).
 Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
+and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J.
+Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
 
@@
 
-Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016b).
+Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 At the time of writing (2024),
-the value of using Git for version control of academic research is well-established (Ram, 2013),
-and the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets \emph{all} of the FAIR principles (\href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
+the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets \emph{all} FAIR principles.
 Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
@@
 the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
-Our suggested workflow can be largely automated in R using the \texttt{theorytools} package but note that our workflow is not contingent upon researchers using R at all.
-To anticipate workflow changes, the package includes a living document with the most recent version of our proposed workflow.
-It can be accessed by running \texttt{vignette("fairtheory",\ package\ =\ "theorytools")} in R.
-We present a brief summary of the instructions at the time of writing here,
+The following conceptual workflow does not require the use of R, but most steps can be automated in R using the \texttt{theorytools} package.
+The package also includes a living document with the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{latest version of the workflow}.
+We present a brief summary of the workflow at the time of writing here,
 to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
@@
 
-If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
+If we compare this to the levels of theory formalization (Guest \& Martin, 2021),
 it is defined at either the ``theory'' or ``specification'' level.
@@
 
-The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved.
-In particular, the practice of version control offers extensive benefits for scientific work (Ram, 2013; Van Lissa et al., 2021).
 To version control our project, we initiate a Git repository in the project folder.
@@
 
-By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
-To document our archive as a FAIR theory requires adding some extra information on Zenodo.
-Supplying the information below helps improve the Findability of a theory.
-See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the resulting FAIR metadata} archived in Zenodo.
+To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
+See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the metadata} associated with our FAIR empirical cycle, as archived in Zenodo.
 
@@
 \item
-  Optionally, add \emph{References} to related works in plain text. For example, here we can provide the full citation of De Groot and Spiekerman:
-  \textgreater{} De Groot, A.
-D., \& Spiekerman, J.
-A.
-A.
-(1969).
-Methodology: Foundations of inference and research in the behavioral sciences.
-De Gruyter Mouton.
-\url{https://doi.org/10.1515/9783112313121}
+  Optionally, add \emph{References} to related works in plain text. For example, we cite De Groot and Spiekerman in this field.
 \end{itemize}
@@
 
-De Groot's empirical cycle has inspired several authors,
-but not all of them have interpreted his work the same.
-For example, Wagenmakers and colleagues' (2018) interpretation of the empirical cycle diverges substantially from De Groot's description.
+Several authors have reinterpreted De Groot's empirical cycle.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-Wagenmakers and colleagues' version can also be implemented as a DOT graph
-to illustrate some clear deviations from the original.
-Notice that, first, the phases of the cycle have been renamed.
+For example, the DOT graph below implements Wagenmakers and colleagues' (2018) interpretation of the empirical cycle.
+First, notice that the phases of the cycle have been renamed.
 This change was not described in the paper.
@@
 then we can represent it by adding labels to the original DOT graph.
-The labels do suggest a focus on empirical psychology not present in De Groot's version.
+These labels suggest a focus on empirical psychology that was not present in De Groot's version.
 Furthermore, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
@@
 This invites the question of what the concrete outcomes of these processes are.
-In other words: what actually changes when going through the cycle, except the scholar's mind.
-To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}, panel (c).
+In other words: what actually changes when going through the cycle, except the scholar's mind?
+To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}c).
 Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
-For example, if observation is theory-laden, it involves induction (Brewer \& Lambert, 2001).
-Deriving hypotheses from theory is not purely deductive either,
-as auxiliary assumptions are often made, which involves induction (e.g., assuming equal variances across groups because groups often have equal variances).
+Theory testing, as takes place in the ``context of justification'', can be said to involve mostly deductive reasoning.
+Theory development and amendment, as takes place in the ``context of discovery'', involves primarily inductive reasoning\footnote{Here, ``induction'' is defined forming general theories based on specific observations.
+Others have used the term ``abduction'' to describe ``inference to the best explanation'' (Peirce, 1960).
+For present purposes, the terms are interchangeable.}.
+However, deriving hypotheses from theory is not purely deductive
+as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
+A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
+Similarly, if we consider the claim that observation is theory-laden, then it too involves induction (Brewer \& Lambert, 2001).
 Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (Peikert, 2023).
@@
 Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
-Figure \ref{fig:figbranch} shows how our example \texttt{empirical\_cycle} repository contains branches with Wagenmakers' and Van Lissa's implementations.
+Figure \ref{fig:figgithub-1} shows how our example \texttt{empirical\_cycle} repository contains branches with Wagenmakers' and Van Lissa's implementations.
 A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
@@
 Both branches and forks can be compared and merged via ``pull requests'', which are essentially a request to incorporate the changes you have made.
-Figure \ref{fig:figcompare} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+Figure \ref{fig:figgithub-2} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
 
 \begin{figure}
-\includegraphics[width=0.5\linewidth]{git_branches} \caption{Branches can contain different versions of a theory.}\label{fig:figbranch}
+\subfloat[Branches contain different versions of a theory\label{fig:figgithub-1}]{\includegraphics[width=0.5\linewidth]{git_branches} }\subfloat[These versions can be compared\label{fig:figgithub-2}]{\includegraphics[width=0.5\linewidth]{git_compare_short} }\caption{FAIR Theories on GitHub}\label{fig:figgithub}
 \end{figure}
-
-\begin{figure}
-\includegraphics[width=0.5\linewidth]{git_compare} \caption{GitHub's compare functionality.}\label{fig:figcompare}
-\end{figure}
-
-\subsection{}\label{section}
 
@@
 the \texttt{theorytools} package contains several vignettes that showcase specific applications.
-At the time of writing, the package includes a \href{}{vignette on theory specification},
-that is to say, moving from a theory represented as prose to a FAIR theory,
-based on the many-theorists project by Glöckner and Fiedler (in preparation).
-Another vignette illustrates the use of FAIR theory for causal inference (Pearl, 1995).
-Furthermore, one vignette illustrates the use of FAIR theory for \href{}{conducting a simulation study}.
+At the time of writing, the package includes a
+vignette \href{https://cjvanlissa.github.io/theorytools/articles/augmented_dags.html}{introducing augmented Directed Acyclic Graphs (aDAGs)} as a format for theory specification that meets the requirements of good psychological theory.
+These aDAGs are X-interoperable for plotting (using \texttt{dagitty} and \texttt{tidySEM}),
+for automatically selecting control variables, and for simulating data (using \texttt{theorytools}).
+Another vignette describes how to take Self-Determination Theory (P.
+A.
+M.
+V.
+Lange, W.Kruglanski, ToryHiggins, Deci, \& Ryan, 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
+Another vignette describes how to take the Dunning-Krüger effect and \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{specify it as a FAIR mathematical formula} (Feld, Sauermann, \& de Grip, 2017).
+Another vignette illustrates the use of FAIR theory for \href{https://cjvanlissa.github.io/theorytools/articles/causal-inference.html}{covariate selection and causal inference} (Pearl, 1995).
 More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
@@
 
+\bibitem[\citeproctext]{ref-feldEstimatingRelationshipSkill2017}
+Feld, J., Sauermann, J., \& de Grip, A.
+(2017).
+Estimating the relationship between skill and overconfidence.
+\emph{Journal of Behavioral and Experimental Economics}, \emph{68}, 18--24.
+\url{https://doi.org/10.1016/j.socec.2017.03.002}
+
 \bibitem[\citeproctext]{ref-frankenhuisStrategicAmbiguitySocial2023}
@@
 
+\bibitem[\citeproctext]{ref-langeSelfDeterminationTheory2012}
+Lange, P.
+A.
+M.
+V., W.Kruglanski, A., ToryHiggins, E., Deci, E.
+L., \& Ryan, R.
+M.
+(2012).
+Self-{Determination Theory}.
+In \emph{Handbook of {Theories} of {Social Psychology}: {Volume} 1} (pp.
+416--437).
+SAGE Publications Ltd.
+\url{https://doi.org/10.4135/9781446249215}
+
 \bibitem[\citeproctext]{ref-lavelleWhenCrisisBecomes2021}
@@
 \url{https://doi.org/10.3390/psych3040053}
+
+\bibitem[\citeproctext]{ref-peirceCollectedPapersCharles1960}
+Peirce, C. (1960). \emph{Collected {Papers Of Charles Peirce}}. Retrieved from \url{http://archive.org/details/collected-papers-of-charles-peirce}
 
@@
 
+\bibitem[\citeproctext]{ref-vandongenPsychoModelsMathematicalComputational2025}
+van Dongen, N.
+N.
+N., \& Volz, L.
+(2025).
+{PsychoModels}: {The} mathematical and computational model repository for psychological science.
+Retrieved March 5, 2025, from \url{https://www.psychomodels.org/}
+
 \bibitem[\citeproctext]{ref-vanlissaUsingEndpointsCheck2023}
@@
 J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
-(2016a).
+(2016).
 The {FAIR Guiding Principles} for scientific data management and stewardship.
 \emph{Scientific Data}, \emph{3}(1, 1), 1--9.
-\url{https://doi.org/10.1038/sdata.2016.18}
-
-\bibitem[\citeproctext]{ref-wilkinsonFAIRGuidingPrinciples2016a}
-Wilkinson, M.
-D., Dumontier, M., Aalbersberg, Ij.
-J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
-(2016b).
-The {FAIR Guiding Principles} for scientific data management and stewardship.
-\emph{Scientific Data}, \emph{3}(1), 160018.
 \url{https://doi.org/10.1038/sdata.2016.18}

### 2025-03-19 manuscript.Rmd
update for submission to pps

@@
 abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
-keywords          : "fairtheory, meta science, theory formation, open science, formal models"
+keywords          : "fairtheory, meta science, theory formation, open science"
 wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
@@
 
-According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
+According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
 a meta-theoretical model of cumulative knowledge acquisition,
@@
 
-```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologieGrondslagenVan1961].'}
+```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologyFoundationsInference1969].'}
 # rsvg::rsvg_pdf("empirical_cycle2.svg", "ec_visio.pdf")
@@
 *"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
-As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologieGrondslagenVan1961].
+As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologyFoundationsInference1969].
 In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.

### 2025-03-19 manuscript.tex
update for submission to pps

@@
 \makeatother
-\keywords{fairtheory, meta science, theory formation, open science, formal models\newline\indent Word count: 9211}
+\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9211}
 \usepackage{lineno}
@@
   pdflang={en-EN},
-  pdfkeywords={fairtheory, meta science, theory formation, open science, formal models},
+  pdfkeywords={fairtheory, meta science, theory formation, open science},
   colorlinks=true,
@@
 
-This is a preprint paper, generated from Git Commit \# 985c9433.
+This is a preprint paper, generated from Git Commit \# 7b710ba5.
 The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
@@
 To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts.
-FAIR theories are Findable in well-established archives, Accessible in practical terms and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable so that they can be iteratively improved through collaborative efforts.
-This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
-We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing.
-By applying the principles and infrastructure that have already revolutionized sharing of data and publications to theory, we establish a sustainable, transparent, and collaborative approach to theory development.
-FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond.
-FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition
+FAIR theories are Findable in well-established archives, Accessible in terms of availability and their ability to be understood, Interoperable for specific purposes, such as selecting control variables, and Reusable so that they can be iteratively improved through collaborative efforts.
+This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary psychological practices, introduces a workflow for FAIRifying theory, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
+FAIR theory constitutes framework for specifying and refining theories, bridging a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond.
+FAIR theory builds on existing open science principles and infrastructures to provide a structured, cumulative framework for theory development, potentially increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
 }
@@
 
-According to the \emph{empirical cycle} (de Groot, 1961),
+According to the \emph{empirical cycle} (De Groot \& Spiekerman, 1969),
 a meta-theoretical model of cumulative knowledge acquisition,
@@
 \includegraphics{ec_visio.pdf}
-\caption{\label{fig:figecs}Three implementations of the ``empirical cycle'' (de Groot, 1961).}
+\caption{\label{fig:figecs}Three implementations of the ``empirical cycle'' (De Groot \& Spiekerman, 1969).}
 \end{figure}
@@
 \emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
-As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (de Groot, 1961).
+As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (De Groot \& Spiekerman, 1969).
 In FAIR theory workshops, we similarly notice a reluctance to reuse and adapt existing theories.
@@
 
-\bibitem[\citeproctext]{ref-degrootMethodologieGrondslagenVan1961}
-de Groot, A.
-D.
-(1961).
-\emph{Methodologie: Grondslagen van onderzoek en denken in de gedragswetenschappen}.
-'s Gravenhage: Uitgeverij Mouton.
-Retrieved from \url{https://books.google.com?id=6hiBDwAAQBAJ}
-
 \bibitem[\citeproctext]{ref-degrootMethodologyFoundationsInference1969}

### 2025-06-06 action_letter.Rmd
update 2025-06-06 10:16:10.132865

---
title             : "To be FAIR: Theory Specification Needs an Update"
authors           : "Caspar van Lissa on behalf of co-authors"
journal           : "Perspectives on Psychological Science"
handling_editor   : "Vrieze"
class             : "draft"
output            : rmarkdown::html_document
---

<!-- revise::revise_letter_pdf -->

Dear Dr `r rmarkdown::metadata$handling_editor`,

Thank you for considering our manuscript for publication at _`r rmarkdown::metadata$journal`_.
We appreciate the feedback that you and the reviewers have provided.
In the following itemised list we respond to each comment point by point.

```{r setup-chunk, include = FALSE}
revise::read_manuscript("manuscript.Rmd")
```

## Editor's Comments

```{asis}
Thank you for submitting Manuscript ID PPS-25-140 entitled "To be FAIR: Theory Specification Needs an Update" to Perspectives on Psychological Science. It now has been reviewed.  The comments of the reviewer(s) are included at the bottom of this letter.

The reviewer(s) have recommended publication, but also suggest some revisions to your manuscript. I have some further comments, which I hope also will be helpful to you in making a revision. Therefore, I invite you to respond to the reviewer(s)' and my comments and to revise your manuscript.

[x] Headings should not be numbered.
[x] Tables and Figures should be uploaded separately and not positioned in the text (although it is fine to indicate where they should be inserted).
[] Figures should be uploaded in their native format (the program they were created in) at 300 dpi. Acceptable figure file formats include MS Office files (DOC, PPT, XLS), Adobe Illustrator (AI), JPG, EPS, and PDF.
[x] References should be done with hanging indents using APA style. 
[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. WHAT??
```

Upload tables separately: tab_meehl.csv
Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png

```{asis}
a) You have a balancing act to strike between providing technical tools (e.g., version control workflows) that can assist scientists in implementing your proposals on the one hand, against a readership and a scientific discipline that likely largely cannot effectively implement such tools. I will give a simple example. The very first thing I read after the title page was "This is a preprint paper, generated from Git Commit # 7b710ba5." I wager that most readers of Perspectives do not know what a git commit is, at least not in any meaningful way. This is a minor example, but illustrates the larger point. Reviewer 2 (Turkheimer) makes this point in his review.
```

This does not directly ask for a change, but summarize relevant changes here.

<!-- AB: -->
1.	I agree that striking the balance between the conceptual and technical details is the very challenge of this paper. I think we should go a route in the revision in which we try to explain technical details in even simpler language. For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects.

```{asis}
b) My interpretation of this manuscript is that it is proposing policies, procedures, and documentation as an important path forward to improve psychology and related soft sciences. One reason I find such proposals interesting is because I don't see them in my other area of expertise, human genetics. No doubt strong conventions develop in genetics, but they seem to develop because scientists writ large working in those fields recognize that those conventions (e.g., publicly available genomic maps organized in a certain way; analytical workflows based on accepted genetic theories) will immediately benefit their own area of research by making new avenues of work feasible, or more efficient. The payoff is obvious. They are "no-brainers" that organically arise. Some conventions in psychology, like registration or even data sharing, do not seem to arise or take hold in quite the same way, and instead rely on top-down requirements by grant funding agencies, journals, etc.

I can definitely see the advantage to policies, procedures, and documentation, if the goal is to identify fraudulent work, because it creates a paper trail to audit. I am more skeptical that these things are crucial to scientific advance in psychology, and there are no empirical results in this paper to the contrary. I believe this echoes Reviewer 3's suggestion to take seriously in your discussion, and I would suggest elsewhere in the paper, the limitations of policies, procedures, and documentation, versus other aspects of the scientific enterprise including culture, norms, and perhaps even the tractability of the subject matter, as in my human genetics example. 
```

<!-- Make comparison to preregistration; that was first proposed and only later evaluated. -->
<!-- Mention how few of preregistered hypotheses are true, and yet how theory is rarely updated? Thus, falsificationism is not working in practice. -->

<!-- <!-- AB: --> -->
<!-- 2.	I think the policies argument is interesting. I agree that currently we simply claim that establish FAIR theories will profit the advancement of science by an argument of analogy (to open data and open code) but ultimately we do not know. For one, I think it's fair to admit this. Second, I would argue that FAIR theories is not a very strict policy (in contrast to other stances in science, which require very specific ways of working, e.g. commiting to falsificationism, adhering to ethical principles, etc.). Really, FAIR theories are no restriction in that sense but more of a protocol for communication, which ultimately does not change your workflow or your methods. -->

We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of open science reforms in psychology, which by-and-large also originated in grass-roots initiatives.

As methodologists, our focus in this paper was on "procedures" rather than policies or documentation. Our goal was to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.

We agree with the Editor that there is no guarantee that *"these things are crucial to scientific advance in psychology"*. We have made an effort to clearly outline the "affordances" of FAIR theory (what does it make possible), but whether or not these affordances will result in actual scientific advancements remains to be seen.

Compare FAIR theory to preregistration: it, too, was introduced first as a good idea, and now - a decade later - empirical studies are being conducted to investigate its effects.

To address this comment, <font color="red">in the Revision:

* We explicitly state that we describe the affordances of FAIR theory and argue for its potential impact by comparing it to other open science methods
* We acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated.

</font>

<!-- However, our paper is a different kind of paper than an empirical study: we provide a reasoned argument to justify FAIR theory, based on philosophy of science   -->

<!--  Unlike more prescriptive norms—such as commitments to falsificationism or methodological registration—FAIR theory does not constrain a researcher’s workflow or analytical choices. It simply makes theoretical assumptions transparent, findable, and versioned. This distinction is important: the goal is not to impose rigid procedures, but to enable a more cumulative and accountable theoretical discourse. Still, we agree with the reviewer that policies and infrastructure alone are insufficient to drive meaningful change. Culture, norms, and the nature of the subject matter all shape how—and whether—reforms take root. In that spirit, we see FAIR theory not as a top-down imposition, but as a bottom-up opportunity: it offers immediate utility to researchers who want their theories to be discoverable, citable, and improvable. Over time, its adoption may well depend not on mandates, but on whether it proves valuable enough to become a “no-brainer” for the psychological sciences as it has in fields like genetics. -->

## Reviewer: 1

```{asis}
In this manuscript, the authors introduce a framework for FAIR theory specification while outlining what makes a theory FAIR, which is accompanied by practical guidance on how to implement this in one’s own work. In my opinion, the paper is relevant for the journal and presents research that is significant for the field of psychology but also adjacent fields that rely on theories similar in nature to ours. FAIR theory is a promising idea that could significantly advance psychological research, especially in terms of transparency and replicability.
```

```{asis}
The abstract provides a (mostly) comprehensive overview of the paper that can be easily followed and understood. I personally felt that the introduction of the theorytools R-package was an important aspect of the paper and should thus briefly be mentioned in the abstract.
```

```{asis}
In the abstract, the authors mention the discussion of “FAIR theories’ potential impact in terms of reducing research waste,” which I couldn’t really find at any point in the manuscript.
```

<!-- AB: -->
3.	Regarding missing reference to "research waste" - yes, this should be picked up. By the way, the latest revision of the Declaration of Helsinki states that ethical science has to avoid "research waste". Pretty cool.

```{asis}
Lines 48–50: I would have benefited from a few more words on the FAIR guiding principles (what they are, where/how they originated, and how they are used).
```

```{asis}
Line 73 onwards: The example of SDT nicely illustrates the authors’ point here!
```

```{asis}
Line 101: In Figure 1a, it would be helpful to see the two phases (currently, those are not clear to me). Also, even though Figure 1c is addressed at a much later point in the manuscript, it would be helpful to briefly mention it here as well, so it doesn’t seem forgotten.
```

```{asis}
Line 102: The corresponding year should be added to “Wagenmakers and colleagues.”
```

```{asis}
Line 147: In the case of “Lamprecht and colleagues,” the year should also be added.
```

```{asis}
In general, I’d move this section up, as the word “theory” has been used a lot already at this point, and readers might benefit from an earlier explanation/definition (for example, directly after the section “The Need for FAIR Theory”).
```

<!-- AB: -->
4.	What is a theory? (Suggestion to move up the section) It seems to always come back to this that readers a confused because everyone has their own idea of what a theory is. We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful 🙂

```{asis}
Line 163: Following APA 7 standards, no “cf.” is added. In the case of three or more authors, it is shortened to “et al.”
```

```{asis}
Line 207: It could be helpful to add a very brief description of the phonological loop in case some readers are not familiar with it and its parameters.
```

```{asis}
Line 350: I think that this was supposed to be displayed as a link: <www.theorymaps.org>
```

```{asis}
Line 390: In the case of a direct quotation, a page number should be added if possible.
```

```{asis}
This is a very nice section; I really enjoyed reading this more practical angle!
```

```{asis}
Line 688: I’d add the year for Van Lissa’s specification.
```

```{asis}
Line 776 onwards: The proposed system of modular publishing seems to have been around for quite some time (since 1998 at least) but is not widely used, which could briefly be addressed as a limitation.
```

```{asis}
Line 800: To make the strengths understandable to all readers, I’d encourage the authors to briefly explain the “open empirical cycle.”
Overall, this is a very “strong” section indeed!
```

```{asis}
In my opinion, it is very important that the learning curve is mentioned; even though I am a big fan of the proposed framework in general, I do feel like it would take some practice and getting used to.
```

```{asis}
Line 852: I’d add a reference for the jingle-jangle fallacy.
```

```{asis}
The affiliations could be changed to follow a common system.
```

We tried to make this suggested change, although there is still some heterogeneity because our different institutions have different preferred / official ways to write the affiliation.

```{asis}
I do understand why the words Findable, Accessible, Interoperable, and Reusable are always written with a majuscule, but it did confuse me a bit during reading at times.
```

Since all of these words have a colloquial meaning in everyday language, as well as a more restrictive operationalization in the literature on the FAIR principles, we think its important to signal the distinction between the formal and colloquial use of the terms. We now clarify this in a footnote, the first time the capitalized terms are introduced:

`r revise::get_revision("faircap")`

```{asis}
The references should use sentence case following APA 7 (e.g., Back to Basics: The Importance of Conceptual Clarification in Psychological Science → Back to basics: The importance of conceptual clarification in psychological science).
```

```{asis}
I’ve mentioned these two already for specific examples, but I’d check them in the whole manuscript again: In the case of more than two authors, use of et al. and the mention of page numbers for direct quotations.
```

```{asis}
Two of the links did not work for me: Line 490 (FAIR metadata example), Line 606 (example of metadata).
```

## Reviewer: 2

```{asis}
Editors generally prefer that reviewers not preface their reviews with declarations of their lack of expertise in the relevant subjects, but I am going to allow myself to do so here, because I already said as much to the editor when I was asked to review the paper. I have no particular expertise in meta-science; I have never (I am somewhat embarrassed to admit) so much as pre-registered an analysis. I am 71, and open science procedures are the blinking 12:00 on my VCR.

I am here as a practitioner of theory, not as a meta-theorist. I have written a few "theoretical" papers over the years, of different kinds that I will get to in a minute.  I want to consider how adopting a FAIR framework would change, improve, or inhibit what I do.

I can think of two kinds of theoretical papers I have written, or perhaps better, two theoretical research programs I have participated in.  One of them is closely tied to empirical results. I am thinking of the Scarr-Rowe hypothesis, something that was developed by Sandra Scarr and David Rowe, and then turned into something resembling a theory by our 2003 paper.   Since then it has been much investigated and argued about.

In this case, I think the FAIR framework would be very helpful. The S-R hypothesis makes a number of empirical claims with respect to theory. These claims have changed in response to empirical results and theoretical disputes over the years.  I have often been frustrated by the difficulty of maintaining meaningful boundaries on the theory and the predictions it makes.
```

```{asis}
Here are two examples. In the months after our paper was published, someone (Nagoshi, C. T., & Johnson, R. C. (2005). Socioeconomic status does not moderate the familiality of cognitive abilities in the Hawaii Family Study of Cognition. Journal of Biosocial Science, 37(6), 773-781.) shot out a "failure to replicate." But it wasn't-- their analysis studied parent-child correlations, not twins, and the S-R hypothesis makes no predictions about them. But despite 20 years of trying to make this clear in the published literature, the study is still cited. An even worse example happened recently, when a group of "race scientists" (Pesta, B. J., Kirkegaard, E. O., te Nijenhuis, J., Lasker, J., & Fuerst, J. G. (2020). Racial and ethnic group differences in the heritability of intelligence: A systematic review and meta-analysis. Intelligence, 78, 101408.) used the S-R hypothesis as a platform for a bunch of unrelated claims about race differences.
```

```{asis}
To be clear, it cuts both ways. It has turned out over the years that the S-R interaction does not occur in Europe, and I have revised the theory to fit the new facts, but I have done so on a strictly ad hoc basis. It would be easy to suggest that I am adding epicycles to it, and having a formalized structure for the theoretic revisions would help keep me honest.  Similarly, when conducting new analyses, I am often confronted with investigator degrees of freedom (should we use parental education or income as an SES indicator) that are very difficult to resolve in the privacy of the lab.  Having a structured platform for doing so would help. (This sounds like pre-registration. I don't think there is a bright line between FAIR theorizing and preregistering.)
```

Discuss relationship betwene FAIR theory and prereg? That sounds like a job for Aaron!

```{asis}
The second kind of theorizing I have done is more philosophical, along the lines of my "Three Laws" paper and related work. Here, I think applying the FAIR framework would be a little more problematic. At times, the current manuscript sounds as though it is trying to make scientific theory machine readable, an idea that has some good points that the authors make clear, but would also eliminate a lot of interesting papers that have been written by human beings from a subjective point of view. What would happen to Meehl's great Sir Karl Sir Ronald paper within FAIR? How much of the importance of that paper would remain if you took Meehl's voice out of it? Is all narrative philosophy "theorizing" that could be augmented or replaced by FAIR? See also: Meehl on "Cliometric metatheory".
```

Discuss relationship between FAIR and prose/author's voice.

```{asis}
These considerations lead me to some recommendations for strengthening the manuscript. The first is to clarify the relationship between FAIR theory and traditional paper-writing. Is the former meant to replace or enhance the latter? It seems to me that using FAIR to enhance the environment created by traditional theoretical papers is much the better option.  That is, everything that currently exists about the S-R hypothesis could continue to exist, and be supplemented by a FAIR framework outlining the theory and its implications. Development of the FAIR model could exist almost independently of the individual papers, analyses, reviews and meta-analyses (see below). Someone could contribute to the literature on a subject by developing a FAIR framework, much as a good review paper does.
```

Yes, this is absolutely how we intended it. Clarify this.

```{asis}
Second, and speaking of Meehl, many of the ideas incorporated here have a relationship to construct validity.  Many of our "theories" of behavior take the form of hypothetical constructs, like intelligence or for that matter the S-R interaction. Much of what is described here seems like a formalization of Meehl's idea of a nomological net-- a network of theoretical connections among abstract entities and empirical observations.
```

This is not exactly true; while that's definitely in line with *my* usage of FAIR theory, other uses are possible.

```{asis}
Finally, and this is the only concern about the paper as a publishable manuscript, the last third of the paper where the authors get into the technical aspects of how one would go about building a FAIR theory, is difficult to read. This is where the manuscript reads more like a software manual than a perspectives paper. It would seem out of place in this journal, and frankly it left me discouraged about the prospect of trying this on my own. In its place, I would recommend working an example for an actual theory, with as much of the specific detail as possible removed from the main text.
```

## Reviewer: 3

```{asis}
I have read through the manuscript on FAIR theories in psychology. I am extremely sympathetic with the arguments in the paper and believe it will make a valuable contribution.

Drawing on previous work on FAIR principles, the authors of this manuscript argue that psychological theories should be findable, accessible, interoperable, and reusable. These principles are clearly part of a broad movement toward improved and open science in disciplines like psychology.

The authors point out that, despite many reforms since the identification of the "replication crisis" meant to improve scientific rigor (e.g. per-registration), the problems lie with deeper root causes, including a problem with theory development. I could not agree more. For example, there is often little connection between theory and hypotheses. In addition, theories are often so ambiguous as to be unfalsifiable. The key issue, according to the authors, lies in where we should focus reform efforts. To date, reform efforts have focused on improving deductive methods, while overlooking the process of theory construction and improvement. I completely agree.

In the rest of the paper, the authors lay out specific aspects of their argument (e.g. what is a theory? what is the role of formalization in theory development? what is modular publishing? what role should version control play in theory development? etc.). They then discuss a specific example of how to implement the FAIR principles in theory construction and development.

As should be obvious, I am a fan of this paper.
```

```{asis}
That said, I wonder if the paper could be improved with a more probing discussion section. If I had one criticism of the paper, I would say that it lets scientists off the hook too easily. An uncritical reader might conclude that the issue is merely one of creating the right set of norms (e.g. version control through Git and archiving through Zenodo). I am doubtful. The sciences (especially physics) have not run into the same problems as psychology and yet they have not always nor always currently practice anything like these FAIR practices. Sure, physics tends to use formal theory, which limits some of the most egregious issues associated with the replication crisis. But that is not true of most theory in biology --- and yet biology has done a better of job of good theory development. I wonder if the authors would be willing to talk more in the discussion about the role of values? We live in a world that is dominated by the logic of incentives: If we just get the incentives right, then we can solve all of our problem. Perhaps. This logic suggests scientists are motivated by mere self-interest (why else would we need the incentives?) and operate in a world of ignorance (e.g. scientists are not aware that their theories are vague and why that is bad for science). I accept that both of these are true, to a large degree. If so, I worry that creating new incentives may only exacerbate problems. It seems to me that we need a change in values. For example, scientists should be embarrassed by publishing vague and ambiguous theories---at least after that has been pointed out to them. I am not trying to be a scold here, but I wonder if there's value in talking about how we might better cultivate norms and values in scientists without requiring more procedures and incentives. After all, why have the problems in the social sciences not so badly infected the sciences? The authors address these issues to some degree (e.g. they mention that they are often asked "who owns a theory?" during seminars. This is precisely the problem! That that question would be asked suggests a deeper problem than incentives---the problem lies in not really understanding what science is or how it works.

Sorry for the rant! I really loved the paper and think it is fine as is. I would love to see a bit more of a discussion about issues like values, but I don't need to see that.
```

Not sure what to do with this! I'm not 100% convinced that the disciplinary differences as the Reviewer describes exist, and I personally do not take a values based approach - I am more interested in "how to" improve social scientific research, and once I know the answer, I am intrinsically motivated to adopt that procedure. I guess I believe that most scholars are doing what they think is best given what they know, and I think that by exposing them to new and better ideas, they will change their ways. I don't see a need to moralize.

\clearpage

# References {-}

::: {#refs custom-style="Bibliography"}
:::

### 2025-06-06 manuscript.Rmd
update 2025-06-06 10:16:10.132865

@@
   - id            : "1"
-    institution   : "Tilburg University, dept. Methodology & Statistics"
+    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
   - id            : "2"
@@
   - id            : "5"
-    institution   : "University of Amsterdam, the Netherlands"
+    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
   - id            : "6"
-    institution   : "Ludwig-Maximilians-Universität München, Germany"
+    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
   - id            : "7"
-    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
+    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
 authornote: |
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
+The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers^[[As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.]{#faircap}].
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
@@
 compliant with the FAIR principles.
-FAIR theory has the potential to improve the efficiency of scholarly communication and
+[FAIR theory has the potential to improve the efficiency of scholarly communication and]{#test2}
 accelerate cumulative knowledge acquisition.
@@
 ```{r tabmeehl}
-data.frame(
+tab_meehl <- data.frame(
   "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
   "X-Interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
-) |>
-papaja::apa_table()
+)
+write.csv(tab_meehl, "tab_meehl.csv")
+papaja::apa_table(tab_meehl)
 ```

### 2025-06-06 manuscript.tex
update 2025-06-06 10:16:10.132865

@@
 \makeatother
-\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9211}
+\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9245}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 7b710ba5.
+This is a preprint paper, generated from Git Commit \# 7d4ab9cc.
 The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept.
-Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Max Planck School of Cognition, Leipzig, Germany\\\textsuperscript{5} University of Amsterdam, the Netherlands\\\textsuperscript{6} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{7} Department of Psychology, MSB Medical School Berlin, Berlin, Germany}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University department of Methodology \& Statistics, Tilburg, The Netherlands\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Max Planck School of Cognition, Leipzig, Germany\\\textsuperscript{5} University of Amsterdam, Amsterdam, The Netherlands\\\textsuperscript{6} Ludwig-Maximilians-Universität München, München, Germany\\\textsuperscript{7} MSB Medical School Berlin Department of Psychology, Berlin, Germany}
 
@@
 D.
-Wilkinson et al., 2016) for both humans and computers.
+Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{testone}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
@@
 compliant with the FAIR principles.
-FAIR theory has the potential to improve the efficiency of scholarly communication and
+\phantomsection\label{test2}{FAIR theory has the potential to improve the efficiency of scholarly communication and}
 accelerate cumulative knowledge acquisition.

### 2025-06-06 manuscript.Rmd
update 2025-06-06 10:16:10.132865

@@
     role:
+      - Conceptualization
       - Formal Analysis
@@
 to select control variables for causal inference [@cinelliCrashCourseGood2022],
-or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
 or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
@@
 For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
+
+An important aspect of Interoperability lies in linking FAIR theories with existing open science methodologies---most notably, preregistration.
+As noted, preregistration alone has not resolved the replication crisis in the social sciences.
+This raises a fundamental question: What additional value does a FAIR theory provide, and how does it complement preregistration?
+
+One distinction is quantitative:
+FAIR theories operate at a higher level of abstraction.
+They articulate general principles and expectations that cut across studies, whereas preregistrations focus on specific applications, analytic strategies, and study designs.
+More fundamentally, however, there is a qualitative difference.
+A FAIR theory is designed to make predictions and be revised in light of empirical results; preregistration, by contrast, prescribes how a particular study will be conducted, regardless of what the outcomes turn out to be.
+
+This difference can be illustrated with a Bayesian analogy.
+A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
+Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data [@peikertWhyDoesPreregistration2023].
+However, after observing data, the theory might require revision.
+In this sense, a theory is first a prior, and then after conducting a specific investigation as prescribed by the preregistration, is refined to become a posterior.
+In short, a preregistration rooted in a FAIR theory not only supports transparency, but also strengthens the epistemic cycle of prediction, testing, and revision---moving us closer to a cumulative science.
 

### 2025-06-06 manuscript.Rmd
add first draft for connection to prereg

--- 2025-03-19 manuscript.Rmd
@@
 For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
+
+An important aspect of Interoperability lies in linking FAIR theories with existing open science methodologies---most notably, preregistration.
+As noted, preregistration alone has not resolved the replication crisis in the social sciences.
+This raises a fundamental question: What additional value does a FAIR theory provide, and how does it complement preregistration?
+
+One distinction is quantitative:
+FAIR theories operate at a higher level of abstraction.
+They articulate general principles and expectations that cut across studies, whereas preregistrations focus on specific applications, analytic strategies, and study designs.
+More fundamentally, however, there is a qualitative difference.
+A FAIR theory is designed to make predictions and be revised in light of empirical results; preregistration, by contrast, prescribes how a particular study will be conducted, regardless of what the outcomes turn out to be.
+
+This difference can be illustrated with a Bayesian analogy.
+A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
+Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data.
+However, after observing data, the theory might require revision.
+In this sense, a theory is first a prior, and then after conducting a specific investigation as prescribed by the preregistration, is refined to become a posterior.
+In short, a preregistration rooted in a FAIR theory not only supports transparency, but also strengthens the epistemic cycle of prediction, testing, and revision---moving us closer to a cumulative science.
 

### 2025-06-06 manuscript.Rmd
add citation
add author role

@@
 to select control variables for causal inference [@cinelliCrashCourseGood2022],
+or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
 or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
@@
 For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].
-
-An important aspect of Interoperability lies in linking FAIR theories with existing open science methodologies---most notably, preregistration.
-As noted, preregistration alone has not resolved the replication crisis in the social sciences.
-This raises a fundamental question: What additional value does a FAIR theory provide, and how does it complement preregistration?
-
-One distinction is quantitative:
-FAIR theories operate at a higher level of abstraction.
-They articulate general principles and expectations that cut across studies, whereas preregistrations focus on specific applications, analytic strategies, and study designs.
-More fundamentally, however, there is a qualitative difference.
-A FAIR theory is designed to make predictions and be revised in light of empirical results; preregistration, by contrast, prescribes how a particular study will be conducted, regardless of what the outcomes turn out to be.
-
-This difference can be illustrated with a Bayesian analogy.
-A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
-Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data.
-However, after observing data, the theory might require revision.
-In this sense, a theory is first a prior, and then after conducting a specific investigation as prescribed by the preregistration, is refined to become a posterior.
-In short, a preregistration rooted in a FAIR theory not only supports transparency, but also strengthens the epistemic cycle of prediction, testing, and revision---moving us closer to a cumulative science.
 

### 2025-06-06 action_letter.Rmd
update 2025-06-06 18:48:11.101405

@@
 ```{r setup-chunk, include = FALSE}
+library(revise)
+revise::read_manuscript("abstract.md")
 revise::read_manuscript("manuscript.Rmd")
@@
 
+We thank the Reviewer for their thoughtful review and detailed comments. We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.
+
 ```{asis}
@@
 
+Thank you for this suggestion; as proposed, we now briefly mention the R-package in the abstract:
+
+`r get_revision("abstractpackage")`
+
 ```{asis}
@@
 ```
+
+nakagawaPoorHypothesesResearch2024
 
@@
 ```
+
+gonzalezExtrinsicConvergentValidity2021
 

### 2025-06-07 manuscript.Rmd

--- 2025-06-06 manuscript.Rmd
@@
 A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
-Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data.
+Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data [@peikertWhyDoesPreregistration2023].
 However, after observing data, the theory might require revision.

### 2025-06-07 manuscript.Rmd
update 2025-06-07 10:38:20.297695

--- 2025-06-06 manuscript.Rmd
@@
 Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
-But, as we argued, this is not the case.
+But, as we argued, contemporary practice falls short of this ideal.
 The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
@@
 to select control variables for causal inference [@cinelliCrashCourseGood2022],
-or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
+or to preregister a study with an explicit derivation chain from theory to hypothesis,
+as well as an inferential procedure that would suggest specific modifications to theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
 or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
@@
 
-One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
-it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [@bringmannBackBasicsImportance2022].
-The "jingle-jangle" fallacy is common in psychology:
-the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
-FAIR theory can help address the measurement crisis:
-since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.
+<span id="fairtheoryprereg">One important future direction is embedding FAIR theories withing existing open science methodologies. 
+For example, consider how FAIR theory relates to preregistration.
+These practices are clearly distinct but complimentary.
+The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
+and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
+The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes [@peikertWhyDoesPreregistration2023].
+These practices differ in their level of abstraction:
+FAIR theories cut across studies on a given phenomenon,
+whereas preregistrations are specific implementations of hypothesis tests,
+with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline. 
+These practices complement each other:
+authors can make the derivation chain from theory to hypothesis more explicit by citing a specific FAIR theory in their preregistration.
+Moreover, it is possible to preregister an inferential procedure that would require revising the theory after observing data,
+or even to have proponents and detractors of a theory review a registered report of such a test.
+In short, combining FAIR theory with preregistration and other existing open science practices has the potential to strengthen the epistemic cycle of prediction, testing, and revision, moving us closer to a cumulative science.</span>
+
+Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
+Since FAIR theories can reference other theories and resources,
+it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
+<span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,
+which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct [@songLiteratureReviewingAddressing2021].
+By explicitly referencing operational definitions in FAIR theories,
+such jingle-jangle fallacies would come to light and could ultimately be resolved.</span>
 

### 2025-06-07 action_letter.Rmd
update 2025-06-07 12:47:56.764452

@@
 
-As methodologists, our focus in this paper was on "procedures" rather than policies or documentation.
-Our goal was to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.
+As methodologists, our focus in this paper is on "procedures", rather than policies or documentation.
+Our goal is to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.
 
 We agree with the Editor that there is no guarantee that *"these things are crucial to scientific advance in psychology"*.
-We have made an effort to clearly outline the "affordances" of FAIR theory (what does it make possible), but whether or not these affordances will result in actual scientific advancements remains to be seen.
+We have, however, provided reasoned arguments to support the "affordances" of FAIR theory: it *makes possible* particular scientific advancements, even if it does not guarantee them.
 
-Compare FAIR theory to preregistration: it, too, was introduced first as a good idea, and now - a decade later - empirical studies are being conducted to investigate its effects.
+Compare FAIR theory to preregistration: it, too, was introduced first as a good idea; now - a decade later - empirical studies are being conducted to investigate its effects.
 
@@
 
+We removed the "cf.", although we couldn't find mention of it in the APA7 style guide. We updated the citation scheme to APA7, so three or more authors will be shortened to "et al.".
+
 ```{asis}
@@
 
+Thank you for pointing this out; we have fixed the formatting.
+
 ```{asis}
@@
 ```
+
+We have added the page numbers to all direct quotations.
 
@@
 
-gonzalezExtrinsicConvergentValidity2021
+We have done as requested; this section now reads:
+
+`r get_revision("jinglejangle")`
 
@@
 
-We tried to make this suggested change, although there is still some heterogeneity because our different institutions have different preferred / official ways to write the affiliation.
+We have tried to make this requested change, although there is some residual heterogeneity because our institutions have different preferred / mandated ways to write the affiliation.
 
@@
 ```
+
+We've updated the reference style to APA7, but note that some journals have fixed the case of specific words in titles (e.g., "FAIR" is fixed to capital letters).
+These remain unchanged.
+We defer to the Editor to indicate whether we should change this, or whether it will be addressed in the typesetting process.
 

### 2025-06-07 manuscript.Rmd
update 2025-06-07 12:47:56.764452

@@
 urlcolor          : blue
+csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
+documentclass     : "apa7"
 output            :
@@
 However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
+
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-
 Psychological theories rarely provide hypotheses that are corroborated by evidence.
 Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
-as the theory's central claims evade falsification.
-A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
-Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
-Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
+rendering them immune to falsification.
+As an example, consider "self-determination theory" (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
+Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness [@ryanSelfdeterminationTheoryFacilitation2000].
 <!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
@@
 -->
-The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
-Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
-Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
+The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
+Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
+Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
 The expanded definition thus enables SDT to absorb potentially falsifying evidence.
@@
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p.
+1].
 The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.
@@
 statements about the functional form of those connections,
-and statements about their specific numerical values [cf. @frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
+and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
 The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
@@
 The term "model" is not uniquely defined within the literature;
-it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
+models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p.
+336].
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and posted many examples on their website <www.theorymaps.org> [@grayHowMapTheory2017].
+and posted many examples on [their website](https://www.theorymaps.org) [@grayHowMapTheory2017].
 Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
@@
 and pertains to a theory's intelligibility to those with practical Access.
-It has been proposed that good theories have the property of "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
+It has been proposed that good theories have the property of "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024, p. 1]. 
 At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
@@
 
-Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016].
+Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 The concept of X-Interoperability was introduced to answer the question: *Interoperable for what?*
-X-Interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-Interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
+X-Interoperability is defined as facilitating "successful communication between machines and between humans and machines [, where] A and B are considered X-Interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024, p.
+5].
 This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
@@
 If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
-*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
+*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008, p. 1].
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologyFoundationsInference1969].
@@
 A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy [@bently2010copyright].
-It explicitly does not
-*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
+Copyright does not
+*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"* (Section 102, Copyright Act).
 Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
@@
 As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*.
+The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
 The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.

### 2025-06-07 manuscript.tex
update 2025-06-07 12:47:56.764452

@@
 \documentclass[
-  man, noextraspace,floatsintext]{apa6}
+  man, noextraspace,floatsintext]{apa7}
 \usepackage{amsmath,amssymb}
@@
 \makeatother
-\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9245}
+\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9491}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 7d4ab9cc.
+This is a preprint paper, generated from Git Commit \# 5d7ad1a4.
 The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
@@
 Caspar J.
-Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S.
+Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Conceptualization, Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S.
 Ernst: Writing -- review \& editing, Investigation; Noah N.N.
@@
 FAIR theories are Findable in well-established archives, Accessible in terms of availability and their ability to be understood, Interoperable for specific purposes, such as selecting control variables, and Reusable so that they can be iteratively improved through collaborative efforts.
-This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary psychological practices, introduces a workflow for FAIRifying theory, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
-FAIR theory constitutes framework for specifying and refining theories, bridging a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond.
+\phantomsection\label{abstractpackage}{This paper adapts the FAIR principles for theory, reflects on FAIR practices in contemporary psychology, introduces a workflow for FAIRifying theory, which is largely automated by the \texttt{theorytools} R-package}, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
+FAIR theory constitutes a protocol for archiving and communicating about theory, addressing a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond.
 FAIR theory builds on existing open science principles and infrastructures to provide a structured, cumulative framework for theory development, potentially increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
@@
 D.
-Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{testone}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
+Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{faircap}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
 Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
@@
 These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
-However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
+However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, et al., 2021).
+
 Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
-
 Psychological theories rarely provide hypotheses that are corroborated by evidence.
 Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
-as the theory's central claims evade falsification.
-A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
-Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
-Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
-The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
-Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
-Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
+rendering them immune to falsification.
+As an example, consider ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
+Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness (Ryan \& Deci, 2000).
+The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
+Under the original definition, one would hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
+Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
 The expanded definition thus enables SDT to absorb potentially falsifying evidence.
 
-Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
+Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh et al., 2021).
 One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
@@
 Firstly, because hypothesis-testing research is over-represented in the literature:
-According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
+According to Kühberger et al. (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
+Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, et al., 2021).
 Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
@@
 Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
-Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
+in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi et al., 2024).
+Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer et al., 2004),
 but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
@@
 The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978, p.
+1).
 The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.
@@
 statements about the functional form of those connections,
-and statements about their specific numerical values (cf. Frankenhuis, Panchanathan, \& Smaldino, 2023; Guest, 2024).
+and statements about their specific numerical values (Frankenhuis et al., 2023; Guest, 2024).
 The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
@@
 The term ``model'' is not uniquely defined within the literature;
-it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020).
+models have been described as ``specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020, p. 336).
 This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
@@
 Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
-with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
+with adequate metadata that is indexed in standardized repositories (Van De Sompel et al., 2004).
 Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
@@
 It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
-Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
+Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert et al., 2021; Van Lissa et al., 2021).
 Git provides a useful framework for developing FAIR theory,
@@
 This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
-It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco, Uggerslev, \& Steel, 2017),
-ontology (Guyon, Kop, Juhel, \& Falissard, 2018),
+It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs (Bosco et al., 2017),
+ontology (Guyon et al., 2018),
 or catalog of psychological phenomena.
@@
 For example, Gray and colleagues introduced a format for representing theories,
-and posted many examples on their website \textless www.theorymaps.org\textgreater{} (Gray, 2017).
+and posted many examples on \href{https://www.theorymaps.org}{their website} (Gray, 2017).
 Similarly, \href{http://psychomodels.org/}{PsychoModels} seeks to inventorize theories and models in psychology (van Dongen \& Volz, 2025).
@@
 Making theories Accessible also allows stakeholders (e.g., practitioners, policymakers, advocates) to inform themselves of the current scientific understanding of specific phenomena.
-While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet, Smith, Boraud, \& Gonon, 2017),
+While isolated empirical findings can appear fragmented and contradictory (Dumas-Mallet et al., 2017),
 theories offer a top-down, big-picture representation of the phenomena studied in a field.
@@
 and pertains to a theory's intelligibility to those with practical Access.
-It has been proposed that good theories have the property of ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024).
+It has been proposed that good theories have the property of ``discursive survival {[}\ldots{]}, the ability to be understood'' (Guest, 2024, p. 1).
 At present, psychological theories are often ambiguous, rendering them difficult to understand (Frankenhuis et al., 2023).
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J.
-Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).
+and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J. Lange et al., 2025).
 
@@
 
-Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016).
+Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016, p. 2).
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 While a theory represented as a bitmap image is not very Interoperable,
-the same image represented in the DOT language ({``{DOT Language},''} 2024)
+the same image represented in the DOT language (\emph{{DOT Language}}, 2024)
 for representing graphs does meet this ideal (an example of such a DOT representation is given below).
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by cross-referencing the DOI of those nested theories ({``Contributing {Citations} and {References},''} 2024).
+This can be achieved by cross-referencing the DOI of those nested theories (\emph{Contributing {Citations} and {References}}, 2024).
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
@@
 The concept of X-Interoperability was introduced to answer the question: \emph{Interoperable for what?}
-X-Interoperability is defined as facilitating ``successful communication between machines and between humans and machines {[}, where{]} A and B are considered X-Interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024).
+X-Interoperability is defined as facilitating ``successful communication between machines and between humans and machines {[}, where{]} A and B are considered X-Interoperable if a common operation X exists that can be applied to both'' (Vogt et al., 2024, p.
+5).
 This revised definition makes it possible to outline a theory's affordances in terms of X-Interoperability.
@@
 Kurt Lewin's (1943) adage ``there's nothing as practical as a good theory'' paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
-But, as we argued, this is not the case.
+But, as we argued, contemporary practice falls short of this ideal.
 The examples of X-Interoperability offered in Table \ref{tab:tabmeehl} illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
 For example, Interoperable theory could be used
-to select control variables for causal inference (Cinelli, Forney, \& Pearl, 2022),
-or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data (Peikert et al., 2021),
+to select control variables for causal inference (Cinelli et al., 2022),
+or to preregister a study with an explicit derivation chain from theory to hypothesis,
+as well as an inferential procedure that would suggest specific modifications to theory after analyzing empirical data (Peikert et al., 2021),
 or to derive machine-readable hypotheses (Lakens \& DeBruine, 2021) which could be automatically evaluated through integration testing (Van Lissa, 2023).
 Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
-For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris, Silk, Steinberg, Myers, \& Robinson, 2007).
+For example, it should be possible to embed a theory about emotion regulation (e.g., Gross, 2015) within a theory of emotion regulation development (Morris et al., 2007).
 
@@
 Two considerations are important when determining what license is appropriate for theory.
-A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy (Bently, Davis, \& Ginsburg, 2010).
+A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy (Bently et al., 2010).
 It explicitly does not
@@
 V.
-Lange, W.Kruglanski, ToryHiggins, Deci, \& Ryan, 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
-Another vignette describes how to take the Dunning-Krüger effect and \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{specify it as a FAIR mathematical formula} (Feld, Sauermann, \& de Grip, 2017).
+Lange et al., 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
+Another vignette describes how to take the Dunning-Krüger effect and \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{specify it as a FAIR mathematical formula} (Feld et al., 2017).
 Another vignette illustrates the use of FAIR theory for \href{https://cjvanlissa.github.io/theorytools/articles/causal-inference.html}{covariate selection and causal inference} (Pearl, 1995).
@@
 
-Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck, Ryan, Robinaugh, Waldorp, \& Borsboom, 2021).
+Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck et al., 2021).
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
@@
 The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
-Another limitation of scope is that FAIR theory does not resolve other related issues in psychology, such as the measurement crisis (Bringmann, Elmer, \& Eronen, 2022) and lack of standardized ontologies for psychological constructs (Bosco et al., 2017).
+Another limitation of scope is that FAIR theory does not resolve other related issues in psychology, such as the measurement crisis (Bringmann et al., 2022) and lack of standardized ontologies for psychological constructs (Bosco et al., 2017).
 However, our work here provides a template for addressing such challenges,
@@
 
-One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
-Aside from the aforementioned ``theory crisis'', there has been talk of a ``measurement crisis'':
-it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties (Bringmann et al., 2022).
-The ``jingle-jangle'' fallacy is common in psychology:
-the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
-FAIR theory can help address the measurement crisis:
-since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.
+\phantomsection\label{fairtheoryprereg}{One important future direction is embedding FAIR theories withing existing open science methodologies.
+For example, consider how FAIR theory relates to preregistration.
+These practices are clearly distinct but complimentary.
+The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
+and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
+The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes (Peikert et al., 2023).
+These practices differ in their level of abstraction:
+FAIR theories cut across studies on a given phenomenon,
+whereas preregistrations are specific implementations of hypothesis tests,
+with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline.
+These practices complement each other:
+authors can make the derivation chain from theory to hypothesis more explicit by citing a specific FAIR theory in their preregistration.
+Moreover, it is possible to preregister an inferential procedure that would require revising the theory after observing data,
+or even to have proponents and detractors of a theory review a registered report of such a test.
+In short, combining FAIR theory with preregistration and other existing open science practices has the potential to strengthen the epistemic cycle of prediction, testing, and revision, moving us closer to a cumulative science.}
+
+Another future direction is the intersection between the aforementioned ``theory crisis'' and the related ``measurement crisis'' pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs (Bringmann et al., 2022).
+Since FAIR theories can reference other theories and resources,
+it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
+\phantomsection\label{jinglejangle}{FAIR theories can also help address ``jingle- and jangle fallacies'' in psychology,
+which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct (\textbf{songLiteratureReviewingAddressing2021?}).
+By explicitly referencing operational definitions in FAIR theories,
+such jingle-jangle fallacies would come to light and could ultimately be resolved.}
 
@@
 Aalbersberg, Ij.
-J., Appleyard, T., Brookhart, S., Carpenter, T., Clarke, M., Curry, S., \ldots{} Vazire, S.
+J., Appleyard, T., Brookhart, S., Carpenter, T., Clarke, M., Curry, S., Dahl, J., DeHaven, A.
+C., Eich, E., Franko, M., Freedman, L., Graf, C., Grant, S., Hanson, B., Joseph, H., Kiermer, V., Kramer, B., Kraut, A., Karn, R.
+K., \ldots{} Vazire, S.
 (2018).
@@
 \bibitem[\citeproctext]{ref-ContributingCitationsReferences2024}
-Contributing {Citations} and {References}. (2024). Retrieved December 5, 2024, from \url{https://support.datacite.org/docs/data-citation}
+\emph{Contributing {Citations} and {References}}. (2024). DataCite Support. \url{https://support.datacite.org/docs/data-citation}
 
@@
 \bibitem[\citeproctext]{ref-DOTLanguage2024}
-{DOT Language}. (2024). Retrieved December 5, 2024, from \url{https://graphviz.org/doc/info/lang.html}
+\emph{{DOT Language}}. (2024). Graphviz. \url{https://graphviz.org/doc/info/lang.html}
 
@@
 \bibitem[\citeproctext]{ref-ehmkeContributorCovenantCode2014}
-Ehmke, C. (2014). Contributor {Covenant}: {A Code} of {Conduct} for {Open Source} and {Other Digital Commons Communities}. Retrieved April 3, 2024, from \url{https://www.contributor-covenant.org/}
+Ehmke, C. (2014). \emph{Contributor {Covenant}: {A Code} of {Conduct} for {Open Source} and {Other Digital Commons Communities}}. \url{https://www.contributor-covenant.org/}
 
@@
 391--408).
-Thousand Oaks: Sage.
+Sage.
 
@@
 \emph{Journal of Crime and Justice}.
-Retrieved from \url{https://www.tandfonline.com/doi/abs/10.1080/0735648X.2008.9721251}
+\url{https://www.tandfonline.com/doi/abs/10.1080/0735648X.2008.9721251}
 
@@
 \bibitem[\citeproctext]{ref-kuhnStructureScientificRevolutions2009}
-Kuhn, T. S. (2009). \emph{The structure of scientific revolutions} (3. ed., {[}Nachdr.{]}). Chicago: Univ. of Chicago Press.
+Kuhn, T. S. (2009). \emph{The structure of scientific revolutions} (3. ed., {[}Nachdr.{]}). Univ. of Chicago Press.
 
@@
 91--136).
-Dordrecht: Springer Netherlands.
+Springer Netherlands.
 \url{https://doi.org/10.1007/978-94-010-3142-4_7}
@@
 \bibitem[\citeproctext]{ref-lamprechtFAIRPrinciplesResearch2019}
-Lamprecht, A.-L., Garcia, L., Kuzak, M., Martinez, C., Arcila, R., Martin Del Pico, E., \ldots{} Capella-Gutierrez, S.
+Lamprecht, A.-L., Garcia, L., Kuzak, M., Martinez, C., Arcila, R., Martin Del Pico, E., Dominguez Del Angel, V., van de Sandt, S., Ison, J., Martinez, P.
+A., McQuilton, P., Valencia, A., Harrow, J., Psomopoulos, F., Gelpi, J.
+Ll., Chue Hong, N., Goble, C., \& Capella-Gutierrez, S.
 (2019).
@@
 \bibitem[\citeproctext]{ref-langeChecklistIncentivizingFacilitating2025a}
-Lange, J., Freyer, N., Musfeld, P., Schönbrodt, F., \& Leising, D. (2025, January 10). A checklist for incentivizing and facilitating good theory building. \url{https://doi.org/10.31219/osf.io/7qvfz}
+Lange, J., Freyer, N., Musfeld, P., Schönbrodt, F., \& Leising, D.
+(2025, January 10).
+\emph{A checklist for incentivizing and facilitating good theory building}.
+\url{https://doi.org/10.31219/osf.io/7qvfz}
 
@@
 B., Chow, E.
-C., Gonzalez-Marquez, P., Loukras, E., \ldots{} Volodko, K.
+C., Gonzalez-Marquez, P., Loukras, E., Maus, A., O'Mahony, A., Pomareda, C., Primbs, M.
+A., Sackman, S.
+L., Smithson, C.
+J.
+R., \& Volodko, K.
 (2021).
@@
 \bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
-Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}
+Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}
 
@@
 \emph{Capturing {Causal Claims}: {A Fine Tuned Text Mining Model} for {Extracting Causal Sentences} from {Social Science Papers}}.
-Retrieved from \url{https://osf.io/kwtpm/download}
+\url{https://osf.io/kwtpm/download}
 
@@
 D., Breckler, S.
-J., \ldots{} Yarkoni, T.
+J., Buck, S., Chambers, C.
+D., Chin, G., Christensen, G., Contestabile, M., Dafoe, A., Eich, E., Freese, J., Glennerster, R., Goroff, D., Green, D.
+P., Hesse, B., Humphreys, M., \ldots{} Yarkoni, T.
 (2015).
@@
 \bibitem[\citeproctext]{ref-peikertTransparencyOpenScience2023}
-Peikert, A. (2023). \emph{Towards {Transparency} and {Open Science}} (doctoralThesis, Humboldt-Universität zu Berlin). Humboldt-Universität zu Berlin. \url{https://doi.org/10.18452/27056}
+Peikert, A. (2023). \emph{Towards {Transparency} and {Open Science}} {[}doctoralThesis, Humboldt-Universität zu Berlin{]}. \url{https://doi.org/10.18452/27056}
+
+\bibitem[\citeproctext]{ref-peikertWhyDoesPreregistration2023}
+Peikert, A., Ernst, M.
+S., \& Brandmaier, A.
+M.
+(2023, February 17).
+\emph{Why does preregistration increase the persuasiveness of evidence?
+{A Bayesian} rationalization} {[}Preprint{]}.
+\url{https://doi.org/10.31234/osf.io/cs8wb}
 
@@
 \bibitem[\citeproctext]{ref-peirceCollectedPapersCharles1960}
-Peirce, C. (1960). \emph{Collected {Papers Of Charles Peirce}}. Retrieved from \url{http://archive.org/details/collected-papers-of-charles-peirce}
+Peirce, C. (1960). \emph{Collected {Papers Of Charles Peirce}}. \url{http://archive.org/details/collected-papers-of-charles-peirce}
 
 \bibitem[\citeproctext]{ref-popperLogicScientificDiscovery2002}
-Popper, K. R. (2002). \emph{The logic of scientific discovery}. Retrieved from \url{http://public.ebookcentral.proquest.com/choice/publicfullrecord.aspx?p=254228}
+Popper, K. R. (2002). \emph{The logic of scientific discovery}. \url{http://public.ebookcentral.proquest.com/choice/publicfullrecord.aspx?p=254228}
 
@@
 \bibitem[\citeproctext]{ref-vandongenPsychoModelsMathematicalComputational2025}
-van Dongen, N.
-N.
-N., \& Volz, L.
-(2025).
-{PsychoModels}: {The} mathematical and computational model repository for psychological science.
-Retrieved March 5, 2025, from \url{https://www.psychomodels.org/}
+van Dongen, N. N. N., \& Volz, L. (2025). \emph{{PsychoModels}: {The} mathematical and computational model repository for psychological science.} \url{https://www.psychomodels.org/}
 
 \bibitem[\citeproctext]{ref-vanlissaUsingEndpointsCheck2023}
-Van Lissa, C. J. (2023). Using {Endpoints} to {Check Reproducibility} {[}Package Documentation{]}. Retrieved March 21, 2024, from \url{https://cjvanlissa.github.io/worcs/articles/endpoints.html}
+Van Lissa, C.
+J.
+(2023).
+\emph{Using {Endpoints} to {Check Reproducibility}} {[}Package Documentation{]}.
+worcs 0.1.15.2 package documentation.
+\url{https://cjvanlissa.github.io/worcs/articles/endpoints.html}
 
@@
 (2024, May 6).
-{FAIR} 2.0: {Extending} the {FAIR Guiding Principles} to {Address Semantic Interoperability}.
-Retrieved November 20, 2024, from \url{http://arxiv.org/abs/2405.03345}
+\emph{{FAIR} 2.0: {Extending} the {FAIR Guiding Principles} to {Address Semantic Interoperability}}.
+\url{http://arxiv.org/abs/2405.03345}
 
@@
 D., Dumontier, M., Aalbersberg, Ij.
-J., Appleton, G., Axton, M., Baak, A., \ldots{} Mons, B.
+J., Appleton, G., Axton, M., Baak, A., Blomberg, N., Boiten, J.-W., da Silva Santos, L.
+B., Bourne, P.
+E., Bouwman, J., Brookes, A.
+J., Clark, T., Crosas, M., Dillo, I., Dumon, O., Edmunds, S., Evelo, C.
+T., Finkers, R., \ldots{} Mons, B.
 (2016).
@@
 R., Kinoshita, B.
-de P., Gadelha, L., et al.others.
+de P., Gadelha, L., Garijo, D., Gustafsson, O.
+J.
+R., Juty, N., Kanwal, S., et al.
 (2024).
 \emph{Applying the {FAIR} principles to computational workflows}.
-Retrieved from \url{https://arxiv.org/abs/2410.03490}
+\url{https://arxiv.org/abs/2410.03490}
 
@@
 Metaphysics Research Lab, Stanford University.
-Retrieved from \url{https://plato.stanford.edu/archives/spr2021/entries/structure-scientific-theories/}
+\url{https://plato.stanford.edu/archives/spr2021/entries/structure-scientific-theories/}
 

### 2025-06-07 action_letter.Rmd
update 2025-06-07 14:19:27.889095

@@
 ```{asis}
-This is a very nice section; I really enjoyed reading this more practical angle!
+Line 688: I’d add the year for Van Lissa’s specification.
 ```
 
-```{asis}
-Line 688: I’d add the year for Van Lissa’s specification.
-```
+Done.
 
@@
 ```
+
+We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of articles in 2010-2019.
+Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
+
+`r get_revision("datasharing")`
 
@@
 ```{asis}
+I submit this review signed as Eric Turkheimer.
+
 Editors generally prefer that reviewers not preface their reviews with declarations of their lack of expertise in the relevant subjects, but I am going to allow myself to do so here, because I already said as much to the editor when I was asked to review the paper.
@@
 I have often been frustrated by the difficulty of maintaining meaningful boundaries on the theory and the predictions it makes.
-```
 
-```{asis}
 Here are two examples.
@@
 Intelligence, 78, 101408.) used the S-R hypothesis as a platform for a bunch of unrelated claims about race differences.
-```
 
-```{asis}
 To be clear, it cuts both ways.
@@
 Having a structured platform for doing so would help.
-(This sounds like pre-registration.
-I don't think there is a bright line between FAIR theorizing and preregistering.)
 ```
 
-Discuss relationship betwene FAIR theory and prereg? That sounds like a job for Aaron!
+Dear prof. Turkheimer,
+
+on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments.
+This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician.
+All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it.
+One author proposed to invite you as a co-author, others were concerned that this might constitute a conflict of interest.
+
+We took the initiative to draft a paragraph that summarizes your comments on FAIR theory from the review, and mention your contribution in the Acknowledgements.
+We would be equally happy to credit your contribution with a co-authorship, depending on your - and the Editor's - preferences.
+
+Our draft paragraph now reads:
+
+`r get_revision("turkheimer")`
+
+The Acknowledgement now reads:
+
+> We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
+
+```{asis}
+(This sounds like pre-registration. I don't think there is a bright line between FAIR theorizing and preregistering.)
+```
+
+We think that this comment signals a need to better explain the important distinctions between these two practices.
+To address this comment, we have included the following section in the revision:
+
+`r get_revision("fairtheoryprereg")`
 
@@
 Discuss relationship between FAIR and prose/author's voice.
+
+`r get_revision("modularpublisingdetract")`
 

### 2025-06-07 manuscript.Rmd
update 2025-06-07 14:19:27.889095

@@
 authornote: |
-  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.
+  We are particularly endebted to our Reviewer, Dr.
+Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
+We also thank the Editor, Dr.
+Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
+  This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.
 The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
@@
 
-We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
-In most fields, the primary unit of scientific communication is the academic paper.
-A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
-Modular publishing is the practice of making each of these resources available as independent citable *information artifacts*,
-with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
-Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
-Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
+We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
+The primary unit of scientific communication has long been the academic paper.
+In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
+These resources are often merely described in papers (if at all), and not made available for reuse.
+Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
+facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
+<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>
+
+Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
 which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
 To maintain a persistent record of scholarly communication,
-Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
-which is used by many academic publishers.
-Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
+Zenodo mints DOIs for information artifacts.
+<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
+<!-- which is used by many academic publishers. -->
+The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
 <!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
-For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
-and cross-reference the published paper with `relationType: IsSupplementTo`.
-Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
-Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
+For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
+If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
+Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
+Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
 the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
@@
 If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
-<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->
+
+<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication.
+</span>
 
@@
 In other words: what actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
+To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
 Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
@@
 Modular publishing capitalizes on these changing metrics,
-and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory [@kirczModularityNextForm1998].
+and FAIR theory allows scholars to be credited for theoretical contributions [@kirczModularityNextForm1998].
 Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
@@
 For example, consider how FAIR theory relates to preregistration.
-These practices are clearly distinct but complimentary.
+These practices are distinct but complimentary.
 The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
 and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
-The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes [@peikertWhyDoesPreregistration2023].
-These practices differ in their level of abstraction:
-FAIR theories cut across studies on a given phenomenon,
-whereas preregistrations are specific implementations of hypothesis tests,
-with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline. 
+The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in the outcomes of a specific empirical study [@peikertWhyDoesPreregistration2023].
+FAIR theories are specified at a level of abstraction that transcends individual studies.
+FAIR theories can inform - and be informed by - both quantitative and qualitative research.
+Preregistrations, by contrast, are specific implementations of quantitative hypothesis tests,
+within the context of a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline.
 These practices complement each other:
@@
 closing a critical gap in the scientific process.
-FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
+
+<span id = "turkheimer">We invite the reader to ask the rhetorical question raised by our reviewer, prof.
+Turkheimer, who - reflecting on a long career as a practitioner of theory in psychology - considered "how adopting a FAIR framework would change, improve, or inhibit what I do".
+We paraphrase his answer here^[Edited for brevity.], because his lived experience complements our youthful "techno-enthusiasm".
+Prof. Turkheimer remarked that a FAIR framework would have been very helpful in maintaining meaningful boundaries between a theory and the predictions it makes,
+in tracking how empirical findings result in changes to empirical claims made by a theory,
+and in resolving theoretical disputes.
+*"FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
+Everything published about a theory can continue to exist in its original form,
+while a FAIR representation supplements that literature by explicitly outlining the theory’s assumptions, logical structure, and testable implications.
+Importantly, FAIR theory can evolve semi-independently from individual papers, offering a persistent and collaborative object that others can reuse, cite, and refine. 
+Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability."*</span>
+
+We envision a future where application of the FAIR principles makes theories more *useful*, living up to Herbert Simon's ideal,
+enabling scholars to incorporate theory into their workflows in a tanglible way,
+providing explicit derivation chains for hypotheses,
+applying transparent rules to select the right control variables for causal inference,
+and proposing specific changes to existing theories based on empirical results.
 This paves the way for more theory-driven scholarship

### 2025-06-07 action_letter.Rmd
update 2025-06-07 15:14:35.120732

@@
 See also: Meehl on "Cliometric metatheory".
-```
 
-Discuss relationship between FAIR and prose/author's voice.
-
-`r get_revision("modularpublisingdetract")`
-
-```{asis}
 These considerations lead me to some recommendations for strengthening the manuscript.
@@
 
-Yes, this is absolutely how we intended it. Clarify this.
+We agree, this is exactly how we envision the role of FAIR theory.
+We think this comment can be concisely addressed by emphasizing that modular publishing does not replace the traditional paper, but adds value to resources created in the process of its writing:
+
+`r get_revision("modularvalue")`
+
+And:
+
+`r get_revision("modularpublisingdetract")`
 
@@
 
-This is not exactly true; while that's definitely in line with *my* usage of FAIR theory, other uses are possible.
+We partly agree; we recognize two distinct lines of thought in this comment.
+
+The first is that FAIR theory relates to construct validity. To address this point, we have rewritten the section on FAIR theory and the "measurement crisis":
+
+`r get_revision("measurementcrisis")`
+
+The second is that much of what we describe looks like Meehl's nomological nets.
+This is definitely true for the way that *we* are using FAIR theory (e.g., in the online examples), as we are inspired by Meehl,
+but it's not necessarily true of FAIR theory in general.
+We're not sure if further changes need to be made to address this.
 

### 2025-06-07 manuscript.Rmd
update 2025-06-07 15:14:35.120732

@@
 The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
-
 abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
@@
 Concern that the abundance of non-replicable findings 
-was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
+was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015].
 These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
@@
 Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
-but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
+but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
 Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
@@
 We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
-The primary unit of scientific communication has long been the academic paper.
+<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
 In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
@@
 Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
-facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
+facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
 <span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>
@@
 If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
-
-<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication.
+<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for argumentation,
+nuance,
+and the author's voice.
 </span>
@@
 
-Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
+<span id="measurementcrisis">Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
 Since FAIR theories can reference other theories and resources,
-it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
+it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.</span>
 <span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,

### 2025-06-07 manuscript.tex
update 2025-06-07 15:14:35.120732

@@
 \makeatother
-\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9491}
+\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9794}
 \usepackage{lineno}
@@
 
-This is a preprint paper, generated from Git Commit \# 5d7ad1a4.
-The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
+We are particularly endebted to our Reviewer, Dr.~Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
+We also thank the Editor, Dr.~Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
 
@@
 
-We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
-In most fields, the primary unit of scientific communication is the academic paper.
-A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
-Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
-with adequate metadata that is indexed in standardized repositories (Van De Sompel et al., 2004).
-Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
-Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
+We envision FAIR theory as an instance of modular publishing (Kircz, 1998).
+\phantomsection\label{modularvalue}{The primary unit of scientific communication has long been the academic paper.
+In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
+These resources are often merely described in papers (if at all), and not made available for reuse.
+Modular publishing is the practice of making each of these resources available as independent \emph{information artifacts},
+facilitating their reuse and making them citable (Van De Sompel et al., 2004).}
+\phantomsection\label{datasharing}{Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology (Tedersoo et al., 2021).}
+
+Modular publishing can be achieved by archiving specific resources (including theory) in repositories like \href{https://zenodo.org/}{Zenodo},
 which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
 To maintain a persistent record of scholarly communication,
-Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
-which is used by many academic publishers.
-Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
-For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
-and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
-Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}.
-Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
+Zenodo mints DOIs for information artifacts.
+The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources (DataCite Metadata Working Group, 2024).
+For example, data can be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset}.
+If the data were collected for a specific paper, that relationship can be cross-referenced with \texttt{relationType:\ IsSupplementTo}.
+Similarly, a FAIR theory can be connected to a specific paper by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}.
+Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
 the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
@@
 If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.
+\phantomsection\label{modularpublisingdetract}{Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for argumentation,
+nuance,
+and the author's voice.
+}
 
@@
 If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
-\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
+\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008, p. 1).
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (De Groot \& Spiekerman, 1969).
@@
 A first consideration is that copyright law protects authors' rights according to the idea-expression dichotomy (Bently et al., 2010).
-It explicitly does not
-\emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''}.
+Copyright does not
+\emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''} (Section 102, Copyright Act).
 Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
@@
 As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
-The authors explicitly mention a second change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}.
+The authors explicitly mention a second change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''} (p.~423).
 The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
@@
 In other words: what actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}c).
+To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}c).
 Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
@@
 Modular publishing capitalizes on these changing metrics,
-and publishing theories as citeable artifacts allows scholars to be credited for contributions to theory (Kircz, 1998).
+and FAIR theory allows scholars to be credited for theoretical contributions (Kircz, 1998).
 Journals that publish theoretical papers could require authors to additionally publish their theories in a FAIR format, cross-referencing the paper,
@@
 For example, consider how FAIR theory relates to preregistration.
-These practices are clearly distinct but complimentary.
+These practices are distinct but complimentary.
 The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
 and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
-The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes (Peikert et al., 2023).
-These practices differ in their level of abstraction:
-FAIR theories cut across studies on a given phenomenon,
-whereas preregistrations are specific implementations of hypothesis tests,
-with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline.
+The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in the outcomes of a specific empirical study (Peikert et al., 2023).
+FAIR theories are specified at a level of abstraction that transcends individual studies.
+FAIR theories can inform - and be informed by - both quantitative and qualitative research.
+Preregistrations, by contrast, are specific implementations of quantitative hypothesis tests,
+within the context of a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline.
 These practices complement each other:
@@
 
-Another future direction is the intersection between the aforementioned ``theory crisis'' and the related ``measurement crisis'' pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs (Bringmann et al., 2022).
+\phantomsection\label{measurementcrisis}{Another future direction is the intersection between the aforementioned ``theory crisis'' and the related ``measurement crisis'' pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs (Bringmann et al., 2022).
 Since FAIR theories can reference other theories and resources,
-it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
+it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.}
 \phantomsection\label{jinglejangle}{FAIR theories can also help address ``jingle- and jangle fallacies'' in psychology,
-which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct (\textbf{songLiteratureReviewingAddressing2021?}).
+which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct (Song et al., 2021).
 By explicitly referencing operational definitions in FAIR theories,
@@
 closing a critical gap in the scientific process.
-FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
+
+\phantomsection\label{turkheimer}{We invite the reader to ask the rhetorical question raised by our reviewer, prof.
+Turkheimer, who - reflecting on a long career as a practitioner of theory in psychology - considered ``how adopting a FAIR framework would change, improve, or inhibit what I do''.
+We paraphrase his answer here\footnote{Edited for brevity.}, because his lived experience complements our youthful ``techno-enthusiasm''.
+Prof.~Turkheimer remarked that a FAIR framework would have been very helpful in maintaining meaningful boundaries between a theory and the predictions it makes,
+in tracking how empirical findings result in changes to empirical claims made by a theory,
+and in resolving theoretical disputes.
+\emph{``FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
+Everything published about a theory can continue to exist in its original form,
+while a FAIR representation supplements that literature by explicitly outlining the theory's assumptions, logical structure, and testable implications.
+Importantly, FAIR theory can evolve semi-independently from individual papers, offering a persistent and collaborative object that others can reuse, cite, and refine.
+Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability.''}}
+
+We envision a future where application of the FAIR principles makes theories more \emph{useful}, living up to Herbert Simon's ideal,
+enabling scholars to incorporate theory into their workflows in a tanglible way,
+providing explicit derivation chains for hypotheses,
+applying transparent rules to select the right control variables for causal inference,
+and proposing specific changes to existing theories based on empirical results.
 This paves the way for more theory-driven scholarship
@@
 
-\bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015a}
+\bibitem[\citeproctext]{ref-nosekPromotingOpenResearch2015}
 Nosek, B.
@@
 
+\bibitem[\citeproctext]{ref-songLiteratureReviewingAddressing2021}
+Song, Y., Watson, R.
+T., \& Zhao, X.
+(2021).
+\emph{Literature {Reviewing}: {Addressing} the {Jingle} and {Jangle Fallacies} and {Jungle Conundrum Using Graph Theory} and {NLP}}.
+Forty-{Second International Conference} on {Information Systems}.
+
 \bibitem[\citeproctext]{ref-szollosiArrestedTheoryDevelopment2021}
@@
 Taylor, S. (2022). The psychology of pandemics. \emph{Annual Review of Clinical Psychology}, \emph{18}(1), 581--609.
+
+\bibitem[\citeproctext]{ref-tedersooDataSharingPractices2021}
+Tedersoo, L., Küngas, R., Oras, E., Köster, K., Eenmaa, H., Leijen, Ä., Pedaste, M., Raju, M., Astapova, A., Lukner, H., Kogermann, K., \& Sepp, T.
+(2021).
+Data sharing practices and data availability upon request differ across scientific disciplines.
+\emph{Scientific Data}, \emph{8}(1), 192.
+\url{https://doi.org/10.1038/s41597-021-00981-0}
 
@@
 
-\bibitem[\citeproctext]{ref-vanlissaTeacherCornerEvaluating2020}
+\bibitem[\citeproctext]{ref-vanlissaTeachersCornerEvaluating2020}
 Van Lissa, C.

### 2025-06-07 test.Rmd
manuscript.rmd not working, try test.rmd instead

--- 2025-06-07 manuscript.Rmd
@@
       - Supervision
-      - Writing – original draft
+      - Writing - original draft
       - Writing – review & editing
@@
     institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
+
 authornote: |
@@
 The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
+
 abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
+
 keywords          : "fairtheory, meta science, theory formation, open science"
@@
 bibliography      : "fair_theory.bib"
+
 floatsintext      : yes
@@
 classoption       : "man, noextraspace"
+
 urlcolor          : blue

### 2025-06-07 test.tex
manuscript.rmd not working, try test.rmd instead

--- 2025-06-07 manuscript.tex
@@
 We also thank the Editor, Dr.~Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
+This version of the paper was generated from Git Commit \# 0df377d6.
+The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
 
@@
 Caspar J.
-Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Conceptualization, Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S.
+Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing - original draft, Writing -- review \& editing; Aaron Peikert: Conceptualization, Formal Analysis, Writing -- original draft, Writing -- review \& editing; Maximilian S.
 Ernst: Writing -- review \& editing, Investigation; Noah N.N.

### 2025-06-07 manuscript.Rmd
whoops, confused Herbert Simon and Kurt Lewin

@@
 
-We envision a future where application of the FAIR principles makes theories more *useful*, living up to Herbert Simon's ideal,
+We envision a future where application of the FAIR principles makes theories more *useful*, living up to Kurt Lewin's ideal,
 enabling scholars to incorporate theory into their workflows in a tanglible way,

### 2025-06-07 action_letter.Rmd
update 2025-06-07 16:43:02.696576

@@
 
+To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
+We also already have two worked examples for actual theories, which we now point out in the text (with links to the relevant vignettes).
+
 ## Reviewer: 3

### 2025-06-07 test.Rmd
update 2025-06-07 16:43:02.696576

@@
 we propose an applied workflow for making theory FAIR.
-The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
+This conceptual workflow is software-agnostic,
+but for R-users, most steps have been automated in the `theorytools` package.
+We briefly introduce the conceptual workflow here, in terms of general principles that can be implemented using other open science infrastructures.
+The R-package includes a living document with the [latest version of the workflow in tutorial format](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html),
+using the specification and FAIRification of De Groot's "empirical cycle" as a guiding example.
+Two other vignettes apply these processes to theories of greater substantive interest: Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000], and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of emotions in children [@morrisRoleFamilyContext2007].
+
+Our approach builds upon existing successful open science infrastructures to the greatest possible extent.
 At the time of writing (2024),
@@
 a popular platform in psychology.
-Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory amongst users of that platform,
+Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory among users of that platform,
 while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
 Note that open science infrastructure is an area of active development, and as such,
-the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
-The following conceptual workflow does not require the use of R, but most steps can be automated in R using the `theorytools` package.
-The package also includes a living document with the [latest version of the workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
-We present a brief summary of the workflow at the time of writing here,
-to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
+workflows might change as new tools or databases are developed or existing tools and database change over time.
 
@@
 it is defined at either the "theory" or "specification" level.
-We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
-The implementation below describes the model as a directed graph (see also Figure \@ref(fig:figecs)a).
-Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
-and the second half describes their proposed interrelations.
-This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
-We can save this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.
-
-```
-digraph {
-
-  observation;
+This theory can be FAIRified in its present form by archiving a text document containing five natural language statements.
+However, it is also possible to increase the level of formalization,
+and present an "implementation" of the theory as a directed graph (note: the implementation printed here is simplified; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details).
+This fulfills criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
+Given the cyclical nature of the conceptual model, such an implementation might look like this:
+
+```
   induction;
   deduction;
+  observation;
   test;
@@
   evaluation -> observation;
-
-}
-```
-
-## 2. Creating a Project Folder
-
-Create a new folder and copy the theory file from the previous step into it.
+```
+
+Note that the first part of the implementation constitutes an ontology - it specifies the entities comprised in the theory.
+The second part of the implementation describes the flow of information from phase to phase.
+Figure \@ref(fig:figecs)a) shows what this implementation looks like when plotted.
+Regardless of which implementation we prefer - we can save it to a text file.
+This text file is the "information artifact" containing our theory.
+
+## 2. Create a Theory Archive
+
 To help meet the Interoperability and Reusability criteria,
-add two more files:
-A README.md file with instructions for future users of your theory,
-and a LICENSE file with the legal conditions for reuse.
-For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html).
-We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
-
-<!-- ### What's in a README? -->
-
-<!-- The readme should contain information to help people get started with using your FAIR theory. -->
-<!-- We suggest the following elements: -->
-
-<!-- * Title, prefaced with `# FAIR theory: The Theory's Name` -->
-<!-- * Description: A plain-text description of the theory and its scope -->
-<!-- * Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use".
-From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)".
-We propose explicitly addressing the theory's X-Interoperability, telling users exactly what they can use the theory for, and how.
-For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
--->
-<!-- * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*. -->
-<!-- * License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*. -->
-<!-- * Citing this work: Tell users how to cite the theory.
-Note that this section is redundant with the Zenodo archive, which has a preferred citation field.
-The disadvantage of redundant information is that you may have to maintain this section of the README going forward.
-The advantage is that documenting related works in the README makes it more readily accessible to users.
-We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
--->
-<!-- * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in.
-Again, this is redundant with metadata entered in Zenodo (step 5).
-We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward.
-For example, we referenced the original empirical cycle paper here: -->
-
-<!-- ``` -->
-<!-- This repository contains an implementation of the "empirical cycle", -->
-<!-- a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works. -->
-
-<!-- > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: -->
-<!-- Foundations of inference and research in the behavioral sciences. -->
-<!-- De Gruyter Mouton. https://doi.org/10.1515/9783112313121 -->
-<!-- ``` -->
-
-## 3. Version Control the Repository
-
-<!-- The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved. -->
-<!-- In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021]. -->
-To version control our project, we initiate a Git repository in the project folder.
-<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
-<!-- opening a terminal in the project folder, -->
-<!-- and running the command `git init`. -->
-We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
-<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`.
--->
-<!-- The default settings for the repository are fine; -->
-<!-- for the next steps of the tutorial it  -->
-Note that the repository
-must be set to "Public" to take advantage of GitHub's Zenodo integration.
-Push the local files to the Git remote repository, and keep them synchronized going forward.
+it is important to bundle the theory file with additional files containing documentation.
+Firstly, add a README.md file with instructions for future users of your theory.
+The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) with guidance on writing a README file for theory.
+Secondly, add a LICENSE file with the legal conditions for reuse.
+As explained previously, we recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
+Place all files in a folder together - this folder will become the theory archive.
+
+## 3. Version Control the Archive
+
+To track all changes to our theory, the previously created folder can be version controlled.
+[Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) is well-suited for this purpose.
+To invite reuse and collaboration, it is further recommended to create a backup in the cloud on a service like [GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/).
 
@@
 
-First, create a Zenodo account with your existing GitHub account.
-Then in Zenodo, go to the GitHub section under your account.
-Following the instructions on the page, activate Zenodo for your theory repository.
-Then, create a new release of the GitHub repository.
-Choose a tag and release title using our adapted semantic versioning, starting with version 1.0.0, if you intend to share your theory with the broader scientific community.
-After publishing the release,
-you should be able to see the archived version in your Zenodo account,
-along with a DOI.
+Aside from tracking changes, it is also important to retain a permanent archive of major versions of a theory,
+so that they can be referenced by future research in perpetuity.
+Zenodo is a FAIR compliant repository for archiving information artifacts, with GitHub integration.
 
@@
 
-To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
-See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle, as archived in Zenodo.
-
-- Set the *resource type* to `Model`; this ensures proper archival in Zenodo
-- Verify that the *title* is prefaced with `FAIR theory:`; this allows sentient readers to recognize the work as a FAIR theory
-- Add the *keyword* `fairtheory`; this aids search engine indexation
-- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
-- List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
-    + `Is documented by` can be used to reference a theory paper you wrote, in which you introduce this FAIR theory
-    + `Is derived from` could be used to reference a paper or book chapter that introduced an existing theory that was not previously made FAIR.
-We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
-- Optionally, add *References* to related works in plain text. For example, we cite De Groot and Spiekerman in this field.
-
-## Automating these Steps
-
-R-users can use the `theorytools` package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):
-
-```
-install.packages("theorytools")
-library(theorytools)
-# Use worcs to check if GitHub permissions are set:
-library(worcs)
-check_git()
-check_github()
-# Create the theory repository:
-fair_theory(path = "c:/theoryfolder/empirical_cycle",
-            title = "The Empirical Cycle",
-            theory_file = "empirical_cycle.dot",
-            remote_repo = "empirical_cycle",
-            add_license = "cc0")
-```
-The first two lines install and load the `theorytools` package.
-Lines 4--6 use the `worcs` package to check whether git for version control is installed and a connection to github can be established to publish the theory.
-Line 8 calls the main entry function `fair_theory` for a local repository in a specified local `path` with a given `title` and the aforementioned dot `theory_file` containing the theory, uploading it to the user's remote github repository with name given in `remote_repo` and a given license specified in `add_license`.
+Once archived, documenting a FAIR theory with relevant metadata improves its Findability.
+We recommend using a standardized metadata schema, such as DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
+Within this schema, set the *resource type* to `Model`, add the words `FAIR theory:` to the title so that sentient readers to recognize the work as a FAIR theory, and add `fairtheory` as a keyword to aid search engine indexation.
+Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
+As mentioned before, [these five steps are partly automated for R-users](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
+See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle.
 
@@
 
-Several authors have reinterpreted De Groot's empirical cycle.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-For example, the DOT graph below implements Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle.
-First, notice that the phases of the cycle have been renamed.
-This change was not described in the paper.
-If we assume that the labels are meant to illustrate the phases,
+Elaborating on our running example, several authors have reinterpreted De Groot's empirical cycle.
+For example, Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle differs from De Groot's in several ways.
+First, the phases of the cycle were renamed, and this change was not described in the paper.
+Assuming that Wagenmakers' new labels are meant to illustrate the phases,
 not substantially change the ontology,
-then we can represent it by adding labels to the original DOT graph.
+we could incorporate this change by adding labels to the original ontology.
 These labels suggest a focus on empirical psychology that was not present in De Groot's version.
@@
 The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
-The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
-
-```
-digraph {
-
-  subgraph cluster_discovery {
+We could implement this change to the original implementation by grouping the respective phases of the cycle;
+a minor and tractable change.
+
+```
+  {
     label="Discovery";
@@
   observation  [label="Old knowledge and old data"];
-  subgraph cluster_justification {
+  {
     label="Justification";
@@
 
-}
 ```
@@
 Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
-<!-- In our representation, -->
-<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
-<!-- For example, the refutation of a hypothesized effect, -->
-<!-- or the serendipitous observation of some pattern in data, -->
-<!-- might be a reason to revise or construct theory. -->
-<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
-<!-- in the form of the inductive bias of methods used to perform the test, -->
-<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
-<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced.
--->
-<!-- ``` -->
 Theory testing, as takes place in the "context of justification", can be said to involve mostly deductive reasoning.
@@
 For present purposes, the terms are interchangeable.].
-<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
-Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
-NvD dissents: -->
-<!-- NvD: I cannot accept this footnote as an adequate characterization.
-Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more).
-See for instance Brian Haig's 2005 paper.
-Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions".
-At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree.
-I am fine with the last sentence though ;-) My suggestion is to write the following: -->
-<!-- NvD: We use induction to describe inferences from specific observations to general theories.
-In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations.
-We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable.
--->
-<!-- Peirce: CP 8.233 Cross-Ref:†† -->
-<!-- 233. This is not essentially different from induction. Only it is most -->
-<!-- usually an induction from instances which are not discrete and numerable. I now -->
-<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
-<!-- line of procedure, or at least not to distinguish it sharply. -->
 However, deriving hypotheses from theory is not purely deductive
@@
 ```
-digraph {
 
@@
 
-}
-```
-
-The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
-If you start with an existing GitHub repository, and wish to make some changes to it,
-this is commonly done by [creating a "branch"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository).
-A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
-Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
-Figure \@ref(fig:figgithub-1) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
-A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
-If you wish to develop a new version of someone else's FAIR theory,
-it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
-This creates a copy of their repository onto your GitHub account.
-Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
-Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
-
-```{r figgithub, out.width = "50%", fig.show='hold', fig.cap='FAIR Theories on GitHub', fig.subcap=c('Branches contain different versions of a theory', 'These versions can be compared'), fig.ncol = 2}
+```
+
+The FAIR theory workflow offers concrete ways to make changes to a FAIR theory object and to compare different versions,
+as explained in [this vignette](ADD ME).
+For example, Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+
+```{r figgithub, out.width = "50%", fig.show='hold', fig.cap='FAIR Theories on GitHub', fig.subcap=c('Different versions of a theory are archived as distinct branches', 'These versions can be explicitly compared'), fig.ncol = 2}
 #par(mfrow=c(1,2))
@@
 More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
-<!-- Some have argued that good theories involve *causal explanations* [@meehlAppraisingAmendingTheories1990]. -->
-<!-- Explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995]. -->
-<!-- Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference, -->
-<!-- and could even be integrated into the analysis environment. -->
-
-<!-- NOTES -->
-
-<!-- * Theory is the vehicle of cumulative knowledge acquisition -->
-<!-- * According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights.
-When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena.
--->
-<!-- * At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory. -->
-<!-- * One source of potential improvements of theory methodology that has not been previously considered is computer science. -->
-<!-- * The process of "iteratively improving" digital objects - in this case, computer code - is well understood. -->
-<!-- * Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well. -->
-<!-- * This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
--->
-<!-- * We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories -->
-<!-- * FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner.
-Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.
--->
-<!-- * FAIR theory allows for algorithmic derivation of hypotheses implied by the theory. -->
-<!-- * FAIR theory enables integration testing: researchers can build a "test suite" of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite.
--->
-<!-- * To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science.
--->
 

### 2025-06-07 test.tex
update 2025-06-07 16:43:02.696576

@@
 We also thank the Editor, Dr.~Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
-This version of the paper was generated from Git Commit \# 0df377d6.
+This version of the paper was generated from Git Commit \# cfb33cc6.
 The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
@@
 we propose an applied workflow for making theory FAIR.
+The workflow does not require the use of R,
+but for R-users, most steps have been automated in the \texttt{theorytools} package.
+We present a brief summary of the conceptual workflow here,
+to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
+The R-package includes a living document with the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{latest version of the workflow in tutorial format},
+to guide users in making their theories FAIR.
 The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
@@
 Note that open science infrastructure is an area of active development, and as such,
-the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
-The following conceptual workflow does not require the use of R, but most steps can be automated in R using the \texttt{theorytools} package.
-The package also includes a living document with the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{latest version of the workflow}.
-We present a brief summary of the workflow at the time of writing here,
-to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
+workflows might change as new tools or databases are developed or existing tools and database change over time.
 
@@
 it is defined at either the ``theory'' or ``specification'' level.
-We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
-The implementation below describes the model as a directed graph (see also Figure \ref{fig:figecs}a).
-Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
-and the second half describes their proposed interrelations.
-This follows the first two properties of good theory according to Meehl (Meehl, 1990).
-We can save this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.
+This theory can be FAIRified in its present form by archiving a text document containing five natural language statements.
+However, it is also possible to increase the level of formalization,
+and present an ``implementation'' of the theory as a directed graph (note: the implementation printed here is simplified; see the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{tutorial} for technical details).
+This fulfills criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
+Given the cyclical nature of the conceptual model, such an implementation might look like this:
 
 \begin{verbatim}
-digraph {
-
-  observation;
   induction;
   deduction;
+  observation;
   test;
@@
   evaluation -> observation;
-
-}
 \end{verbatim}
 
-\subsection{2. Creating a Project Folder}\label{creating-a-project-folder}
-
-Create a new folder and copy the theory file from the previous step into it.
+Note that the first part of the implementation constitutes an ontology - it specifies the entities comprised in the theory.
+The second part of the implementation describes the flow of information from phase to phase.
+Figure \ref{fig:figecs}a) shows what this implementation looks like when plotted.
+Regardless of which implementation we prefer - we can save it to a text file.
+This text file is the ``information artifact'' containing our theory.
+
+\subsection{2. Create a Theory Archive}\label{create-a-theory-archive}
+
 To help meet the Interoperability and Reusability criteria,
-add two more files:
-A README.md file with instructions for future users of your theory,
-and a LICENSE file with the legal conditions for reuse.
-For guidance on writing the README file, see \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{this vignette}.
-We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
-
-\subsection{3. Version Control the Repository}\label{version-control-the-repository}
-
-To version control our project, we initiate a Git repository in the project folder.
-We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
-Note that the repository
-must be set to ``Public'' to take advantage of GitHub's Zenodo integration.
-Push the local files to the Git remote repository, and keep them synchronized going forward.
+it is important to bundle the theory file with additional files containing documentation.
+Firstly, add a README.md file with instructions for future users of your theory.
+The \texttt{theorytools} package contains a \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{vignette} with guidance on writing a README file for theory.
+Secondly, add a LICENSE file with the legal conditions for reuse.
+As explained previously, we recommend explicitly waiving copyright with the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
+Place all files in a folder together - this folder will become the theory archive.
+
+\subsection{3. Version Control the Archive}\label{version-control-the-archive}
+
+To track all changes to our theory, the previously created folder can be version controlled with \href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{Git}.
+To invite reuse and collaboration, it is further recommended to create a backup in the cloud on a service like \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{GitHub}.
 
@@
 
-First, create a Zenodo account with your existing GitHub account.
-Then in Zenodo, go to the GitHub section under your account.
-Following the instructions on the page, activate Zenodo for your theory repository.
-Then, create a new release of the GitHub repository.
-Choose a tag and release title using our adapted semantic versioning, starting with version 1.0.0, if you intend to share your theory with the broader scientific community.
-After publishing the release,
-you should be able to see the archived version in your Zenodo account,
-along with a DOI.
+Zenodo is a FAIR compliant repository for archiving information artifacts with GitHub integration.
+Whereas GitHub facilitates collaboration - it is insufficient for archival purposes.
+Major versions of a theory, developed on GitHub, should be archived on Zenodo.
+This process can be fully automated.
 
@@
 
-To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
-See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the metadata} associated with our FAIR empirical cycle, as archived in Zenodo.
-
-\begin{itemize}
-\tightlist
-\item
-  Set the \emph{resource type} to \texttt{Model}; this ensures proper archival in Zenodo
-\item
-  Verify that the \emph{title} is prefaced with \texttt{FAIR\ theory:}; this allows sentient readers to recognize the work as a FAIR theory
-\item
-  Add the \emph{keyword} \texttt{fairtheory}; this aids search engine indexation
-\item
+Providing relevant metadata improves the Findability of a FAIR theory archived on Zenodo.
+See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the metadata} associated with our FAIR empirical cycle.
+We recommend setting the \emph{resource type} to \texttt{Model}, adding the words \texttt{FAIR\ theory:} to the title so that sentient readers to recognize the work as a FAIR theory, and adding \texttt{fairtheory} as a keyword to aid search engine indexation.
 Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
-\item
-  List the DOIs/identifiers of \emph{related works}. Use the \texttt{Relation} field as appropriate. For example:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \texttt{Is\ documented\ by} can be used to reference a theory paper you wrote, in which you introduce this FAIR theory
-  \item
-    \texttt{Is\ derived\ from} could be used to reference a paper or book chapter that introduced an existing theory that was not previously made FAIR.
-We used \texttt{Is\ derived\ from} to reference De Groot and Spiekerman's empirical cycle.
-  \end{itemize}
-\item
-  Optionally, add \emph{References} to related works in plain text. For example, we cite De Groot and Spiekerman in this field.
-\end{itemize}
-
-\subsection{Automating these Steps}\label{automating-these-steps}
-
-R-users can use the \texttt{theorytools} package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):
-
-\begin{verbatim}
-install.packages("theorytools")
-library(theorytools)
-# Use worcs to check if GitHub permissions are set:
-library(worcs)
-check_git()
-check_github()
-# Create the theory repository:
-fair_theory(path = "c:/theoryfolder/empirical_cycle",
-            title = "The Empirical Cycle",
-            theory_file = "empirical_cycle.dot",
-            remote_repo = "empirical_cycle",
-            add_license = "cc0")
-\end{verbatim}
-
-The first two lines install and load the \texttt{theorytools} package.
-Lines 4--6 use the \texttt{worcs} package to check whether git for version control is installed and a connection to github can be established to publish the theory.
-Line 8 calls the main entry function \texttt{fair\_theory} for a local repository in a specified local \texttt{path} with a given \texttt{title} and the aforementioned dot \texttt{theory\_file} containing the theory, uploading it to the user's remote github repository with name given in \texttt{remote\_repo} and a given license specified in \texttt{add\_license}.
+As mentioned before, \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{these five steps are partly automated for R-users}.
 
@@
 
-Several authors have reinterpreted De Groot's empirical cycle.
 An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-For example, the DOT graph below implements Wagenmakers and colleagues' (2018) interpretation of the empirical cycle.
-First, notice that the phases of the cycle have been renamed.
-This change was not described in the paper.
-If we assume that the labels are meant to illustrate the phases,
+Elaborating on our running example, several authors have reinterpreted De Groot's empirical cycle.
+For example, Wagenmakers and colleagues' (2018) interpretation of the empirical cycle differs from De Groot's in several ways.
+First, the phases of the cycle were renamed, and this change was not described in the paper.
+Assuming that Wagenmakers' new labels are meant to illustrate the phases,
 not substantially change the ontology,
-then we can represent it by adding labels to the original DOT graph.
+we could incorporate this change by adding labels to the original ontology.
 These labels suggest a focus on empirical psychology that was not present in De Groot's version.
@@
 The authors explicitly mention a second change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''} (p.~423).
-The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.
+We could implement this change to the original implementation by grouping the respective phases of the cycle;
+a minor and tractable change.
 
 \begin{verbatim}
-digraph {
-
-  subgraph cluster_discovery {
+  {
     label="Discovery";
@@
   evaluation -> observation  [label="Knowledge accumulation"];
-
-}
 \end{verbatim}
@@
 \begin{verbatim}
-digraph {
 
@@
   results -> theory [label="interpretation and generalization"];
-
-}
 \end{verbatim}
 
-The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
-If you start with an existing GitHub repository, and wish to make some changes to it,
-this is commonly done by \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository}{creating a ``branch''}.
-A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
-Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
-Figure \ref{fig:figgithub-1} shows how our example \texttt{empirical\_cycle} repository contains branches with Wagenmakers' and Van Lissa's implementations.
-A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
-If you wish to develop a new version of someone else's FAIR theory,
-it is possible to \href{https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo}{``fork their repository''}.
-This creates a copy of their repository onto your GitHub account.
-Both branches and forks can be compared and merged via ``pull requests'', which are essentially a request to incorporate the changes you have made.
-Figure \ref{fig:figgithub-2} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+The FAIR theory workflow offers concrete ways to make changes to a FAIR theory object and to compare different versions,
+as explained in \href{ADD\%20ME}{this vignette}.
+For example, Figure \ref{fig:figgithub-2} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
 
 \begin{figure}
-\subfloat[Branches contain different versions of a theory\label{fig:figgithub-1}]{\includegraphics[width=0.5\linewidth]{git_branches} }\subfloat[These versions can be compared\label{fig:figgithub-2}]{\includegraphics[width=0.5\linewidth]{git_compare_short} }\caption{FAIR Theories on GitHub}\label{fig:figgithub}
+\subfloat[Different versions of a theory are archived as distinct branches\label{fig:figgithub-1}]{\includegraphics[width=0.5\linewidth]{git_branches} }\subfloat[These versions can be explicitly compared\label{fig:figgithub-2}]{\includegraphics[width=0.5\linewidth]{git_compare_short} }\caption{FAIR Theories on GitHub}\label{fig:figgithub}
 \end{figure}
@@
 
-We envision a future where application of the FAIR principles makes theories more \emph{useful}, living up to Herbert Simon's ideal,
+We envision a future where application of the FAIR principles makes theories more \emph{useful}, living up to Kurt Lewin's ideal,
 enabling scholars to incorporate theory into their workflows in a tanglible way,

### 2025-06-08 action_letter.Rmd
update 2025-06-08 09:16:52.945266

@@
 
+We are grateful for this thoughtful and supportive review.
+The Reviewer's clear articulation of the central goals of our paper, and their agreement with the need to (re)focus reform efforts on theory development, is affirming and encouraging.
+
 ```{asis}
@@
 
-Not sure what to do with this!
-I'm not 100% convinced that the disciplinary differences as the Reviewer describes exist, and I personally do not take a values based approach - I am more interested in "how to" improve social scientific research, and once I know the answer, I am intrinsically motivated to adopt that procedure.
-I guess I believe that most scholars are doing what they think is best given what they know, and I think that by exposing them to new and better ideas, they will change their ways.
-I don't see a need to moralize.
+While we are completely sympathetic with the Reviewer's considerations about the roles of norms and values,
+we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
+While incentives are briefly mentioned in the paper, our focus as methodologists is on the "affordances" of FAIR theory: what types of academic work does FAIR theory make possible, and what problems could it solve.
+We believe that meta-scientists and sociologists of science would be better positioned to engage with the topics of norms and values,
+and that such a topic would be better suited to an empirical enquiry (i.e., establishing what the most widely held norms and values are, and how these relate to endorsements of academic practices).
 

### 2025-06-08 test.Rmd
update 2025-06-08 09:16:52.945266

@@
 If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
-<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for argumentation,
-nuance,
-and the author's voice.
+<span id="modularpublisingdetract">
+Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for greater nuance and the author's voice.
+Theories published in traditional papers can be supplemented by FAIR versions that live independently, evolve collaboratively, and feed into reproducible workflows.
 </span>
@@
 
-To concretize the FAIR principles,
-we propose an applied workflow for making theory FAIR.
-This conceptual workflow is software-agnostic,
-but for R-users, most steps have been automated in the `theorytools` package.
-We briefly introduce the conceptual workflow here, in terms of general principles that can be implemented using other open science infrastructures.
-The R-package includes a living document with the [latest version of the workflow in tutorial format](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html),
-using the specification and FAIRification of De Groot's "empirical cycle" as a guiding example.
-Two other vignettes apply these processes to theories of greater substantive interest: Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000], and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of emotions in children [@morrisRoleFamilyContext2007].
-
-Our approach builds upon existing successful open science infrastructures to the greatest possible extent.
-At the time of writing (2024),
+We propose a practical, software-agnostic workflow for making theory FAIR.
+While these steps can be implemented using a variety of tools,
+R users can streamline the process using the `theorytools` package, which automates most steps.
+We will illustrate the workflow using De Groot's *empirical cycle*,
+a meta-theory of theory construction,
+as a running example for this tutorial.
+The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
+
+The R-package includes further examples for FAIR theory creation in tutorial form,
+including a [worked example for implementing this workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
+Two other vignettes apply the workflow to substantive theories:
+Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000],
+and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of children's emotions [@morrisRoleFamilyContext2007].
+
+Our workflow builds upon existing open science infrastructures to the greatest possible extent.
+At the time of writing (2025),
 the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* FAIR principles.
@@
 
-## 1. Implementing the Theory
-
-We will use the *empirical cycle* as a running example for this tutorial.
-The empirical cycle, described on page 28 of @degrootMethodologyFoundationsInference1969,
-is a meta-theory of theory construction.
-The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
-<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
-<!-- We can thus consider it a meta-theory of theory construction. -->
-<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
-The original theory consists of a series of natural language statements:
+## 1. Implement the Theory
+
+Begin by expressing the theory as a structured information artifact.
+This can range from plain-text natural language statements to a more formal representation, such as a directed graph.
+For example, the empirical cycle was originally described on page 28 of @degrootMethodologyFoundationsInference1969 as a series of natural language statements:
 
@@
 
-If we compare this to the levels of theory formalization [@guestHowComputationalModeling2021],
-it is defined at either the "theory" or "specification" level.
-This theory can be FAIRified in its present form by archiving a text document containing five natural language statements.
-However, it is also possible to increase the level of formalization,
-and present an "implementation" of the theory as a directed graph (note: the implementation printed here is simplified; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details).
-This fulfills criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
+Implementing the theory as an information artifact can be as simple as isolating these statements from the paper,
+and saving them to a plain text file.
+
+However, we can also make an additional attempt to formalize the theory.
+According to the levels of theory formalization,
+the empirical cycle is presently defined at either the "theory" or "specification" level [@guestHowComputationalModeling2021].
+We can further formalize the theory to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
+This fulfills criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
 Given the cyclical nature of the conceptual model, such an implementation might look like this:
@@
 Figure \@ref(fig:figecs)a) shows what this implementation looks like when plotted.
-Regardless of which implementation we prefer - we can save it to a text file.
-This text file is the "information artifact" containing our theory.
+Regardless of which implementation we prefer,
+we can save it to a plain text file - this is the "information artifact" containing our theory.
 
@@
 
-To help meet the Interoperability and Reusability criteria,
-it is important to bundle the theory file with additional files containing documentation.
+To meet the Interoperability and Reusability criteria,
+it is important to properly document the theory file.
 Firstly, add a README.md file with instructions for future users of your theory.
-The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) with guidance on writing a README file for theory.
+The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) on writing README files for theory.
 Secondly, add a LICENSE file with the legal conditions for reuse.
-As explained previously, we recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
-Place all files in a folder together - this folder will become the theory archive.
+We recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
+Place all files in a folder together - this folder becomes the theory archive.
 
@@
 
-To track all changes to our theory, the previously created folder can be version controlled.
+To track all changes to our theory, the theory archive can be version controlled.
 [Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) is well-suited for this purpose.
-To invite reuse and collaboration, it is further recommended to create a backup in the cloud on a service like [GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/).
+Hosting a backup in the cloud on a platform like [GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/) additionally makes the theory publicly accessible and facilitates community engagement.
 
@@
 
-Aside from tracking changes, it is also important to retain a permanent archive of major versions of a theory,
-so that they can be referenced by future research in perpetuity.
-Zenodo is a FAIR compliant repository for archiving information artifacts, with GitHub integration.
+Archiving major versions of a theory in a FAIR-compliant repository that issues a persistent identifier (DOI) improves their Findability and allows them to be referenced in perpetuity.
+Zenodo is a FAIR compliant repository with GitHub- and OSF integration.
 
@@
 
-Once archived, documenting a FAIR theory with relevant metadata improves its Findability.
-We recommend using a standardized metadata schema, such as DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
-Within this schema, set the *resource type* to `Model`, add the words `FAIR theory:` to the title so that sentient readers to recognize the work as a FAIR theory, and add `fairtheory` as a keyword to aid search engine indexation.
-Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
-As mentioned before, [these five steps are partly automated for R-users](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
+When archiving a FAIR theory, documenting it with relevant metadata improves its Findability.
+We recommend using a standardized metadata schema like DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
 See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle.
+Within this schema,
+
+* Set the *resource type* to the category `Model`,
+* Add the words `FAIR theory:` to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words `meta-analysis` in the title),
+* Add `fairtheory` to the keywords to aid search engine indexation.
+* Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
 

### 2025-06-08 action_letter.Rmd
update 2025-06-08 09:35:02.868689

--- 2025-06-07 action_letter.Rmd
@@
 
-Not sure what to do with this!
-I'm not 100% convinced that the disciplinary differences as the Reviewer describes exist, and I personally do not take a values based approach - I am more interested in "how to" improve social scientific research, and once I know the answer, I am intrinsically motivated to adopt that procedure.
-I guess I believe that most scholars are doing what they think is best given what they know, and I think that by exposing them to new and better ideas, they will change their ways.
-I don't see a need to moralize.
+We appreciate this thoughtful and provocative comment, which explores the broader sociological and cultural dimensions of scientific reform.
+We are sympathetic to the reviewer’s view that the challenges facing psychology are not solely technical or procedural, but likely also reflect deeper issues of norms and values.
+
+That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
+As methodologists, our primary aim in this paper is to articulate what FAIR theory affords: a framework that supports better theory specification, sharing, and reuse.
+We believe this contribution is most effective when kept sharply focused on the practical infrastructure and its scientific use cases.
+
+We do not mean to suggest that infrastructure alone is sufficient, nor do we discount the importance of cultivating scientific values.
+However, we believe that meta-scientists and sociologists of science would be better positioned to engage with the topics of norms and values,
+and that these topics are better suited to empirical enquiry - i.e., establishing what the most widely held norms and values are, and how these relate to endorsements of academic practices.
+
+We are sympathetic to the Reviewer's meta-scientific considerations,
+and we hope that the Reviewer will agree that - while there is clearly a need for enquiry into norms and values - the present paper might not be the best avenue for it.
 

### 2025-06-08 action_letter.Rmd
update 2025-06-08 13:02:50.777903

@@
 class             : "draft"
-output            : rmarkdown::html_document
+bibliography      : "fair_theory.bib"
+output            : revise::revise_letter_pdf
 ---
-
-<!-- revise::revise_letter_pdf -->
 
@@
 library(revise)
+rm(".revise_manuscripts")
 revise::read_manuscript("abstract.md")
-revise::read_manuscript("manuscript.Rmd")
+revise::read_manuscript("test.Rmd")
 ```
@@
 For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects.
+@wilkinsonFAIRGuidingPrinciples2016
 
@@
 
-We thank the Reviewer for their thoughtful review and detailed comments. We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.
+We thank the Reviewer for their thoughtful and encouraging review.
+We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory—not only within psychology, but also in related disciplines.
+We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.
 
@@
 
-Thank you for this suggestion; as proposed, we now briefly mention the R-package in the abstract:
+As suggested, we now briefly mention the R-package in the abstract:
 
@@
 
-nakagawaPoorHypothesesResearch2024
+We thank the Reviewer for pointing this out; we have included the reference to the paper making this argument in the following section:
+
+`r get_revision("waste")`
 
@@
 (Suggestion to move up the section) It seems to always come back to this that readers a confused because everyone has their own idea of what a theory is.
-We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful 🙂
+We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful
 
@@
 To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
-We also already have two worked examples for actual theories, which we now point out in the text (with links to the relevant vignettes).
+There are already two worked examples for actual theories in the package vignettes,
+which we now reference in the text.
+
+The resulting shortened section reads as follows:
+
+`r get_revision("tutorial")`
 
@@
 
-We appreciate this thoughtful and provocative comment, which explores the broader sociological and cultural dimensions of scientific reform.
+We appreciate this thought-provoking comment, which explores the broader sociological and cultural dimensions of scientific reform.
 We are sympathetic to the reviewer’s view that the challenges facing psychology are not solely technical or procedural, but likely also reflect deeper issues of norms and values.
@@
 That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
-As methodologists, our primary aim in this paper is to articulate what FAIR theory affords: a framework that supports better theory specification, sharing, and reuse.
-We believe this contribution is most effective when kept sharply focused on the practical infrastructure and its scientific use cases.
+As methodologists, our primary aim in this paper is to articulate the affordances of FAIR theory:
+a framework that supports better theory specification, sharing, and reuse.
+We believe our contribution will be most effective if it remains sharply focused on the practical infrastructure and its scientific use cases.
 
@@
 
-We are sympathetic to the Reviewer's meta-scientific considerations,
+We share the Reviewer's interest in the meta-scientific perspective,
 and we hope that the Reviewer will agree that - while there is clearly a need for enquiry into norms and values - the present paper might not be the best avenue for it.
+
+We did make one minor change to address this comment, and "punched up" the section on Theory and Scientific Progress, with a reference to scientific norms as documented in ethical guidelines:
+
+`r get_revision("norms")`
 

### 2025-06-08 test.Rmd
update 2025-06-08 13:02:50.777903

@@
 this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
-<!-- by successively increasing the empirical content of a theory. -->
-<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
-<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
-This needs more work! -->
-<!-- NvD: SOmething like?
-"In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories.
--->
 There are, however, indications that contemporary psychology falls short of this ideal.
-Firstly, because hypothesis-testing research is over-represented in the literature: 
-According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
-Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
-The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
-Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
-Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
-but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
-Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-<!-- NvD: I don't think that statement completely covers it.
-How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.).
-However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena.
--->
-<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
-Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].
-
+Meehl observed that theories in psychology "lack the cumulative character of scientific knowledge.
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p.
+1].
+Recent empirical findings confirm this view.
+Firstly, because hypothesis-testing research is vastly over-represented in the literature, amounting to 89.6% of published papers [@kuhbergerPublicationBiasPsychology2014].
+Closer examination of such studies reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
+Only 15% of hypothesis-testing studies referenced any theory,
+and rarely in direct relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
 Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
 The majority of papers ostensibly test hypotheses,
-but these are rarely connected to, let alone derived from, theory.
-Moreover, test results do not routinely contribute to the improvement or rejection of theories.
-The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p.
-1].
-The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.
+but these are rarely connected to theory.
+
+Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
+in which case it would be important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
+Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
+such as straw-man null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
+<span id = "waste">Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
+does not advance our principled understanding of psychological phenomena,
+and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].</span>
+<span id = "norms">Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021]; inefficient at best, and more likely, futile.
+As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must "avoid research waste",
+our field should take seriously its ethical responsibility to develop procedures to reduce it.</span>
+The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory; 
+instantiating theory as a digital "object" that scholars can access, reuse, and update in their daily workflows.
 
@@
 
-We propose a practical, software-agnostic workflow for making theory FAIR.
+<span id="tutorial">We propose a practical, software-agnostic workflow for making theory FAIR.
 While these steps can be implemented using a variety of tools,
@@
 * Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
+</span>
 

### 2025-06-08 action_letter.Rmd
update 2025-06-08 14:18:50.386429

@@
 
-<!-- AB: -->
-3.	Regarding missing reference to "research waste" - yes, this should be picked up.
-By the way, the latest revision of the Declaration of Helsinki states that ethical science has to avoid "research waste".
-Pretty cool.
-
 ```{asis}
@@
 
-```{asis}
-Line 73 onwards: The example of SDT nicely illustrates the authors’ point here!
-```
+As requested, we have added a few more words - but we also want to avoid redundancy, as the FAIR principles are fairly (pun intended) common sense, and well-documented in other publications which are all cited in the opening paragraph.
+In the body of the text, we explicitly engage with each of the FAIR (sub-) principles, so all information is already contained within the paper.
+
+To address the comment, we have rewritten the opening paragraph as follows:
+
+`r get_revision("whatsfair")`
 
@@
 
+Done.
+
 ```{asis}
@@
 ```
+
+Done.
 
@@
 
-<!-- AB: -->
-4.	What is a theory?
-(Suggestion to move up the section) It seems to always come back to this that readers a confused because everyone has their own idea of what a theory is.
-We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful
+We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific paradigm:
+
+`r get_revision("alldefinitions")`
 
@@
 
-We removed the "cf.", although we couldn't find mention of it in the APA7 style guide. We updated the citation scheme to APA7, so three or more authors will be shortened to "et al.".
+We removed "cf." here, although we couldn't find mention of it in the APA7 style guide.
+We use "cf." to indicate when a specific reference *contradicts* a point; e.g. "all theories should be formal (Smaldino, 2017; cf. Oude-Maatman, 2024)": Oude-Maatman disagrees with this statement.
+We have also updated the citation scheme to APA7, so three or more authors will be shortened to "et al.".
+Note that some journals prescribe citation styles that (partly) conflict with APA7; we defer to the Editor on how to handle such cases.
 
@@
 
-We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of articles in 2010-2019.
+We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of psychological articles in 2010-2019, and more (~80%) in other disciplines.
 Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
@@
 
+We have rewritten this section slightly, including a brief explanation of the "open empirical cycle" and how the present work relates to it:
+
+`r get_revision("openec")`
+
 ```{asis}
@@
 ```
+
+We had mentioned the learning curve in the Limitations section, and have expanded this discussion to address the present comment:
+
+`r get_revision("learningcurve")`
 

### 2025-06-08 test.Rmd
update 2025-06-08 14:18:50.386429

@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers^[[As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.]{#faircap}].
-Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
-This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
+<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced during the course of scholarly work
+by making them Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016]^[[As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.]{#faircap}].
+Since the FAIR principles' inception, 
+they have become a widely adopted standard for archival of academic output,
+representing an estimated tens of billions of dollars in reuse value [@vogtFAIR20Extending2024].
+Scholars have demonstrated their relevance for making other digital objects more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].</span>
+The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
-a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
-compliant with the FAIR principles.
-[FAIR theory has the potential to improve the efficiency of scholarly communication and]{#test2}
-accelerate cumulative knowledge acquisition.
+a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially connected to - the scientific paper.
+<span id="alldefinitions">Definitions of theory are abundant and hotly debated,
+but many of them are compatible with the FAIR principles.</span>
+FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
 We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.
@@
 Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
-One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
-<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
-<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning).
-However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains.
-E, M, and C stand form something is E = MC<sup>2</sup>).
--->
-When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
-it becomes difficult do derive precise predictions,
-and therefore hard to falsify the theory. 
-A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
+One frequently raised concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
+When theories are ambiguous,
+precise predictions cannot be derived from them without
+resorting to subjective interpretation or invoking additional assumptions,
+which makes them harder to falsify.
+A second concern has received less attention,
+is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
 Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
 The equally critical processes of theory construction and improvement have been largely overlooked.
-<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences.
-Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].
-NvD dissents: -->
-<!-- NvD: I cannot accept this footnote as an adequate characterization.
-Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more).
-See for instance Brian Haig's 2005 paper.
-Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions".
-At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree.
-I am fine with the last sentence though ;-) My suggestion is to write the following: -->
-<!-- NvD: We use induction to describe inferences from specific observations to general theories.
-In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations.
-We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable.
--->
-<!-- Peirce: CP 8.233 Cross-Ref:†† -->
-<!-- 233. This is not essentially different from induction. Only it is most -->
-<!-- usually an induction from instances which are not discrete and numerable. I now -->
-<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
-<!-- line of procedure, or at least not to distinguish it sharply. -->
 The present paper addresses this knowledge gap by applying, for the first time,
@@
 
+## What is Theory?
+
+Given that a pluriformity of definitions are consistent with FAIR theory principles,
+we do not limited ourselves to any one particular definition - although, at times, our writing will inevitably reveal a particular vantage point.
+Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
+The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
+acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
+We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain (Table \@ref(tab:tabmeehl)):
+statements about the types of entities postulated (i.e., ontology),
+statements about causal connections between those entities,
+statements about the functional form of those connections,
+and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
+The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
+It shifts the focus from theories as collections of sentences to mathematical models.
+The term "model" is not uniquely defined within the literature;
+models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p.
+336].
+This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
+The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
+It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.
+
 ## Theory and Scientific Progress
@@
 research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs)a.
-In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data.
+In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data [-@wagenmakersCreativityVerificationCyclePsychological2018].
 In the other half of the cycle (the "Context of Discovery"),
@@
 to be open, theory should adhere to established open science standards for specification and archival.
-We propose to implement theories as information artifacts,
+We propose to implement theories as digital objects,
 and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
 Metadata are "data about the data". 
-They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
+They provide information about the nature and content of a digital object and are stored in the repository where the version of record of the FAIR theory is deposited.
 FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
@@
 <!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
-Following the original proposal of Lamprecht and colleagues,
+Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019s],
 we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
-We reflect on the necessary changes (which are minor),
+We reflect on the necessary (minor) changes,
 as well as on the current state and future of FAIR theory in psychology.
-The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
+The resulting principles provide guidance for instantiating theory as a FAIR digital object,
 and we provide worked examples to encourage their adoption.
@@
 
-## What is Theory?
-
-<!-- We don't explicitly define theory, but discuss the most important ideas of what theory might be, and advocate for a pluralism of information that can be tracked as FAIR theory (e.g.
-also such things as scientific practice (experimental design, materials)).
--->
-<!-- The nature of the tracked information can change over time, as a scientific theory matures. -->
-<!-- The discussion about the distinction between FAIR theories and models is removed, as I believe it might be confusing for first-time readers and not really necessary for the bigger picture of the paper (but of course could be re-added).
--->
-<!-- NvD: I wonder if we need what is stated about the syntactic, semantic, and pragmatic perspectives.
-We can leave it at the characterization that theories are (typically) concerned with the explanation and understanding of what can be observed (directly or indirectly) tested (in the laboratory or natural experiments) and manipulated.
-But, this can come in many forms and we do not exclude the possibility that there are acceptions to this characterization that we could still consider a scientific theory.
-We can then state that we use Meehl's view for convenience in this paper, but that it is compatible other perspectives (or something.
--->
-Definitions of theory are abundant, and are the subject of extensive scholarly debate.
-Given that a pluriformity of definitions are consistent with FAIR theory principles,
-our suggested approach is not limited to any one particular definition.
-Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
-The syntactic view describes theories as "sets of sentences in a given logical domain language" [@scientifictheories, ch. 2],
-acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's [-@meehlAppraisingAmendingTheories1990] hierarchy of ever-more specific "statements" a theory might contain:
-statements about the types of entities postulated (i.e., ontology),
-statements about causal connections between those entities,
-statements about the functional form of those connections,
-and statements about their specific numerical values [@frankenhuisStrategicAmbiguitySocial2023; @guestWhatMakesGood2024].
-The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
-It shifts the focus from theories as collections of sentences to mathematical models.
-The term "model" is not uniquely defined within the literature;
-models have been described as "specific instantiations of theories, narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020, p.
-336].
-This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf.
-Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#).
--->
-The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
-It also argues that nonformal aspects (e.g. commonly used analogies) and practices (e.g. experimental designs) can be an important part of scientific theories.
-
-It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
+### What to Archive?
+
+It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be archived as FAIR theory. 
 As the practice of FAIRification becomes more embedded,
-we expect that it will become increasingly clear what kind and form of information is useful.
+we expect that it will become increasingly clear what kind of information is useful.
 As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
@@
 
-Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
-<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.
--->
-More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+Concerns about the state of psychological theory have motivated increasing calls for greater theory "formalization" [@smaldinoModelsAreStupid2017; cf. @oudemaatmanPsychologysTheoryCrisis2021].
+Formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
@@
 
+Crucially for the present paper, the issue of formalization is orthogonal to FAIRification.
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
-rather, it increases the fidelity and ease with which they are communicated.
-Thus, FAIR theory does not require theories to be formal.
-At the same time, formal theories are not automatically FAIR.
+rather, it pertains to rigorous and transparent archival and communication about theories, with the aim of enhancing their reusability.
+FAIR theory does not require formalization, and formal theories are not automatically FAIR.
 The FAIR principles
-<!-- pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility. -->
-<!-- They  -->
-can be applied to theories representated in natural language,
+apply to theories representated in natural language,
 as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
@@
 These resources are often merely described in papers (if at all), and not made available for reuse.
-Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
+Modular publishing is the practice of making each of these resources available as independent *digital objects*,
 facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
@@
 To maintain a persistent record of scholarly communication,
-Zenodo mints DOIs for information artifacts.
+Zenodo mints DOIs for digital objects.
 <!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
 <!-- which is used by many academic publishers. -->
-The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
+The DataCite Metadata Schema offers a standard way to document digital objects with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
 <!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
@@
 
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
+We can take inspiration from the field of computer science for well-established processes for iteratively improving digital objects.
 Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
@@
 In computer science, these challenges are navigated by assigning version numbers.
-Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to information artifacts.
+Specifically, *semantic versioning* comprises a simple set of rules for assigning version numbers to digital objects.
 Whereas version control tracks changes,
@@
 First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
-Of the many services that provide DOIs for scientific information artifacts,
+Of the many services that provide DOIs for archived objects,
 Zenodo and the Open Science Framework are commonly used in psychology.
@@
 Metadata should also include identifiers for all the versions of the theory it describes	(F3);
-Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
+Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of that artifact's versions.
 
@@
 This is a known problem in relation to research data and software [@katzSpecialIssueSoftware2024].
-Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
+Regrettably, most academic search engines only index traditional print publications, not other digital objects.
 Since the status quo is to publish theories in papers,
 the FAIR requirements are met if scholars continue to do so,
-and additionally publish theories as separate information artifacts.
+and additionally publish theories as separate digital objects.
 The `"fairtheory"` keyword can also be used to signal the presence of theory within a paper.
-<!-- An ad-hoc solution is to publish a paper as documentation for the information artifact [@mcgillivrayDeepImpactStudy2022]. -->
+<!-- An ad-hoc solution is to publish a paper as documentation for the digital object [@mcgillivrayDeepImpactStudy2022]. -->
 <!-- There is even a journal - the [Journal of Open Source Software](https://joss.theoj.org/) - that publishes software documentation as a paper so that it will be indexed by search engines. -->
@@
 In the longer term, it may not be necessary to write a paper for each theory.
-If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
-Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
+If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
+Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index digital objects stored in Zenodo.
 
@@
 
-Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
+Interoperability pertains to the property of digital objects to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016, p. 2].
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 be instantiated in a human- and machine-readable datatype,
-as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
+as should all digital objects created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
 Depending on the level of formalization of the theory,
@@
 
-Begin by expressing the theory as a structured information artifact.
+Begin by expressing the theory as a structured digital object.
 This can range from plain-text natural language statements to a more formal representation, such as a directed graph.
@@
 
-Implementing the theory as an information artifact can be as simple as isolating these statements from the paper,
+Implementing the theory as a digital object can be as simple as isolating these statements from the paper,
 and saving them to a plain text file.
@@
 Regardless of which implementation we prefer,
-we can save it to a plain text file - this is the "information artifact" containing our theory.
+we can save it to a plain text file - this is the "digital object" containing our theory.
 
@@
 We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
-We demonstrated how to create, version-control, and archive theories as digital information artifacts.
+We demonstrated how to create, version-control, and archive theories as digital objects.
 We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
@@
 One important strength of FAIR theory is that it provides much-needed open science methods for the underemphasized inductive phase of the empirical cycle.
-Most extant open science methods focus on increased rigor in testing, but provide little guidance as to what to do with the newly collected empirical evidence.
-By providing much-needed open science methods for theory construction,
-FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the "open empirical cycle" [@hoijtinkOpenEmpiricalCycle2023].
+<span id= "openec">Recently, the "open empirical cycle" was introduced, arguing that each phase in De Groot's model of cumulative  knowledge generation via scientific research can be supported by specific open science practices to increase the transparency, quality, trustworthiness, and replicability of research [@hoijtinkOpenEmpiricalCycle2023].
+As we identified, however, most existing open science methods focus on rigor in testing (phases 2-4 of the cycle),
+but few provide guidance on how to derive hypotheses from theory (phase 1), or how to relate empirical findings back to theory (phase 5), leaving a gap in the cycle.
+By instantiating theory as a FAIR digital object,
+we provide much-needed open science infrastructure for transparently deriving hypotheses and modifying theory,
+thus contributing to closing the "open empirical cycle".</span>.
 
@@
 Given the current emphasis on empirical papers [@mcphetresDecadeTheoryReflected2021], theoretical papers can be hard to publish.
-FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
+FAIR theories, by contrast, can be readily disseminated as citable digital objects, thus changing the incentive structure to favor theory development.
 
@@
 
-Another limitation is the learning curve associated with tools like Git and Zenodo.
+<span id="learningcurve">Another limitation is the learning curve associated with tools like Git and Zenodo.
 The `theorytools` R-package mitigates this limitation for R-users by automating key steps in the process.
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
-One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
+One final way to address the learning curve is via specialization and collaboration, or team science [@vanlissaVisionTeamScience2024]:
+as scientific workflows increase in sophistication, it is increasingly difficult for any one scholar to master all skills involved.
+In relation to FAIR theory, we see unique opportunities for intergenerational collaboration and knowledge exchange,
+as theoreticians tend to be seasoned experts, whereas open science literacy is more commonly found among early career scholars.</span>
+
+One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
 Education might help address this limitation; with this in mind,

### 2025-06-11 action_letter.Rmd
update 2025-06-08 14:19:28.538087
Update action_letter.Rmd
removed "WHAT??"
Update action_letter.Rmd
Addition to Reviewer 2's comment on Nomological Nets

@@
 
+<font color = "red"> TO DO </font>
+
 ## Reviewer: 2
@@
 but it's not necessarily true of FAIR theory in general. 
+We agree that several psychological theories can be specified as a nomological network of abstract entities and empirical predictions/observations.
+However, not all theories have been specified to such a level or need to be specified in such a way. 
+As such, FAIR Theory accommodates nomological nets and construct validation endeavours, but is not identical to them. 
+It is interesting to note that Cronbach and Meehl (1955) did talk about (proto)FAIR requirements:
+"Whether or not an interpretation of a test’s properties or relations involves questions of construct validity is to be decided by examining the entire body of evidence offered, together with what is asserted about the test in the context of this evidence." (p.4)
+"Hence, the investigator who proposes to establish a test as a measure of a construct must specify his network or theory sufficiently clearly that others can accept or reject it." (p.13)
+"A consumer of the test who rejects the author’s theory cannot accept the author’s validation.
+He must validate the test for himself, if he wishes to show that it represents the construct as he defines it."
 We're not sure if further changes need to be made to address this.

### 2025-06-12 action_letter.Rmd
update 2025-06-12 16:23:17.396093

--- 2025-06-08 action_letter.Rmd
@@
 All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it.
-One author proposed to invite you as a co-author, others were concerned that this might constitute a conflict of interest.
+We invite your, and the Editor's, advice on how to best go about it.
 
-We took the initiative to draft a paragraph that summarizes your comments on FAIR theory from the review, and mention your contribution in the Acknowledgements.
-We would be equally happy to credit your contribution with a co-authorship, depending on your - and the Editor's - preferences.
+As a starting point, we drafted a paragraph that summarizes your comments on FAIR theory from the review, and mentioned your contribution in the Acknowledgements.
+We are open to incorporating your thoughts and crediting your contribution in other ways as well, depending on what seems appropriate to you and the Editor.
 
@@
 
-To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
-There are already two worked examples for actual theories in the package vignettes,
-which we now reference in the text.
+To address this comment, we have removed as much of the specific detail as possible from the main text, relegating it to the package vignettes instead.
+To make this possible, we have substantially expanded the existing package vignettes and created one additional vignette.
+
+There are now three worked examples for actual theories in the package vignettes:
+[For Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html), the [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html), and [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html),
+which we reference in the text.
 

### 2025-06-12 test.Rmd
update 2025-06-12 16:23:17.396093

@@
 
-Begin by expressing the theory as a structured digital object.
-This can range from plain-text natural language statements to a more formal representation, such as a directed graph.
-For example, the empirical cycle was originally described on page 28 of @degrootMethodologyFoundationsInference1969 as a series of natural language statements:
+Begin by creating an empty folder to hold all files associated with the theory - this folder will become the theory archive.
+The first file to create is the theory itself.
+This could be a plain-text file containing natural language statements,
+or a more formal representation, such as a directed graph.
+For example, the empirical cycle was originally described as a series of natural language statements [@degrootMethodologyFoundationsInference1969, p. 28]:
 
@@
 
-However, we can also make an additional attempt to formalize the theory.
-According to the levels of theory formalization,
+Optionally, we can formalize the theory further.
+According to a taxonomy of different levels of theory formalization,
 the empirical cycle is presently defined at either the "theory" or "specification" level [@guestHowComputationalModeling2021].
-We can further formalize the theory to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
-This fulfills criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
+To fulfill criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)),we can further formalize it to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
 Given the cyclical nature of the conceptual model, such an implementation might look like this:
@@
 
-## 2. Create a Theory Archive
+## 2. Document the Theory
 
@@
 We recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
-Place all files in a folder together - this folder becomes the theory archive.
-
-## 3. Version Control the Archive
+
+## 3. Version Control the Theory Archive
 

### 2025-06-19 action_letter.Rmd
update 2025-06-19 09:59:24.514362

@@
 We appreciate the feedback that you and the reviewers have provided.
-In the following itemised list we respond to each comment point by point.
+In this action letter, we respond to each comment point by point.
 
@@
 
+We have 
 Upload tables separately: tab_meehl.csv
@@
 
-This does not directly ask for a change, but summarize relevant changes here.
+We understand that striking the balance between the conceptual and technical details is the very challenge of this paper.
+In the revision we have tried to explain technical details in even simpler language.
+The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
+We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails, and because these general principles are not tied to the R-pacakge and our specific implementation (they could be implemented using other infrastructures).
 
-<!-- AB: -->
-1.	I agree that striking the balance between the conceptual and technical details is the very challenge of this paper.
-I think we should go a route in the revision in which we try to explain technical details in even simpler language.
-For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects.
-@wilkinsonFAIRGuidingPrinciples2016
+With regard to your specific example of technical language that might elude the readership, the GitHub reference in the author note now reads: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX."
 
@@
 
-As methodologists, our focus in this paper is on "procedures", rather than policies or documentation.
-Our goal is to position FAIR theory as a communication protocol: a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.
+To clarify our perspective: As methodologists, our focus is on "procedures", rather than policies or documentation.
+Our goal in this paper is to position FAIR theory as a set of standards, grounded in philosophy of science and implemented in existing infrastructure, that facilitates theory specification, reuse, and collaborative refinement.
 
@@
 
-Compare FAIR theory to preregistration: it, too, was introduced first as a good idea; now - a decade later - empirical studies are being conducted to investigate its effects.
+Compare FAIR theory to preregistration: it, too, was introduced first as a good idea.
+Only now - a decade later - are empirical studies are being conducted to investigate its effects.on scientific practices.
 
@@
 
+**DO THIS**
+
 ```{asis}
@@
 
+**DO THIS, ANDREAS?**
+
 ```{asis}
@@
 
-We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, up from ~7% of articles in 2000–2009 to ~50% of psychological articles in 2010-2019, and more (~80%) in other disciplines.
+We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, in psychology the percentage of articles that share data has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
+In other disciplines, the increase has been even larger (up to ~80%).
 Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
@@
 
-<font color = "red"> TO DO </font>
+We apologize for the oversight, and have fixed these hyperlinks.
 
@@
 
-on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments.
+on behalf of all co-authors, let me express our sincere appreciation for your thoughtful comments.
 This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician.
-All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it.
+We have attempted to address your comments in the Revision, as described below.
+In addition, all authors agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not entirely sure how to incorporate it.
 We invite your, and the Editor's, advice on how to best go about it.
@@
 
-We think that this comment signals a need to better explain the important distinctions between these two practices.
+We respectfully disagree that FAIR theory is like preregistration, and we think that this comment signals a need to better explain the important distinctions between these two practices.
 To address this comment, we have included the following section in the revision:
@@
 There are now three worked examples for actual theories in the package vignettes:
-[For Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html), the [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html), and [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html),
-which we reference in the text.
+
+* [Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html)
+* The [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html)
+* [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html)
 
@@
 a framework that supports better theory specification, sharing, and reuse.
-We believe our contribution will be most effective if it remains sharply focused on the practical infrastructure and its scientific use cases.
+We believe we can make this contribution most effectively if we remain sharply focused on the practical infrastructure and its scientific use cases.
 

### 2025-06-19 test.Rmd
update 2025-06-19 09:59:24.514362

@@
 Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
-This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.
+Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #`r substr(gert::git_commit_id(),1,8)`.
 The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
@@
 and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
-[This is an example of FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) extracted from Zenodo.
+[This is an example of FAIR metadata](https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json) extracted from Zenodo.
 
@@
 We recommend using a standardized metadata schema like DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
-See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle.
+See here for [an example of the metadata](https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json) associated with our FAIR empirical cycle.
 Within this schema,

### 2025-06-19 test.Rmd
update 2025-06-19 15:08:14.390215

@@
 Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #`r substr(gert::git_commit_id(),1,8)`.
-The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
+The first and senior author positions were fixed, and remaining author positions randomized.
 
@@
 
-<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced during the course of scholarly work
+<span id="whatsfair">The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced in the course of scholarly work
 by making them Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016]^[[As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.]{#faircap}].
@@
 representing an estimated tens of billions of dollars in reuse value [@vogtFAIR20Extending2024].
-Scholars have demonstrated their relevance for making other digital objects more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].</span>
+Scholars have demonstrated their relevance for making other digital objects more open, including research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying; @vanlissaWORCSWorkflowOpen2021].</span>
 The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
 To this end, we introduce "FAIR theory":
-a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially connected to - the scientific paper.
-<span id="alldefinitions">Definitions of theory are abundant and hotly debated,
-but many of them are compatible with the FAIR principles.</span>
+a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially associated with - the scientific paper.
+<span id="alldefinitions">Definitions of theory abound and hotly debated,
+but as many of them are compatible with the FAIR principles,
+this paper is not limited to a specific definition.</span>
 FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
@@
 rendering them immune to falsification.
-As an example, consider "self-determination theory" (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Consider "self-determination theory" [SDT, @deciSelfDeterminationTheory2012], one of the most widely cited social psychological theories, which we formalized and FAIRified in [this vignette](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
+SDT emphasizes the role of intrinsic and extrinsic motivation in human behavior.
 Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
 Over time, however, this definition expanded to include motivations driven by the fulfillment of basic psychological needs for autonomy, competence, and relatedness [@ryanSelfdeterminationTheoryFacilitation2000].
-<!-- Consider, for instance, a research scientist working in a highly competitive laboratory.
-Under the original definition, their intrinsic motivation would be limited to the joy of discovery alone.
-Yet, under the expanded framework, factors that might initially appear extrinsic can be reconceptualized as intrinsic: securing grant funding enhances their autonomy by granting them the freedom to pursue innovative ideas; grappling with deadlines bolsters their sense of competence; and praise from their peers fulfills their need for relatedness.
--->
 The implications of these shifting definitions becomes clear when deriving hypotheses about the type of motivation involved in changing an infant's dirty diaper.
@@
 Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
-The expanded definition thus enables SDT to absorb potentially falsifying evidence.
+Expanding the definition thus enables SDT to absorb potentially falsifying evidence.
 
@@
 open science principles to psychological theory.
-We apply introduce *FAIR theory* to 
-facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
+We introduce *FAIR theory* as a methodology that can facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-we do not limited ourselves to any one particular definition - although, at times, our writing will inevitably reveal a particular vantage point.
+we do not limited ourselves to any one particular definition - although, at times, our writing inevitably reveals a particular vantage point.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic [@scientifictheories].
@@
 Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
-Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
+in which case it would be useful to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
+Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched cultural practices [@gigerenzerNullRitualWhat2004],
 such as straw-man null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
 <span id = "waste">Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
-does not advance our principled understanding of psychological phenomena,
-and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].</span>
+cannot advance our principled understanding of psychological phenomena,
+and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].
 <span id = "norms">Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021]; inefficient at best, and more likely, futile.
 As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must "avoid research waste",
-our field should take seriously its ethical responsibility to develop procedures to reduce it.</span>
+our field should take seriously its ethical responsibility to develop procedures to reduce it.</span></span>
 The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory; 
@@
 <!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
-Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019s],
+Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019],
 we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
@@
 Going even further, a *generative/computational model* could be specified,
-which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+which is completely parameterized (e.g., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
 Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
-Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).
-
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
+Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement instruments (e.g., different questionnaires used to assess the same construct), or information about participant recruitment- and retention strategies.
+
+Theories can include or reference other theories.
+For example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
 such as adherence to infection prevention protocols (e.g., social distancing),
@@
 [@taylor2022psychology].
-The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+Such a theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 
@@
 
-Crucially for the present paper, the issue of formalization is orthogonal to FAIRification.
+Crucially for the present paper, formalization is orthogonal to FAIRification.
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
@@
 FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).
+This principle is illustrated in our [vignette on FAIRifying De Groot's empirical cycle](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html):
+it is equally possible to FAIRify the theory in its original formulation by archiving a text document with five plain-langues propositions,
+or to formalize the theory and represent it as a human- and machine-readable diagram before FAIRifying it.
 
@@
 
+<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
+Yet scholars often produce many other valuable resources in the process of writing papers,
+including instruments, materials, data, code, and theory.
+These resources are often merely described in papers and not made available for reuse.
+Modular publishing is the practice of making each of these resources available as independent *digital objects*,
+facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
 We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
-<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
-In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
-These resources are often merely described in papers (if at all), and not made available for reuse.
-Modular publishing is the practice of making each of these resources available as independent *digital objects*,
-facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
-<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>
+</span>
+<span id="datasharing">At the time of writing, some modular publishing practices are already widely adopted;
+data sharing, for example, has become the de-facto standard in psychology in the past decade [@tedersooDataSharingPractices2021].</span>
 
@@
 
-We can take inspiration from the field of computer science for well-established processes for iteratively improving digital objects.
-Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
+The field of computer science provides inspiration for well-established processes for iteratively improving digital objects.
+Version control systems, like Git, are used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
 Git tracks line-by-line changes to text-based files,
-and maintains a complete history of those changes.
-It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
-Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
+and maintains a complete annotated history of those changes.
+It has previously been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
+For example, Git can facilitate reproducible research, manage distributed collaboration, and improve preregistrations [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
 Git provides a useful framework for developing FAIR theory,
 because it enables explicitly comparing versions of a file (or: theory),
+documenting why changes were made,
 incorporating changes by different authors,
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
-Note that archival of the version of record is *not* a function of Git(Hub).
-While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.
+Note that, while cloud archives associated with Git (e.g., GitHub) facilitate collaborative theory development, 
+they are not suitable for archiving the version of record due to a lack of FAIR-compliance.
+Thus, theory development may take place on GitHub, but versions of record should be archived on a platform like like Zenodo, with appropriate metadata.
 
@@
 guides the social process of theory development, and signals how much a theory has been changed.
-We propose the following adaptation of semantic versioning for theories:
-
-Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:
-
-- MAJOR version when you commit backwards incompatible changes, i.e., the
-      theory now contains empirical statements that are at odds with
-      a previous version of the theory
-- MINOR version when you expand the set of empirical statements in a
-      backward compatible manner (i.e., the previous version is
-      subsumed within the new version)
-- PATCH version when you make backward compatible bug fixes, cosmetic
-      changes, fix spelling errors, or add clarifications
+
+We propose adaptating semantic versioning for theories by assigning a version number in the format `MAJOR.MINOR.PATCH` (e.g., 0.1.0), where the `MAJOR` number is incremented when backwards incompatible changes are made.
+For example, if the theory now contains empirical statements that are at odds with a previous version of the theory.
+The `MINOR` number should be incremented when the set of empirical statements are expanded in a backward compatible manner (i.e., the previous version is subsumed within the new version).
+The `PATCH` number should be incremented when making backward compatible bug fixes, cosmetic changes, fix spelling errors, or add clarifications.
 
@@
 Metadata should also include identifiers for all the versions of the theory it describes	(F3);
-Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of that artifact's versions.
+Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of its subversions.
 
@@
 Zenodo stands out in this respect.
-Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
-the version of record should be archived in a FAIR repository like Zenodo.
 Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
@@
 If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
-Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index digital objects stored in Zenodo.
+Conversely, as organizations (e.g., Google Scholar, Web of Science, Pure, ORCID) begin to recognize other forms of academic output beyond papers, they may begin to index digital objects stored in Zenodo.
 
@@
 For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
-Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
+The data engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
 although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.
@@
 At present, there are several impediments to theories' Accessibility.
-One impediment is that, 
-when theories are published in paywalled journal articles,
+First, when theories are published in paywalled journal articles,
 they might not be practically Accessible to all,
-even if they are in principle Accessible to those who can pay the fee.
+even if they are in principle Accessible to paying readers.
 Open Access publishing increases the practical Accessibility of all academic output, including theory.
-
-A second impediment to Accessibility is more indirect
+A second impediment is more indirect
 and pertains to a theory's intelligibility to those with practical Access.
@@
 Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
-Shared background knowledge can come from paradigms held by members of a scientific community [@kuhnStructureScientificRevolutions2009],
-from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
-Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
-At the same time, it is important to acknowledge the *indeterminacy of translation*^[Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
-has multiple alternative translations, with no objective means of choosing the correct one.],
-which implies that it is not possible to remove *all* ambiguity when communicating an idea [@quineReasonsIndeterminacyTranslation1970].
+This can come from shared paradigms [@kuhnStructureScientificRevolutions2009],
+from education, and from the available methods and instrumentation - or it can be problematically absent.
+Accessibility is improved by explicitly referring to sources of assumed background knowledge, and by reducing unnecessary ambiguity.
+At the same time, it is important to acknowledge that it is impossible to remove *all* ambiguity when communicating an idea.
+The *indeterminacy of translation* holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula) has multiple alternative translations,
+with no objective means of choosing the correct one [@quineReasonsIndeterminacyTranslation1970].
 This places a theoretical upper bound on theories' ability to be understood.
@@
 In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
-It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
-Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+While the *indeterminacy of translation* places an upper bound on interpretability,
+scholars should nevertheless strive to reduce unnecessary ambiguity to the greatest possible extent.
 It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by cross-referencing the DOI of those nested theories [@ContributingCitationsReferences2024].
+This can be achieved by cross-referencing the DOI of those nested theories [@dataciteContributingCitationsReferences2024].
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
@@
 If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
-*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008, p. 1].
+"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's" [@mischelToothbrushProblem2008, p. 1].
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress [@degrootMethodologyFoundationsInference1969].
@@
 Copyright does not
-*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"* (Section 102, Copyright Act).
+"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery" (Section 102, Copyright Act).
 Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
@@
 
-# Making a Theory FAIR
-
-<span id="tutorial">We propose a practical, software-agnostic workflow for making theory FAIR.
+# FAIR Theory Workflow
+
+<span id="tutorial">We present a conceptual workflow for making theory FAIR,
+to give readers some sense of the steps involved. 
 While these steps can be implemented using a variety of tools,
-R users can streamline the process using the `theorytools` package, which automates most steps.
-We will illustrate the workflow using De Groot's *empirical cycle*,
-a meta-theory of theory construction,
-as a running example for this tutorial.
-The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
-
-The R-package includes further examples for FAIR theory creation in tutorial form,
-including a [worked example for implementing this workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
-Two other vignettes apply the workflow to substantive theories:
-Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000],
-and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of children's emotions [@morrisRoleFamilyContext2007].
-
-Our workflow builds upon existing open science infrastructures to the greatest possible extent.
+the `theorytools` R-package automates most steps.
+This package includes a [worked example for implementing this workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) which, as a living document,
+can be kept up-to-date with changing infrastructures.
+The package further includes tutorial examples for FAIR theory creation based on existing substantive theories, including
+an example of how to formalize and FAIRify Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000],
+how to FAIRify and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of children's emotions [@morrisRoleFamilyContext2007] and use it for causal inference,
+and an example of how to [FAIRify a mathematical model based on the Dunning-Kruger effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].
+
+To prevent the emergence of an open science "cottage industry",
+we recommend using existing open science infrastructures to the greatest possible extent.
 At the time of writing (2025),
@@
 
+Imagine that one would want to FAIRify De Groot's *empirical cycle*,
+a meta-theory of theory construction.
 Begin by creating an empty folder to hold all files associated with the theory - this folder will become the theory archive.
@@
 
-Implementing the theory as a digital object can be as simple as isolating these statements from the paper,
-and saving them to a plain text file.
+Implementing the theory as a digital object can be as simple as saving these statements to a plain text file.
 
 Optionally, we can formalize the theory further.
-According to a taxonomy of different levels of theory formalization,
-the empirical cycle is presently defined at either the "theory" or "specification" level [@guestHowComputationalModeling2021].
-To fulfill criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)),we can further formalize it to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
+According to a taxonomy of different levels of theory formalization [@guestHowComputationalModeling2021],
+the empirical cycle is currently defined at either the "theory" or "specification" level.
+To fulfill criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)), we can further formalize it to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
 Given the cyclical nature of the conceptual model, such an implementation might look like this:
+
+<!-- ``` -->
+<!--   induction   [label="formulation of hypotheses"]; -->
+<!--   deduction   [label="derivation of specific consequences from the hypotheses, -->
+<!--                       in the form of testable predictions"]; -->
+<!--   observation [label="collection and grouping of empirical materials; -->
+<!--                       (tentative) formation of hypotheses"]; -->
+<!--   test        [label="of the hypotheses against new empirical materials, by way -->
+<!--                       of checking whether or not the predictions are fulfilled"]; -->
+<!--   evaluation  [label="of the outcome of the testing procedure with respect to -->
+<!--                       the hypotheses or theories stated, as well as with a view -->
+<!--                       to subsequent, continued or related, investigations"]; -->
+
+<!--   observation -> induction; -->
+<!--   induction -> deduction; -->
+<!--   deduction -> test; -->
+<!--   test -> evaluation; -->
+<!--   evaluation -> observation; -->
+<!-- ``` -->
 
@@
 
-* Set the *resource type* to the category `Model`,
-* Add the words `FAIR theory:` to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words `meta-analysis` in the title),
-* Add `fairtheory` to the keywords to aid search engine indexation.
+* Set the `resource type` to `Model`,
+* Add the words `FAIR theory:` to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses should use the words `meta-analysis` in the title),
+* Add `fairtheory` to the keywords to aid search engine indexation (using "fair theory" causes some search engines to look for either the word "fair" or "theory", which would be overly inclusive).
 * Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
 </span>
+
+The FAIR implementation of De Groot's empirical cycle that resulted from the lead author implementing this workflow is available at <https://doi.org/10.5281/zenodo.14552329>.
 
@@
 for automatically selecting control variables, and for simulating data (using `theorytools`).
-Another vignette describes how to take Self-Determination Theory [@langeSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
+Another vignette describes how to take Self-Determination Theory [@deciSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
 Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].

### 2025-06-19 test.tex
update 2025-06-19 15:08:14.390215

@@
 We also thank the Editor, Dr.~Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
-This version of the paper was generated from Git Commit \# cfb33cc6.
-The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.
+Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id \#6e9f28c9.
+The first and senior author positions were fixed, and remaining author positions randomized.
 
@@
 
-The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M.
+\phantomsection\label{whatsfair}{The FAIR Guiding Principles (hereafter: FAIR principles) were established by a diverse consortium of stakeholders to improve the reusability of research data and other resources produced in the course of scholarly work
+by making them Findable, Accessible, Interoperable and Reusable (M.
 D.
-Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{faircap}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
-Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S.
+Wilkinson et al., 2016)\footnote{\phantomsection\label{faircap}{As the colloquial use of these terms differs from their definition according to the FAIR principles, we capitalize these terms when referring to specific FAIR principles.}}.
+Since the FAIR principles' inception,
+they have become a widely adopted standard for archival of academic output,
+representing an estimated tens of billions of dollars in reuse value (Vogt et al., 2024).
+Scholars have demonstrated their relevance for making other digital objects more open, including research software (Lamprecht et al., 2019) and computational workflows (Van Lissa et al., 2021; S.
 R.
-Wilkinson et al., 2024).
-This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
+Wilkinson et al., 2024).}
+The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
 To this end, we introduce ``FAIR theory'':
-a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
-compliant with the FAIR principles.
-\phantomsection\label{test2}{FAIR theory has the potential to improve the efficiency of scholarly communication and}
-accelerate cumulative knowledge acquisition.
+a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially associated with - the scientific paper.
+\phantomsection\label{alldefinitions}{Definitions of theory abound and hotly debated,
+but as many of them are compatible with the FAIR principles,
+this paper is not limited to a specific definition.}
+FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
 We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.
@@
 rendering them immune to falsification.
-As an example, consider ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
+Consider ``self-determination theory'' (SDT, Deci \& Ryan, 2012), one of the most widely cited social psychological theories, which we formalized and FAIRified in \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{this vignette}.
+SDT emphasizes the role of intrinsic and extrinsic motivation in human behavior.
 Intrinsic motivation was initially defined as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
@@
 Under the expanded definition, one would hypothesize the opposite, as the act fulfills the need for relatedness.
-The expanded definition thus enables SDT to absorb potentially falsifying evidence.
+Expanding the definition thus enables SDT to absorb potentially falsifying evidence.
 
 Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh et al., 2021).
-One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
-When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
-it becomes difficult do derive precise predictions,
-and therefore hard to falsify the theory.
-A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
+One frequently raised concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
+When theories are ambiguous,
+precise predictions cannot be derived from them without
+resorting to subjective interpretation or invoking additional assumptions,
+which makes them harder to falsify.
+A second concern has received less attention,
+is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
 Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
@@
 open science principles to psychological theory.
-We apply introduce \emph{FAIR theory} to
-facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
-
-\subsection{Theory and Scientific Progress}\label{theory-and-scientific-progress}
-
-According to the \emph{empirical cycle} (De Groot \& Spiekerman, 1969),
-a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}a.
-In one half of the cycle, labeled the ``Context of Justification'' by Wagenmakers and colleagues, hypotheses derived from theory are tested on data.
-In the other half of the cycle (the ``Context of Discovery''),
-patterns observed in data are generalized to theoretical principles, Figure \ref{fig:figecs}b.
-In this model, theories are the vehicle of scientists' understanding of phenomena.
-Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
-
-\begin{figure}
-\centering
-\includegraphics{ec_visio.pdf}
-\caption{\label{fig:figecs}Three implementations of the ``empirical cycle'' (De Groot \& Spiekerman, 1969).}
-\end{figure}
-
-In a progressive research program (Lakatos, 1971),
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
-There are, however, indications that contemporary psychology falls short of this ideal.
-Firstly, because hypothesis-testing research is over-represented in the literature:
-According to Kühberger et al. (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
-Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, et al., 2021).
-Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
-The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
-Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
-in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi et al., 2024).
-Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer et al., 2004),
-but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
-Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
-Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).
-
-Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
-The majority of papers ostensibly test hypotheses,
-but these are rarely connected to, let alone derived from, theory.
-Moreover, test results do not routinely contribute to the improvement or rejection of theories.
-The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge.
-They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978, p.
-1).
-The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.
-
-\subsection{Making Theory FAIR}\label{making-theory-fair}
-
-Merely publishing theory in a journal article does not make it open;
-to be open, theory should adhere to established open science standards for specification and archival.
-We propose to implement theories as information artifacts,
-and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
-Metadata are ``data about the data''.
-They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
-FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
-\emph{Accessible} in a machine- and human-readable filetype;
-\emph{Interoperable} for specific purposes, for example, within the data analysis environment;
-and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
-Following the original proposal of Lamprecht and colleagues,
-we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
-We reflect on the necessary changes (which are minor),
-as well as on the current state and future of FAIR theory in psychology.
-The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
-and we provide worked examples to encourage their adoption.
+We introduce \emph{FAIR theory} as a methodology that can facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.
 
@@
 
-Definitions of theory are abundant, and are the subject of extensive scholarly debate.
 Given that a pluriformity of definitions are consistent with FAIR theory principles,
-our suggested approach is not limited to any one particular definition.
+we do not limited ourselves to any one particular definition - although, at times, our writing inevitably reveals a particular vantage point.
 Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021).
@@
 acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
-We recognize the syntactic view in Meehl's (1990) hierarchy of ever-more specific ``statements'' a theory might contain:
+We recognize the syntactic view in Meehl's (1990) hierarchy of ever-more specific ``statements'' a theory might contain (Table \ref{tab:tabmeehl}):
 statements about the types of entities postulated (i.e., ontology),
@@
 
-It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
+\subsection{Theory and Scientific Progress}\label{theory-and-scientific-progress}
+
+According to the \emph{empirical cycle} (De Groot \& Spiekerman, 1969),
+a meta-theoretical model of cumulative knowledge acquisition,
+research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}a.
+In one half of the cycle, labeled the ``Context of Justification'' by Wagenmakers and colleagues, hypotheses derived from theory are tested on data (2018).
+In the other half of the cycle (the ``Context of Discovery''),
+patterns observed in data are generalized to theoretical principles, Figure \ref{fig:figecs}b.
+In this model, theories are the vehicle of scientists' understanding of phenomena.
+Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
+
+\begin{figure}
+\centering
+\includegraphics{ec_visio.pdf}
+\caption{\label{fig:figecs}Three implementations of the ``empirical cycle'' (De Groot \& Spiekerman, 1969).}
+\end{figure}
+
+In a progressive research program (Lakatos, 1971),
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+There are, however, indications that contemporary psychology falls short of this ideal.
+Meehl observed that theories in psychology ``lack the cumulative character of scientific knowledge.
+They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978, p.
+1).
+Recent empirical findings confirm this view.
+Firstly, because hypothesis-testing research is vastly over-represented in the literature, amounting to 89.6\% of published papers (Kühberger et al., 2014).
+Closer examination of such studies reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, et al., 2021).
+Only 15\% of hypothesis-testing studies referenced any theory,
+and rarely in direct relation to the hypothesis (McPhetres et al., 2021).
+Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
+The majority of papers ostensibly test hypotheses,
+but these are rarely connected to theory.
+
+Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
+in which case it would be useful to make these explicit (Fried, 2020; Norouzi et al., 2024).
+Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched cultural practices (Gigerenzer et al., 2004),
+such as straw-man null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
+\phantomsection\label{waste}{Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
+cannot advance our principled understanding of psychological phenomena,
+and consequently contributes to research waste (Nakagawa et al., 2024).
+\phantomsection\label{norms}{Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021); inefficient at best, and more likely, futile.
+As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must ``avoid research waste'',
+our field should take seriously its ethical responsibility to develop procedures to reduce it.}}
+The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory;
+instantiating theory as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.
+
+\subsection{Making Theory FAIR}\label{making-theory-fair}
+
+Merely publishing theory in a journal article does not make it open;
+to be open, theory should adhere to established open science standards for specification and archival.
+We propose to implement theories as digital objects,
+and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
+Metadata are ``data about the data''.
+They provide information about the nature and content of a digital object and are stored in the repository where the version of record of the FAIR theory is deposited.
+FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
+\emph{Accessible} in a machine- and human-readable filetype;
+\emph{Interoperable} for specific purposes, for example, within the data analysis environment;
+and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
+Following the original proposal of Lamprecht and colleagues (2019),
+we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
+We reflect on the necessary (minor) changes,
+as well as on the current state and future of FAIR theory in psychology.
+The resulting principles provide guidance for instantiating theory as a FAIR digital object,
+and we provide worked examples to encourage their adoption.
+
+\subsubsection{What to Archive?}\label{what-to-archive}
+
+It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be archived as FAIR theory.
 As the practice of FAIRification becomes more embedded,
-we expect that it will become increasingly clear what kind and form of information is useful.
+we expect that it will become increasingly clear what kind of information is useful.
 As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
@@
 Going even further, a \emph{generative/computational model} could be specified,
-which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
+which is completely parameterized (e.g., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
 Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
-Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement tools (e.g., different questionnaires used to assess the same construct), or study subjects (e.g., specific strains of rats).
-
-As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
+Examples include experimental designs (e.g., longitudinal designs observing change over time), measurement instruments (e.g., different questionnaires used to assess the same construct), or information about participant recruitment- and retention strategies.
+
+Theories can include or reference other theories.
+For example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
 such as adherence to infection prevention protocols (e.g., social distancing),
@@
 (Taylor, 2022).
-The theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
+Such a theory may encompass a particular transmission \emph{model} for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.
 
@@
 
-Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
-More rigorous formalization increases theories' falsifiability (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
+Concerns about the state of psychological theory have motivated increasing calls for greater theory ``formalization'' (Smaldino, 2017; cf. Oude Maatman, 2021).
+Formalization increases theories' falsifiability (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
 For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
@@
 
+Crucially for the present paper, formalization is orthogonal to FAIRification.
 FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
-rather, it increases the fidelity and ease with which they are communicated.
-Thus, FAIR theory does not require theories to be formal.
-At the same time, formal theories are not automatically FAIR.
+rather, it pertains to rigorous and transparent archival and communication about theories, with the aim of enhancing their reusability.
+FAIR theory does not require formalization, and formal theories are not automatically FAIR.
 The FAIR principles
-can be applied to theories representated in natural language,
+apply to theories representated in natural language,
 as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
@@
 FAIR theory is thus consistent with, but does not require, formalization (also see the section on \emph{Accessibility} below).
+This principle is illustrated in our \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{vignette on FAIRifying De Groot's empirical cycle}:
+it is equally possible to FAIRify the theory in its original formulation by archiving a text document with five plain-langues propositions,
+or to formalize the theory and represent it as a human- and machine-readable diagram before FAIRifying it.
 
@@
 
+\phantomsection\label{modularvalue}{The primary unit of scientific communication has long been the academic paper.
+Yet scholars often produce many other valuable resources in the process of writing papers,
+including instruments, materials, data, code, and theory.
+These resources are often merely described in papers and not made available for reuse.
+Modular publishing is the practice of making each of these resources available as independent \emph{digital objects},
+facilitating their reuse and making them citable (Van De Sompel et al., 2004).
 We envision FAIR theory as an instance of modular publishing (Kircz, 1998).
-\phantomsection\label{modularvalue}{The primary unit of scientific communication has long been the academic paper.
-In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
-These resources are often merely described in papers (if at all), and not made available for reuse.
-Modular publishing is the practice of making each of these resources available as independent \emph{information artifacts},
-facilitating their reuse and making them citable (Van De Sompel et al., 2004).}
-\phantomsection\label{datasharing}{Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology (Tedersoo et al., 2021).}
+}
+\phantomsection\label{datasharing}{At the time of writing, some modular publishing practices are already widely adopted;
+data sharing, for example, has become the de-facto standard in psychology in the past decade (Tedersoo et al., 2021).}
 
@@
 To maintain a persistent record of scholarly communication,
-Zenodo mints DOIs for information artifacts.
-The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources (DataCite Metadata Working Group, 2024).
+Zenodo mints DOIs for digital objects.
+The DataCite Metadata Schema offers a standard way to document digital objects with relevant metadata which increases their findability and documents their relationships to other resources (DataCite Metadata Working Group, 2024).
 For example, data can be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset}.
@@
 If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.
-\phantomsection\label{modularpublisingdetract}{Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for argumentation,
-nuance,
-and the author's voice.
+\phantomsection\label{modularpublisingdetract}{
+Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for greater nuance and the author's voice.
+Theories published in traditional papers can be supplemented by FAIR versions that live independently, evolve collaboratively, and feed into reproducible workflows.
 }
@@
 
-We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
-Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
+The field of computer science provides inspiration for well-established processes for iteratively improving digital objects.
+Version control systems, like Git, are used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
 Git tracks line-by-line changes to text-based files,
-and maintains a complete history of those changes.
-It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
-Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert et al., 2021; Van Lissa et al., 2021).
+and maintains a complete annotated history of those changes.
+It has previously been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
+For example, Git can facilitate reproducible research, manage distributed collaboration, and improve preregistrations (Peikert et al., 2021; Van Lissa et al., 2021).
 Git provides a useful framework for developing FAIR theory,
 because it enables explicitly comparing versions of a file (or: theory),
+documenting why changes were made,
 incorporating changes by different authors,
@@
 This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
-Note that archival of the version of record is \emph{not} a function of Git(Hub).
-While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.
+Note that, while cloud archives associated with Git (e.g., GitHub) facilitate collaborative theory development,
+they are not suitable for archiving the version of record due to a lack of FAIR-compliance.
+Thus, theory development may take place on GitHub, but versions of record should be archived on a platform like like Zenodo, with appropriate metadata.
 
@@
 In computer science, these challenges are navigated by assigning version numbers.
-Specifically, \emph{semantic versioning} comprises a simple set of rules for assigning version numbers to information artifacts.
+Specifically, \emph{semantic versioning} comprises a simple set of rules for assigning version numbers to digital objects.
 Whereas version control tracks changes,
@@
 guides the social process of theory development, and signals how much a theory has been changed.
-We propose the following adaptation of semantic versioning for theories:
-
-Given a version number in the format MAJOR.MINOR.PATCH (where MAJOR, MINOR, and PATCH are placeholders for positive integer numbers including zero), increment the:
-
-\begin{itemize}
-\tightlist
-\item
-  MAJOR version when you commit backwards incompatible changes, i.e., the
-  theory now contains empirical statements that are at odds with
-  a previous version of the theory
-\item
-  MINOR version when you expand the set of empirical statements in a
-  backward compatible manner (i.e., the previous version is
-  subsumed within the new version)
-\item
-  PATCH version when you make backward compatible bug fixes, cosmetic
-  changes, fix spelling errors, or add clarifications
-\end{itemize}
+
+We propose adaptating semantic versioning for theories by assigning a version number in the format \texttt{MAJOR.MINOR.PATCH} (e.g., 0.1.0), where the \texttt{MAJOR} number is incremented when backwards incompatible changes are made.
+For example, if the theory now contains empirical statements that are at odds with a previous version of the theory.
+The \texttt{MINOR} number should be incremented when the set of empirical statements are expanded in a backward compatible manner (i.e., the previous version is subsumed within the new version).
+The \texttt{PATCH} number should be incremented when making backward compatible bug fixes, cosmetic changes, fix spelling errors, or add clarifications.
 
@@
 Metadata should also include identifiers for all the versions of the theory it describes (F3);
-Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.
+Zenodo handles this by default by providing an overarching DOI for a digital object which subsumes the DOIs of its subversions.
 
@@
 Zenodo stands out in this respect.
-Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work (Ram, 2013),
-the version of record should be archived in a FAIR repository like Zenodo.
 Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
@@
 This is a known problem in relation to research data and software (Katz \& Chue Hong, 2024).
-Regrettably, most academic search engines only index traditional print publications, not other information artifacts.
+Regrettably, most academic search engines only index traditional print publications, not other digital objects.
 Since the status quo is to publish theories in papers,
 the FAIR requirements are met if scholars continue to do so,
-and additionally publish theories as separate information artifacts.
+and additionally publish theories as separate digital objects.
 The \texttt{"fairtheory"} keyword can also be used to signal the presence of theory within a paper.
 In the longer term, it may not be necessary to write a paper for each theory.
-If Zenodo becomes more recognized as a centralized repository for information artifacts, researchers may begin to search there more regularly.
-Conversely, as organizations begin to recognize the value in tracking academic output other than papers, repositories may begin to index information artifacts stored in Zenodo.
+If Zenodo becomes more recognized as a centralized repository for digital objects, researchers may begin to search there more regularly.
+Conversely, as organizations (e.g., Google Scholar, Web of Science, Pure, ORCID) begin to recognize other forms of academic output beyond papers, they may begin to index digital objects stored in Zenodo.
 
@@
 For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
-Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
+The data engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
 although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.
@@
 At present, there are several impediments to theories' Accessibility.
-One impediment is that,
-when theories are published in paywalled journal articles,
+First, when theories are published in paywalled journal articles,
 they might not be practically Accessible to all,
-even if they are in principle Accessible to those who can pay the fee.
+even if they are in principle Accessible to paying readers.
 Open Access publishing increases the practical Accessibility of all academic output, including theory.
-
-A second impediment to Accessibility is more indirect
+A second impediment is more indirect
 and pertains to a theory's intelligibility to those with practical Access.
@@
 Successful communication requires shared background knowledge between sender and receiver (Vogt et al., 2024).
-Shared background knowledge can come from paradigms held by members of a scientific community (Kuhn, 2009),
-from education, and from the available instrumentation for observation, measurement, and analysis - or it can be problematically absent.
-Accessibility is improved by explicitly referring to sources of assumed backround knowledge, and by reducing unnecessary ambiguity.
-At the same time, it is important to acknowledge the \emph{indeterminacy of translation}\footnote{Every communicative utterance (e.g., a statement in natural language, a mathematical formula, et cetera)
-  has multiple alternative translations, with no objective means of choosing the correct one.},
-which implies that it is not possible to remove \emph{all} ambiguity when communicating an idea (Quine, 1970).
+This can come from shared paradigms (Kuhn, 2009),
+from education, and from the available methods and instrumentation - or it can be problematically absent.
+Accessibility is improved by explicitly referring to sources of assumed background knowledge, and by reducing unnecessary ambiguity.
+At the same time, it is important to acknowledge that it is impossible to remove \emph{all} ambiguity when communicating an idea.
+The \emph{indeterminacy of translation} holds that every communicative utterance (e.g., a statement in natural language, a mathematical formula) has multiple alternative translations,
+with no objective means of choosing the correct one (Quine, 1970).
 This places a theoretical upper bound on theories' ability to be understood.
@@
 In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
-It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
-Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
+While the \emph{indeterminacy of translation} places an upper bound on interpretability,
+scholars should nevertheless strive to reduce unnecessary ambiguity to the greatest possible extent.
 It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
@@
 formalization,
-and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J. Lange et al., 2025).
+and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (Lange et al., 2025).
 
@@
 
-Interoperability pertains to the property of information artifacts to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016, p. 2).
+Interoperability pertains to the property of digital objects to ``integrate or work together {[}\ldots{]} with minimal effort'' (M. D. Wilkinson et al., 2016, p. 2).
 Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
@@
 be instantiated in a human- and machine-readable datatype,
-as should all information artifacts created while performing scholarly work (Van Lissa et al., 2021).
+as should all digital objects created while performing scholarly work (Van Lissa et al., 2021).
 Depending on the level of formalization of the theory,
@@
 the same image represented in the DOT language (\emph{{DOT Language}}, 2024)
-for representing graphs does meet this ideal (an example of such a DOT representation is given below).
+for representing graphs does meet this ideal.
 
@@
 for example, a theory that specifies causal relationships between constructs could refer back to an ontological theory from which those constructs are derived.
-This can be achieved by cross-referencing the DOI of those nested theories (\emph{Contributing {Citations} and {References}}, 2024).
+This can be achieved by cross-referencing the DOI of those nested theories (DataCite, 2024).
 The second part of this principle allows for tracing the provenance of a theory; keeping track of its prior versions and other theories that inspired it.
@@
 and this paper takes one step towards establishing more fine-grained community standards for FAIR theory.
-\href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{This is an example of FAIR metadata} extracted from Zenodo.
+\href{https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json}{This is an example of FAIR metadata} extracted from Zenodo.
 
 If we consider the current state of Reusability in psychological theory, there appears to be a norm against theory reuse:
-\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008, p. 1).
+``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's'' (Mischel, 2008, p. 1).
 As cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data, such a norm impedes scientific progress (De Groot \& Spiekerman, 1969).
@@
 Copyright does not
-\emph{``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery''} (Section 102, Copyright Act).
+``extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery'' (Section 102, Copyright Act).
 Copyright thus applies to creative works expressing a theory (e.g., prose, visual illustrations),
@@
 
-\section{Making a Theory FAIR}\label{making-a-theory-fair}
-
-To concretize the FAIR principles,
-we propose an applied workflow for making theory FAIR.
-The workflow does not require the use of R,
-but for R-users, most steps have been automated in the \texttt{theorytools} package.
-We present a brief summary of the conceptual workflow here,
-to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.
-The R-package includes a living document with the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{latest version of the workflow in tutorial format},
-to guide users in making their theories FAIR.
-The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
-At the time of writing (2024),
+\section{FAIR Theory Workflow}\label{fair-theory-workflow}
+
+We present a conceptual workflow for making theory FAIR,
+to give readers some sense of the steps involved.
+While these steps can be implemented using a variety of tools,
+the \texttt{theorytools} R-package automates most steps.
+This package includes a \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{worked example for implementing this workflow} which, as a living document,
+can be kept up-to-date with changing infrastructures.
+The package further includes tutorial examples for FAIR theory creation based on existing substantive theories, including
+an example of how to formalize and FAIRify Decy and Ryan's \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{\emph{Self-Determination Theory}} (Ryan \& Deci, 2000),
+how to FAIRify and Morris' \href{https://cjvanlissa.github.io/theorytools/articles/causal-inference.html}{\emph{Tripartite Model}} of parental socialization of children's emotions (Morris et al., 2007) and use it for causal inference,
+and an example of how to \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{FAIRify a mathematical model based on the Dunning-Kruger effect} (Feld et al., 2017).
+
+To prevent the emergence of an open science ``cottage industry'',
+we recommend using existing open science infrastructures to the greatest possible extent.
+At the time of writing (2025),
 the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets \emph{all} FAIR principles.
@@
 a popular platform in psychology.
-Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory amongst users of that platform,
+Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory among users of that platform,
 while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
@@
 
-\subsection{1. Implementing the Theory}\label{implementing-the-theory}
-
-We will use the \emph{empirical cycle} as a running example for this tutorial.
-The empirical cycle, described on page 28 of De Groot and Spiekerman (1969),
-is a meta-theory of theory construction.
-The resulting FAIR implementation of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
-The original theory consists of a series of natural language statements:
+\subsection{1. Implement the Theory}\label{implement-the-theory}
+
+Imagine that one would want to FAIRify De Groot's \emph{empirical cycle},
+a meta-theory of theory construction.
+Begin by creating an empty folder to hold all files associated with the theory - this folder will become the theory archive.
+The first file to create is the theory itself.
+This could be a plain-text file containing natural language statements,
+or a more formal representation, such as a directed graph.
+For example, the empirical cycle was originally described as a series of natural language statements (De Groot \& Spiekerman, 1969, p. 28):
 
@@
 
-If we compare this to the levels of theory formalization (Guest \& Martin, 2021),
-it is defined at either the ``theory'' or ``specification'' level.
-This theory can be FAIRified in its present form by archiving a text document containing five natural language statements.
-However, it is also possible to increase the level of formalization,
-and present an ``implementation'' of the theory as a directed graph (note: the implementation printed here is simplified; see the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{tutorial} for technical details).
-This fulfills criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
+Implementing the theory as a digital object can be as simple as saving these statements to a plain text file.
+
+Optionally, we can formalize the theory further.
+According to a taxonomy of different levels of theory formalization (Guest \& Martin, 2021),
+the empirical cycle is currently defined at either the ``theory'' or ``specification'' level.
+To fulfill criterion I1 of the FAIR principles: using a formal language for knowledge representation \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{see Supplemental Table S1}), we can further formalize it to the ``implementation'' level by specifying it in the DOT language for describing directed graphs\footnote{Presented here in a simplified form; see the \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{tutorial} for technical details}.
 Given the cyclical nature of the conceptual model, such an implementation might look like this:
@@
 \begin{verbatim}
-  induction;
-  deduction;
-  observation;
-  test;
-  evaluation;
+  induction   [label="formulation of hypotheses"];
+  deduction   [label="derivation of specific consequences from the hypotheses,
+                      in the form of testable predictions"];
+  observation [label="collection and grouping of empirical materials;
+                      (tentative) formation of hypotheses"];
+  test        [label="of the hypotheses against new empirical materials, by way
+                      of checking whether or not the predictions are fulfilled"];
+  evaluation  [label="of the outcome of the testing procedure with respect to
+                      the hypotheses or theories stated, as well as with a view
+                      to subsequent, continued or related, investigations"];
 
@@
 Figure \ref{fig:figecs}a) shows what this implementation looks like when plotted.
-Regardless of which implementation we prefer - we can save it to a text file.
-This text file is the ``information artifact'' containing our theory.
-
-\subsection{2. Create a Theory Archive}\label{create-a-theory-archive}
-
-To help meet the Interoperability and Reusability criteria,
-it is important to bundle the theory file with additional files containing documentation.
+Regardless of which implementation we prefer,
+we can save it to a plain text file - this is the ``digital object'' containing our theory.
+
+\subsection{2. Document the Theory}\label{document-the-theory}
+
+To meet the Interoperability and Reusability criteria,
+it is important to properly document the theory file.
 Firstly, add a README.md file with instructions for future users of your theory.
-The \texttt{theorytools} package contains a \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{vignette} with guidance on writing a README file for theory.
+The \texttt{theorytools} package contains a \href{https://cjvanlissa.github.io/theorytools/articles/readme.html}{vignette} on writing README files for theory.
 Secondly, add a LICENSE file with the legal conditions for reuse.
-As explained previously, we recommend explicitly waiving copyright with the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
-Place all files in a folder together - this folder will become the theory archive.
-
-\subsection{3. Version Control the Archive}\label{version-control-the-archive}
-
-To track all changes to our theory, the previously created folder can be version controlled with \href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{Git}.
-To invite reuse and collaboration, it is further recommended to create a backup in the cloud on a service like \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{GitHub}.
+We recommend explicitly waiving copyright with the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.
+
+\subsection{3. Version Control the Theory Archive}\label{version-control-the-theory-archive}
+
+To track all changes to our theory, the theory archive can be version controlled.
+\href{https://git-scm.com/book/en/v2/Getting-Started-Installing-Git}{Git} is well-suited for this purpose.
+Hosting a backup in the cloud on a platform like \href{https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/}{GitHub} additionally makes the theory publicly accessible and facilitates community engagement.
 
@@
 
-Zenodo is a FAIR compliant repository for archiving information artifacts with GitHub integration.
-Whereas GitHub facilitates collaboration - it is insufficient for archival purposes.
-Major versions of a theory, developed on GitHub, should be archived on Zenodo.
-This process can be fully automated.
+Archiving major versions of a theory in a FAIR-compliant repository that issues a persistent identifier (DOI) improves their Findability and allows them to be referenced in perpetuity.
+Zenodo is a FAIR compliant repository with GitHub- and OSF integration.
 
@@
 
-Providing relevant metadata improves the Findability of a FAIR theory archived on Zenodo.
-See here for \href{https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json}{an example of the metadata} associated with our FAIR empirical cycle.
-We recommend setting the \emph{resource type} to \texttt{Model}, adding the words \texttt{FAIR\ theory:} to the title so that sentient readers to recognize the work as a FAIR theory, and adding \texttt{fairtheory} as a keyword to aid search engine indexation.
+When archiving a FAIR theory, documenting it with relevant metadata improves its Findability.
+We recommend using a standardized metadata schema like DataCite (DataCite Metadata Working Group, 2024).
+See here for \href{https://github.com/cjvanlissa/fair_theory/blob/3b4894da576cb76d19e911a05dd513d5172058ec/example_metadata.json}{an example of the metadata} associated with our FAIR empirical cycle.
+Within this schema,
+
+\begin{itemize}
+\tightlist
+\item
+  Set the \emph{resource type} to the category \texttt{Model},
+\item
+  Add the words \texttt{FAIR\ theory:} to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words \texttt{meta-analysis} in the title),
+\item
+  Add \texttt{fairtheory} to the keywords to aid search engine indexation.
+\item
   Optionally, submit the theory to the \href{https://zenodo.org/communities/fairtheory}{``FAIR Theory Community''} to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
-As mentioned before, \href{https://cjvanlissa.github.io/theorytools/articles/fair-theory.html}{these five steps are partly automated for R-users}.
+\end{itemize}
+
+The FAIR implementation of De Groot's empirical cycle that resulted from the lead author implementing this workflow is available at \url{https://doi.org/10.5281/zenodo.14552329}.
 
@@
 for automatically selecting control variables, and for simulating data (using \texttt{theorytools}).
-Another vignette describes how to take Self-Determination Theory (P.
-A.
-M.
-V.
-Lange et al., 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
+Another vignette describes how to take Self-Determination Theory (Deci \& Ryan, 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
 Another vignette describes how to take the Dunning-Krüger effect and \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{specify it as a FAIR mathematical formula} (Feld et al., 2017).
@@
 We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
-We demonstrated how to create, version-control, and archive theories as digital information artifacts.
+We demonstrated how to create, version-control, and archive theories as digital objects.
 We introduced the \texttt{theorytools} R-package to partly automate these processes, reducing the barrier of entry for researchers,
@@
 One important strength of FAIR theory is that it provides much-needed open science methods for the underemphasized inductive phase of the empirical cycle.
-Most extant open science methods focus on increased rigor in testing, but provide little guidance as to what to do with the newly collected empirical evidence.
-By providing much-needed open science methods for theory construction,
-FAIR theory helps restore the balance between inductive and deductive research and contributes to closing the ``open empirical cycle'' (Hoijtink et al., 2023).
+\phantomsection\label{openec}{Recently, the ``open empirical cycle'' was introduced, arguing that each phase in De Groot's model of cumulative knowledge generation via scientific research can be supported by specific open science practices to increase the transparency, quality, trustworthiness, and replicability of research (Hoijtink et al., 2023).
+As we identified, however, most existing open science methods focus on rigor in testing (phases 2-4 of the cycle),
+but few provide guidance on how to derive hypotheses from theory (phase 1), or how to relate empirical findings back to theory (phase 5), leaving a gap in the cycle.
+By instantiating theory as a FAIR digital object,
+we provide much-needed open science infrastructure for transparently deriving hypotheses and modifying theory,
+thus contributing to closing the ``open empirical cycle''.}.
 
@@
 Given the current emphasis on empirical papers (McPhetres et al., 2021), theoretical papers can be hard to publish.
-FAIR theories, by contrast, can be readily disseminated as citable information artifacts, thus changing the incentive structure to favor theory development.
+FAIR theories, by contrast, can be readily disseminated as citable digital objects, thus changing the incentive structure to favor theory development.
 
@@
 
-Another limitation is the learning curve associated with tools like Git and Zenodo.
+\phantomsection\label{learningcurve}{Another limitation is the learning curve associated with tools like Git and Zenodo.
 The \texttt{theorytools} R-package mitigates this limitation for R-users by automating key steps in the process.
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.
-One barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
+One final way to address the learning curve is via specialization and collaboration, or team science (Van Lissa et al., 2024):
+as scientific workflows increase in sophistication, it is increasingly difficult for any one scholar to master all skills involved.
+In relation to FAIR theory, we see unique opportunities for intergenerational collaboration and knowledge exchange,
+as theoreticians tend to be seasoned experts, whereas open science literacy is more commonly found among early career scholars.}
+
+One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
 Education might help address this limitation; with this in mind,
@@
 
-\bibitem[\citeproctext]{ref-ContributingCitationsReferences2024}
-\emph{Contributing {Citations} and {References}}. (2024). DataCite Support. \url{https://support.datacite.org/docs/data-citation}
-
 \bibitem[\citeproctext]{ref-cramerMajorDepressionComplex2016}
@@
 
+\bibitem[\citeproctext]{ref-dataciteContributingCitationsReferences2024}
+DataCite. (2024). \emph{Contributing {Citations} and {References}}. DataCite Support. \url{https://support.datacite.org/docs/data-citation}
+
 \bibitem[\citeproctext]{ref-datacitemetadataworkinggroupDataCiteMetadataSchema2024}
@@
 Deci, E. L. (1971). Effects of externally mediated rewards on intrinsic motivation. \emph{Journal of Personality and Social Psychology}, \emph{18}(1), 105--115. \url{https://doi.org/10.1037/h0030644}
+
+\bibitem[\citeproctext]{ref-deciSelfDeterminationTheory2012}
+Deci, E.
+L., \& Ryan, R.
+M.
+(2012).
+Self-{Determination Theory}.
+In P.
+A.
+M.
+V.
+Lange, A.
+W.Kruglanski, \& E.
+ToryHiggins (Eds.), \emph{Handbook of {Theories} of {Social Psychology}: {Volume} 1} (pp.
+416--437).
+SAGE Publications Ltd.
+\url{https://doi.org/10.4135/9781446249215}
 
@@
 
-\bibitem[\citeproctext]{ref-langeSelfDeterminationTheory2012}
-Lange, P.
-A.
-M.
-V., W.Kruglanski, A., ToryHiggins, E., Deci, E.
-L., \& Ryan, R.
-M.
-(2012).
-Self-{Determination Theory}.
-In \emph{Handbook of {Theories} of {Social Psychology}: {Volume} 1} (pp.
-416--437).
-SAGE Publications Ltd.
-\url{https://doi.org/10.4135/9781446249215}
-
 \bibitem[\citeproctext]{ref-lavelleWhenCrisisBecomes2021}
@@
 
+\bibitem[\citeproctext]{ref-nakagawaPoorHypothesesResearch2024}
+Nakagawa, S., Armitage, D., Froese, T., Yang, Y., \& Lagisz, M.
+(2024, May 27).
+\emph{Poor hypotheses and research waste in biology: Learning from a theory crisis in psychology}.
+\url{https://doi.org/10.32942/X2S03W}
+
 \bibitem[\citeproctext]{ref-norouziCapturingCausalClaims2024}
@@
 
+\bibitem[\citeproctext]{ref-oudemaatmanPsychologysTheoryCrisis2021}
+Oude Maatman, F. (2021, July 12). \emph{Psychology's {Theory Crisis}, and {Why Formal Modelling Cannot Solve It}}. \url{https://doi.org/10.31234/osf.io/puqvs}
+
 \bibitem[\citeproctext]{ref-pearlCausalDiagramsEmpirical1995}
@@
 \url{https://doi.org/10.1177/1745691620966795}
+
+\bibitem[\citeproctext]{ref-smaldinoModelsAreStupid2017}
+Smaldino, P.
+E.
+(2017).
+Models {Are Stupid}, and {We Need More} of {Them}.
+In R.
+R.
+Vallacher, S.
+J.
+Read, \& A.
+Nowak (Eds.), \emph{Computational {Social Psychology}} (1st ed., pp.
+311--331).
+Routledge.
+\url{https://doi.org/10.4324/9781315173726-14}
 

### 2025-06-19 test.Rmd
update 2025-06-19 15:53:45.231145

@@
 These aDAGs are X-interoperable for plotting (using `dagitty` and `tidySEM`),
-for automatically selecting control variables, and for simulating data (using `theorytools`).
-Another vignette describes how to take Self-Determination Theory [@deciSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
-Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].
-Another vignette illustrates the use of FAIR theory for [covariate selection and causal inference](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].
-More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
+for automatically selecting control variables, and for simulating data (using `theorytools`),
+as described in [this vignette](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].
+Other vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
 
@@
 access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
-Whereas the idea of theory can be quite nebulous to empirical psychologists,
-FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
+Whereas "theory" can be a nebulous concept to empirical psychologists,
+FAIR theory makes it practical and tangible, incorporating theory into scholarly workflows.
 Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
@@
 we provide much-needed open science infrastructure for transparently deriving hypotheses and modifying theory,
-thus contributing to closing the "open empirical cycle".</span>.
-
-Our approach aligns closely with contemporary developments in open science,
+thus contributing to closing the "open empirical cycle".</span>
+
+Our approach aligns closely with recent developments in open science,
 such as modular publishing, interdisciplinarity, meta-research, and team science.
@@
 FAIR theories, by contrast, can be readily disseminated as citable digital objects, thus changing the incentive structure to favor theory development.
-
-Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling [@haslbeckModelingPsychopathologyData2021].
+FAIR theory's Accessibility can advance interdisciplinarity,
+for example because theoretical frameworks can be reused, adapted, or used for analogical modeling across different fields [@haslbeckModelingPsychopathologyData2021].
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
@@
 One important limitation of the present work is that,
-while we build on well-established information architecture like Zenodo,
-it is unlikely that the proposed metadata standard is definitive.
+while we build on well-established infrastructures like Zenodo,
+it is unlikely that the proposed workflow is definitive.
 Community adoption can reveal areas of further improvement.
@@
 
-<span id="learningcurve">Another limitation is the learning curve associated with tools like Git and Zenodo.
+<span id="learningcurve">Another limitation is the learning curve associated with tools and infrastructures like Git and Zenodo.
 The `theorytools` R-package mitigates this limitation for R-users by automating key steps in the process.
@@
 
-One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the "toothbrush problem".
+One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories (the "toothbrush problem").
 Education might help address this limitation; with this in mind,
@@
 <span id="measurementcrisis">Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
-Since FAIR theories can reference other theories and resources,
-it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.</span>
+FAIR theories could reference specific measurement instruments, or even theories of measurement,
+when operationalizing constructs named in a theory.</span>
 <span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,
@@
 
-FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
-It provides much-needed open science methods for the inductive phase of the empirical cycle,
-closing a critical gap in the scientific process.
-
-<span id = "turkheimer">We invite the reader to ask the rhetorical question raised by our reviewer, prof.
-Turkheimer, who - reflecting on a long career as a practitioner of theory in psychology - considered "how adopting a FAIR framework would change, improve, or inhibit what I do".
-We paraphrase his answer here^[Edited for brevity.], because his lived experience complements our youthful "techno-enthusiasm".
+<span id = "turkheimer">We invite the reader to ask the rhetorical question raised by one of our reviewers, prof.
+Turkheimer, who - reflecting on a career as a practitioner of theory in psychology - considered "how adopting a FAIR framework would change, improve, or inhibit what I do".
+<!-- We paraphrase his answer here^[Edited for brevity.], because his lived experience complements our youthful "techno-enthusiasm". -->
 Prof. Turkheimer remarked that a FAIR framework would have been very helpful in maintaining meaningful boundaries between a theory and the predictions it makes,
@@
 and in resolving theoretical disputes.
-*"FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
+"FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
 Everything published about a theory can continue to exist in its original form,
@@
 Importantly, FAIR theory can evolve semi-independently from individual papers, offering a persistent and collaborative object that others can reuse, cite, and refine. 
-Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability."*</span>
-
-We envision a future where application of the FAIR principles makes theories more *useful*, living up to Kurt Lewin's ideal,
+Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability."</span>
+
+We envision a future where application of the FAIR principles makes theories more *useful*,
+living up to Kurt Lewin's ideal,
 enabling scholars to incorporate theory into their workflows in a tanglible way,
@@
 and proposing specific changes to existing theories based on empirical results.
+FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
+It provides much-needed open science methods for the inductive phase of the empirical cycle,
+closing a critical gap in the scientific process.
 This paves the way for more theory-driven scholarship

### 2025-06-19 test.Rmd
update 2025-06-19 16:20:01.419007

@@
 and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
-<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
-<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
 Following the original proposal of Lamprecht and colleagues [-@lamprechtFAIRPrinciplesResearch2019],

### 2025-06-19 action_letter.Rmd
update 2025-06-19 22:16:44.792218

@@
 
-<!-- Make comparison to preregistration; that was first proposed and only later evaluated. -->
-<!-- Mention how few of preregistered hypotheses are true, and yet how theory is rarely updated? Thus, falsificationism is not working in practice. -->
-
-<!-- <!-- AB: --> -->
-<!-- 2.	I think the policies argument is interesting.
-I agree that currently we simply claim that establish FAIR theories will profit the advancement of science by an argument of analogy (to open data and open code) but ultimately we do not know.
-For one, I think it's fair to admit this.
-Second, I would argue that FAIR theories is not a very strict policy (in contrast to other stances in science, which require very specific ways of working, e.g.
-commiting to falsificationism, adhering to ethical principles, etc.).
-Really, FAIR theories are no restriction in that sense but more of a protocol for communication, which ultimately does not change your workflow or your methods.
--->
-
 We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of open science reforms in psychology, which by-and-large also originated in grass-roots initiatives.
@@
 We agree with the Editor that there is no guarantee that *"these things are crucial to scientific advance in psychology"*.
-We have, however, provided reasoned arguments to support the "affordances" of FAIR theory: it *makes possible* particular scientific advancements, even if it does not guarantee them.
+We have, however, provided reasoned arguments to support the *potential impact* of FAIR theory: it makes particular scientific advancements possible.
+It is up to the scholarly community whether this potential impact will be realized.
 
-Compare FAIR theory to preregistration: it, too, was introduced first as a good idea.
-Only now - a decade later - are empirical studies are being conducted to investigate its effects.on scientific practices.
+To address this comment, in the Revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic nor guaranteed.
+Furthermore, we acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated:
 
-To address this comment, <font color="red">in the Revision:
-
-* We explicitly state that we describe the affordances of FAIR theory and argue for its potential impact by comparing it to other open science methods
-* We acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated.
-
-</font>
-
-<!-- However, our paper is a different kind of paper than an empirical study: we provide a reasoned argument to justify FAIR theory, based on philosophy of science   -->
-
-<!--  Unlike more prescriptive norms—such as commitments to falsificationism or methodological registration—FAIR theory does not constrain a researcher’s workflow or analytical choices.
-It simply makes theoretical assumptions transparent, findable, and versioned.
-This distinction is important: the goal is not to impose rigid procedures, but to enable a more cumulative and accountable theoretical discourse.
-Still, we agree with the reviewer that policies and infrastructure alone are insufficient to drive meaningful change.
-Culture, norms, and the nature of the subject matter all shape how—and whether—reforms take root.
-In that spirit, we see FAIR theory not as a top-down imposition, but as a bottom-up opportunity: it offers immediate utility to researchers who want their theories to be discoverable, citable, and improvable.
-Over time, its adoption may well depend not on mandates, but on whether it proves valuable enough to become a “no-brainer” for the psychological sciences as it has in fields like genetics.
--->
+`r get_revision("evaluateimpact")`
 

### 2025-06-20 action_letter.Rmd
Update action_letter.Rmd

@@
 [x] References should be done with hanging indents using APA style. 
-[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. WHAT??
+[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. 
 ```
 
-We have 
-Upload tables separately: tab_meehl.csv
-Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png
+In response to your request, we have: 
+- uploaded tables separately: tab_meehl.csv
+- uploaded figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png
 
@@
 The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
-We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails, and because these general principles are not tied to the R-pacakge and our specific implementation (they could be implemented using other infrastructures).
+We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails.
+Note that these general principles are neither tied to the specific R-pacakge not to our specific implementation.
+Importantly, they could be implemented using an entirely different infrastructure.
 
@@
 
-We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of open science reforms in psychology, which by-and-large also originated in grass-roots initiatives.
+We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of Open Science reforms in Psychology, which by-and-large also originated in grass-roots initiatives.
 
@@
 
-To address this comment, in the Revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic nor guaranteed.
+To address this comment, in the revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic or guaranteed.
 Furthermore, we acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated:
@@
 We thank the Reviewer for their thoughtful and encouraging review.
-We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory—not only within psychology, but also in related disciplines.
+We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory --- not only within psychology, but also in related disciplines.
 We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.
@@
 
-As suggested, we now briefly mention the R-package in the abstract:
+As suggested, we now briefly mention the R package in the abstract:
 
@@
 
-We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific paradigm:
+We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific philosophical paradigm:
 

### 2025-06-20 manuscript.Rmd
Update manuscript.Rmd
extended info on phonological loop

@@
 More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
+<span id="phonoloop">Take for example the phonological loop in Baddeley's working memory model [@baddeley1992].
+The idea behind it is that verbal information is kept in memory via a rehearsal loop with limited capacity.
+The verbal description of the phonological loop in Baddeley's theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate of information, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].</span>
 Without committing to specific implementations a-priori,

### 2025-06-20 action_letter.Rmd
added link to revision regarding phonological loop
Update action_letter.Rmd

@@
 
-**DO THIS, ANDREAS?**
+We added a brief description and a reference as suggested:
+
+`r get_revision("phonoloop")`
 
@@
 
-We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, in psychology the percentage of articles that share data has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
+We respectfully disagree; Open Data - one example of modular publishing - has increased dramatically in recent decades.
+In Psychology the percentage of articles that share data openly has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
 In other disciplines, the increase has been even larger (up to ~80%).
-Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
+Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in Psychology:
 
@@
 
-> We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
+> We are particularly indebted to one of our reviewers, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
 

### 2025-06-20 action_letter.Rmd
Update action_letter.Rmd
Update to the construct validity and nomological net comment

@@
 We agree that several psychological theories can be specified as a nomological network of abstract entities and empirical predictions/observations.
-However, not all theories have been specified to such a level or need to be specified in such a way. 
-As such, FAIR Theory accommodates nomological nets and construct validation endeavours, but is not identical to them. 
+However, not all theories have been specified in such a way or to such a level. 
+Nor do we think that this is what we should strive for categorically.
+As such, FAIR Theory accommodates nomological nets and construct validation endeavours, 
+but is not identical to them. 
+Thus, we decided not to mention nomological nets with respect to theory specification.
 It is interesting to note that Cronbach and Meehl (1955) did talk about (proto)FAIR requirements:
@@
 "A consumer of the test who rejects the author’s theory cannot accept the author’s validation.
-He must validate the test for himself, if he wishes to show that it represents the construct as he defines it."
-We're not sure if further changes need to be made to address this.
+He must validate the test for himself, if he wishes to show that it represents the construct as he defines it." (p.13)
+We have incorporated these quotes as a footnote at the end of Section "The Need for FAIR Theory", which reads as follows:
 

### 2025-06-20 manuscript.Rmd
Update manuscript.Rmd
R-package -> R package

@@
 
-## The Need for FAIR theory
+## The Need for FAIR Theory
 
@@
 We demonstrated how to create, version-control, and archive theories as digital information artifacts.
-We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
+We introduced the `theorytools` R package to partly automate these processes, reducing the barrier of entry for researchers,
 and creating a FAIR resource for theory construction tools and documentation that can be updated as best practices continue to develop.
@@
 For this reason, we advocate the use of Git for version control, Zenodo for archiving, and DataCite for standardized metadata.
-Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R-package.
+Barriers to entry can also be lowered by simplifying workflows, which is the goal of the `theorytools` R package.
 Fourth, the availability of Open Educational Materials (OEM) about theory development contributes to doctoral socialization.
@@
 Another limitation is the learning curve associated with tools like Git and Zenodo.
-The `theorytools` R-package mitigates this limitation for R-users by automating key steps in the process.
+The `theorytools` R package mitigates this limitation for R-users by automating key steps in the process.
 Moreover, the initial investment in time can be offset by long-term productivity gains and increased impact of FAIR theory.

### 2025-06-20 action_letter.Rmd

@@
 
-We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of Open Science reforms in Psychology, which by-and-large also originated in grass-roots initiatives.
+We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of Open Science reforms in psychology, which by-and-large also originated in grass-roots initiatives.
 
@@
 We respectfully disagree; Open Data - one example of modular publishing - has increased dramatically in recent decades.
-In Psychology the percentage of articles that share data openly has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
+In psychology, the percentage of articles that share data openly has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
 In other disciplines, the increase has been even larger (up to ~80%).
-Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in Psychology:
+Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
 

### 2025-06-20 test.Rmd

@@
 The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
-To this end, we introduce "FAIR theory":
+We introduce "FAIR theory":
 a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially associated with - the scientific paper.
 <span id="alldefinitions">Definitions of theory abound and hotly debated,
-but as many of them are compatible with the FAIR principles,
-this paper is not limited to a specific definition.</span>
+but as many are compatible with the FAIR principles,
+this paper is not limited to one particular definition.</span>
 FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
@@
 a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases, see Figure \@ref(fig:figecs)a.
-In one half of the cycle, labeled the "Context of Justification" by Wagenmakers and colleagues, hypotheses derived from theory are tested on data [-@wagenmakersCreativityVerificationCyclePsychological2018].
+research ideally follows a cyclical process (Figure \@ref(fig:figecs)a).
+Naturalistic observations or patterns identified in data give rise to preliminary hypotheses via induction.
+Then, deduction is used to derive testable predictions from these hypotheses.
+Finally, the outcomes of tests are evaluated with regard to their implications for theory.
+Wagenmakers and colleagues divided this cycle into two phases [-@wagenmakersCreativityVerificationCyclePsychological2018, Figure \@ref(fig:figecs)b].
+In one half of the cycle, which they labeled the "Context of Justification",
+hypotheses derived from theory are tested on data.
 In the other half of the cycle (the "Context of Discovery"),
-patterns observed in data are generalized to theoretical principles, Figure \@ref(fig:figecs)b.
+patterns observed in data are generalized to theoretical principles.
+<!-- Hier iets over mijn interpretatie -->
 In this model, theories are the vehicle of scientists' understanding of phenomena.
-Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
 
@@
 In a progressive research program [@lakatosHistoryScienceIts1971],
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena via deductive testing and inductive theory construction.
 There are, however, indications that contemporary psychology falls short of this ideal.
@@
 and ground their hypotheses in established theoretical foundations.
-It further increases the impact and reuse potential of theories across disciplines,
+It could increase the impact and reuse potential of theories across disciplines,
 either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
@@
 If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
-we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \@ref(tab:tabmeehl)).
+we see how each of these properties makes a specific type of interoperability possible (Table \@ref(tab:tabmeehl)).
 
@@
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-Elaborating on our running example, several authors have reinterpreted De Groot's empirical cycle.
-For example, Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle differs from De Groot's in several ways.
-First, the phases of the cycle were renamed, and this change was not described in the paper.
-Assuming that Wagenmakers' new labels are meant to illustrate the phases,
-not substantially change the ontology,
-we could incorporate this change by adding labels to the original ontology.
+Elaborating on our running example, we can implement Wagenmakers and colleagues' take on the empirical cycle [-@wagenmakersCreativityVerificationCyclePsychological2018, Figure \@ref(fig:figecs)b).
+Their interpretation differs from De Groot's in several ways:
+First, they refer to the phases by different names, and this change was not described in the paper.
+Assuming that these new names are merely intended to be illustrative and not ontologically distinct,
+we can incorporate this change by adding labels to the original ontology.
 These labels suggest a focus on empirical psychology that was not present in De Groot's version.
-Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
-As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
 The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
@@
 
-The first author was inspired by De Groot as well,
-but again specified the empirical cycle differently.
+In implementing these different versions,
+the first author realized that their work is guided by yet another different interpretation of the empirical cycle.
 First, in De Groot's formulation, each stage describes a process.
 This invites the question of what the concrete outcomes of these processes are.
-In other words: what actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
-Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
-Theory testing, as takes place in the "context of justification", can be said to involve mostly deductive reasoning.
-Theory development and amendment, as takes place in the "context of discovery", involves primarily inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations.
+For example, note the label "knowledge accumulation" in Wagenmakers' version.
+This label invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
+What actually changes when going through the cycle, except the scholar's mind?
+To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, and the edges connecting the nodes describe processes acting upon those deliverables (Figure \@ref(fig:figecs)c).
+In this specification, `theory` can be understood as FAIR theory.
+
+A second distinction is that the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
+Theory testing, as takes place in the "context of justification", can be said to involve *mostly* deductive reasoning.
+Theory development and amendment, as takes place in the "context of discovery", involves mostly inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations.
 Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960].
 For present purposes, the terms are interchangeable.].
-However, deriving hypotheses from theory is not purely deductive
+However, deriving hypotheses from theory is not *entirely* deductive
 as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
-A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
-Similarly, if we consider the claim that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
-Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
+A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
+Similarly, if we accept that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
+Similarly, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
 These alterations result in the following implementation of the empirical cycle:
@@
 ```
-
 theory;
@@
 results -> theory [label="interpretation and generalization"];
-
 ```
@@
 as explained in [this vignette](ADD ME).
-For example, Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
 
@@
 
-<!-- FAIR theory incorporates theory into open science workflows, -->
-<!-- facilitates scholarly communication about theories, -->
-<!-- making it easier to share theories with less opportunity for ambiguity and misunderstanding. -->
-<!-- FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories. -->
-<!-- More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition, -->
-<!-- removes barriers to knowledge exchange with the global scholarly community, -->
-<!-- opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration. -->
-
 Another future direction for FAIR theory is as an instrument of science communication.
@@
 By providing a mechanism for iterative improvement based on emerging evidence, FAIR theory also supports effective evidence-based decision making.
+
+<span id="evaluateimpact">Finally, while this paper has discussed the potential impact of FAIR theory in addressing contemporary challenges in psychology,
+future research is needed to evaluate whether this potential is realized.
+Registered reports were first introduced to psychology about a decade ago [@nosekRegisteredReportsMethod2014],
+and are now becoming commonplace, as are studies evaluating their adoption and impact on scientific practices and research results [@scheelExcessPositiveResults2021].
+We envision a similar trajectory for FAIR theory.</span>
 

### 2025-06-20 test.tex

@@
 We also thank the Editor, Dr.~Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
-Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id \#6e9f28c9.
+Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id \#a5acd015.
 The first and senior author positions were fixed, and remaining author positions randomized.
@@
 FAIR theories are Findable in well-established archives, Accessible in terms of availability and their ability to be understood, Interoperable for specific purposes, such as selecting control variables, and Reusable so that they can be iteratively improved through collaborative efforts.
-\phantomsection\label{abstractpackage}{This paper adapts the FAIR principles for theory, reflects on FAIR practices in contemporary psychology, introduces a workflow for FAIRifying theory, which is largely automated by the \texttt{theorytools} R-package}, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
-FAIR theory constitutes a protocol for archiving and communicating about theory, addressing a critical gap in open scholarly practices and supporting the renewed interest in theory development in psychology and beyond.
-FAIR theory builds on existing open science principles and infrastructures to provide a structured, cumulative framework for theory development, potentially increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
+This paper adapts the FAIR principles for theory, reflects on current FAIR practices in relation to psychological theory, and discusses FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on theories' structure and development, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies.
+\phantomsection\label{abstractpackage}{We present a conceptual workflow for FAIRifying theory in this paper, which builds on existing open science principles and infrastructures.
+More detailed tutorials, worked examples, and convenience functions to automate this workflow are available in the \texttt{theorytools} R-package.} FAIR theory constitutes a structured protocol for archiving, communicating about, and iteratively improving theory, addressing a critical gap in open scholarly practices and potentially increasing the efficiency of cumulative knowledge acquisition in psychology and beyond.
 }
@@
 The present paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
-To this end, we introduce ``FAIR theory'':
+We introduce ``FAIR theory'':
 a digital instantiation of scientific theory, published as a self-contained and citable digital object distinct from - but potentially associated with - the scientific paper.
 \phantomsection\label{alldefinitions}{Definitions of theory abound and hotly debated,
-but as many of them are compatible with the FAIR principles,
-this paper is not limited to a specific definition.}
+but as many are compatible with the FAIR principles,
+this paper is not limited to one particular definition.}
 FAIR theory can potentially improve the transparency and efficiency of scholarly communication, reduce research waste, and accelerate cumulative knowledge acquisition.
@@
 a meta-theoretical model of cumulative knowledge acquisition,
-research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}a.
-In one half of the cycle, labeled the ``Context of Justification'' by Wagenmakers and colleagues, hypotheses derived from theory are tested on data (2018).
+research ideally follows a cyclical process (Figure \ref{fig:figecs}a).
+Naturalistic observations or patterns identified in data give rise to preliminary hypotheses via induction.
+Then, deduction is used to derive testable predictions from these hypotheses.
+Finally, the outcomes of tests are evaluated with regard to their implications for theory.
+Wagenmakers and colleagues divided this cycle into two phases (2018, Figure \ref{fig:figecs}b).
+In one half of the cycle, which they labeled the ``Context of Justification'',
+hypotheses derived from theory are tested on data.
 In the other half of the cycle (the ``Context of Discovery''),
-patterns observed in data are generalized to theoretical principles, Figure \ref{fig:figecs}b.
+patterns observed in data are generalized to theoretical principles.
 In this model, theories are the vehicle of scientists' understanding of phenomena.
-Ideally, they are iteratively updated based on deductive testing and inductive theory construction.
 
@@
 In a progressive research program (Lakatos, 1971),
-this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
+this cycle is regularly completed to iteratively advance our understanding of the studied phenomena via deductive testing and inductive theory construction.
 There are, however, indications that contemporary psychology falls short of this ideal.
@@
 If we consider Meehl's nine properties of strong theories (properties 3-8 are grouped because they all refer to functional form),
-we see how each of these properties incurs certain affordances in terms of X-Interoperability (Table \ref{tab:tabmeehl}).
+we see how each of these properties makes a specific type of interoperability possible (Table \ref{tab:tabmeehl}).
 
@@
 \begin{verbatim}
-  induction   [label="formulation of hypotheses"];
-  deduction   [label="derivation of specific consequences from the hypotheses,
-                      in the form of testable predictions"];
-  observation [label="collection and grouping of empirical materials;
-                      (tentative) formation of hypotheses"];
-  test        [label="of the hypotheses against new empirical materials, by way
-                      of checking whether or not the predictions are fulfilled"];
-  evaluation  [label="of the outcome of the testing procedure with respect to
-                      the hypotheses or theories stated, as well as with a view
-                      to subsequent, continued or related, investigations"];
+induction;
+deduction;
+observation;
+test;
+evaluation;
 
@@
 \item
-  Set the \emph{resource type} to the category \texttt{Model},
+  Set the \texttt{resource\ type} to \texttt{Model},
 \item
-  Add the words \texttt{FAIR\ theory:} to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words \texttt{meta-analysis} in the title),
+  Add the words \texttt{FAIR\ theory:} to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses should use the words \texttt{meta-analysis} in the title),
 \item
-  Add \texttt{fairtheory} to the keywords to aid search engine indexation.
+  Add \texttt{fairtheory} to the keywords to aid search engine indexation (using ``fair theory'' causes some search engines to look for either the word ``fair'' or ``theory'', which would be overly inclusive).
 \item
@@
 We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
-Elaborating on our running example, several authors have reinterpreted De Groot's empirical cycle.
-For example, Wagenmakers and colleagues' (2018) interpretation of the empirical cycle differs from De Groot's in several ways.
-First, the phases of the cycle were renamed, and this change was not described in the paper.
-Assuming that Wagenmakers' new labels are meant to illustrate the phases,
-not substantially change the ontology,
-we could incorporate this change by adding labels to the original ontology.
+Elaborating on our running example, we can implement Wagenmakers and colleagues' take on the empirical cycle {[}-Wagenmakers et al. (2018), Figure \ref{fig:figecs}b).
+Their interpretation differs from De Groot's in several ways:
+First, they refer to the phases by different names, and this change was not described in the paper.
+Assuming that these new names are merely intended to be illustrative and not ontologically distinct,
+we can incorporate this change by adding labels to the original ontology.
 These labels suggest a focus on empirical psychology that was not present in De Groot's version.
-Furthermore, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
-As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
 The authors explicitly mention a second change: \emph{``We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''} (p.~423).
@@
 
-The first author was inspired by De Groot as well,
-but again specified the empirical cycle differently.
+In implementing these different versions,
+the first author realized that their work is guided by yet another different interpretation of the empirical cycle.
 First, in De Groot's formulation, each stage describes a process.
 This invites the question of what the concrete outcomes of these processes are.
-In other words: what actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \ref{fig:figecs}c).
-Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
-Theory testing, as takes place in the ``context of justification'', can be said to involve mostly deductive reasoning.
-Theory development and amendment, as takes place in the ``context of discovery'', involves primarily inductive reasoning\footnote{Here, ``induction'' is defined forming general theories based on specific observations.
+For example, note the label ``knowledge accumulation'' in Wagenmakers' version.
+This label invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
+What actually changes when going through the cycle, except the scholar's mind?
+To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, and the edges connecting the nodes describe processes acting upon those deliverables (Figure \ref{fig:figecs}c).
+In this specification, \texttt{theory} can be understood as FAIR theory.
+
+A second distinction is that the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
+Theory testing, as takes place in the ``context of justification'', can be said to involve \emph{mostly} deductive reasoning.
+Theory development and amendment, as takes place in the ``context of discovery'', involves mostly inductive reasoning\footnote{Here, ``induction'' is defined forming general theories based on specific observations.
 Others have used the term ``abduction'' to describe ``inference to the best explanation'' (Peirce, 1960).
 For present purposes, the terms are interchangeable.}.
-However, deriving hypotheses from theory is not purely deductive
+However, deriving hypotheses from theory is not \emph{entirely} deductive
 as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
-A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
-Similarly, if we consider the claim that observation is theory-laden, then it too involves induction (Brewer \& Lambert, 2001).
-Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (Peikert, 2023).
+A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
+Similarly, if we accept that observation is theory-laden, then it too involves induction (Brewer \& Lambert, 2001).
+Similarly, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well (Peikert, 2023).
 These alterations result in the following implementation of the empirical cycle:
@@
 \begin{verbatim}
-
 theory;
@@
 as explained in \href{ADD\%20ME}{this vignette}.
-For example, Figure \ref{fig:figgithub-2} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
+Figure \ref{fig:figgithub-2} shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.
 
@@
 These aDAGs are X-interoperable for plotting (using \texttt{dagitty} and \texttt{tidySEM}),
-for automatically selecting control variables, and for simulating data (using \texttt{theorytools}).
-Another vignette describes how to take Self-Determination Theory (Deci \& Ryan, 2012), a theory originally represented as prose, and \href{https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html}{specify it as a FAIR aDAG}.
-Another vignette describes how to take the Dunning-Krüger effect and \href{https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html}{specify it as a FAIR mathematical formula} (Feld et al., 2017).
-Another vignette illustrates the use of FAIR theory for \href{https://cjvanlissa.github.io/theorytools/articles/causal-inference.html}{covariate selection and causal inference} (Pearl, 1995).
-More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
+for automatically selecting control variables, and for simulating data (using \texttt{theorytools}),
+as described in \href{https://cjvanlissa.github.io/theorytools/articles/causal-inference.html}{this vignette} (Pearl, 1995).
+Other vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
 
@@
 access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
-Whereas the idea of theory can be quite nebulous to empirical psychologists,
-FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
+Whereas ``theory'' can be a nebulous concept to empirical psychologists,
+FAIR theory makes it practical and tangible, incorporating theory into scholarly workflows.
 Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
@@
 we provide much-needed open science infrastructure for transparently deriving hypotheses and modifying theory,
-thus contributing to closing the ``open empirical cycle''.}.
-
-Our approach aligns closely with contemporary developments in open science,
+thus contributing to closing the ``open empirical cycle''.}
+
+Our approach aligns closely with recent developments in open science,
 such as modular publishing, interdisciplinarity, meta-research, and team science.
@@
 FAIR theories, by contrast, can be readily disseminated as citable digital objects, thus changing the incentive structure to favor theory development.
-
-Interdisciplinarity benefits from FAIR theory's Accessibility across different fields; thus, theoretical frameworks can be reused, adapted, or used for analogical modeling (Haslbeck et al., 2021).
+FAIR theory's Accessibility can advance interdisciplinarity,
+for example because theoretical frameworks can be reused, adapted, or used for analogical modeling across different fields (Haslbeck et al., 2021).
 Meta-research benefits from the fact that FAIR theory enables studying the structure, content, and development of theories over time.
@@
 
-\phantomsection\label{learningcurve}{Another limitation is the learning curve associated with tools like Git and Zenodo.
+\phantomsection\label{learningcurve}{Another limitation is the learning curve associated with tools and infrastructures like Git and Zenodo.
 The \texttt{theorytools} R-package mitigates this limitation for R-users by automating key steps in the process.
@@
 
-One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories, also known as the ``toothbrush problem''.
+One potential barrier to adopting FAIR theory is cultural resistance to sharing and modifying theories (the ``toothbrush problem'').
 Education might help address this limitation; with this in mind,
@@
 \phantomsection\label{measurementcrisis}{Another future direction is the intersection between the aforementioned ``theory crisis'' and the related ``measurement crisis'' pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs (Bringmann et al., 2022).
-Since FAIR theories can reference other theories and resources,
-it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.}
+FAIR theories could reference specific measurement instruments, or even theories of measurement,
+when operationalizing constructs named in a theory.}
 \phantomsection\label{jinglejangle}{FAIR theories can also help address ``jingle- and jangle fallacies'' in psychology,
@@
 
+\phantomsection\label{evaluateimpact}{Finally, while this paper has discussed the potential impact of FAIR theory in addressing contemporary challenges in psychology,
+future research is needed to evaluate whether this potential is realized.
+Registered reports were first introduced to psychology about a decade ago (\textbf{nosekRegisteredReportsMethod2014?}),
+and are now becoming commonplace, as are studies evaluating their adoption and impact on scientific practices and research results (Scheel, Schijen, et al., 2021).
+We envision a similar trajectory for FAIR theory.}
+
 \subsection{Conclusion}\label{conclusion}
 
-FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
-It provides much-needed open science methods for the inductive phase of the empirical cycle,
-closing a critical gap in the scientific process.
-
-\phantomsection\label{turkheimer}{We invite the reader to ask the rhetorical question raised by our reviewer, prof.
-Turkheimer, who - reflecting on a long career as a practitioner of theory in psychology - considered ``how adopting a FAIR framework would change, improve, or inhibit what I do''.
-We paraphrase his answer here\footnote{Edited for brevity.}, because his lived experience complements our youthful ``techno-enthusiasm''.
+\phantomsection\label{turkheimer}{We invite the reader to ask the rhetorical question raised by one of our reviewers, prof.
+Turkheimer, who - reflecting on a career as a practitioner of theory in psychology - considered ``how adopting a FAIR framework would change, improve, or inhibit what I do''.
 Prof.~Turkheimer remarked that a FAIR framework would have been very helpful in maintaining meaningful boundaries between a theory and the predictions it makes,
@@
 and in resolving theoretical disputes.
-\emph{``FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
+``FAIR theory provides a structured, machine- and human-readable framework that distills and formalizes the core components of a theory.
 Everything published about a theory can continue to exist in its original form,
@@
 Importantly, FAIR theory can evolve semi-independently from individual papers, offering a persistent and collaborative object that others can reuse, cite, and refine.
-Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability.''}}
-
-We envision a future where application of the FAIR principles makes theories more \emph{useful}, living up to Kurt Lewin's ideal,
+Contributing to the development of a FAIR theory thus becomes a new way of contributing to the theoretical literature - much like writing a high-quality review article, but with the added benefit of interoperability, versioning, and reusability.''}
+
+We envision a future where application of the FAIR principles makes theories more \emph{useful},
+living up to Kurt Lewin's ideal,
 enabling scholars to incorporate theory into their workflows in a tanglible way,

### 2025-06-21 action_letter.Rmd
update 2025-06-21 06:45:52.895789

--- 2025-06-20 action_letter.Rmd
@@
 [x] References should be done with hanging indents using APA style. 
-[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. 
+[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. WHAT??
 ```
 
-In response to your request, we have: 
-- uploaded tables separately: tab_meehl.csv
-- uploaded figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png
+We have 
+Upload tables separately: tab_meehl.csv
+Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png
 
@@
 The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
-We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails.
-Note that these general principles are neither tied to the specific R-pacakge not to our specific implementation.
-Importantly, they could be implemented using an entirely different infrastructure.
+We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails, and because these general principles are not tied to the R-pacakge and our specific implementation (they could be implemented using other infrastructures).
 
@@
 
-We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of Open Science reforms in Psychology, which by-and-large also originated in grass-roots initiatives.
+We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of open science reforms in psychology, which by-and-large also originated in grass-roots initiatives.
 
@@
 
-To address this comment, in the revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic or guaranteed.
+To address this comment, in the Revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic nor guaranteed.
 Furthermore, we acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated:
@@
 We thank the Reviewer for their thoughtful and encouraging review.
-We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory --- not only within psychology, but also in related disciplines.
+We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory—not only within psychology, but also in related disciplines.
 We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.
@@
 
-As suggested, we now briefly mention the R package in the abstract:
+As suggested, we now briefly mention the R-package in the abstract:
 
@@
 
-**DO THIS**
+We have rewritten this section so that it is clear that the two phases were not part of the original empirical cycle, but introduced by Wagenmakers (1b).
+
+We have also made the two phases visually more distinct by adding a different background to each in 1b.
+
+Finally, we now briefly introduce 1c as well:
+
+`r get_revision("introduce1c")`
 
@@
 
-We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific philosophical paradigm:
+We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific paradigm:
 
@@
 
-We added a brief description and a reference as suggested:
-
-`r get_revision("phonoloop")`
+**DO THIS, ANDREAS?**
 
@@
 
-We respectfully disagree; Open Data - one example of modular publishing - has increased dramatically in recent decades.
-In Psychology the percentage of articles that share data openly has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
+We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, in psychology the percentage of articles that share data has increased from ~7% in 2000–2009 to ~50% in 2010-2019.
 In other disciplines, the increase has been even larger (up to ~80%).
-Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in Psychology:
+Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:
 
@@
 
-> We are particularly indebted to one of our reviewers, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
+> We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
 

### 2025-06-21 test.Rmd
update 2025-06-21 06:45:52.895789

@@
 According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
-a meta-theoretical model of cumulative knowledge acquisition,
+a meta-theory of cumulative knowledge acquisition,
 research ideally follows a cyclical process (Figure \@ref(fig:figecs)a).
@@
 Wagenmakers and colleagues divided this cycle into two phases [-@wagenmakersCreativityVerificationCyclePsychological2018, Figure \@ref(fig:figecs)b].
-In one half of the cycle, which they labeled the "Context of Justification",
-hypotheses derived from theory are tested on data.
-In the other half of the cycle (the "Context of Discovery"),
-patterns observed in data are generalized to theoretical principles.
-<!-- Hier iets over mijn interpretatie -->
-In this model, theories are the vehicle of scientists' understanding of phenomena.
+One phase is the "Context of Justification",
+in which hypotheses derived from theory are tested on data.
+The other phase is the "Context of Discovery",
+in which patterns observed in data are generalized to theoretical principles.
+<span id="introduce1c">In these interpretations of the empirical cycle, each stage describes a process.
+This invites the question what the concrete outcomes of these processes are.
+How does knowledge accumulate when iterating through the cycle; what actually changes, other than the scholar's mind?
+To address these questions, the lead author specified a revised version of the empirical cycle (see "Changing a Theory")
+in which the nodes refer to specific deliverables,
+and the edges describe processes acting upon those deliverables (Figure \@ref(fig:figecs)c).
+In this specification, `theory` refers to FAIR theory,
+`hypothesis` to machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021],
+`test` refers to a preregistered inferential procedure [Preregistration-As-Code, @peikertReproducibleResearchTutorial2021],
+data refers to FAIR data,
+and results refer to (preprint) papers, ideally supplemented with a comprehensive reproducible research archive [@vanlissaWORCSWorkflowOpen2021].
+In this model, theories are the vehicle of scientists' understanding of phenomena;
+they are what changes when iterating through the cycle.</span>
 
@@
 ```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologyFoundationsInference1969].'}
-# rsvg::rsvg_pdf("empirical_cycle2.svg", "ec_visio.pdf")
+# rsvg::rsvg_pdf("empirical_cycle3.svg", "ec_visio.pdf")
 knitr::include_graphics("ec_visio.pdf")
@@
 
-In implementing these different versions,
-the first author realized that their work is guided by yet another different interpretation of the empirical cycle.
-First, in De Groot's formulation, each stage describes a process.
-This invites the question of what the concrete outcomes of these processes are.
-For example, note the label "knowledge accumulation" in Wagenmakers' version.
-This label invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
-What actually changes when going through the cycle, except the scholar's mind?
-To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, and the edges connecting the nodes describe processes acting upon those deliverables (Figure \@ref(fig:figecs)c).
-In this specification, `theory` can be understood as FAIR theory.
-
-A second distinction is that the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
+The present work is guided by an interpretation of the empirical cycle in which nodes refer to specific FAIR digital objects (FAIR theory, machine-readable hypotheses, FAIR data, et cetera),
+and edges connecting the nodes refer to processes acting on (X-operating with) those digital objects (Figure \@ref(fig:figecs)c).
+This implementation addresses an important ambiguity of prior versions of the empirical cycle.
+De Groot describes how the phase of evaluation involves "formation of new [theories]";
+evaluating results from one iteration of the cycle thus become observations (Phase 1) that initiate a new iteration.
+De Groot stops short of detailing the process of theory construction or revision,
+and in Wagenmakers' version, the term "knowledge accumulation" also remains unspecified.
+Thus in both prior versions, it is unclear *how* knowledge accumulates in the evaluation phase.
+In Van Lissa's specification (2025), `theory` is understood as FAIR theory,
+which has clear procedures for making changes and connecting these changes to empirical evidence.
+
+A second departure from De Groot is that the processes of induction and deduction are intentionally abandoned as distinct phases.
 Theory testing, as takes place in the "context of justification", can be said to involve *mostly* deductive reasoning.
@@
 However, deriving hypotheses from theory is not *entirely* deductive
-as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
-A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
-Similarly, if we accept that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
-Similarly, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
+as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction [@peikertTransparencyOpenScience2023].
+A common example is assuming equal variances across groups when testing a hypothesis about a mean difference between groups.
+One might justify this assumption because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
+Additionally, if we accept that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
 These alterations result in the following implementation of the empirical cycle:
@@
 ```
-theory;
-prediction;
-data;
-test;
-results;
-
-theory -> prediction [label="deduction"];
-prediction -> test [label = "implement inferential procedure"];
-data -> results;
-test -> results [label = "apply to data"];
-results -> theory [label="interpretation and generalization"];
+theory [label="fair theory"];
+prediction [label="preregistration"];
+data [label="fair data"];
+test [label="reproducible analysis pipeline"];
+results [label="preprint, reproducible research archive"];
+
+theory -> prediction [label="derive hypothesis, refer to specific fair theory"];
+prediction -> test [label = "implement inferential procedure as code"];
+data -> results [label="apply preregistered analysis to data to obtain results"];
+test -> results [label = "write preprint, create research archive"];
+results -> theory [label="interpret findings and modify fair theory"];
 ```

### 2025-06-21 test.Rmd
update 2025-06-21 06:55:29.337682

@@
 authornote: |
-  We are particularly endebted to our Reviewer, Dr.
+  We are particularly indebted to one of our Reviewers, Dr.
 Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.
@@
 Formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
-For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
+<span id="phonoloop">Take for example the phonological loop in Baddeley's working memory model [@baddeleyWorkingMemory1992].
+The idea behind it is that verbal information is kept in memory via a rehearsal loop with limited capacity.
+The verbal description of the phonological loop in Baddeley's theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate of information, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].</span>
 Without committing to specific implementations a-priori,
